# Introduction {#intro}


## Installation  

Install the development version from GitHub using `remotes` R package.  

```{r eval=TRUE, message=FALSE}
# install remotes 
#install.packages("remotes")
remotes::install_github("FelixErnst/mia")
```

## Package data  

The datasets in `mia` are conversions of the `phyloseq` datasets `GlobalPatterns` `enterotype`, `esophagus` and `soilrep`. 
Taxa contains example taxonomic data derived from the metagenomeFeatures package.

```{r}
library(mia)
# Example how to load data
data("GlobalPatterns")

GlobalPatterns
```

## Tools    

1. `mia`    : Microbiome analysis tools   
2. `miaViz` : Visualization support for `mia`   


**See also:**    

[`TreeSummarizedExperiment`](https://www.bioconductor.org/packages/release/bioc/html/TreeSummarizedExperiment.html)  
[`MicrobiomeExperiment`](https://github.com/FelixErnst/MicrobiomeExperiment)    [`SummarizedExperiment`](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html)  
[`microbiome`](https://bioconductor.org/packages/devel/bioc/html/microbiome.html)    




