# Basic concepts {#concepts}


## Prevalence

Population prevalence (frequency) at a 1% relative abundance threshold:

```{r concepts_prevalence, warning=FALSE, message=FALSE}
library(mia)
data(GlobalPatterns)
me <- GlobalPatterns # Rename data for generic examples
head(mia::getPrevalence(me, detection = 1/100, sort = TRUE, as_relative = TRUE))
```


Population prevalence (frequency) at the absolute abundance threshold at read count 1:

```{r concepts_prevalence2}
head(mia::getPrevalence(me, detection = 1, sort = TRUE, abund_values = "counts", as_relative = FALSE))
```



## Prevalent microbiota analysis

If you only need the names of the prevalent taxa, do as follows. This
returns the taxa that exceed the given prevalence and detection
thresholds.

```{r core-members, message=FALSE, warning=FALSE, eval = FALSE}
prev <- mia::getPrevalentTaxa(me, detection = 0, prevalence = 50/100)
```

See also related functions for the analysis of rare and variable taxa (rareMembers; rareAbundance; lowAbundance).  

