# (PART) Introduction {-}

# Data Infrastructure {#data-introduction}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

The
[`SummarizedExperiment`](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html)
(`SE`) is a widely used class for analyzing data obtained by common sequencing
techniques. `SE` is common structure for several Bioconductor packages that are
used for analyzing RNAseq, ChIp-Seq data. `SE` class is also used in R packages
for analyzing microarrays, flow cytometry, proteomics, single-cell sequencing
data and many more. The single-cell analysis is facilitated by
[SingelCellExperiment
class](https://bioconductor.org/packages/release/bioc/vignettes/SingleCellExperiment/inst/doc/intro.html)
(`SCE`), which allows the user to store results of dimensionality reduction and
alternative experiments. Alternative experiments (`altExps`) can be differently
processed data within the analysis workflows.

Recently,
[`TreeSummarizedExperiment`](https://www.bioconductor.org/packages/release/bioc/html/TreeSummarizedExperiment.html)
(`TSE`)
was developed to extend the `SE` and `SCE` class for incorporating hierarchical
information (including phylogenetic tree) and reference sequences.

The `mia` package implements tools using these classes for analysis of
microbiome sequencing data.

## Installation

Install the development version from GitHub using `remotes` R package.  

```{r eval=FALSE, message=FALSE}
# install remotes 
#install.packages("remotes")
BiocManager::install("FelixErnst/mia")
```

### Packages    

1. `mia`    : Microbiome analysis tools   
2. `miaViz` : Microbiome analysis specific visualization

**See also:**    

[`microbiome`](https://bioconductor.org/packages/devel/bioc/html/microbiome.html)


## Background

The widely used `phyloseq` package and class were around before the [`SummarizedExperiment`](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html)  
and the derived 
[`TreeSummarizedExperiment`](https://www.bioconductor.org/packages/release/bioc/html/TreeSummarizedExperiment.html) 
class. Many methods for taxonomic profiling data are readily available for the  `phyloseq` class structure. 

In order to facilitate the transition, we provide here a short description how `phyloseq` and `*Experiment` classes relate to 
each other.

`assays`     : This slot is similar to `otu_table` in `phyloseq`. In a `SummarizedExperiment`
               object multiple assays, raw counts, transformed counts can be stored. See also 
               [`MultiAssayExperiment`](https://bioconductor.org/packages/release/bioc/html/MultiAssayExperiment.html) for storing data from multiple `experiments` such as RNASeq, Proteomics, etc.       
`rowData`    : This slot is similar to `tax_table` in `phyloseq` to store taxonomic information.     
`colData`    : This slot is similar to `sample_data` in `phyloseq` to store information related to samples.    
`rowTree`    : This slot is similar to `phy_tree` in `phyloseq` to store phylogenetic tree.     

In this book, you will come across terms like `FeatureIDs` and `SampleIDs`.   
`FeatureIDs` : These are basically OTU/ASV ids which are row names in `assays` and `rowData`.    
`SampleIDs`  : As the name suggests, these are sample ids which are column names in `assays` and row names in `colData`.  

`FeatureIDs` and `SampleIDs` are used but the technical terms `rownames` and 
`colnames` are encouraged to be used, since they relate to actual objects we 
work with.

<img src="https://raw.githubusercontent.com/FelixErnst/TreeSummarizedExperiment
/2293440c6e70ae4d6e978b6fdf2c42fdea7fb36a/vignettes/tse2.png" width="100%"/>

**Figure sources:** 

**Original article**
-   Huang R _et al_. (2021) [TreeSummarizedExperiment: a S4 class 
for data with hierarchical structure](https://doi.org/10.12688/
f1000research.26669.2). F1000Research 9:1246.

**Reference Sequence slot extension**
- Lahti L _et al_. (2020) [Upgrading the R/Bioconductor ecosystem for microbiome 
research](https://doi.org/10.7490/
f1000research.1118447.1) F1000Research 9:1464 (slides).

## Loading experimental microbiome data


### Importing data from external files

Microbiome (taxonomic) profiling data is commonly distributed in
various file formats. You can import such external data files as a
(Tree)SummarizedExperiment object but the details depend on the file
format. Here, we provide examples for common formats.

**CSV data tables** can be imported with the standard R functions,
  then converted to the desired format. For detailed examples, you can
  check the [Bioconductor course
  material](https://bioconductor.org/help/course-materials/2019/BSS2019/04_Practical_CoreApproachesInBioconductor.html)
  by Martin Morgan. The following example reads abundance tables,
  taxonomic mapping tables, and sample metadata, assuming that the
  input data files are properly prepared with appropriate row and
  column names.

```{r importingcsv, message=FALSE, eval=FALSE}
counts <- read.csv(count_file)   # Abundance table (e.g. ASV data; to assay data)
tax <- read.csv(tax_file)        # Taxonomy table (to rowData)
samples <- read.csv(sample_file) # Sample data (to colData)
se <- SummarizedExperiment(assays = list(counts = counts),
                           colData = samples,
                           rowData = tax)
```

Specific import functions are provided for:

-   Biom files (see `help(mia::loadFromBiom)`)
-   QIIME2 files (see `help(mia::loadFromQIIME2)`)
-   Mothur files (see `help(mia::loadFromMothur)`)

#### Example for importing Biom files

This example shows how Biom files are imported into a TreeSummarizedExperiment object. 

The data is from following publication: 
Tengeler AC _et al._ (2020) [**Gut microbiota from persons with
attention-deficit/hyperactivity disorder affects the brain in
mice**](https://doi.org/10.1186/s40168-020-00816-x). 

The data set consists of 3 files:

-   biom file: abundance table and taxonomy information
-   csv file: sample metadata
-   tree file: phylogenetic tree

Store the data in your desired local directory (for instance, _data/_ under the
working directory), and define source file paths

```{r}
biom_file_path <- "data/Aggregated_humanization2.biom"
sample_meta_file_path <- "data/Mapping_file_ADHD_aggregated.csv"
tree_file_path <- "data/Data_humanization_phylo_aggregation.tre"
```  

Now we can load the biom data into a SummarizedExperiment (SE) object.

```{r}
library(mia)

# Imports the data
se <- loadFromBiom(biom_file_path)

# Check
se
```  

The `assays` slot includes a list of abundance tables. The imported abundance table is named as "counts".
Let us inspect only the first cols and rows.

```{r}
assays(se)$counts[1:3, 1:3]
```

The `rowdata` includes taxonomic information from the biom file. The `head()` command
shows just the beginning of the data table for an overview.

`knitr::kable()` is for printing the information more nicely.

```{r}
head(rowData(se))
```

These taxonomic rank names (column names) are not real rank names. Let’s replace them with real rank names.

In addition to that, the taxa names include, e.g., '"k__' before the name, so let's
make them cleaner by removing them. 

```{r}
names(rowData(se)) <- c("Kingdom", "Phylum", "Class", "Order", 
                        "Family", "Genus")

# Goes through the whole DataFrame. Removes '.*[kpcofg]__' from strings, where [kpcofg] 
# is any character from listed ones, and .* any character.
rowdata_modified <- BiocParallel::bplapply(rowData(se), 
                                           FUN = stringr::str_remove, 
                                           pattern = '.*[kpcofg]__')

# Genus level has additional '\"', so let's delete that also
rowdata_modified <- BiocParallel::bplapply(rowdata_modified, 
                                           FUN = stringr::str_remove, 
                                           pattern = '\"')

# rowdata_modified is a list, so it is converted back to DataFrame format. 
rowdata_modified <- DataFrame(rowdata_modified)

# And then assigned back to the SE object
rowData(se) <- rowdata_modified

# Now we have a nicer table
head(rowData(se))
```

We notice that the imported biom file did not contain the sample meta data
yet, so it includes an empty data frame.

```{r}
head(colData(se))
```

Let us add a sample metadata file.

```{r}
# We use this to check what type of data it is
# read.table(sample_meta_file_path)

# It seems like a comma separated file and it does not include headers
# Let us read it and then convert from data.frame to DataFrame
# (required for our purposes)
sample_meta <- DataFrame(read.table(sample_meta_file_path, sep = ",", header = FALSE))

# Add sample names to rownames
rownames(sample_meta) <- sample_meta[,1]

# Delete column that included sample names
sample_meta[,1] <- NULL

# We can add headers
colnames(sample_meta) <- c("patient_status", "cohort", "patient_status_vs_cohort", "sample_name")

# Then it can be added to colData
colData(se) <- sample_meta
```

Now `colData` includes the sample metadata.

```{r}
head(colData(se))
```

Now, let's add a phylogenetic tree.

The current data object, se, is a SummarizedExperiment object. This
does not include a slot for adding a phylogenetic tree. In order to do
this, we can convert the SE object to an extended TreeSummarizedExperiment
object which includes also a `rowTree` slot.

```{r}
tse <- as(se, "TreeSummarizedExperiment")

# tse includes same data as se
tse
```

Next, let us read the tree data file and add it to the R data object (tse).

```{r}
# Reads the tree file
tree <- ape::read.tree(tree_file_path)

# Add tree to rowTree
rowTree(tse) <- tree

# Check
tse
```

Now `rowTree` includes a phylogenetic tree:

```{r, eval=FALSE}
head(rowTree(tse))
```

### Conversions between data formats in R

If the data has already been imported in R in another format, it
can be readily converted into `TreeSummarizedExperiment`, as shown in our next
example. Note that similar conversion functions to
`TreeSummarizedExperiment` are available for multiple data formats via
the `mia` package (see makeTreeSummarizedExperimentFrom* for phyloseq,
Biom, and DADA2).

```{r, message=FALSE}
library(mia)

# phyloseq example data
data(GlobalPatterns, package="phyloseq") 
GlobalPatterns_phyloseq <- GlobalPatterns
GlobalPatterns_phyloseq
```

```{r, message=FALSE}
# convert phyloseq to TSE
GlobalPatterns_TSE <- makeTreeSummarizedExperimentFromPhyloseq(GlobalPatterns_phyloseq) 
GlobalPatterns_TSE
```

We can also convert `TreeSummarizedExperiment` objects into `phyloseq`
with respect to the shared components that are supported by both
formats (i.e. taxonomic abundance table, sample metadata, taxonomic
table, phylogenetic tree, sequence information). This is useful for
instance when additional methods are available for `phyloseq`.

```{r, message=FALSE}
# convert TSE to phyloseq
GlobalPatterns_phyloseq2 <- makePhyloseqFromTreeSummarizedExperiment(GlobalPatterns_TSE) 
GlobalPatterns_phyloseq2
```

## Metadata

## Microbiome and tree data specific aspects

```{r, message=FALSE}
library(mia)
data("GlobalPatterns", package = "mia")
se <- GlobalPatterns 
se
```

### Assays  

The `assays` slot contains the experimental data as count matrices. Multiple 
matrices can be stored the result of `assays` is actually a list of matrices.

```{r}
assays(se)
```

Individual assays can be accessed via `assay`

```{r}
assay(se, "counts")[1:5,1:7]
```

To illustrate the use of multiple assays, the relative abundance data can be 
calcualted and stored along the original count data using `relAbundanceCounts`.

```{r}
se <- relAbundanceCounts(se)
assays(se)
```

Now there are two assays available in the `se` object, `counts` and 
`relabundance`.

```{r}
assay(se, "relabundance")[1:5,1:7]
```

### colData

`colData` contains data on the samples.

```{r coldata}
colData(se)
```

### rowData

`rowData` contains data on the features of the analyzed samples. Of particular
interest for the microbiome field this is used to store taxonomic information.

```{r rowdata}
rowData(se)
```

### rowTree  

Phylogenetic trees also play an important role for the microbiome field. The 
`TreeSummarizedExperiment` class is able to keep track of feature and node
relations via two functions, `rowTree` and `rowLinks`.

A tree can be accessed via `rowTree` as `phylo` object.       
```{r rowtree}
rowTree(se)
```

The links to the individual features are available through `rowLinks`.

```{r rowlinks}
rowLinks(se)
```

Please note that there can be a 1:1 relationship between tree nodes and 
features, but this is not a must have. This means there can be features, which
are not linked to nodes, and nodes, which are not linked to features. To change
the links in an existing object, the `changeTree` function is available.

## Data conversion

Sometimes custom solutions are needed for analyzing the data. `mia` contains a 
few functions to help in these situations.

### Tidy data

For several custom analysis and visualization packages, such as those from the 
`tidyverse`, the `SE` data can be converted to long data.frame format with 
`meltAssay`.    

```{r}
library(mia)
molten_se <- meltAssay(se,
                       add_row_data = TRUE,
                       add_col_data = TRUE,
                       abund_values = "relabundance")
molten_se
```

## Conclusion

Some wrapping up...

## Additional Reading

### Lecture slides

Introduction to microbiome data science [lecture slides](https://github.com/microbiome/course_2021_radboud/tree/main/slides).

### R programming resources

 * R programming basics: [Base R](https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf)
 * Basics of R programming: [Base R](https://raw.githubusercontent.com/rstudio/cheatsheets/master/base-r.pdf)
 * [R cheat sheets](https://www.rstudio.com/resources/cheatsheets/)
 * R visualization with [ggplot2](https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf) 
 * [R graphics cookbook](http://www.cookbook-r.com/Graphs/)

Rmarkdown

* [Rmarkdown tips](https://rmarkdown.rstudio.com/)


RStudio

* [RStudio cheat sheet](https://www.rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf) 

### Resources for TreeSummarizedExperiment

 * SingleCellExperiment
   + [Publication](https://bioconductor.org/packages/release/bioc/vignettes/SingleCellExperiment/inst/doc/intro.html)
   + [Project page](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html)
 * SummarizedExperiment
   + [Publication](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html)
   + [Project page](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html)
 * TreeSummarizedExperiment
   + [Publication](https://f1000research.com/articles/9-1246)
   + [Project page](https://www.bioconductor.org/packages/release/bioc/html/TreeSummarizedExperiment.html)
   
### Resources for phyloseq

 * [List of R tools for microbiome analysis](https://microsud.github.io/Tools-Microbiome-Analysis/)
 * [phyloseq](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0061217)
 * [microbiome tutorial](http://microbiome.github.io/tutorials/)
 * [microbiomeutilities](https://microsud.github.io/microbiomeutilities/)
 * Bioconductor Workflow for Microbiome Data Analysis: from raw reads to community analyses ([Callahan et al. F1000, 2016](https://f1000research.com/articles/5-1492/v2)).




### Further reading


* [Data Analysis and Visualization in R for Ecologists](https://datacarpentry.org/R-ecology-lesson/) by Data Carpentry

* [Modern Statistics for Modern Biology. Holmes & Huber (2018)](http://web.stanford.edu/class/bios221/book/) for background in statistical analysis

* [Microbiome Data Science. Shetty & Lahti, 2019](https://openresearchlabs.github.io/publications/papers/2018-Shetty-Lahti-MDS.pdf)

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
