# Packages {#packages}


```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```


## Package installation

Several R packages provide methods for the analysis and manipulation
of `SummarizedExperiment` and related data containers. One of these is
`mia`. The installation for this and other packages has the following
procedure.

Stable Biocondcuctor release version can be installed with:

```{r eval=FALSE, message=FALSE}
BiocManager::install("microbiome/mia")
```

Biocondcuctor development version requires the installation of the
latest R beta version, and this is primarily recommended for those who
already have solid experience with R/Bioconductor and need access to
the latest experimental updates.

```{r eval=FALSE, message=FALSE}
BiocManager::install("microbiome/mia", version="devel")
```

The bleeding edge (and potentially unstable) development version lives
in Github:

```{r eval=FALSE, message=FALSE}
devtools::install_github("microbiome/mia")
```



## Some available packages

Some of the R packages supporting the framework include:

- [mia](microbiome.github.io/mia): Microbiome analysis tools [@R-mia]
- [miaViz](microbiome.github.io/miaViz): Microbiome analysis specific visualization [@Ernst2022]
- [miaSim](microbiome.github.io/miaSim): Microbiome data simulations [@Simsek2021]
- [miaTime](microbiome.github.io/miaTime): Microbiome time series analysis [@Lahti2021]
- philr (external, @Silverman2017)




## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
