# Packages {#packages}


```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```


## Package installation

Several R packages provide methods for the analysis and manipulation
of `SummarizedExperiment` and related data containers. One of these is
`mia`. The installation for this and other packages has the following
procedure.

Stable Biocondcuctor release version can be installed with:

```{r eval=FALSE, message=FALSE}
BiocManager::install("microbiome/mia")
```

Biocondcuctor development version requires the installation of the
latest R beta version, and this is primarily recommended for those who
already have solid experience with R/Bioconductor and need access to
the latest experimental updates.

```{r eval=FALSE, message=FALSE}
BiocManager::install("microbiome/mia", version="devel")
```

The bleeding edge (and potentially unstable) development version lives
in Github:

```{r eval=FALSE, message=FALSE}
devtools::install_github("microbiome/mia")
```



## Some available packages

Some of the R packages supporting the _TreeSummarizedExperiment_ framework include:

### mia family of methods

- [mia](microbiome.github.io/mia) : generic microbiome analysis tools   
- [miaViz](microbiome.github.io/miaViz) : microbiome data visualization
- [miaSim](microbiome.github.io/miaSim) : microbiome data simulation
- [miaTime](microbiome.github.io/miaTime) : microbiome time series analysis

### Tree-based methods

- [philr](http://bioconductor.org/packages/devel/bioc/html/philr.html) (external)

### Differential abundance

- [benchdamic](https://bioconductor.org/packages/release/bioc/vignettes/benchdamic/inst/doc/intro.html) for benchmarking differential abundance methods
- [ANCOMBC](https://bioconductor.org/packages/devel/bioc/html/ANCOMBC.html) for differential abundance analysis

### Manipulation

- [MicrobiotaProcess](https://bioconductor.org/packages/release/bioc/html/MicrobiotaProcess.html) for analyzing microbiome and other ecological data within the tidy framework

### Data

- [curatedMetagenomicData](https://bioconductor.org/packages/release/data/experiment/html/curatedMetagenomicData.html) a large collection of curated human microbiome data sets
- [microbiomeDataSets](https://bioconductor.org/packages/release/data/experiment/html/microbiomeDataSets.html) microbiome demo data sets



## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
