# Taxonomic information {#taxonomic-information}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r, message=FALSE}
library(mia)
data("GlobalPatterns")
se <- GlobalPatterns 
```

## Data agglomeration {#data-agglomeration}

Agglomerate at a specific taxonomic rank. If multiple assays (counts and relabundance) exists, both will be agglomerated.

```{r}
altExp(se, "Family") <- agglomerateByRank(se, rank = "Family",
                                          agglomerateTree = TRUE)

se
altExp(se, "Family")
```

```{r}
altExp(se, "Family") <- relAbundanceCounts(altExp(se, "Family"))
assay(altExp(se, "Family"), "relabundance")[1:5,1:7]
```
  
```{r}
assay(altExp(se, "Family"), "counts")[1:5,1:7]
```

These newly created data can also be kept in `altExp` a feature provided by 
`SingleCellExperiment`.    

`altExpNames` now consists of `family` level data. This can be extended to use 
any level present in `r taxonomyRanks(se)`.   

## Get unique  

Get which Phyla are present.  
```{r unique-1}
head(unique(rowData(se)[,"Phylum"]))
```

## Pick specific  

Retrieving of specific elements are required for specific analysis. For
instance, extracting abundances for a specific taxa in all samples or all taxa 
in one sample.  

### Abundances of all taxa in specific sample 
```{r}
taxa.abund.cc1 <- getAbundanceSample(se, 
                                     sample_id = "CC1",
                                     abund_values = "counts")

taxa.abund.cc1[1:10]
```

### Abundances of specific taxa in all samples   

```{r}
taxa.abundances <- getAbundanceFeature(se, 
                                      feature_id = "255340",
                                       abund_values = "counts")
taxa.abundances[1:10]
```

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
