# (PART) Focus Topics {-}

# Taxonomic Information {#taxonomic-information}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r, message=FALSE}
library(mia)
data("GlobalPatterns", package="mia")
se <- GlobalPatterns 
```

Taxonomic information is a key part of analyzing microbiome data and without
it, any type of data analysis probably will not make much sense. However,
the degree of detail of taxonomic information differs depending on the dataset
and annotation data used.

Therefore, the mia package expects a loose assembly of taxonomic information
and assumes certain key aspects:

* Taxonomic information is given as character vectors or factors in the 
`rowData` of a `SummarizedExperiment` object.
* The columns containing the taxonomic information must be named `domain`,
`kingdom`, `phylum`, `class`, `order`, `family`, `genus`, `species` or with
a capital first letter.
* the columns must be given in the order shown above
* column can be omited, but the order must remain

## Assigning taxonomic information.

There are a number of methods to assign taxonomic information. We like to give
a short introduction about the methods available without ranking one over the 
other. This has to be your choice based on the result for the individual 
dataset.

### dada2

The dada2 package [@R-dada2] implements the `assignTaxonomy` function, which 
takes as input the ASV sequences associated with each row of data and a training
dataset. For more information visit the 
[dada2 homepage](https://benjjneb.github.io/dada2/assign.html).

### DECIPHER

The DECIPHER package [@R-DECIPHER] implements the `IDTAXA` algorithm to assign
either taxonomic information or function information. For `mia`
only the first option is of interest for now and more information can be
found on the [DECIPHER website](http://www2.decipher.codes/Classification.html).

## Functions to access taxonomic information

`checkTaxonomy` checks whether the taxonomic information is usable for `mia`

```{r}
checkTaxonomy(se)
```

Since the `rowData` can contain other data, `taxonomyRanks` will return the 
columns `mia` assumes to contain the taxonomic information.

```{r}
taxonomyRanks(se)
```

This can then be used to subset the `rowData` to columns needed.

```{r}
rowData(se)[,taxonomyRanks(se)]
```

`taxonomyRankEmpty` checks for empty values in the given `rank` and returns a 
logical vector of `length(x)`.

```{r}
all(!taxonomyRankEmpty(se, rank = "Kingdom"))
table(taxonomyRankEmpty(se, rank = "Genus"))
table(taxonomyRankEmpty(se, rank = "Species"))
```

`getTaxonomyLabels` is a multi-purpose function, which turns taxonomic
information into a character vector of `length(x)`

```{r}
head(getTaxonomyLabels(se))
```

By default, this will use the lowest non-empty information to construct a
string with the following scheme `level:value`. If all levels are the same,
this part is omitted, but can be added by setting `with_rank = TRUE`.

```{r}
phylum <- !is.na(rowData(se)$Phylum) & 
    vapply(data.frame(apply(rowData(se)[,taxonomyRanks(se)[3:7]],1L,is.na)),all,logical(1))
head(getTaxonomyLabels(se[phylum,]))
head(getTaxonomyLabels(se[phylum,], with_rank = TRUE))
```

By default the return value of `getTaxonomyLabels` contains only unique elements
by passing it through `make.unique`. This step can be omitted by setting 
`make_unique = FALSE`.

```{r}
head(getTaxonomyLabels(se[phylum,], with_rank = TRUE, make_unique = FALSE))
```

To apply the loop resolving function `resolveLoop` from the
`TreeSummarizedExperiment` package [@R-TreeSummarizedExperiment] within
`getTaxonomyLabels`, set `resolve_loops = TRUE`.

### Generate a taxonomic tree on the fly

To create a taxonomic tree, `taxonomyTree` used the information and returns a
`phylo` object. Duplicate information from the `rowData` is removed.

```{r}
taxonomyTree(se)
```

```{r}
se <- addTaxonomyTree(se)
se
```

The implementation is based on the `toTree` function from the
`TreeSummarizedExperiment` package [@R-TreeSummarizedExperiment].

## Data agglomeration {#data-agglomeration}

One of the main applications of taxonomic information in regards to count data
is to agglomerate count data on taxonomic levels and track the influence of 
changing conditions through these levels. For this `mia` contains the
`agglomerateByRank` function. The ideal location to store the agglomerated data
is as an alternative experiment.

```{r}
se <- relAbundanceCounts(se)
altExp(se, "Family") <- agglomerateByRank(se, rank = "Family",
                                          agglomerateTree = TRUE)
altExp(se, "Family")
```

If multiple assays (counts and relabundance) exist, both will be agglomerated.

```{r}
assayNames(se)
assayNames(altExp(se, "Family"))
```

```{r}
assay(altExp(se, "Family"), "relabundance")[1:5,1:7]
```
  
```{r}
assay(altExp(se, "Family"), "counts")[1:5,1:7]
```

`altExpNames` now consists of `Family` level data. This can be extended to use 
any level present in `r mia::taxonomyRanks(se)`.   

## Data transformation

Data transformation is a very common procedure in microbiome analysis. 
In transformation, each data point is replaced with transformed value that is 
calculated by applying transformation formula to the data point. Transformation 
can be used, for example, to normalize skewed data, or to reduce weight of bigger 
values compared to smaller values. 

In mia package, transformations are applied to abundance data. The transformed 
abundance table is stored back to 'assays'. mia includes transformation 
functions for sample-wise or column-wise transformation ('transformSamples()'), 
and for feature-wise or row-wise transformation ('transformFeatures()'). 

For complete list of available transformations and parameters, see function 
[help](https://microbiome.github.io/mia/reference/transformCounts.html).

```{r}
se <- transformSamples(x = se, abund_values = "counts", method = "clr", 
                       pseudocount = 1, name = "clr_transformation")

head(assay(se, "clr_transformation"))
```

-   In 'pa' transformation, 'threshold' specifies the value that divides observations to
be absent or present. By default, it is 0.

```{r}
se <- transformFeatures(se, method = "pa", threshold = 10)

head(assay(se, "pa"))
```

```{r}
# list of abundance tables that assays slot contains
assays(se)
```

## Pick specific  

Retrieving of specific elements that are required for specific analysis. For
instance, extracting abundances for a specific taxa in all samples or all taxa 
in one sample.  

### Abundances of all taxa in specific sample 
```{r}
taxa.abund.cc1 <- getAbundanceSample(se, 
                                     sample_id = "CC1",
                                     abund_values = "counts")
taxa.abund.cc1[1:10]
```

### Abundances of specific taxa in all samples   

```{r}
taxa.abundances <- getAbundanceFeature(se, 
                                      feature_id = "Phylum:Bacteroidetes",
                                      abund_values = "counts")
taxa.abundances[1:10]
```


## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
