# Quality Control {#quality-control}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r, message=FALSE}
library(mia)
data("GlobalPatterns")
se <- GlobalPatterns 
```

## Get top taxa and taxonomy   

### Features  

The `getTopTaxa` can be used for identifying top taxa in the data.   
```{r top-feature-taxo}
top_features <- getTopTaxa(se, method="median", top=10)
tax_data <- rowData(se)[top_features,taxonomyRanks(se)]
tax_data
```

## Library size   

The total counts/sample can be calculated using the
`perCellQCMetrics`/`addPerCellQC` from the `scater` package. The first one 
just calculates the values, whereas the latter one directly adds them to the
`colData`.

```{r lib-size}
library(scater)
perCellQCMetrics(se)
se <- addPerCellQC(se)
colData(se)
```

The calcualted library sizes can be visualized using the `plotColData` function.

```{r viz-lib-size-1, fig.width=8, fig.height=4}
library(ggplot2)
plotColData(se,"sum","X.SampleID", colour_by = "X.SampleID") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r viz-lib-size-2, fig.width=8, fig.height=4}
plotColData(se,"sum","SampleType", colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
