# Quality Control {#quality-control}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r, message=FALSE}
library(mia)
data("GlobalPatterns", package="mia")
se <- GlobalPatterns 
```

## Get top taxa and taxonomy   

### Features  

The `getTopTaxa` can be used for identifying top taxa in the data.   
```{r top-feature-taxo}
top_features <- getTopTaxa(se, method="median", top=10)
tax_data <- rowData(se)[top_features,taxonomyRanks(se)]
tax_data
```

## Library size   

The total counts/sample can be calculated using the
`perCellQCMetrics`/`addPerCellQC` from the `scater` package. The former one 
just calculates the values, whereas the latter one directly adds them to the
`colData`.

```{r lib-size}
library(scater)
perCellQCMetrics(se)
se <- addPerCellQC(se)
colData(se)
```

The distribution of calculated library sizes can be visualized as a
histogram (left), or by sorting the samples by library size (right).

```{r plot-viz-lib-size-1, fig.width=8, fig.height=4, fig.cap="Library size distribution."}
library(ggplot2)

p1 <- ggplot(as.data.frame(colData(se))) +
        geom_histogram(aes(x = sum), color = "black", fill = "gray", bins = 30) +
        labs(x = "Library size", y = "Frequency (n)") + 
        # scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x), 
        # labels = scales::trans_format("log10", scales::math_format(10^.x))) +
        theme_bw() +
        theme(panel.grid.major = element_blank(), # Removes the grid
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black")) # Adds y-axis

library(dplyr)
df <- as.data.frame(colData(se)) %>%
        arrange(sum) %>%
        mutate(index = 1:n())
p2 <- ggplot(df, aes(y = index, x = sum/1e6)) +
        geom_point() +	
        labs(x = "Library size (million reads)", y = "Sample index") +	
        theme_bw() +
        theme(panel.grid.major = element_blank(), # Removes the grid
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black")) # Adds y-axis

library(patchwork)
p1 + p2
```

Library sizes - and other variables from `colData` - can be also visualized by using 
specified function called `plotColData`.

```{r plot-viz-lib-size-2, fig.width=8, fig.height=4, fig.cap="Library sizes per sample."}
library(ggplot2)
# Sort samples by read count, order the factor levels, and store back to se as DataFrame
# TODO: plotColData could include an option for sorting samples based on colData variables
colData(se) <- as.data.frame(colData(se)) %>%
                 arrange(X.SampleID) %>%
        	 mutate(X.SampleID = factor(X.SampleID, levels=X.SampleID)) %>%
		 DataFrame
plotColData(se,"sum","X.SampleID", colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1)) +
    labs(y = "Library size (N)", x = "Sample ID") 	    
```

```{r plot-viz-lib-size-3, fig.width=8, fig.height=4, fig.cap="Library sizes per sample type."}
plotColData(se,"sum","SampleType", colour_by = "SampleType") + 
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```


## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
