# Microbiome Diversity {#microbiome-diversity}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

Diversity measurements are an important topic, when analyzing microbiome data. 
Typically three levels of diversity measurements exist, which are trying to
put a number on different aspects of the questions associated with diversity 
[@Whittaker1690].

Many different ways for estimating such diversity measurements have been 
described in the literature. Which measurement is best or applicable for your 
samples, is not the aim of the following sections.

```{r}
library(mia)
data("GlobalPatterns")
se <- GlobalPatterns
```

## Alpha diversity

_Alpha diversity_, also sometimes interchangeably used with the terms 
_species diversity_ or _species richness_, summarizes the number different
species and their abundance per sample into a single number.

For this purpose wrapper functions are available interacting with other packages
implementing the calculation, such as _vegan_ [@R-vegan].

`estimateAlphaDiversity` is the main function for this purpose, which calculates
the selected diversity based on the selected assay data and adds it to the 
`colData` of the `SummarizedExperiment` under the given column `name`.

```{r}
se <- estimateAlphaDiversity(se, abund_values = "counts",
                             index = "shannon", name = "shannon")
head(colData(se)$shannon)
```

This allows the values to analyzed directly from the `colData`, for example
by plotting them using `plotColData` from the _scater_ package [@R-scater].

```{r plot-div-shannon, message=FALSE, fig.cap="Shannon diversity estimates plotted grouped by sample type."}
library(scater)
plotColData(se, "shannon", "SampleType", colour_by = "SampleType") +
    theme(axis.text.x = element_text(angle=45,hjust=1))
```

By default `estimateAlphaDiversity` will calculate all available estimates...

```{r}
se <- estimateAlphaDiversity(se)
```

.. and a plot comparing them can then be constructed directly.

```{r}
plots <- lapply(c("shannon","simpson","inv_simpson", "richness", "chao1", "ACE"),
                plotColData,
                object = se,
                x = "SampleType",
                colour_by = "SampleType")
plots <- lapply(plots,"+", theme(axis.text.x = element_text(angle=45,hjust=1)))
ggpubr::ggarrange(plotlist = plots, nrow = 2, ncol = 3, common.legend = TRUE, legend = "right")
```

## Beta diversity

In comparison to alpha diversities, beta diversities describe inter-microbiome
(dis-)similarites. Technically speaking, beta diversities are usually
represented as `dist` objects, which contain triangular data describing the
distance between each pair of samples. These distances are than subjected to
multi-dimensional scaling, also known as PCoA, to retrieve reduced dimensions
for further evaluation or visualization.

### Calculating beta diversity

In the following examples distances are calculated by variable functions
supplied to the `FUN` argument. The function can defined by the usre, but it
must return a `dist` function, which can then be used to calculate reduced
dimension either via MDS or NMDS and the results can be stored in the
`reducedDim`. The whole process is wrapped in the `runMDS` and `runNMDS`
functions.

```{r, message=FALSE}
se <- runMDS2(se, FUN = vegan::vegdist, name = "MDS_BC")
```

To visualize reduced dimensions, `scater` implements the `plotReducedDim` 
function, which can enrich a plot with additional information as colour, shape 
or size.

```{r plot-mds-bray-curtis, fig.cap="MDS plot based on the Bray-Curtis distances on the GlobalPattern dataset."}
plotReducedDim(se, "MDS_BC", colour_by = "SampleType")
```

With additional tools from the `ggplot2` universe, comparisons can be 
performed informing on the applicability to visualize the samples in a 
meaningful way.

```{r plot-mds-nmds-comparison, fig.cap="Comparison of MDS and NMDS plots based on the Bray-Curtis or euclidean distances on the GlobalPattern dataset.", message=FALSE}
se <- runMDS2(se, FUN = vegan::vegdist, name = "MDS_euclidean",
             method = "euclidean")
se <- runNMDS(se, FUN = vegan::vegdist, name = "NMDS_BC")
se <- runNMDS(se, FUN = vegan::vegdist, name = "NMDS_euclidean",
              method = "euclidean")
plots <- lapply(c("MDS_BC","MDS_euclidean","NMDS_BC","NMDS_euclidean"),
                plotReducedDim, object = se, colour_by = "SampleType")
ggpubr::ggarrange(plotlist = plots, nrow = 2, ncol = 2, common.legend = TRUE,
                  legend = "right")
```

The _UniFrac_ is a special case, since it requires data on the relationship
of features in form on a `phylo` tree. `calculateUniFrac` performs the 
calculation to return a `dist` object, which can again be used within `runMDS`.

```{r}
se <- runMDS2(se, FUN = calculateUniFrac, name = "UniFrac",
              tree = rowTree(se),
              ntop = nrow(se))
```

Plotting can be performed again using `plotReducedDim`.

```{r plot-unifrac}
plotReducedDim(se, "UniFrac", colour_by = "SampleType")
```

### Other ordination methods

Other dimension reduction methods, such as `PCA`, `t-SNE` and `UMAP` are 
inherited directly from the `scater` package.

```{r}
se <- runPCA(se, name = "PCA", exprs_values = "counts", ncomponents = 10)
```

```{r plot-pca, fig.cap="PCA plot on the GlobalPatterns data set containing sample from different sources."}
plotReducedDim(se, "PCA", colour_by = "SampleType")
```

As mentioned before, applicability of the different methods depends on your
sample set.

```{r}
se <- runTSNE(se, name = "TSNE", exprs_values = "counts", ncomponents = 3)
```

```{r plot-tsne, fig.cap="t-SNE plot on the GlobalPatterns data set containing sample from different sources."}
plotReducedDim(se, "TSNE", colour_by = "SampleType", ncomponents = c(1:3))
```

## Community comparisons

PERMANOVA and other methods.

## Further reading

In certain settings beta diversities might be used to group samples without
prior knowledge. For this we want to point to excellent resources on 
[how to extract information from the clusters](http://bioconductor.org/books/release/OSCA/clustering.html).

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
