# Microbiome Diversity {#microbiome-diversity}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

Diversity measurements are an important topic, when analyzing microbiome data. 
Typically three levels of diversity measurements exist, which are trying to
put a number on different aspects of the questions associated with diversity.

Many different ways for estimating such diversity measurements have been 
described in the literature. Which measurement is best or applicable for your 
samples, is not the aim of the following sections.

```{r}
data("GlobalPatterns")
se <- GlobalPatterns
```

## Alpha diversity

_Alpha diversity_, also sometimes interchangeably used with the terms 
_species diversity_ or _species richness_, summarizes the number different
species and their abundance per sample into a single number.

For this purpose wrapper functions are available interacting with other packages
implementing the calculation, such as _vegan_ [@R-vegan].

`estimateAlphaDiversity` is the main function for this purpose, which calculates
the selected diversity based on the selected assay data and adds it to the 
`colData` of the `SummarizedExperiment` under the given column `name`.

```{r}
se <- estimateAlphaDiversity(se, abund_values = "counts",
                             index = "shannon", name = "shannon")
head(colData(se)$shannon)
```

This allows the values to analyzed directly from the `colData`, for example
by plotting them using `plotColData` from the _scater_ package [@R-scater].

```{r plot-div-shannon, message=FALSE, fig.cap="Shannon diversity estimates plotted grouped by sample typo."}
library(scater)
plotColData(se, "shannon", "SampleType", colour_by = "SampleType")
```

By default `estimateAlphaDiversity` will calculate all available estimates...

```{r}
se <- estimateAlphaDiversity(se)
```

A plot comparing them can then be constructed directly.

```{r}
plots <- lapply(c("shannon","simpson","inv_simpson", "richness", "chao1", "ACE"),
                plotColData,
                object = se,
                x = "SampleType",
                colour_by = "SampleType")
plots <- lapply(plots,"+", theme(axis.text.x = element_text(angle=45,hjust=1)))
ggpubr::ggarrange(plotlist = plots, nrow = 1, common.legend = TRUE, legend = "right")
```

## Beta diversity


### Community comparisons

PERMANOVA and other methods.


### Visualizing beta diversity

Ordination with PCoA, UMAP, and other methods.


## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
