# Microbiome Community {#microbiome-community}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r load-pkg-data}
library(mia)
data("GlobalPatterns")
tse <- GlobalPatterns
```


## Community composition

Heatmaps


## Community typing

### Dirichlet Multinomial Mixtures (DMM)

This section focus on DMM analysis. 

One technique that allows to search for groups of samples that are
similar to each other is the [Dirichlet-Multinomial Mixture
Model](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0030126). In
DMM, we first determine the number of clusters (k) that best fit the
data (model evidence) using Laplace approximation. After fitting the
model with k clusters, we obtain for each sample k probabilities that
reflect the probability that a sample belongs to the given cluster.

Let's cluster the data with DMM clustering. 

```{r dmm}
# Runs model and calculates the most likely number of clusters from 1 to 7.
# Since this is a large dataset it takes long computational time.
# For this reason we use only a subset of the data; agglomerated by Phylum as a rank.
tse <- agglomerateByRank(tse, rank = "Phylum", agglomerateTree=TRUE)
tse_dmn <- runDMN(tse, name = "DMN", k = 1:7)
```

```{r}
# It is stored in metadata
tse_dmn
```

Return information on metadata that the object contains.

```{r}
names(metadata(tse_dmn))
```

This returns a list of DMN objects for a closer investigation.

```{r}
getDMN(tse_dmn)
```


Show Laplace approximation (model evidence) for each model of the k models.

```{r}
library(miaViz)
plotDMNFit(tse_dmn, type = "laplace")
```

Return the model that has the best fit.

```{r}
getBestDMNFit(tse_dmn, type = "laplace")
```
### PCoA for ASV-level data with Bray-Curtis; with DMM clusters shown with colors

Group samples and return DMNGroup object that contains a summary.
Patient status is used for grouping.

```{r}
dmn_group <- calculateDMNgroup(tse_dmn, variable = "SampleType",  exprs_values = "counts",
                               k = 2)

dmn_group
```

Mixture weights  (rough measure of the cluster size).


```{r}
DirichletMultinomial::mixturewt(getBestDMNFit(tse_dmn))
```


Samples-cluster assignment probabilities / how probable it is that sample belongs
to each cluster

```{r}
head(DirichletMultinomial::mixture(getBestDMNFit(tse_dmn)))
```

Contribution of each taxa to each component

```{r}
head(DirichletMultinomial::fitted(getBestDMNFit(tse_dmn)))
```
Get the assignment probabilities


```{r}
prob <- DirichletMultinomial::mixture(getBestDMNFit(tse_dmn))
# Add column names
colnames(prob) <- c("comp1", "comp2")

# For each row, finds column that has the highest value. Then extract the column 
# names of highest values.
vec <- colnames(prob)[max.col(prob,ties.method = "first")]

```

Computing the euclidean PCoA and storing it as dataframe

```{r}
# Does clr transformation. Pseudocount is added, because data contains zeros.
tse <- transformCounts(tse, method = "clr", pseudocount = 1)

# Gets clr table
clr_assay <- assays(tse)$clr

# Transposes it to get taxa to columns
clr_assay <- t(clr_assay)

# Calculates Euclidean distances between samples. Because taxa is in columns,
# it is used to compare different samples.
euclidean_dist <- vegan::vegdist(clr_assay, method = "euclidean")

# Does principal coordinate analysis
euclidean_pcoa <- ecodist::pco(euclidean_dist)

# Creates a data frame from principal coordinates
euclidean_pcoa_df <- data.frame(pcoa1 = euclidean_pcoa$vectors[,1], 
                                pcoa2 = euclidean_pcoa$vectors[,2])

```

```{r}
# Creates a data frame that contains principal coordinates and DMM information
euclidean_dmm_pcoa_df <- cbind(euclidean_pcoa_df,
                               dmm_component = vec)
# Creates a plot
euclidean_dmm_plot <- ggplot(data = euclidean_dmm_pcoa_df, 
                             aes(x=pcoa1, y=pcoa2,
                                 color = dmm_component)) +
  geom_point() +
  labs(x = "Coordinate 1",
       y = "Coordinate 2",
       title = "PCoA with Aitchison distances") +  
  theme(title = element_text(size = 12)) # makes titles smaller

euclidean_dmm_plot
```


## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
