# Cross-correlation {#cross-correlation}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
if(!require(ecodist)){
    install.packages("ecodist")
    library(ecodist)
}
```

```{r load-pkg-data}
library(mia)
data("GlobalPatterns", package="mia")
tse <- GlobalPatterns
```

With cross-correlation analysis, we can analyze how strongly and how differently
variables are associated between each other. For instance, we can analyze if 
higher presence of a specific taxon equals to higher levels of a biomolecule. 

Here, we analyze associations between taxa correlate and lipids. Data is from 
following publication Lahti _et al_. (2015) [Associations between the human intestinal 
microbiota, Lactobacillus rhamnosus GG and serum lipids indicated by 
integrated analysis of high-throughput profiling 
data](https://peerj.com/articles/32/).

```{r cross-correlation1}
# Imports the data
tse <- microbiomeDataSets::peerj32()

# Microbiome data
tse[[1]] 
```

```{r cross-correlation2}
# Lipid data
tse[[2]]
```

```{r cross-correlation3}
if(!require(microbiome)){
    BiocManager::install("microbiome")
}

# Does log10 transform for microbiome data
tse[[1]] <- transformSamples(tse[[1]], method = "log10", pseudocount = 1)

# Gets microbiome and lipid data to cross-correlate
x <- t(assay(tse[[1]], "log10"))
y <- t(assay(tse[[2]], "counts"))

# Cross correlates data sets
correlation_table <- microbiome::associate(x, y, method = "spearman", mode = "table", 
                                           p.adj.threshold = 0.05, n.signif = 1)

knitr::kable(head(correlation_table))
```

Manipulates and reorders the table

```{r cross-correlation4}
if(!require(reshape2)){
    install.packages("reshape2")
}
if(!require(dplyr)){
    install.packages("dplyr")
}
# Gets taxa that have at least one statistically significant correlation
taxa <- correlation_table %>% dplyr::filter(p.adj < 0.05 & abs(Correlation) > 0) %>% 
  dplyr::select(X1) %>% unique %>% sapply(as.character)
# Gets taxa that have at least one statistically significant correlation
lipids <- correlation_table %>% dplyr::filter(p.adj < 0.05 & abs(Correlation) > 0) %>% 
  dplyr::select(X2) %>% unique %>% sapply(as.character)

# Takes only those taxa and lipids that have statistically significant values
correlation_table <- correlation_table[correlation_table[["X1"]] %in% taxa & 
                                           correlation_table[["X2"]] %in% lipids, ]

# Converts data to matrix, correlations as values
mat <- reshape2::acast(correlation_table, X2 ~ X1, value.var = "Correlation")

# Hierarchical clustering, gets the order of taxa and lipids
taxa_indices <- hclust(as.dist(1 - cor(mat, use="pairwise.complete.obs")))$order
order_taxa <- colnames(mat)[taxa_indices]
lipids_indices <- hclust(as.dist(1 - cor(t(mat), use="pairwise.complete.obs")))$order
order_lipids <- rownames(mat)[lipids_indices]

# Converts taxa and lipids columns to factor so that they have the desired order
correlation_table[["X1"]] <- factor(correlation_table[["X1"]], levels = order_taxa)
correlation_table[["X2"]] <- factor(correlation_table[["X2"]], levels = order_lipids)
```

Creates the heatmap

```{r cross-correlation5, fig.height = 8, fig.width = 6}
# Determines the scaling of colors
limits <- c(-1, 1)
breaks <- seq(from = min(limits), to = max(limits), by = 0.2)
colours <- c("darkblue", "blue", "white", "red", "darkred")

# Which observation have p-value under 0.05? --> creates a subset
cor_table_sub <- correlation_table[which(correlation_table[["Correlation"]] < 0.05), ]

# Creates a ggplot object
ggplot(correlation_table, aes(x = X1, y = X2, fill = Correlation)) +
  geom_tile() +
  
  scale_fill_gradientn(name = "Correlation", 
                       breaks = breaks, limits = limits, colours = colours) + 
  
  # Adds label to those observations that have p-value under 0.05
  geom_text(data = cor_table_sub, aes(x = X1, y = X2, label = "+")) +
  
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1),
        legend.key.size = unit(1, "cm")) +
  labs(x = "Taxa", y = "Lipids")
```


## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```

