# Community typing (clustering) {#clustering}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r load-pkg-data1}
library(mia)
data("GlobalPatterns", package="mia")
tse <- GlobalPatterns
```

Clustering is an unsupervised machine learning technique. The idea of
it is to find clusters from the data. A cluster is a group of
features/samples that share pattern.  For example, with clustering, we
can find group of samples that share similar community
composition. There are multiple clustering algorithms available.

## Hierarchical clustering

Hiearchical clustering aims to find hiearchy between
samples/features. There are to approaches: agglomerative ("bottom-up")
and divisive ("top-down").

In agglomerative approach, each observation is first unique cluster.
Algorithm continues by agglomerating similar clusters. Divisive
approach starts with one cluster that contains all the
observations. Clusters are splitted recursively to clusters that
differ the most. Clustering ends when each cluster contains only one
observation.

Hiearchical clustering can be visualized with dendrogram tree. In each
splitting point, the three is divided into two clusters leading to
hierarchy.

Let's load data from mia package.

```{r hclust1}
library(mia)
library(vegan)

# Load experimental data
data(peerj13075)
(tse <- peerj13075)
```

Hierarchical clustering requires 2 steps. In the fist step, dissimilarities are 
calculated. In prior to that, data transformation is applied if needed. Since
sequencing data is compositional, relative transformation is applied.
In the second step, clustering is performed based on dissimilarities. 

```{r hclust2}
if( !require(NbClust) ){install.packages("NbClust"); library(NbClust)}
if( !require(cobiclust) ){install.packages("cobiclust"); library(cobiclust)}

# Apply transformation
tse <- transformCounts(tse, method = "relabundance")
# Get the assay
assay <- assay(tse, "relabundance")
# Transpose assay --> samples are now in rows --> we are clustering samples
assay <- t(assay)

# Calculate distances
diss <- vegdist(assay, method = "bray")

# Perform hierarchical clustering
hc <- hclust(diss, method = "complete")

# To visualize, convert hclust object into dendrogram object
dendro <- as.dendrogram(hc)

# Plot dendrogram
plot(dendro)
```

We can use dendrogram to determine the number of clusters. Usually the
tree is splitted where the branch length is the largest. However, as
we can see from the dendrogram, clusters are not clear. Algorithms are
available to identify the optimal number of clusters.

```{r hclust3}
# Determine the optimal number of clusters
res <- NbClust(diss = diss, distance = NULL, method = "ward.D2",
               index = "silhouette")

res$Best.nc
```

Based on the result, let's divide observations into 15 clusters.

```{r hclust4}
if( !require(dendextend) ){
    install.packages("dendextend")
    library(dendextend)
}

# Find clusters
cutree(hc, k = 15) 

# Making colors for 6 clusters
col_val_map <- randomcoloR::distinctColorPalette(15) %>%
     as.list() %>% setNames(paste0("clust_",seq(15)))

dend <- color_branches(dendro, k=15, col=unlist(col_val_map))
labels(dend) <- NULL
plot(dend)

```

## K-means clustering

Hierarchical clustering did not yield clusters. Let's try k-means
clustering instead. Here observations are divided into clusters so
that the distances between observations and cluster centers are
minimized; an observation belongs to cluster whose center is the
nearest.

The algorithm starts by dividing observation to random clusters whose
number is defined by user. The centroids of clusters are then
calculated. After that, observations' allocation to clusters are
updated so that the means are minimized. Again, centroid are
calculated, and algorithm continues iteratively until the assignments
do not change.

The number of clusters can be determined based on algorithm. Here we
utilize silhouette analysis.

```{r kmeans1}
if( !require(factoextra) ){
    install.packages("factoextra")
    library(factoextra)
}

# Convert dist object into matrix
diss <- as.matrix(diss)
# Perform silhouette analysis and plot the result
fviz_nbclust(diss, kmeans, method = "silhouette")
```

Based on the result of silhouette analysis, we choose 3 to be the number of clusters
in k-means clustering.

```{r kmeans2}
library(scater)

# The first step is random, add seed for reproducibility
set.seed(15463)
# Perform k-means clustering with 3 clusters
km <- kmeans(diss, 3, nstart = 25)
# Add the result to colData
colData(tse)$clusters <- as.factor(km$cluster)

# Perform PCoA so that we can visualize clusters
tse <- runMDS(tse, assay.type = "relabundance", FUN = vegan::vegdist, method = "bray")

# Plot PCoA and color clusters
plotReducedDim(tse, "MDS", colour_by = "clusters")
```

## Dirichlet Multinomial Mixtures (DMM)

This section focus on DMM analysis. 

One technique that allows to search for groups of samples that are
similar to each other is the [Dirichlet-Multinomial Mixture
Model](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0030126). In
DMM, we first determine the number of clusters (k) that best fit the
data (model evidence) using Laplace approximation. After fitting the
model with k clusters, we obtain for each sample k probabilities that
reflect the probability that a sample belongs to the given cluster.

Let's cluster the data with DMM clustering. 

```{r dmm1}
# Runs model and calculates the most likely number of clusters from 1 to 7.
# Since this is a large dataset it takes long computational time.
# For this reason we use only a subset of the data; agglomerated by Phylum as a rank.
tse <- GlobalPatterns
tse <- agglomerateByRank(tse, rank = "Phylum", agglomerateTree=TRUE)
```

```{r dmm2}
tse_dmn <- mia::runDMN(tse, name = "DMN", k = 1:7)
```

```{r}
# It is stored in metadata
tse_dmn
```

Return information on metadata that the object contains.

```{r}
names(metadata(tse_dmn))
```

This returns a list of DMN objects for a closer investigation.

```{r}
getDMN(tse_dmn)
```


Show Laplace approximation (model evidence) for each model of the k models.

```{r}
library(miaViz)
plotDMNFit(tse_dmn, type = "laplace")
```

Return the model that has the best fit.

```{r}
getBestDMNFit(tse_dmn, type = "laplace")
```

### PCoA for ASV-level data with Bray-Curtis; with DMM clusters shown with colors

Group samples and return DMNGroup object that contains a summary.
Patient status is used for grouping.

```{r}
dmn_group <- calculateDMNgroup(tse_dmn, variable = "SampleType",  assay.type = "counts",
                               k = 2, seed=.Machine$integer.max)

dmn_group
```

Mixture weights  (rough measure of the cluster size).


```{r}
DirichletMultinomial::mixturewt(getBestDMNFit(tse_dmn))
```


Samples-cluster assignment probabilities / how probable it is that sample belongs
to each cluster

```{r}
head(DirichletMultinomial::mixture(getBestDMNFit(tse_dmn)))
```

Contribution of each taxa to each component

```{r}
head(DirichletMultinomial::fitted(getBestDMNFit(tse_dmn)))
```
Get the assignment probabilities


```{r}
prob <- DirichletMultinomial::mixture(getBestDMNFit(tse_dmn))
# Add column names
colnames(prob) <- c("comp1", "comp2")

# For each row, finds column that has the highest value. Then extract the column 
# names of highest values.
vec <- colnames(prob)[max.col(prob,ties.method = "first")]

```

Computing the euclidean PCoA and storing it as a data frame

```{r}
# Does clr transformation. Pseudocount is added, because data contains zeros.
assay(tse, "pseudo") <- assay(tse, "counts") + 1
tse <- transformCounts(tse, assay.type = "pseudo", method = "relabundance")
tse <- transformCounts(tse, "relabundance", method = "clr")

library(scater)

# Does principal coordinate analysis
df <- calculateMDS(tse, assay.type = "clr", method = "euclidean")

# Creates a data frame from principal coordinates
euclidean_pcoa_df <- data.frame(pcoa1 = df[,1], 
                                pcoa2 = df[,2])

```

```{r}
# Creates a data frame that contains principal coordinates and DMM information
euclidean_dmm_pcoa_df <- cbind(euclidean_pcoa_df,
                               dmm_component = vec)
# Creates a plot
euclidean_dmm_plot <- ggplot(data = euclidean_dmm_pcoa_df, 
                             aes(x=pcoa1, y=pcoa2,
                                 color = dmm_component)) +
  geom_point() +
  labs(x = "Coordinate 1",
       y = "Coordinate 2",
       title = "PCoA with Aitchison distances") +  
  theme(title = element_text(size = 12)) # makes titles smaller

euclidean_dmm_plot
```

## Community Detection

Another approach for discovering communities within the samples of the
data, is to run community detection algorithms after building a
graph. The following demonstration builds a graph based on the k
nearest-neighbors and performs the community detection on the fly.

_`bluster`_ [@R-bluster] package offers several clustering methods,
among which graph-based are present, enabling the community detection
task.

Installing package:

```{r}
if(!require(bluster)){
  BiocManager::install("bluster")
}
```

The algorithm used is "short random walks" [@Pons2006]. Graph is
constructed using different k values (the number of nearest neighbors
to consider during graph construction) using the robust centered log
ratio (rclr) assay data. Then plotting the communities using UMAP
[@McInnes2018] ordination as a visual exploration aid.  In the
following demonstration we use the `enterotype` dataset from the
[@R-mia] package.

```{r, message=FALSE, warning=FALSE}
library(bluster)
library(patchwork) # For arranging several plots as a grid
library(scater)

data("enterotype", package="mia")
tse <- enterotype
tse <- transformCounts(tse, method = "rclr")

# Performing and storing UMAP
tse <- runUMAP(tse, name="UMAP", assay.type="rclr")

k <- c(2,3,5,10)
ClustAndPlot <- function(x) {
  # Creating the graph and running the short random walks algorithm  
  graph_clusters <- clusterRows(t(assays(tse)$rclr), NNGraphParam(k=x))
  
  # Results of the clustering as a color for each sample
  plotUMAP(tse, colour_by = I(graph_clusters)) +
    labs(title = paste0("k = ", x))
}

# Applying the function for different k values
plots <- lapply(k,ClustAndPlot)

# Displaying plots in a grid
(plots[[1]] + plots[[2]]) / (plots[[3]] + plots[[4]])
```

Similarly, the _`bluster`_ [@R-bluster] package offers clustering
diagnostics that can be used for judging the clustering quality (see
[Assorted clustering
diagnostics](http://bioconductor.org/packages/release/bioc/vignettes/bluster/inst/doc/diagnostics.html)).
In the following, Silhouette width as a diagnostic tool is computed
and results are visualized for each case presented earlier. For more
about Silhouettes read [@Rousseeuw1987].

```{r, message=FALSE, warning=FALSE}

ClustDiagPlot <- function(x) {
  # Getting the clustering results
  graph_clusters <- clusterRows(t(assays(tse)$rclr), NNGraphParam(k=x))
  
  # Computing the diagnostic info
  sil <- approxSilhouette(t(assays(tse)$rclr), graph_clusters)
  
  # Plotting as a boxlpot to observe cluster separation
  boxplot(split(sil$width, graph_clusters), main=paste0("k = ", x))
  
}
# Applying the function for different k values
res <- lapply(k,ClustDiagPlot)
```

## Biclustering

Biclustering methods cluster rows and columns simultaneously in order
to find subsets of correlated features/samples.

Here, we use following packages:

-   [biclust](https://cran.r-project.org/web/packages/biclust/index.html)
-   [cobiclust](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.13582)

_cobiclust_ is especially developed for microbiome data whereas _biclust_ is more
general method. In this section, we show three different cases and example 
solutions to apply biclustering to them. 

1.   Taxa vs samples
2.   Taxa vs biomolecule/biomarker
3.   Taxa vs taxa

Biclusters can be visualized using heatmap or boxplot, for
instance. For checking purposes, also scatter plot might be valid
choice.

Check more ideas for heatmaps from chapters \@ref(viz-chapter) and
\@ref(microbiome-community.

### Taxa vs samples

When you have microbial abundance matrices, we suggest to use
_cobiclust_ which is designed for microbial data.

Load example data
```{r load-pkg-data2}
library(mia)
data("HintikkaXOData")
mae <- HintikkaXOData
```

Only the most prevalent taxa are included in analysis. 

```{r cobiclust_1}
# Subset data in the first experiment
mae[[1]] <- subsetByPrevalentTaxa(mae[[1]], rank = "Genus", prevalence = 0.2, detection = 0.001)
# clr-transform in the first experiment
mae[[1]] <- transformCounts(mae[[1]], method = "relabundance")
mae[[1]] <- transformCounts(mae[[1]], "relabundance", method = "rclr")
```

_cobiclust_ takes counts table as an input and gives _cobiclust_ object as an output.
It includes clusters for taxa and samples. 

```{r cobiclust_2}
# Do clustering; use counts table´
clusters <- cobiclust(assay(mae[[1]], "counts"))

# Get clusters
row_clusters <- clusters$classification$rowclass
col_clusters <- clusters$classification$colclass

# Add clusters to rowdata and coldata
rowData(mae[[1]])$clusters <- factor(row_clusters)
colData(mae[[1]])$clusters <- factor(col_clusters)

# Order data based on clusters
mae[[1]] <- mae[[1]][order(rowData(mae[[1]])$clusters), order(colData(mae[[1]])$clusters)]

# Print clusters
clusters$classification
```

Next we can plot clusters. Annotated heatmap is a common choice.

```{r cobiclust_3, fig.width=14, fig.height=12}

if(!require(pheatmap)){
    install.packages("pheatmap")
    library(pheatmap)
}
# z-transform for heatmap
mae[[1]] <- transformCounts(mae[[1]], assay.type = "rclr",
                            MARGIN = "features",
                            method = "z", name = "clr_z")

# Create annotations. When column names are equal, they should share levels. 
# Here samples include 3 clusters, and taxa 2. That is why we have to make 
# column names unique. 
annotation_col <- data.frame(colData(mae[[1]])[, "clusters", drop = F])
colnames(annotation_col) <- "col_clusters"

annotation_row <- data.frame(rowData(mae[[1]])[, "clusters", drop = F])
colnames(annotation_row) <- "row_clusters"
```


Plot the heatmap.

```{r cobiclust_3b, fig.width=14, fig.height=12}
pheatmap(assay(mae[[1]], "clr_z"), cluster_rows = F, cluster_cols = F, 
         annotation_col = annotation_col,
         annotation_row = annotation_row)
```


Boxplot is commonly used to summarize the results:

```{r cobiclust_4}
if(!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}
if(!require(patchwork)){
    install.packages("patchwork")
    library(patchwork)
}

# ggplot requires data in melted format
melt_assay <- meltAssay(mae[[1]], assay.type = "rclr", add_col_data = T, add_row_data = T)

# patchwork two plots side-by-side
p1 <- ggplot(melt_assay) +
  geom_boxplot(aes(x = clusters.x, y = rclr)) +
  labs(x = "Taxa clusters")

p2 <- ggplot(melt_assay) +
  geom_boxplot(aes(x = clusters.y, y = rclr)) +
  labs(x = "Sample clusters")

p1 + p2
```

### Taxa vs biomolecules

Here, we analyze cross-correlation between taxa and metabolites. This
is a case, where we use _biclust_ method which is suitable for numeric
matrices in general.

```{r biclust_1}
# Samples must be in equal order 
# (Only 1st experiment  was ordered in cobiclust step leading to unequal order)
mae[[1]] <- mae[[1]][ , colnames(mae[[2]]) ]

# Make rownames unique since it is require by other steps
rownames(mae[[1]]) <- make.unique(rownames(mae[[1]]))
# Calculate correlations
corr <- getExperimentCrossCorrelation(mae, 1, 2, 
                                      assay.type1 = "rclr", 
                                      assay.type2 = "nmr", 
                                      mode = "matrix", 
                                      cor_threshold = 0.2)

```


_biclust_ takes matrix as an input and returns _biclust_ object. 

```{r biclust_2}
# Load package
if(!require(biclust)){
    install.packages("biclust")
    library(biclust)
}

# Set seed for reproducibility
set.seed(3973)

# Find biclusters
bc <- biclust(corr, method=BCPlaid(), fit.model = y ~ m,
              background = TRUE, shuffle = 100, back.fit = 0, max.layers = 10,
              iter.startup = 10, iter.layer = 100, verbose = FALSE)

bc
```

The object includes cluster information. However compared to
_cobiclust_, _biclust_ object includes only information about clusters
that were found, not general cluster.

Meaning that if one cluster size of 5 features was found out of 20 features, 
those 15 features do not belong to any cluster. That is why we have to create an
additional cluster for features/samples that are not assigned into any cluster.

```{r biclust_3}
# Functions for obtaining biclust information

# Get clusters for rows and columns
.get_biclusters_from_biclust <- function(bc, assay){
  # Get cluster information for columns and rows
  bc_columns <- t(bc@NumberxCol)
  bc_columns <- data.frame(bc_columns)
  bc_rows <- bc@RowxNumber
  bc_rows <- data.frame(bc_rows)
  
  # Get data into right format
  bc_columns <- .manipulate_bc_data(bc_columns, assay, "col")
  bc_rows <- .manipulate_bc_data(bc_rows, assay, "row")
  
  return(list(bc_columns = bc_columns, bc_rows = bc_rows))
}

# Input clusters, and how many observations there should be, i.e.,
# the number of samples or features
.manipulate_bc_data <- function(bc_clusters, assay, row_col){
  # Get right dimension
  dim <- ifelse(row_col == "col", ncol(assay), nrow(assay))
  # Get column/row names
  if( row_col == "col" ){
    names <- colnames(assay)
  } else{
    names <- rownames(assay)
  }
  
  # If no clusters were found, create one. Otherwise create additional
  # cluster which
  # contain those samples that are not included in clusters that were found.
  if( nrow(bc_clusters) != dim ){
      bc_clusters <- data.frame(cluster = rep(TRUE, dim))
  } else {
      # Create additional cluster that includes those samples/features that
      # are not included in other clusters.
      vec <- ifelse(rowSums(bc_clusters) > 0, FALSE, TRUE)
      # If additional cluster contains samples, then add it
      if ( any(vec) ){
          bc_clusters <- cbind(bc_clusters, vec)
      }
  }
  # Adjust row and column names
  rownames(bc_clusters) <- names
  colnames(bc_clusters) <- paste0("cluster_", 1:ncol(bc_clusters))
  return(bc_clusters)
}
```


```{r biclust_4}
# Get biclusters
bcs <- .get_biclusters_from_biclust(bc, corr)

bicluster_rows <- bcs$bc_rows
bicluster_columns <- bcs$bc_columns

# Print biclusters for rows
head(bicluster_rows)
```

Let's collect information for the scatter plot. 

```{r biclust_5}
# Function for obtaining sample-wise sum, mean, median, and mean variance
# for each cluster

.sum_mean_median_var <- function(tse1, tse2, assay.type1, assay.type2, clusters1, clusters2){
  
  list <- list()
  # Create a data frame that includes all the information
  for(i in 1:ncol(clusters1) ){
    # Subset data based on cluster
    tse_subset1 <- tse1[clusters1[,i], ]
    tse_subset2 <- tse2[clusters2[,i], ]
    # Get assay
    assay1 <- assay(tse_subset1, assay.type1)
    assay2 <- assay(tse_subset2, assay.type2)
    # Calculate sum, mean, median, and mean variance
    sum1 <- colSums2(assay1, na.rm = T)
    mean1 <- colMeans2(assay1, na.rm = T)
    median1 <- colMedians(assay1, na.rm = T)
    var1 <- colVars(assay1, na.rm = T)
    
    sum2 <- colSums2(assay2, na.rm = T)
    mean2 <- colMeans2(assay2, na.rm = T)
    median2 <- colMedians(assay2, na.rm = T)
    var2 <- colVars(assay2, na.rm = T)
    
    list[[i]] <- data.frame(sample = colnames(tse1), sum1, sum2, mean1, mean2, 
                     median1, median2, var1, var2)
  }

  return(list)
}

# Calculate info
df <- .sum_mean_median_var(mae[[1]], mae[[2]], "rclr", "nmr", bicluster_rows, bicluster_columns)
```

Now we can create a scatter plot. X-axis includes median clr abundance
of microbiome and y-axis median absolute concentration of each
metabolite. Each data point represents a single sample.

From the plots, we can see that there is low negative correlation in
both cluster 1 and 3.  This means that when abundance of bacteria
belonging to cluster 1 or 3 is higher, the concentration of
metabolites of cluster 1 or 3 is lower, and vice versa.

```{r biclust_6, fig.width=14, fig.height=6, fig.show="keep", out.width="33%"}
pics <- list()
for(i in seq_along(df)){
  pics[[i]] <- ggplot(df[[i]])  +
      geom_point(aes(x = median1, y = median2)) + 
      labs(title = paste0("Cluster ", i),
           x = "Taxa (rclr median)",
           y = "Metabolites (abs. median)")
  print(pics[[i]])
}
# pics[[1]] + pics[[2]] + pics[[3]]
```

_pheatmap_ does not allow boolean values, so they must be converted into factors.

```{r biclust_7}
bicluster_columns <- data.frame(apply(bicluster_columns, 2, as.factor))
bicluster_rows <- data.frame(apply(bicluster_rows, 2, as.factor))
```

Again, we can plot clusters with heatmap.

```{r biclust_8, fig.width=10, fig.height=10}
# Adjust colors for all clusters
if( ncol(bicluster_rows) > ncol(bicluster_columns) ){
  cluster_names <- colnames(bicluster_rows)
} else {
  cluster_names <- colnames(bicluster_columns)
}
annotation_colors <- list()
for(name in cluster_names){
  annotation_colors[[name]] <- c("TRUE" = "red", "FALSE" = "white")
}

# Create a heatmap
pheatmap(corr, cluster_cols = F, cluster_rows = F,
         annotation_col = bicluster_columns, 
         annotation_row = bicluster_rows,
         annotation_colors = annotation_colors)
```

### Taxa vs taxa

Third and final example deals with situation where we want to analyze
correlation between taxa. _biclust_ is suitable for this.

```{r biclust_9}
# Calculate cross-correlation
corr <- getExperimentCrossCorrelation(mae, 1, 1, 
                                      assay.type1 = "rclr", assay.type2 = "rclr", 
                                      mode = "matrix",
                                      cor_threshold = 0.2, verbose = F, show_warning = F)

# Find biclusters
bc <- biclust(corr, method=BCPlaid(), fit.model = y ~ m,
              background = TRUE, shuffle = 100, back.fit = 0, max.layers = 10,
              iter.startup = 10, iter.layer = 100, verbose = FALSE)
```

```{r biclust_10}
# Get biclusters
bcs <- .get_biclusters_from_biclust(bc, corr)

bicluster_rows <- bcs$bc_rows
bicluster_columns <- bcs$bc_columns
```


```{r biclust_11}
# Create a column that combines information
# If row/column includes in multiple clusters, cluster numbers are separated with "_&_"
bicluster_columns$clusters <- apply(bicluster_columns, 1, 
                                    function(x){paste(paste(which(x)), collapse = "_&_") })
bicluster_columns <- bicluster_columns[, "clusters", drop = FALSE]

bicluster_rows$clusters <- apply(bicluster_rows, 1, 
                                 function(x){paste(paste(which(x)), collapse = "_&_") })
bicluster_rows <- bicluster_rows[, "clusters", drop = FALSE]
```

```{r biclust_12, fig.width=14, fig.height=12}
# Convert boolean values into factor
bicluster_columns <- data.frame(apply(bicluster_columns, 2, as.factor))
bicluster_rows <- data.frame(apply(bicluster_rows, 2, as.factor))

pheatmap(corr, cluster_cols = F, cluster_rows = F,
         annotation_col = bicluster_columns, 
         annotation_row = bicluster_rows)
```

## Additional Community Typing

For more community typing techniques applied to the 'SprockettTHData'
data set, see the attached .Rmd file.

Link:

   * [Rmd](add-comm-typing.Rmd)
