---
title: ""
output: html_notebook
---

# Visualization

```{r setup, echo=FALSE, results="asis"}
#library(rebook)
#chapterPreamble()
```

Whether a data set contains information on a microbial community or it originates from a different source, the way that data are visualized inevitably shapes how they will be interpreted, and motivates the next steps of the analysis.

A large variety of graphing methods belong to microbial analysis, but only few are the choices that will return useful answers about the data. Therefore, knowledge on the available tools and their possible applications plays an important role in selecting the most suitable method for the asked question.

This chapter introduces the reader to a number of visualization techniques found in this book, such as:

* bar plots
* box plots
* heat maps
* ordination charts
* regression charts
* trees

The toolkit which provides the essential plotting functionality includes the following packages:

* `patchwork`, `cowplot`, `ggpubr` and `gridExtra`: plot layout and multi-panel plotting
* `miaViz`: specific visualization tools for `TreeSummaizedExperiment` objects
* `scater`: specific visualization tools for `SingleCellExperiment` objects
* `ggplot2`, `pheatmap`, `ggtree` and `sechm`: composition heat maps
* `ANCOMBC`, `ALDEx2` and `map2`: visual differential abundance
* `fido` and `pibble`: tree-based methods for differential abundance
* `plotly`: animated and 3D plotting

For systematic and extensive tutorials on the visual tools available in `phyloseq` and `mia`, readers can refer to the following material:

* [phyloseq tutorials](http://joey711.github.io/phyloseq/tutorials-index)
* [microbiome tutorials](https://microbiome.github.io/tutorials/)

```{r, include = FALSE}
library(ggplot2)
theme_set(theme_classic())
library(mia)
library(scater)
library(patchwork)
# essential data
data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns
# additional data for pre-analysis exploration
tse <- addPerCellQC(tse)
tse <- addPerFeatureQC(tse)
# additional data for alpha diversity
tse <- mia::estimateDiversity(tse, 
                              abund_values = "counts",
                              index = "shannon", 
                              name = "shannon")
tse <- mia::estimateFaith(tse,
                          abund_values = "counts")
# additional data for beta diversity
tse <- runMDS(tse,
              FUN = vegan::vegdist,
              method = "bray",
              name = "MDS",
              exprs_values = "counts",
              ncomponents = 3)
tse <- runNMDS(tse,
               FUN = vegan::vegdist,
               name = "NMDS")
```

## Pre-analysis exploration

### Accessing row and column data

`SCE` and `TSE` objects contain multiple layers of information in the form of rows, columns and meta data. The `scater` package supports in accessing, modifying and graphing the meta data related to features as well as samples.

```{r}
# list row meta data
names(rowData(tse))
# list column meta data
names(colData(tse))
```

Such meta data can be directly plotted with the functions `plotRowData` and `plotColData`.

```{r, warning = FALSE}
# plot QC Mean against Species
plotRowData(tse, "mean", "Species") +
  theme(axis.text.x = element_blank()) +
  labs(x = "Species", y = "QC Mean")
# plot QC Sum against Sample ID, colour-labeled by Sample Type
plotColData(tse, "sum", "X.SampleID", colour_by = "SampleType") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Sample ID", y = "QC Sum")
```

Alternatively, they can be converted to a `data.frame` object and passed to `ggplot`.

```{r}
# store colData into a data frame
coldata <- as.data.frame(colData(tse))
# plot Number of Samples against Sampling Site
ggplot(coldata, aes(x = SampleType)) +
  geom_bar(width = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Sampling Site",
       y = "Number of Samples")
```

Further methods of application can be found in the chapters \@ref(qc) and \@ref(richness) and in a few [external tutorials](https://github.com/davismcc/scater_tutorials_open_data) with open data. Additionally, `rowData` and `colData` allow manipulation and subsetting of large data sets into smaller units, as explained in chapter \@ref(datamanipulation).

### Viewing abundance and prevalence patterns

Prior-to-analysis exploration may involve questions such as how microorganisms are distributed across samples (abundance) and what microorganisms are present in most of the samples (prevalence). The information on abundance and prevalence can be summarized into a **jitter** or **density plot** and a **tree**, respectively, with the `miaViz` package.

Specifically, the functions `plotAbundance`, `plotAbundanceDensity` and `plotRowTree` are used, and examples on their usage are discussed throughout chapter \@ref(quality-control).

## Diversity estimation

Alpha diversity is commonly measured as one of the diversity indices explained in chapter \@ref(community-diversity). Because the focus lies on each sample separately, one-dimensional plots, such as **scatter**, **violin** and **box plots**, are suitable. 

Beta diversity is generally evaluated as one of the dissimilarity indices reported in chapter \@ref(community-similarity). Unlike alpha diversity, samples are compared collectively to estimate the heterogeneity across them, therefore multidimensional plots, such as **Shepard** and **ordination plots** are suitable.

|                         | alpha diversity            | beta diversity            |
|:-----------------------:|:--------------------------:|:-------------------------:|
| used metrics            | diversity indices          | dissimilarity indices     |
|                         |                            |                           | 
| metric dimensionality   | one-dimensional            | multidimensional          |
|                         |                            |                           | 
| suitable visualization  | scatter, violin, box plots | Shepard, ordination plots |

As a conclusion, visualization techniques for alpha and beta diversity significantly differ from one another.

### Alpha diversity with scatter, violin and box plots

The basic method to visualize the diversity values assigned to the different samples in a `TSE` object includes the following, where each data point represents one sample:

```{r}
# plot shannon diversity index, colour-labeled by Sample Type
plotColData(tse, "shannon", colour_by = "SampleType")
```

The several indices available for the evaluation of alpha diversity often return slightly divergent results, which can be visually compared with a multiple violin or box plot. For this purpose, `plotColData` (for violin plots) or `ggplot` (for box plots) are recursively applied to a number of diversity indices with the function `lapply` and the multi-panel plotting functionality of the `patchwork` package is then exploited.

```{r}
# store colData into a data frame
coldata <- as.data.frame(colData(tse))
# generate plots for shannon and faith indices
# and store them into a list
plots <- lapply(c("shannon", "faith"),
                function(i) ggplot(coldata, aes_string(y = i)) +
                  geom_boxplot() +
                  theme(axis.text.x = element_blank(),
                        axis.ticks.x = element_blank()))
# combine plots with patchwork
plots[[1]] + plots[[2]]
```
The analogous output in the form of a violin plot is obtained in chapter \@ref(faith-diversity). In addition, box plots that group samples according to certain information, such as origin, sex, age and health condition, can be labeled with p-values for significant differences with the package `ggsignif` package, as shown in chapter \@ref(estimate-diversity).

### Beta diversity with Shepard and coordination plots

The `scater` package offers the general function `plotReducedDim`, which is able to produce plots from the following coordination methods:

* `runMDS`
* `runNMDS`
* `runPCA`
* `runTSNE`
* `runUMAP`

```{r}
p1 <- plotReducedDim(tse, "NMDS", colour_by = "SampleType")
```

Multiple combinations of coordinates or dimensions can also be integrated into a multi-panel arrangement.

```{r}
p2 <- plotReducedDim(tse, "MDS", ncomponents = c(1:3), colour_by = "SampleType")
```

Similarly to iterating `plotColData` over indices of alpha diversity, `lapply` and `patchwork` can be used in combination to recursively apply `plotReducedDim` and visually compare results among various coordination methods.

```{r}
plots <- lapply(c("MDS", "NMDS"),
                plotReducedDim,
                object = tse)
plots[[1]] + plots[[2]]
```

## Statistical analysis

### community composition, biclustering and multi-assay analysis with heat maps

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
#prettySessionInfo()
```