# Community Similarity {#sec-community-similarity}

```{r setup, echo = FALSE, results = "asis"}
library(rebook)
chapterPreamble()
```

```{r include = FALSE}
# global knitr options
knitr::opts_chunk$set(
  message = FALSE,
  fig.width = 10,
  dpi = 300,
  dev = "png",
  dev.args = list(type = "cairo-png")
)
```
Beta diversity quantifies the dissimilarity between communities (multiple samples), as opposed to alpha diversity which focuses on variation within a community (one sample). In microbiome research, commonly used metrics of beta diversity include:

- Bray-Curtis index (for compositional data)
- Jaccard index (for presence/absence data, ignoring abundance information)
- Aitchison distance (Euclidean distance for clr transformed abundances, aiming to avoid the compositionality bias)
- Unifrac distance (takes into account the phylogenetic tree information).

Notably, only some of these measures are actual
_distances_, as this is a mathematical concept whose definition is not satisfied
by certain ecological measure, such as the Bray-Curtis index. Therefore, the terms
dissimilarity and beta diversity are preferred.

| Method description          | Assay type          | Beta diversity metric |
|:---------------------------:|:-------------------:|:---------------------:|
| Quantitative profiling      | Absolute counts     | Bray-Curtis           | 
| Relative profiling          | Relative abundances | Bray-Curtis           |
| Aitchison distance          | Absolute counts     | Aitchison             |
| Aitchison distance          | clr                 | Euclidean             |
| Robust Aitchison distance   | rclr                | Euclidean             |
| Presence/Absence similarity | Relative abundances | Jaccard               |
| Presence/Absence similarity | Absolute counts     | Jaccard               |
| Phylogenetic distance       | Rarefied counts     | Unifrac               |

In practice, beta diversity is usually represented as a `dist` object. Such an 
object is a triangular matrix where the distance between each pair of samples is encoded by a specific cell. 
This distance matrix can then undergo ordination, which is an
important ecological tool to reduce the dimensionality of data for a more
efficient analysis and visualization. Ordination techniques aim to capture as
much essential information from the data as possible and turn it into a lower
dimensional representation. Dimension reduction is bound to lose information but
commonly used ordination techniques can preserve relevant information of sample
similarities in an optimal way, which is defined in different ways by different
methods.

Based on the type of algorithm, ordination methods in microbiome research can
be generally divided in two categories: 

- unsupervised ordination
- supervised ordination

The former includes Principal Coordinate Analysis (PCoA), Principal Component
Analysis (PCA) and Uniform Manifold Approximation and Projection for Dimension
Reduction (UMAP), whereas the latter is mainly represented by distance-based
Redundancy Analysis (dbRDA). We will first discuss unsupervised ordination
methods and then proceed to supervised ones.

Let us now prepare some demonstration data for the practical examples.

```{r prep-tse}
# Load mia and import sample dataset
library(mia)
data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns

# Beta diversity metrics like Bray-Curtis are often 
# applied to relabundances
tse <- transformAssay(tse,
                      assay.type = "counts",
                      method = "relabundance")

# Other metrics like Aitchison to clr-transformed data
tse <- transformAssay(tse,
                      assay.type = "relabundance",
                      method = "clr",
                      pseudocount = TRUE)

# Add group information Feces yes/no
tse$Group <- tse$SampleType == "Feces"
```



## Divergence

Divergence measure refers to a difference in community composition
between the given sample(s) and a reference sample. This can be
evaluated with `addDivergence`. Reference and algorithm for the
calculation of divergence can be specified as `reference` and `FUN`,
respectively.

```{r}
tse <- mia::addDivergence(tse,
                               assay.type = "counts",
                               reference = "median",
                               FUN = vegan::vegdist)
```


## Unsupervised ordination {#sec-unsupervised-ordination}

Unsupervised ordination methods variation in the data without additional
information on covariates or other supervision of the model. Among the different
approaches, Multi-Dimensional Scaling (MDS) and non-metric MDS (NMDS) can be
regarded as the standard. They are jointly referred to as PCoA. For this
demonstration we will analyse beta diversity in GlobalPatterns, and observe the
variation between stool samples and those with a different origin.


### Comparing communities by beta diversity analysis

A typical comparison of community compositions starts with a visual
representation of the groups by a 2D ordination. Then we estimate relative
abundances and MDS ordination based on Bray-Curtis index between the groups,
and visualize the results.

In the following examples dissimilarity is calculated with the function supplied
to the `FUN` argument. Several metrics of beta diversity are defined by the `vegdist`
function of the vegan package, which is often used in this context. However, such
custom functions created by the user also work, as long as they return a `dist`
object. In either case, this function is then applied to calculate reduced
dimensions via an ordination method, the results of which can be stored in the
`reducedDim` slot of the TreeSE. This entire process is contained by the `runMDS`
and `runNMDS` functions.

```{r runMDS}
# Load package to plot reducedDim
library(scater)

# Run PCoA on relabundance assay with Bray-Curtis distances
tse <- runMDS(tse,
              FUN = vegan::vegdist,
              method = "bray",
              assay.type = "relabundance",
              name = "MDS_bray")
```

Sample dissimilarity can be visualized on a lower-dimensional display (typically
2D) using the `plotReducedDim` function from the scater package. This also
provides tools to incorporate additional information encoded by color, shape,
size and other aesthetics. Can you find any difference between the groups?

```{r plot-mds-bray-curtis, fig.cap = "MDS plot based on the Bray-Curtis distances on the GlobalPattern dataset."}
# Create ggplot object
p <- plotReducedDim(tse, "MDS_bray",
                    colour_by = "Group")

# Calculate explained variance
e <- attr(reducedDim(tse, "MDS_bray"), "eig")
rel_eig <- e / sum(e[e > 0])

# Add explained variance for each axis
p <- p + labs(x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1),
                        "%", ")", sep = ""),
              y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), 
                        "%", ")", sep = ""))

p
```

A few combinations of beta diversity metrics and assay types are typically
used. For instance, Bray-Curtis dissimilarity and Euclidean distance are often
applied to the relative abundance and the clr assays, respectively. Besides
**beta diversity metric** and **assay type**, the **PCoA algorithm** is also a
variable that should be considered. Below, we show how the choice of these three
factors can affect the resulting lower-dimensional data.

```{r mds-nmds-comparison, results='hide'}
# Run NMDS on relabundance assay with Bray-Curtis distances
tse <- runNMDS(tse,
               FUN = vegan::vegdist,
               method = "bray",
               assay.type = "relabundance",
               name = "NMDS_bray")

# Run MDS on clr assay with Aitchison distances
tse <- runMDS(tse,
              FUN = vegan::vegdist,
              method = "euclidean",
              assay.type = "clr",
              name = "MDS_aitchison")

# Run NMDS on clr assay with Euclidean distances
tse <- runNMDS(tse,
               FUN = vegan::vegdist,
               method = "euclidean",
               assay.type = "clr",
               name = "NMDS_aitchison")
```

Multiple ordination plots are combined into a multi-panel plot with the
patchwork package, so that different methods can be compared to find similarities
between them or select the most suitable one to visualize beta diversity in the
light of the research question.

```{r, fig.cap = "Comparison of MDS and NMDS plots based on the Bray-Curtis or Aitchison distances on the GlobalPattern dataset."}
# Load package for multi-panel plotting
library(patchwork)

# Generate plots for all 4 reducedDims
plots <- lapply(c("MDS_bray", "MDS_aitchison",
                  "NMDS_bray", "NMDS_aitchison"),
                plotReducedDim,
                object = tse,
                colour_by = "Group")

# Generate multi-panel plot
wrap_plots(plots) +
  plot_layout(guides = "collect")
```

The _Unifrac_ method is a special case, as it requires data on the
relationship of features in form on a `phylo` tree. `calculateUnifrac`
performs the calculation to return a `dist` object, which can again be
used within `runMDS`.

```{r plot-unifrac, fig.cap = "Unifrac distances scaled by MDS of the GlobalPattern dataset."}
tse <- runMDS(tse,
              FUN = mia::calculateUnifrac,
              name = "Unifrac",
              tree = rowTree(tse),
              ntop = nrow(tse),
              assay.type = "counts")

plotReducedDim(tse, "Unifrac",
               colour_by = "Group")
```

### Other ordination methods {#sec-other-ord-methods}

Other dimension reduction methods, such as PCA and UMAP, are inherited from the
scater package.

```{r plot-pca, fig.cap = "PCA plot on the GlobalPatterns data set containing sample from different sources."}
tse <- runPCA(tse,
              name = "PCA",
              assay.type = "counts",
              ncomponents = 10)

plotReducedDim(tse, "PCA",
               colour_by = "Group")
```

As mentioned before, applicability of the different methods depends on your
sample set and research question.

```{r plot-umap, fig.cap = "UMAP plot on the GlobalPatterns data set containing sample from different sources."}
tse <- runUMAP(tse,
               name = "UMAP",
               assay.type = "counts",
               ncomponents = 3)

plotReducedDim(tse, "UMAP",
               colour_by = "Group",
               ncomponents = c(1:3))
```

### Explained variance

The percentage of explained variance is typically shown for PCA
ordination plots. This quantifies the proportion of overall variance
in the data that is captured by the PCA axes, or how well the
ordination axes reflect the original distances.

Sometimes a similar measure is shown for MDS/PCoA. The interpretation
is generally different, however, and hence we do not recommend using
it. PCA is a special case of PCoA with Euclidean distances.  With
non-Euclidean dissimilarities PCoA uses a trick where the pointwise
dissimilarities are first cast into similarities in a Euclidean space
(with some information loss i.e. stress) and then projected to the
maximal variance axes. In this case, the maximal variance axes do not
directly reflect the correspondence of the projected distances and
original distances, as they do for PCA.

In typical use cases, we would like to know how well the ordination
reflects the original similarity structures; then the quantity of
interest is the so-called "stress" function, which measures the
difference in pairwise similarities between the data points in the
original (high-dimensional) vs. projected (low-dimensional) space.

Hence, we propose that for PCoA and other ordination methods, users
would report relative stress, which varies in the unit interval and is better
if smaller. This can be calculated as shown below.

```{r relstress}
# Load vegan package
library(vegan)

# Quantify dissimilarities in the original feature space
# Pick relabunance assay separately
x <- assay(tse, "relabundance") 
d0 <- as.matrix(vegdist(t(x), "bray"))

# PCoA Ordination
pcoa <- as.data.frame(cmdscale(d0, k = 2))
names(pcoa) <- c("PCoA1", "PCoA2")

# Quantify dissimilarities in the ordination space
dp <- as.matrix(dist(pcoa))

# Calculate stress i.e. relative difference 
# in the original and projected dissimilarities
stress <- sum((dp - d0)^2) / sum(d0^2)
```

A Shepard plot visualizes the original versus the ordinated dissimilarity
between the observations.

```{r shepard}
ord <- order(as.vector(d0))
df <- data.frame(d0 = as.vector(d0)[ord],
                 dmds = as.vector(dp)[ord])

ggplot(df, aes(x = d0, y = dmds)) +
  geom_smooth() +
  geom_point() +    
  labs(title = "Shepard plot",
       x = "Original distance",
       y = "MDS distance",   
       subtitle = paste("Stress:", round(stress, 2))) +
  theme_bw()
```


## Supervised / constrained ordination

dbRDA is a supervised counterpart of PCoA. It maximize the variance
with respect to the covariates provided by the user. This can be used
to quantify associations between each covariate and community
composition (beta diversity). The table below summarizes the relations
between the supervised and unsupervised ordination methods.


|                          | supervised ordination  | unsupervised ordination  |
|:------------------------:|:----------------------:|:------------------------:|
| Euclidean distance       | RDA                    | PCA                      |
| non-Euclidean distance   | dbRDA                  | PCoA/MDS, NMDS, UMAP     | 


In summary, the dbRDA is the more general method that allows a wider
variety dissimilarity, or beta diversity, indices. This method is
available via `mia::runRDA`, which calls `vegan::dbrda`. By default,
this uses Euclidean distances, which is equivalent to the ordinary
RDA. However, the dbRDA method (`mia::runRDA`) allows the use of other
dissimilarity indices as well.


Let us next demonstrate dbRDA with the enterotype dataset. Here
samples correspond to patients. The `colData` lists the clinical
status of each patient and a few covariates such as gender and
age.


```{r import-rda-dataset}
# Load data
data("enterotype", package = "mia")
tse2 <- enterotype

# Apply relative transform
tse2 <- transformAssay(tse2,
                       method = "relabundance")
```

dbRDA can be perfomed with the `runRDA` function. In addition to the arguments
previously defined for [unsupervised ordination](#unsupervised-ordination), this
function takes a formula to control for variables and an action to treat missing
values. Along with clinical status, which is the main outcome, we control for
gender and age, and exclude observations where one of these variables is missing.

```{r run-rda}
# Perform RDA
tse2 <- runRDA(tse2,
               assay.type = "relabundance",
               formula = assay ~ ClinicalStatus + Gender + Age,
               distance = "bray",
               na.action = na.exclude)

# Store results of PERMANOVA test
rda_info <- attr(reducedDim(tse2, "RDA"), "significance")
```

The importance of each variable on the similarity between samples can be
assessed from the results of PERMANOVA, automatically provided by the `runRDA`
function. We see that both clinical status and age explain more than 10% of the
variance, but only age has statistical significance.

```{r rda-permanova-res}
rda_info$permanova |>
  knitr::kable()
```

To ensure that the homogeneity assumption holds, we retrieve the corresponding
information from the results of RDA. None of the p-values is lower
than the significance threshold, and thus homogeneity is observed.

```{r rda-homogeneity-res}
rda_info$homogeneity |>
  knitr::kable()
```

Next, we proceed to visualize the weight and significance of each variable on
the similarity between samples with an RDA plot, which can be generated with
the `plotRDA` function from the miaViz package.

```{r plot-rda}
# Load packages for plotting function
library(miaViz)

# Generate RDA plot coloured by clinical status
plotRDA(tse2, "RDA", colour_by = "ClinicalStatus")
```

From above, we can see that only age significantly describes
differences between the microbial profiles of different samples. Such visual
approach complements the previous results obtained with PERMANOVA.


## Case studies

#### Visualizing the most dominant genus on PCoA {#sec-pcoa-genus}

In this section, we visualize the most dominant genus on PCoA. A similar
visualization was proposed by [-@Salosensaari2021]. First, let us agglomerate the
data at the Genus level and identify the dominant taxa for each sample.

```{r}
# Agglomerate to genus level
tse_genus <- mergeFeaturesByRank(tse,
                                 rank = "Genus")

# Convert to relative abundances
tse_genus <- transformAssay(tse,
                            method = "relabundance",
                            assay.type = "counts")

# Add info on dominant genus per sample
tse_genus <- addDominant(tse_genus,
                         assay.type = "relabundance",
                         name = "dominant_taxa")
# Overview
summarizeDominance(tse_genus, rank = "Genus", digits = 3, name = "dominant_taxa")
```

Next, we perform PCoA with Bray-Curtis dissimilarity.

```{r}
tse_genus <- runMDS(tse_genus,
                    FUN = vegan::vegdist,
                    name = "PCoA_BC",
                    assay.type = "relabundance")
```

Finally, we get the top taxa and and visualize their abundances on PCoA. Note
that A 3D interactive version of the plot below can be found in [@sec-extras].

```{r}
# Getting the top taxa
top_taxa <- getTop(tse_genus,
                           top = 6,
                           assay.type = "relabundance")

# Naming all the rest of non top-taxa as "Other"
most_abundant <- lapply(colData(tse_genus)$dominant_taxa,
                        function(x) {if (x %in% top_taxa) {x} 
                          else {"Other"}})

# Storing the previous results as a new column within colData
colData(tse_genus)$most_abundant <- as.character(most_abundant)

# Calculating percentage of the most abundant
most_abundant_freq <- table(as.character(most_abundant))
most_abundant_percent <- round(most_abundant_freq / 
                                 sum(most_abundant_freq) * 100, 1)

# Retrieving the explained variance
e <- attr(reducedDim(tse_genus, "PCoA_BC"), "eig")
var_explained <- e / sum(e[e > 0]) * 100

# Define colors for visualization
my_colors <- c("black", "blue", "lightblue", "darkgray",
               "magenta", "darkgreen", "red")

# Visualization
plot <-plotReducedDim(tse_genus, "PCoA_BC",
                      colour_by = "most_abundant") +
  scale_colour_manual(values = my_colors,
                      labels = paste0(names(most_abundant_percent), "(", most_abundant_percent, "%)")) +
  labs(x = paste("PC 1 (", round(var_explained[1], 1), "%)"),
       y = paste("PC 2 (", round(var_explained[2], 1), "%)"),
       color = "")

plot
```

Similarly, we visualize and compare the sub-population.

```{r}
# Calculating the frequencies and percentages for both categories
freq_TRUE  <- table(as.character(most_abundant[tse_genus$Group]))
freq_FALSE <- table(as.character(most_abundant[!tse_genus$Group]))
percent_TRUE  <- round(freq_TRUE  / sum(freq_TRUE)  * 100, 1)
percent_FALSE <- round(freq_FALSE / sum(freq_FALSE) * 100, 1)

# Visualization
plotReducedDim(tse_genus[ , colData(tse_genus)$Group == TRUE], "PCoA_BC",
               colour_by = "most_abundant") +
  scale_colour_manual(values = my_colors,
                      labels = paste0(names(percent_TRUE), "(", percent_TRUE, "%)")) +
  labs(x = paste("PC 1 (", round(var_explained[1], 1), "%)"),
       y = paste("PC 2 (", round(var_explained[2], 1), "%)"),
       title = "Group = TRUE", color = "")

plotReducedDim(tse_genus[ , colData(tse_genus)$Group == FALSE], "PCoA_BC",
               colour_by = "most_abundant") +
  scale_colour_manual(values = my_colors,
                      labels = paste0(names(percent_FALSE), "(", percent_FALSE, "%)")) +
  labs(x = paste("PC 1 (", round(var_explained[1], 1), "%)"),
       y = paste("PC 2 (", round(var_explained[2], 1), "%)"),
       title = "Group = FALSE", color = "")
```

### Testing differences in community composition between sample groups {#sec-dbrda-workflow}

Permutational Analysis of Variance (PERMANOVA; [-@Anderson2001]) is a
widely used non-parametric multivariate method that aims to estimate
the actual statistical significance of differences in the observed
community composition between two groups of samples.

PERMANOVA tests the hypothesis that the centroids and dispersion of
the community are equivalent between the compared groups. A p-value
smaller than the significance threshold indicates that the groups have
a different community composition. This method is implemented with the
[`adonis2`](https://www.rdocumentation.org/packages/vegan/versions/2.4-2/topics/adonis)
function from the vegan package.

By default, the argument `by` is set to `"terms"`, in which the order
of variables in the formula matters. In this case, each variable is
analyzed sequentially, and the result is different when more than 1
variable is introduced and their order differs.  Therefore, it is
recommended to set `by = "margin"`, which specifies that the marginal
effect of each variable is analyzed individually. You can view a
comparison between the two designs in chapter [PERMANOVA comparison](https://microbiome.github.io/OMA/docs/devel/pages/97_extra_materials.html#compare-permanova).

We can perform PERMANOVA either with `adonis2` function or by first
performing dbRDA and then applying permutational test its results. An
advantage of the latter approach is that by doing so we can get
coefficients: how much each taxa affects the variation between
communities.

```{r}
# Agglomerate data to Species level
tse <- mergeFeaturesByRank(tse,
                           rank = "Species")

# Set seed for reproducibility
set.seed(1576)
# We choose 99 random permutations. Consider applying more (999 or 9999) in your
# analysis. 
permanova <- adonis2(t(assay(tse, "relabundance")) ~ Group,
                     by = "margin", # each term (here only 'Group') analyzed individually
                     data = colData(tse),
                     method = "euclidean",
                     permutations = 99)

# Set seed for reproducibility
set.seed(1576)
# Perform dbRDA
dbrda <- dbrda(t(assay(tse,"relabundance")) ~ Group, 
               data = colData(tse))
# Perform permutational analysis
permanova2 <- anova.cca(dbrda,
                        by = "margin", # each term (here only 'Group') analyzed individually
                        method = "euclidean",
                        permutations = 99)

# Get p-values
p_values <- c(permanova["Group", "Pr(>F)"], permanova2["Group", "Pr(>F)"])
p_values <-as.data.frame(p_values)
rownames(p_values) <- c("adonis2", "dbRDA+anova.cca")
p_values
```

As we can see, the community composition is significantly different
between the groups (p < 0.05), and these two methods give equal p-values.

Let us visualize the model coefficients for species that exhibit the
largest differences between the groups. This gives some insights into
how the groups tend to differ from each other in terms of community
composition.

```{r}
# Add taxa info
sppscores(dbrda) <- t(assay(tse, "relabundance"))
# Get coefficients
coef <- dbrda$CCA$v
# Get the taxa with biggest weights
top.coef <- head(coef[rev(order(abs(coef))), , drop = FALSE], 20)
# Sort weights in increasing order
top.coef <- top.coef[order(top.coef), ]
# Get top names
top_names <- names(top.coef)[order(abs(top.coef), decreasing = TRUE)]
```

```{r plot-top-coef-anova, fig.cap = ""}
df <- data.frame(x = top.coef,
                 y = factor(names(top.coef), unique(names(top.coef))))

ggplot(df, aes(x = x, y = y)) +
  geom_bar(stat = "identity") +
  labs(x = "", y= "", title = "Top Taxa") +
  theme_bw()
```

In the example above, the largest differences between the two groups
can be attributed to _`r top_names[1] `_ (elevated in the first
group) and _`r top_names[2] `_ (elevated in the second
group), and many other co-varying species.


### Retrieving feature loadings

The functions `vegan::dbrda` and `vegan::capscale` are exchangeable
with `vegan::adonis2` with certain settings, and can be used to
retrieve feature loadings.


### Checking the homogeneity condition 

It is important to note that the application of PERMANOVA assumes
homogeneous group dispersions (variances). This can be tested with the
PERMDISP2 method [@Anderson2006] by using the same assay and distance
method than in PERMANOVA.

```{r}
anova(betadisper(vegdist(t(assay(tse, "counts"))), colData(tse)$Group))
```

If the groups have similar dispersion, PERMANOVA can be seen as an
appropriate choice for comparing community compositions.


## Summary


As a final note, we provide a comprehensive list of functions for the evaluation
of dissimilarity indices available in the mia and scater packages. The
`calculate` methods return a reducedDim object as an output, whereas the `run`
methods store the reducedDim object into the specified TreeSE.

* Canonical Correspondence Analysis (CCA): `getCCA` and `runCCA`
* dbRDA: `getRDA` and `runRDA`; our recommended default method
  to assess differences in community composition (beta diversity)
* Double Principal Coordinate Analysis (DPCoA): `getDPCoA` and `runDPCoA`
* Jensen-Shannon Divergence (JSD): `calculateJSD` and `runJSD`
* MDS: `calculateMDS` and `runMDS`
* NMDS: `getNMDS` and `runNMDS`
* Overlap: `calculateOverlap` and `runOverlap`
* PERMANOVA: (e.g. from `vegan::adonis2`) can be used to assess
  significance when comparing community composition between
  groups. Retrieving the loadings and components is more tricky,
  however.
* t-distributed Stochastic Neighbor Embedding (t-SNE): `calculateTSNE` and `runTSNE`
* UMAP: `calculateUMAP` and `runUMAP`


For more information on sample clustering, you can refer to:

* [How to extract information from clusters](http://bioconductor.org/books/release/OSCA/clustering.html)  
* Chapter [@sec-clustering] on community typing  




