# Community Typing {#sec-clustering}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

Community typing in microbial ecology involves identifying distinct microbial 
communities by recognizing patterns in the data.The community types are 
described based on representative taxa characteristics, distinguished 
from the characteristics of the other community types. The samples are described 
by community type assignments, which are defined by the ecosystem features in 
the sample. Community typing techniques can typically be divided into unsupervised 
clustering and dimensionality reduction techniques, where clustering is more 
commonly in use.

In this chapter, we first walk you through some common clustering schemes in use
for microbial community ecology, and then focus in more detailed community typing
techniques based on data dimensionality reduction.

## Clustering 

Clustering techniques aim to find groups,  called clusters, that share a pattern
in the data. In the microbiome context, clustering techniques are included in 
microbiome community typing methods. For example, clustering allow samples to be 
distinguished from each other based on their microbiome community composition.
Clustering scheme consists of two steps, the first is to compute the sample 
dissimilarities with a given distance metrics, and the second is to form the 
clusters based on the dissimilarity matrix. There are multiple clustering 
algorithms available, some of which are introduced in this 
chapter.The data can be clustered either based on features or samples. The examples below 
are focused on sample clustering.

*bluster* is a Bioconductor package providing tools for clustering data in the 
`SummarizedExperiment` container. It offers multiple algorithms such 
as hierarchical clustering, DBSCAN, and K-means. 

```{r load_bluster}
# Load dependencies
library(bluster)
library(kableExtra)
```

In the examples of this chapter we use 'peerj13075' data for microbiota community 
typing. This chapter illustrates how different results can be obtained depending 
on the choice of the algorithm. To reduce calculation time we decided to 
agglomerate taxa onto 'Class' level and filter out the least prevalent taxa as 
well as less commonly detected within a sample resulting in 25 features in the 
data.

```{r load-pkg-data1}
library(mia)
data("peerj13075", package = "mia")
tse <- peerj13075

# Filter out most taxa for faster calculation
altExp(tse, "prevalent") <- agglomerateByPrevalence(tse, rank = "class", 
                                                    prevalence = 20/100, 
                                                    detection = 1/100)
```

## Hierarchical clustering

The hierarchical clustering algorithm aims to find hierarchy between
samples/features. There are two approaches: agglomerative ("bottom-up")
and divisive ("top-down"). In agglomerative approach, each observation is first 
in a unique cluster, after which the algorithm continues to agglomerate similar 
data points into clusters. The divisive approach, instead, starts with one 
cluster that contains all observations. Clusters are split recursively into 
clusters that differ the most. The clustering can be continued until each 
cluster contains only one observation.

In this example we use `addCluster` function from mia to  cluster the data. 
`addCluster` function allows to choose a clustering algorithm and offers multiple 
parameters to shape the result. HclustParam() parameter is chosen for hierarchical 
clustering. HclustParam() parameter itself has parameters on its own 
[HclustParam documentation](https://rdrr.io/github/LTLA/bluster/man/HclustParam-class.html).
A parameter is `MARGIN` defines whether to cluster features or samples. Here we 
cluster counts data, for which we compute the dissimilarities with the 
Bray-Curtis distance. Here, again, we use `ward.d2` metohod.  Returning 
statistical information can chosen by using 
`full` parameter. Finally, the `clust.col` parameter allows us to choose the 
name of the column in the colData (default name is `clusters`).

```{r hclust1}
library(vegan)

set.seed(123)
altExp(tse, "prevalent") <- addCluster(altExp(tse, "prevalent"),
               assay.type = "counts",
               MARGIN = "samples",
               HclustParam(method = "ward.D2",
                           dist.fun = vegdist,
                           metric = "bray"),
               full = TRUE,
               clust.col = "hclust")

table(altExp(tse, "prevalent")$hclust)
```

Hierarchical clustering suggests six clusters to describe the data.

Now, we visualize the hierarchical structure of the clusters with a dendrogram 
tree. In dendrograms, the tree is split where the branch length is the largest. 
In each splitting point, the tree is divided into two clusters leading to the
hierarchy. In this example, each sample is labelled by their dominant taxon
to visualize ecological differences between the clusters.

```{r hclust3}
library(dendextend)

# Get hclust data from metadata
hclust_data <- metadata(altExp(tse, "prevalent"))$clusters$hclust

# Get the dendrogram object
dendro <- as.dendrogram(hclust_data)
k <- length(unique(altExp(tse, "prevalent")$hclust))

# Color the branches by clusters
cols <- scales::hue_pal()(k)

# Order the colors to be equivalent to the factor levels
cols <- cols[c(1,4,5,3,2,6)]
dend <- color_branches(dendro, k = k, col = unlist(cols))

# Label the samples by their dominant taxon
altExp(tse, "prevalent") <- addDominant(altExp(tse, "prevalent"))
labels(dend) <- altExp(tse, "prevalent")$dominant_taxa
dend <- color_labels(dend, k = k, col = unlist(cols))

# Plot dendrogram
par(mar=c(9, 3, 0.5, 0.5))
dend %>% set("labels_cex", 0.8) %>% plot()
```

In our case, we cut the dendrogram in half by default. To know how many clusters
we have, we can check the colData.

```{r hclust4}
# Get the clusters
summary(colData(altExp(tse, "hclust_full"))$Hclust)
```

We can also visualize the clusters by by projecting the data onto two-dimensional 
surface that captures the most variability in the data. In this example, we use
multi-dimensional scaling (MDS). The visualization demonstrates that 
the clusters can be distinguished on a two-dimensional plot, although the 
technique does not preserve detailed information about the cluster division
process.

```{r hclust3}
library (scater)

# Add the MDS dimensions for plotting
altExp(tse, "prevalent") <- runMDS(altExp(tse, "prevalent"), 
                                    assay.type = "counts", 
                                    FUN = vegan::vegdist, 
                                    method = "bray")
```

```{r hclust4, fig.width=6, fig.height=4}
plotReducedDim(altExp(tse, "prevalent"), "MDS", theme_size = 13) + 
              geom_point(size = 3, alpha = 0.6, 
                      aes(color = altExp(tse, "prevalent")$hclust)) +
              labs(color = "HClust", title = "Hclust")
```

## Dirichlet Multinomial Mixtures (DMM)

This section focuses on [Dirichlet-Multinomial Mixture
Model](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0030126) 
analysis. It is a Bayesian clustering technique that allows to search for sample 
patterns that reflect sample similarity in the data using both prior information
and the data.

In this example, we cluster the data with DMM clustering. In the example below, 
we calculate model fit using Laplace approximation ro reduce the calculation time. 
The cluster information is added in the metadata with an optional `name`. 

```{r dmm1}
# Run the model and calculates the most likely number of clusters from 1 to 6

set.seed(123)
altExp(tse, "prevalent") <- addCluster(altExp(tse, "prevalent"), 
                                      assay.type = "counts", 
                                      name = "DMM", 
                                      DmmParam(k = 1:6, type = "laplace"), 
                                      MARGIN = "samples", 
                                      full = TRUE, 
                                      clust.col = "dmmclust")
```

The plot below represents the Laplace approximation to the model evidence for 
each of the six models. We can see that the best number of clusters is two, as 
the minimum value of Laplace approximation to the negative log model evidence for 
DMM models as a function of k, determines the optimal k. The optimal k suggests 
to fit a model with k mixtures of Dirichlet distributions as a prior. 

```{r dmm2, fig.width=5, fig.height=3}
library(miaViz)
p <- plotDMNFit(altExp(tse, "prevalent"), type = "laplace", name = "DMM")
p + theme_minimal(base_size = 11)
```

The plot above suggests that two clusters describe the data the best. Now 
we can plot, which taxa define the cluster features.

```{r dmm3, fig.width=6, fig.height=2}
# Get the estimates on how much each phyla contributes on each cluster
best_model <- metadata(altExp(tse, "prevalent"))$DMM$dmm[2]
drivers <- as.data.frame(best_model[[1]]@fit$Estimate)

# Clean names
drivers$class <- rownames(drivers)

for (i in 1:2) {
  # Transfer to relative to optimize plotting
  drivers[i] <- drivers[i]/sum(drivers[i])
  
  drivers <- drivers[order(drivers[[i]], decreasing = TRUE),]
  p <- ggplot(head(drivers, 5), 
              aes(x = reorder(head(class, 5), + 
                    head(drivers[[i]], 5)), 
                    y = head(drivers[[i]], 5))) +
              geom_bar(stat = "identity", 
                    fill = scales::hue_pal()(1), 
                    alpha = 0.6) +
                    coord_flip() + 
                    labs(title = paste("Top phyla in group", i)) +
                    theme_light(base_size = 11) + 
                    labs(x="", y="") + 
                    scale_y_continuous(limits=c(0,1))
  print(p)
}
```

As well as in hierarchical clustering, we can also visualize the clusters by 
projecting the data onto two-dimensional surface calculated using MDS. The plot
rather clearly demonstrates the cluster division.

```{r}
# DMM clusters
plotReducedDim(altExp(tse, "prevalent"), "MDS", theme_size = 13) + 
              geom_point(size = 3, alpha = 0.6, 
                      aes(color = altExp(tse, "prevalent")$dmmclust)) +
              labs(color = "DMMclust", title = "DMM") + theme_minimal()
```

## Biclustering

Biclustering methods cluster rows and columns simultaneously in order
to find subsets of correlated features/samples.

Here, we use following packages:

-   [biclust](https://cran.r-project.org/web/packages/biclust/index.html)
-   [cobiclust](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.13582)

_cobiclust_ is especially developed for microbiome data whereas _biclust_ is more
general method. In this section, we show two different cases and example 
solutions to apply biclustering to them. 

1.   Taxa vs samples
2.   Taxa vs biomolecule/biomarker

Biclusters can be visualized using heatmap or boxplot, for
instance. For checking purposes, also scatter plot might be valid
choice.

Check out more ideas for heatmaps from chapters [@sec-viz-chapter] and
[@sec-microbiome-community].

### Taxa vs samples

When you have microbial abundance matrices, we suggest to use
_cobiclust_ which is designed for microbial data.

Load example data
```{r load-pkg-data2}
library(cobiclust)
data("HintikkaXOData")
mae <- HintikkaXOData
```

Only the most prevalent taxa are included in analysis. 

```{r cobiclust_1}
# Subset data in the first experiment
mae[[1]] <- subsetByPrevalent(mae[[1]], rank = "Genus", 
                                      prevalence = 0.2, 
                                      detection = 0.001)

# rclr-transform in the first experiment
mae[[1]] <- transformAssay(mae[[1]], method = "rclr")
```

_cobiclust_ takes counts table as an input and gives _cobiclust_ object as an output.
It includes clusters for taxa and samples. 

```{r cobiclust_2}
# Do clustering using counts table
clusters <- cobiclust(assay(mae[[1]], "counts"))

# Get clusters
row_clusters <- clusters$classification$rowclass
col_clusters <- clusters$classification$colclass

# Add clusters to rowdata and coldata
rowData(mae[[1]])$clusters <- factor(row_clusters)
colData(mae[[1]])$clusters <- factor(col_clusters)

# Order data based on clusters
mae[[1]] <- mae[[1]][order(rowData(mae[[1]])$clusters),
                     order(colData(mae[[1]])$clusters)]

# Print clusters
clusters$classification
```

Next we can plot clusters. Annotated heatmap is a common choice.

```{r cobiclust_3a, fig.width=14, fig.height=12}
library(pheatmap)
# z-transform for heatmap
mae[[1]] <- transformAssay(mae[[1]], assay.type = "rclr",
                            MARGIN = "features", method = "z", name = "rclr_z")

# Create annotations. When column names are equal, they should share levels.
# Here samples include 3 clusters, and taxa 2. That is why we have to make
# column names unique.
annotation_col <- data.frame(colData(mae[[1]])[, "clusters", drop = F])
colnames(annotation_col) <- "col_clusters"

annotation_row <- data.frame(rowData(mae[[1]])[, "clusters", drop = F])
colnames(annotation_row) <- "row_clusters"
```

Plot the heatmap.

```{r cobiclust_3b, fig.width=14, fig.height=12}
pheatmap(assay(mae[[1]], "rclr_z"), cluster_rows = F, cluster_cols = F,
         annotation_col = annotation_col, annotation_row = annotation_row)
```

Boxplot is commonly used to summarize the results:

```{r cobiclust_4}
library(ggplot2)
library(patchwork)

# ggplot requires data in melted format
melt_assay <- meltSE(mae[[1]], assay.type = "rclr", 
                        add_col_data = T, add_row_data = T)

# patchwork two plots side-by-side
p1 <- ggplot(melt_assay) +
    geom_boxplot(aes(x = clusters.x, y = rclr)) +
    labs(x = "Taxa clusters")

p2 <- ggplot(melt_assay) +
    geom_boxplot(aes(x = clusters.y, y = rclr)) +
    labs(x = "Sample clusters")

p1 + p2
```

### Taxa vs biomolecules

Here, we analyze cross-correlation between taxa and metabolites. This
is a case, where we use _biclust_ method which is suitable for numeric
matrices in general. First we pre-process the data.

```{r biclust_1}
# Samples must be in equal order
# (Only 1st experiment was ordered in cobiclust step leading to unequal order)
mae[[1]] <- mae[[1]][, colnames(mae[[2]])]

# Make rownames unique, since it is required by other steps
rownames(mae[[1]]) <- make.unique(rownames(mae[[1]]))

# Transform the metabolites to be in log basis
mae[[2]] <- transformAssay(mae[[2]], assay.type = "nmr", method = "log10")

# Add missing data to the metabolites
replace_na <- function(row) {
    na_indices <- which(is.na(row))
    non_na_values <- row[!is.na(row)]
    row[na_indices] <- sample(non_na_values, length(na_indices), replace = TRUE)
    row
}
assay(mae[[2]], "log10") <- t(apply(assay(mae[[2]], "log10"), 1, replace_na))
```

Next, we compute the spearman correlation matrix.

```{r biclust_2}
# Calculate correlations
corr <- getExperimentCrossCorrelation(mae, 1, 2, assay.type1 = "rclr",
                                      assay.type2 = "log10", mode = "matrix",
                                      correlation = "spearman")
```

_biclust_ takes a matrix as an input and returns a _biclust_ object. 

```{r biclust_3}
library(biclust)
# Set seed for reproducibility
set.seed(3973)

# Find biclusters
bc <- biclust(corr, method = BCPlaid(), verbose = FALSE)

bc
```

The object includes cluster information. However compared to
_cobiclust_, _biclust_ object includes only information about clusters
that were found, not general cluster.

Meaning that if one cluster size of 5 features was found out of 20 features, 
those 15 features do not belong to any cluster. That is why we have to create an
additional cluster for features/samples that are not assigned into any cluster.

```{r biclust_4}
# Functions for obtaining biclust information

# Get clusters for rows and columns
.get_biclusters_from_biclust <- function(bc, assay) {
    # Get cluster information for columns and rows
    bc_columns <- t(bc@NumberxCol)
    bc_columns <- data.frame(bc_columns)
    bc_rows <- bc@RowxNumber
    bc_rows <- data.frame(bc_rows)

    # Get data into right format
    bc_columns <- .manipulate_bc_data(bc_columns, assay, "col")
    bc_rows <- .manipulate_bc_data(bc_rows, assay, "row")
    
    return(list(bc_columns = bc_columns, bc_rows = bc_rows))
}

# Input clusters, and how many observations there should be, i.e.,
# the number of samples or features
.manipulate_bc_data <- function(bc_clusters, assay, row_col) {
    # Get right dimension
    dim <- ifelse(row_col == "col", ncol(assay), nrow(assay))
    # Get column/row names
    if (row_col == "col") {
        names <- colnames(assay)
    } else {
        names <- rownames(assay)
    }

    # If no clusters were found, create one. Otherwise create additional
    # cluster which
    # contain those samples that are not included in clusters that were found.
    if (nrow(bc_clusters) != dim) {
        bc_clusters <- data.frame(cluster = rep(TRUE, dim))
    } else {
        # Create additional cluster that includes those samples/features that
        # are not included in other clusters.
        vec <- ifelse(rowSums(bc_clusters) > 0, FALSE, TRUE)

        # If additional cluster contains samples, then add it
        if (any(vec)) {
            bc_clusters <- cbind(bc_clusters, vec)
        }
    }
    
    # Adjust row and column names
    rownames(bc_clusters) <- names
    colnames(bc_clusters) <- paste0("cluster_", 1:ncol(bc_clusters))
    return(bc_clusters)
}
```

```{r biclust_5}
# Get biclusters
bcs <- .get_biclusters_from_biclust(bc, corr)

bicluster_rows <- bcs$bc_rows
bicluster_columns <- bcs$bc_columns

# Print biclusters for rows
head(bicluster_rows)
```

Let's collect information for the scatter plot. 

```{r biclust_6}
# Function for obtaining sample-wise sum, mean, median, and mean variance
# for each cluster

.sum_mean_median_var <- function(tse1, tse2, assay.type1, assay.type2, clusters1, clusters2) {
    list <- list()
    # Create a data frame that includes all the information
    for (i in 1:ncol(clusters1)) {
        # Subset data based on cluster
        tse_subset1 <- tse1[clusters1[, i], ]
        tse_subset2 <- tse2[clusters2[, i], ]
        # Get assay
        assay1 <- assay(tse_subset1, assay.type1)
        assay2 <- assay(tse_subset2, assay.type2)
        # Calculate sum, mean, median, and mean variance
        sum1 <- colSums2(assay1, na.rm = T)
        mean1 <- colMeans2(assay1, na.rm = T)
        median1 <- colMedians(assay1, na.rm = T)
        var1 <- colVars(assay1, na.rm = T)
        
        sum2 <- colSums2(assay2, na.rm = T)
        mean2 <- colMeans2(assay2, na.rm = T)
        median2 <- colMedians(assay2, na.rm = T)
        var2 <- colVars(assay2, na.rm = T)
        
        list[[i]] <- data.frame(sample = colnames(tse1), sum1, sum2, mean1, 
                                 mean2, median1, median2, var1, var2)
    }
    return(list)
}

# Calculate info
df <- .sum_mean_median_var(mae[[1]], mae[[2]], "rclr", "log10", bicluster_rows, bicluster_columns)
```

Now we can create a scatter plot. X-axis includes median clr abundance
of microbiome and y-axis median absolute concentration of each
metabolite. Each data point represents a single sample.

From the plots, we can see that there is low negative correlation in
both cluster 1 and 3.  This means that when abundance of bacteria
belonging to cluster 1 or 3 is higher, the concentration of
metabolites of cluster 1 or 3 is lower, and vice versa.

```{r biclust_7, fig.width=14, fig.height=6, fig.show="keep", out.width="33%"}
pics <- list()
for (i in seq_along(df)) {
    pics[[i]] <- ggplot(df[[i]]) +
        geom_point(aes(x = median1, y = median2)) +
        labs(title = paste0("Cluster ", i), x = "Taxa (rclr median)",
             y = "Metabolites (abs. median)")
    print(pics[[i]])
}
pics[[1]] + pics[[2]] + pics[[3]]
```

_pheatmap_ does not allow boolean values, so they must be converted into factors.

```{r biclust_8}
bicluster_columns <- data.frame(apply(bicluster_columns, 2, as.factor))
bicluster_rows <- data.frame(apply(bicluster_rows, 2, as.factor))
```

Again, we can plot clusters with heatmap.

```{r biclust_9, fig.width=10, fig.height=10}
# Adjust colors for all clusters
if (ncol(bicluster_rows) > ncol(bicluster_columns)) {
    cluster_names <- colnames(bicluster_rows)
} else {
    cluster_names <- colnames(bicluster_columns)
}
annotation_colors <- list()
for (name in cluster_names) {
    annotation_colors[[name]] <- c("TRUE" = "red", "FALSE" = "white")
}

# Create a heatmap
pheatmap(corr, cluster_cols = F, cluster_rows = F,
         annotation_col = bicluster_columns, annotation_row = bicluster_rows,
         annotation_colors = annotation_colors)
```

## Additional Community Typing

For more community typing techniques applied to the 'SprockettTHData'
data set, see the attached .Rmd file.

Link:

   * [Rmd](add-comm-typing.Rmd)
