# Differential Abundance {#sec-differential-abundance}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

This section provides an introduction to Differential Abundance Analysis (DAA),
which is used to identify differences in the abundances of individual taxa (at
any taxonomic level) between two or more groups, such as treatment and control.
Here, we demonstrate its implementation on [Tengeler2020](https://microbiome.github.io/mia/reference/Tengeler2020.html) data set.

The goal of DAA is to identify biomarkers of a certain phenotype or condition,
and gain understanding of a complex system by looking at its isolated components.
For example, the identification of a bacterial taxon that is differentially
abundant between healthy patients and diseased patients can lead to important
insights into the pathophysiology of the disease. In other words, differentially
abundant taxa can be involved in the dynamics of the disease, which in turn helps
understand the system as a whole. Despite its relevance in current research,
the DAA approach has also been subject to debate [@Quinn2021].

## Statistical challenges of microbiome data 

Microbiome data display unique properties that are exclusively addressed by DAA
tools developed for microbiome analysis. Specifically, microbiome data are
characterized by high variability, zero-inflation and compositionality.
**High variability** expresses that abundance of taxa often varies by several
orders of magnitude from sample to sample. **Zero-inflation** means that
typically more than 70% of the values are zeros, which could be due to either
physical absence (structural zeros) or insufficient sampling effort (sampling
zeros). **Compositionality** implies that a change in the absolute abundance of
one taxon will lead to apparent variations in the relative abundances of other
taxa in the same sample. If neglected, such properties may cause significant
bias in the results of DAA. Therefore, several approaches have been developed to
address the unique properties of microbiome data and provide statistically
useful results.

The first approach to target zero-inflated data consists of specialized models,
such as **over-dispersed count models** and **zero-inflated mixture models**.
DESeq2, edgeR and corncorb are based on over-dispersed count models, whereas
metagenomeSeq, RAIDA, ZIBB and Omnibus implement zero-inflated mixture models to
address zero-inflation. Typically, these models assume a negative binomial,
beta-binomial or normal/log-normal distribution. Alternatively, **zero imputation**
also represents a valid approach to deal with zero-inflated data. ALDEx2 and
eBay apply a Bayesian model to impute the zeros when working with proportion
data, accounting for sampling variability and sequencing depth variation. Other
methods, such as MaAsLin2 and ANCOMBC impute the zeros with a pseudo-count strategy.

Regarding the compositionality of microbiome data, several approaches have been
developed to perform **robust normalization** with methods specifically designed
to reduce the bias found in compositional data. Some examples include trimmed
mean of M-values (TMM) normalization used by edgeR, relative log expression (RLE)
normalization used by DESeq2, cumulative sum scaling (CSS) normalization used
by metagenomeSeq, centered log-ratio transformation (CLR) normalization used
by ALDEx2 and geometric mean of pairwise ratios (GMPR) normalization used by
Omnibus and Wrench normalization [@Kumar2018], which corrects the compositional
bias by an empirical Bayes approach. Other methods to deal with compositional
data entail reference taxa approach used by DACOMP and RAIDA, analyzing the
pattern of pairwise log ratios as done by ANCOM and bias-correction applied by
ANCOMBC.

We recommend to have a look at @Nearing2022. In this study, multiple DAA methods
were applied to 38 different datasets and their results were compared to one
another. Because each method follows a slightly different approach in terms of
assumptions and normalization techniques, it was shown that results on the same
dataset may differ substantially depending on the method. Recently, @Yang2022
comprehensively evaluated DAA methods via a semi-parametric framework and 106
real datasets. This study also concluded that different methods can produce
contradictory results, creating the risk of cherry-picking the most favorable
options for one’s own hypothesis. Therefore, it is highly recommended to perform
DAA with multiple methods to determine whether the findings can be reproduced by
different approaches. Built on the findings of @Calgaro2020, the *benchdamic* [@Calgaro2022] package could offer a valuable support in this regard. Through a comprehensive evaluation process it serves both practitioners by comparing DA methods from existing literature, and method developers by providing an impartial tool to evaluate their new approaches in comparison to what is already available. For details, check its extensive [vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/benchdamic/inst/doc/intro.html).

## Using the tools

In this section we demonstrate the use of four methods that can be
recommended based on recent literature (ANCOM-BC [@ancombc2020],
_ALDEx2_ [@Gloor2016], _Maaslin2_ [@Mallick2020], _LinDA_ [@Zhou2022]
and _ZicoSeq_ [@Yang2022]).

The purpose of this section is to show how to perform DAA in R, not
how to correctly do causal inference. Depending on your experimental
setup and your theory, you must determine how to specify any model
exactly.  E.g., there might be confounding factors that might drive
(the absence of) differences between the shown groups that we ignore
here for simplicity. Or your dataset is repeated sampling design,
matched-pair design or the general longitudianl design.  We will
demonstrate how to include covariates in those models. We picked a
dataset that merely has microbial abundances in a TSE object as well
as a grouping variable in the sample data. We simplify the examples by
only including two of the three groups.

```{r import-daa-data}
library(mia)
library(tidyverse)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Show patient status by cohort
table(tse$patient_status, tse$cohort) %>%
  knitr::kable()
```

### Preparing the data for DAA 

Before starting the analysis, it is recommended to reduce the size and
complexity of the data to make the results more reproducible. For this
purpose, we agglomerate the features by genus and filter them by a prevalence
threshold of 10%.

```{r prep-daa-data}
# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse,
                            rank = "Genus",
                            prevalence = 10 / 100)

# Transform count assay to relative abundances
tse <- transformAssay(tse,
                      assay.type = "counts",
                      method = "relabundance")
```

While some DAA tools provide optional arguments for prevalence filtering, here
we filtered the tse object directly. This way, we ensure that the input data
remains the same when multiple tools are used.

### ALDEx2

In this section, we will show how to perform DAA with ALDEx2, which can be
regarded as the method of choice for its consistency, as it normally identifies features that are also found by complementary methods [@Nearing2022]. A more
extensive introduction to its functionality is available in the
[ALDEx2 vignette](https://bioconductor.org/packages/release/bioc/vignettes/ALDEx2/inst/doc/ALDEx2_vignette.html). 

ALDEx2 estimates technical variation within each sample per taxon by utilizing
the Dirichlet distribution. It furthermore applies the CLR transformation (or
closely related log-ratio transforms). Depending on the experimental setup, it
will perform a two sample Welch's t-test and Wilcoxon test or a one-way ANOVA
and Kruskal-Wallis test. For more complex study designs, there is a possibility
to utilize the `glm` functionality within ALDEx2. The Benjamini-Hochberg
procedure is applied by default to correct for multiple testing.

```{r run-aldex2}
# Load package
library(ALDEx2)

# Generate Monte Carlo samples of the Dirichlet distribution for each sample.
# Convert each instance using the centered log-ratio transform.
# This is the input for all further analyses.
set.seed(123)
x <- aldex.clr(assay(tse), tse$patient_status)     
```

The t-test:

```{r aldex2-ttest}
# calculates expected values of the Welch's t-test and Wilcoxon rank
# test on the data returned by aldex.clr
x_tt <- aldex.ttest(x, paired.test = FALSE, verbose = FALSE)
```

Effect sizes:

```{r aldex2-effect}
# Determines the median clr abundance of the feature in all samples and in
# groups, the median difference between the two groups, the median variation
# within each group and the effect size, which is the median of the ratio
# of the between group difference and the larger of the variance within groups
x_effect <- aldex.effect(x, CI = TRUE, verbose = FALSE)

# combine all outputs 
aldex_out <- data.frame(x_tt, x_effect)
```

Now, we can create a so called Bland-Altman or MA plot (left). It shows the
association between the relative abundance and the magnitude of the difference
per sample. Next to that, we can also create a plot that shows the dispersion
on the x-axis instead of log-ratio abundance. Red dots represent genera that are
differentially abundant ($q \leq 0.1$) between the 2 groups. Black points are
rare taxa and grey ones are abundant taxa. The dashed line represent an effect
size of 1. @Gloor2016 provides more information on these plots.

```{r plot-aldex2}
par(mfrow = c(1, 2))

aldex.plot(aldex_out,
           type = "MA",
           test = "welch",
           xlab = "Log-ratio abundance",
           ylab = "Difference",
           cutoff = 0.05)

aldex.plot(aldex_out,
           type = "MW",
           test = "welch",
           xlab = "Dispersion",
           ylab = "Difference",
           cutoff = 0.05)
```

The evaluation as differential abundant in above plots is based on the
corrected p-value. According to the ALDEx2 developers, the safest
approach is to identify those features where the 95% CI of the effect
size does not cross 0. As we can see in below table, this is not the
case for any of the identified genera (see overlap column, which
indicates the proportion of overlap). Also, the authors recommend to
focus on effect sizes and CIs rather than interpreting the p-value. To
keep the comparison simple, we will here use the p-value as decision
criterion. But please be aware that the effect size together with the
CI is a better answer to the question we are typically interested in.

```{r aldex2-res}
aldex_out %>%
  rownames_to_column(var = "Genus") %>%
  # here we choose the wilcoxon output rather than t-test output
  filter(wi.eBH <= 0.05)  %>%
  dplyr::select(Genus, we.eBH, wi.eBH, effect, overlap) %>%
  knitr::kable()
```

### ANCOM-BC

The analysis of composition of microbiomes with bias correction
(ANCOM-BC) [@ancombc2020] is a recently developed method for differential
abundance testing. It is based on an earlier published approach
[@Mandal2015].  The previous version of ANCOM was among the methods
that produced the most consistent results and is probably a
conservative approach [@Nearing2022].  However, the new ANCOM-BC
method operates quite differently compared to the former ANCOM method.

As the only method, ANCOM-BC incorporates the so called *sampling
fraction* into the model. The latter term could be empirically
estimated by the ratio of the library size to the microbial
load. According to the authors, ignoring variations
in this sampling fraction would bias DAA results. Furthermore,
this method provides p-values and confidence intervals for each
taxon. It also controls the FDR and it is computationally simple to
implement.

Note that the original method was implemented in the `ancombc()` function (see 
[extended tutorial](https://www.bioconductor.org/packages/release/bioc/vignettes/ANCOMBC/inst/doc/ANCOMBC.html)).
The method has since then been updated and new features have been added to enable
multi-group comparisons and repeated measurements among other improvements. 
We do not cover the more advanced features of ANCOMBC in this tutorial 
as these features are documented in detail in this 
[tutorial](https://www.bioconductor.org/packages/release/bioc/vignettes/ANCOMBC/inst/doc/ANCOMBC2.html).

We now proceed with a simple example.  First, we specify a formula. In this 
formula, other covariates could potentially be included to adjust for 
confounding. We show this further below. Again, please make sure to check the 
[function documentation](https://rdrr.io/github/FrederickHuangLin/ANCOMBC/man/ancombc.html)
as well as the linked tutorials to learn about the additional arguments 
that we specify.

```{r run-ancombc, warning=FALSE}
# Load package
library(ANCOMBC)

# Run ANCOM-BC at the genus level and only including the prevalent genera
ancombc2_out <- ancombc2(data = tse,
                         assay.type = "counts",
                         fix_formula = "patient_status",
                         p_adj_method = "fdr",
                         prv_cut = 0,
                         group = "patient_status",
                         struc_zero = TRUE,
                         neg_lb = TRUE,
                         # multi group comparison is deactivated automatically
                         global = TRUE)
```

The object `out` contains all model output. Again, see the
[documentation of the
function](https://rdrr.io/github/FrederickHuangLin/ANCOMBC/man/ancombc.html)
under **Value** for details. Our question whether taxa are
differentially abundant can be answered by looking at the `res`
object, which contains dataframes with the coefficients, standard
errors, p-values and q-values. Below we show the first entries of this
dataframe.

```{r ancombc-res}
# store the FDR adjusted results 
ancombc2_out$res %>%
  dplyr::select(taxon, lfc_patient_statusControl, q_patient_statusControl) %>%
  filter(q_patient_statusControl < 0.05) %>%
  arrange(q_patient_statusControl) %>%
  head() %>%
  knitr::kable()
```

### MaAsLin2 

Let us next illustrate MaAsLin2 [@Mallick2020]. This method is based on
generalized linear models and flexible for different study designs
and covariate structures. For details, check their
[Biobakery tutorial](https://github.com/biobakery/biobakery/wiki/maaslin2).

```{r run-maaslin2, warning=FALSE, results="hide"}
# Load package
library(Maaslin2)

# maaslin expects features as columns and samples as rows 
# for both the abundance table as well as metadata 

# We can specify different GLMs/normalizations/transforms.
# specifying a ref is especially important if you have more than 2 levels
maaslin2_out <- Maaslin2(input_data = as.data.frame(t(assay(tse))),
                         input_metadata = as.data.frame(colData(tse)),
                         output = "DAA example",
                         transform = "AST",
                         fixed_effects = "patient_status",
                         # you can also fit MLM by specifying random effects
                         # random_effects = c(...),
                         reference = "patient_status,Control",
                         normalization = "TSS",
                         standardize = FALSE,
                         # filtering was previously performed
                         min_prevalence = 0)
```

Which genera are identified as differentially abundant? (leave out "head" to see all).

```{r, maaslin2-res}
maaslin2_out$results %>%
  filter(qval < 0.05) %>%
  knitr::kable()
```

This will create a folder that is called like in the output specified
above. It contains also figures to visualize difference between
significant taxa.

### LinDA 

Lastly, we cover linear models for differential abundance analysis of
microbiome compositional data (@Zhou2022). This is very similar to
ANCOMBC with few differences: 1) LinDA corrects for the compositional
bias differently using the mode of all regression coefficients. 2) it
is faster (100x-1000x than ANCOMBC and according to the authors); 3)
it supports hierarchical models. The latest ANCOMBC versions are also
supporting hierarchical models. Nevertheless, LinDA seems a promising
tool that achieves a very good power/fdr trade-off together with
ANCOMBC according to the review. The speed improvements might make it
critical especially for datasets that have higher sample or feature
set sizes.

```{r run-linda}
# Load package
library(MicrobiomeStat)

# Run LinDA
linda_out <- linda(feature.dat = as.data.frame(assay(tse)),
                   meta.dat = as.data.frame(colData(tse)),
                   formula = "~ patient_status",
                   alpha = 0.05,
                   prev.filter = 0,
                   mean.abund.filter = 0)
```

```{r linda-res}
# List genera for which H0 could be rejected:
linda_out$output$patient_statusControl %>%
  filter(reject) %>%
  dplyr::select(stat, padj) %>%
  rownames_to_column(var = "feature") %>%
  knitr::kable()
```

### ZicoSeq

Subsequently, we demonstrate DAA with ZicoSeq, a method based on linear models
and permutation. Further details can be found in
[this tutorial](https://cran.r-project.org/web/packages/GUniFrac/vignettes/ZicoSeq.html).
This approach has been assessed to exhibit high power and a low false discovery
rate, which has the following components:

  1. Winsorization to decrease the influence of outliers;
  2. Posterior sampling based on a beta mixture prior to address
  sampling variability and zero inflation;
  3. Reference-based multiple-stage normalization to address
  compositional effects;

```{r run-zicoseq}
# Load package
library(GUniFrac)

set.seed(123)
zicoseq_out <- ZicoSeq(feature.dat = as.matrix(assay(tse)),
                       meta.dat = as.data.frame(colData(tse)),
                       grp.name = "patient_status",
                       feature.dat.type = "count",
                       return.feature.dat = TRUE,
                       prev.filter = 0,
                       mean.abund.filter = 0,
                       max.abund.filter = 0,
                       perm.no = 999)
```

```{r zicoseq-res}
zicoseq_res <- cbind.data.frame(p.raw = zicoseq_out$p.raw,
                                p.adj.fdr = zicoseq_out$p.adj.fdr)

zicoseq_res %>%
  filter(p.adj.fdr < 0.05) %>%
  arrange(p.adj.fdr) %>%
  knitr::kable()
```


```{r plot-zicoseq}
## x-axis is the effect size: R2 * direction of coefficient
ZicoSeq.plot(ZicoSeq.obj = zicoseq_out,
             pvalue.type = 'p.adj.fdr')
```

### PhILR

PhILR is a tree-based method that tests group-wise associations based on
balances. A detailed introduction to this method is available in
[this Bioconductor tutorial](https://www.bioconductor.org/packages/devel/bioc/vignettes/philr/inst/doc/philr-intro.html). 

### Comparison of methods

Although the methods described above yield unidentical results, they are
expected to agree on a few differentially abundant taxa. To draw more informed
conclusions, it is good practice to compare the outcomes of different methods in terms of found features, their effect sizes and significances, as well as other method-specific aspects. Such comparative approach is outlined in
[this exercise](#compare-daa-methods).

## DAA with confounding

Confounders can be defined as variables that are related to and affect the
apparent dynamics between the response and the main independent variable.
They are common in experimental studies. Generally, they can be classified
into 3 groups:

- Biological confounders, such as age and sex 

- Technical confounders produced during sample collection, processing and analysis

- Confounders resulting from experimental models, such as batch effects and sample history

Controlling for confounders is an important practice to reach an unbiased
conclusion. To perform causal inference, it is crucial that the method
is able to include confounders in the model. This is not possible with
statistical tests of general use, such as the Wilcoxon test. In contrast,
methods that target DAA, such as those described in this chapter, allow
controlling for confounders. In the following examples, we will perform DAA
with a main independent variable and a few confounders.

### Selecting confounders

In addition to patient status, we will now control for two confounders: cohort
and library size. The former is a categorical variable with three factors,
whereas the latter is a discrete numerical variable. Remarkably, most DAA
methods accept these two and several other data types.

For demonstration, library size is treated as a confounder and included
in the formulas of the DAA methods. Although this is a satisfactory approach to
control for uneven sequencing efforts across samples, rarefaction generally represents a better solution [@Schloss2023]. With that said, library size can be readily computed and added to the colData.


```{r daa-data-libsize}
# Compute and store library size in colData
colData(tse)$library_size <- colSums(assay(tse, "counts"))
```

### ANCOM-BC

Here, confounders can be added to the formula along with patient status, the
main outcome variable. This way, the model evaluates whether differentially
abundant taxa are associated with one of the variables when the other two
are kept constant.

```{r run-adj-ancombc, warning=FALSE}
# perform the analysis 
ancombc2_out <- ancombc2(tse,
                         assay.type = "counts",
                         fix_formula = "patient_status + cohort + library_size",
                         p_adj_method = "fdr",
                         lib_cut = 0,
                         group = "patient_status", 
                         struc_zero = TRUE, 
                         neg_lb = TRUE,
                         alpha = 0.05,
                         # multi-group comparison is deactivated automatically
                         global = TRUE)
```

In the output, each taxon is assigned with several effect sizes (lfc, which
stands for log-fold change) and adjusted p-values (q). For categorical variables such as patient status and cohort, the statistics indicate whether the abundance
of a given taxon is significantly different between the specified group (column
name) and the reference group (the group that does not appear in the column
names), whereas for numerical variables such as library size, they indicate
whether the abundance of a given taxon varies with that variable.

```{r adj-ancombc-res}
ancombc2_out$res %>%
  dplyr::select(starts_with(c("taxon", "lfc", "q"))) %>%
  arrange(q_patient_statusControl) %>%
  head() %>%
  knitr::kable()
```

### LinDA

As in the previous method, confounders can be included in the formula with the
main outcome variable.

```{r run-adj-linda}
linda_out <- linda(as.data.frame(assay(tse, "counts")),
                   as.data.frame(colData(tse)),
                   formula = "~ patient_status + cohort + library_size",
                   alpha = 0.05,
                   prev.filter = 0,
                   mean.abund.filter = 0)
```

The model returns an output for every variable included in the formula. Normally,
only the results on the main outcome variable are relevant and can be retrieved
as shown below. However, the statistics on the confounders can be similarly
obtained by accessing the corresponding items from the output object.

```{r adj-linda-res}
# Select results for the patient status
linda_res <- linda_out$output$patient_statusControl

linda_res %>%
  filter(reject) %>%
  dplyr::select(log2FoldChange, stat, padj) %>%
  rownames_to_column(var = "feature") %>%
  head() %>%
  knitr::kable()
```

The output shows effect sizes in terms of log-fold changes and a derived
statistic (stat) as well as the corresponding adjusted p-values for differences
in abundance of each taxon between the control and treated group.

### ZicoSeq

For this method, confounders can be added as a list to the `adj.name` argument.

```{r run-adj-zicoseq}
set.seed(123)
zicoseq_out <- ZicoSeq(feature.dat = as.matrix(assay(tse)),
                       meta.dat = as.data.frame(colData(tse)),
                       grp.name = "patient_status",
                       adj.name = c("cohort", "library_size"), 
                       feature.dat.type = "count",
                       return.feature.dat = TRUE,
                       prev.filter = 0,
                       mean.abund.filter = 0,
                       max.abund.filter = 0,
                       perm.no = 999)
```

The output shows the raw and adjusted p-values for clinical status.

```{r adj-zicoseq-res}
zicoseq_res <- cbind.data.frame(p.raw = zicoseq_out$p.raw,
                                p.adj.fdr = zicoseq_out$p.adj.fdr)

zicoseq_res %>%
  filter(p.adj.fdr < 0.05) %>%
  head() %>%
  knitr::kable()
```

## Additional resources

DAA can be performed by several means. Although most of them provide similar
functionality, some may be more suitable than others given a certain study
design or data type. Commonly used DAA tools include:

- ALDEx2 [@Gloor2016]
- ANCOM [@Mandal2015]
- ANCOMBC [@ancombc2020]
- corncob [@Martin2021]
- DACOMP [@Brill2019]
- DESeq2 [@Love2014]
- eBay [@Liu2020]
- edgeR [@Chen2016]
- fastANCOM [@fastANCOM2022]
- LDM [@Hu2020]
- lefser [@Khlebrodova2021]
- limma [@Ritchie2015]
- LinDA [@Zhou2022]
- MaAsLin2 [@Mallick2020]
- metagenomeSeq [@Paulson2017]
- Omnibus [@Omnibus2018]
- RAIDA [@Sohn2015]
- t-test
- Wilcoxon test
- ZicoSeq [@Yang2022]
- ZINQ [@Ling2021]


