# Differential abundance {#sec-differential-abundance}

```{r echo=FALSE, results="asis"}
#| label: setup

library(rebook)
chapterPreamble()
```

An important component of many microbiome studies is Differential Abundance
Analysis (DAA). It is used to identify differences in the abundances of
individual taxa (at any taxonomic level) between two or more groups, such as
treatment versus control groups.

The goal of DAA is to identify biomarkers of a certain phenotype or condition,
and gain understanding of a complex system by looking at its isolated
components. For example, the identification of a bacterial taxon that is more
abundant in healthy patients compared to diseased patients can lead to important
insights into the underlying mechanisms of the disease. In other words,
differentially abundant taxa can be involved in the dynamics of the disease,
which in turn helps to understand the system as a whole.


## Challenges in differential abundance analysis

As discussed in Section [@sec-stat-challenges], microbiome data display unique
properties, such as compositionality and skewed/zero-inflated distributions.
Evidently, these properties need to be addressed properly also in DAA.
Unfortunately, there is currently no consensus on how that should be done.
Instead, there exist dozens of methods employing varying techniques and,
consequently, providing highly discordant results.

The discordancy of results provided by different DAA methods was recently well
illustrated by @Nearing2022. In this study, multiple DAA methods were applied to
38 different datasets and their results were compared to one another. Similar
findings were also made by @Yang2022 who evaluated DAA methods 106 real datasets
(along with simulated data). They found especially that on many datasets almost
any taxon was detected by at least one DAA method (see Figure 7 in their
article). They thus raised concerns of potential cherry-picking, that is, a
researcher is likely to find the taxa in favor of their hypothesis by just
trying out several different DAA methods.


## Recommendations based on recent literature

Usually, when a new DAA method is introduced, the authors demonstrate it to
perform better to some of the existing alternatives. That is often achieved by
employing simulated data and possibly one or more real datasets. The authors may
have, however, selected their simulation models, example datasets, evaluation
metrics, compared methods, etc. specifically to show the introduced method in a
favorable light. Assessing the performance of different methods based solely on
the studies where they were originally introduced may therefore not be
advisable.

A more neutral evaluation of the performance of different DAA methods can be
found from several independent benchmarking studies (e.g. @Weiss2017,
@Hawinkel2019, @Calgaro2020, @Wallen2021, @Nearing2022, @Cappellato2022,
@Roche2022, @Swift2023). Unfortunately, these studies have ended up with very
discordant conclusions on the methods. However, three very recent benchmark
studies employing mostly real datasets, have ended up with somewhat concordant
conclusions.

-	@Wirbel2024 produced realistic test datasets by first creating random
case-control groups on real datasets (from healthy patients) and then implanting
differential abundance signals to such data. They found that only t-test/linear
regression, the Wilcoxon test, limma [@Ritchie2015], and fastANCOM
[@fastANCOM2022] properly controlled false discoveries at relatively high
sensitivity.

-	@Pelto2025 studied how the results from different DAA methods replicate
between random partitions of datasets and between datasets from independent
studies. They found that when considering consistency along with sensitivity,
the best performance was obtained by analyzing relative (TSS normalized)
abundances with a non-parametric method or (after log transformation) with
linear regression/t-test or analyzing presence/absence of taxa with logistic
regression.

-	@Gamboa-Tuz2025 employed three real datasets to benchmark different DAA
methods against biological ground truth. They found that methods employing the
TSS normalization and RNA-seq-derived methods performed best.

In conclusion, it seems that simple methods perform better (or similarly) than
more elaborate methods. An additional benefit of using simple methods is that
they are intuitive and easy to understand. We therefore recommend performing DAA
by analyzing **relative (TSS normalized) abundances by a non-parametric method**
(e.g. ordinal regression model) or, after a log transformation,
by **t-test/linear regression**.

Furthermore, based on recent studies on classification accuracy of healthy vs.
non-healthy patients, it seems that a high proportion of information in
microbiome data is included in the mere **presence/absence** of taxa
(@Giliberti2022 @Karwowska2025). A viable, and even simpler, alternative to DAA
is thus **differential presence/absence analysis** (DPAA, see also @Pelto2025).

Lastly, we note that the performance of a DAA method may depend
on the characteristics of the analyzed data (@Weiss2017). Consequently,
there surely exist cases where some other methods perform better than the
recommended simple methods. If one wants to evaluate which method may perform
best specifically on their data, one option to use the benchdamic package by
@Calgaro2022 (see vignette
[here](https://www.bioconductor.org/packages/release/bioc/vignettes/benchdamic/inst/doc/intro.html)).


## Performing differential abundance and presence/absence analysis

In this section we show how to perform differential abundance or
presence/absence analysis as recommended above. We illustrate the basic case of
comparing abundances or presence/absences of taxa between two experimental
groups. Furthermore, we illustrate how to add covariates in the analysis and
how to examine the interaction effect of two variables.

We use the Tengler2020 dataset from mia package. The dataset includes samples
from people with ADHD (N = 13) and without ADHD (Control group, N = 14). The
people are from three different cohorts.

```{r}
#| label: load_packages_and_data

library(tidyverse)
library(mia)
library(knitr)
library(miaViz)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Show patient status by cohort
table(tse$patient_status, tse$cohort) |>
  kable()
```


#### Preparing the data

Before starting the analysis, it is recommended to reduce the size and
complexity of the data to make the results more reproducible. For this
purpose, we agglomerate the features by genus and filter them by a prevalence
threshold of 10%. Furthermore, we create assays for relative abundances and presence/absence of
taxa. Lastly, we transfrom variable `patient_status` to a factor with `Control`
as the reference level. (It is a standard practice to use the control group as
the reference group in many statistical analyses.)

```{r}
#| label: prepare_data 

# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse, rank = "Genus", prevalence = 0.10)

# Transform count data and create assay of relative abundances
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")

# Transform count data and create assay of presence/abundance of taxa
tse <- transformAssay(tse, assay.type = "counts", method = "pa")

# Transform patient_status to a factor and set "Control" as the refrence level
tse$patient_status <- factor(tse$patient_status, levels = c('Control', 'ADHD'))

# We also transform cohort as a factor
tse$cohort <- factor(tse$cohort, levels = c('Cohort_1', 'Cohort_2', 'Cohort_3'))
```


### Analyzing relative abundances with ordinal regression model {#sec-daa}

We show here how to perform DAA by analyzing relative abundances with ordinal
regression model (ORM). We have chosen ORM as it makes few assumptions on the
distributions of abundances while it can incorporate covariates (including
interaction terms) in the analysis. As ORM performs very similarly to the
familiar non-parametric Wilcoxon test in the basic two group comparison it can
also be seen as an extension of the Wilcoxon test (@Harrell2001, @Liu2017).

By using ORM to model the skewed and/or zero inflated distributions of microbial
abundances, we can avoid the strong parametric assumptions (e.g. negative
binomial distribution) made by some methods. Furthermore, we do not need to
transform abundances. This good because the choice of the transformation may be
somewhat arbitrary while it may affect the DAA results. Moreover, ORM has been
shown to perform well in practice in DAA when considering its sensitivity and
the consistency of its results (@Pelto2025).

We first create a function that (for each taxon) runs ORM and returns the effect
size estimate and p-value (based on score test) for each variable in the model.

```{r}
#| label: function_orm

# The function uses the orm() function from the rms package 
library(rms)

run_orm <- function(abundance, metadata, formula){
  
  # Create the design matrix of the model
  mm <- model.matrix(formula, metadata) |> 
    cbind(abundance) |> 
    tibble::as_tibble() |> 
    dplyr::select(-"(Intercept)")
  
  inds <- 1:(ncol(mm) - 1)
  vars <- colnames(mm)[inds]
  
  # Fit ordinal regression model with all variables
  # (Intercepts are automatically added)
  fit_1 <- rms::orm(abundance ~ ., data = mm)
  score_1 <- fit_1$stats["Score"]
  
  res <- data.frame(estimate = fit_1$coefficients[vars], p_value = NA)
  
  # Calculate p-value based on score test for each variable
  if(length(inds) == 1){
    res$p_value <- fit_1$stats["Score P"]
  }else{
    for(i in inds){
      fit_0 <- rms::orm(abundance ~ ., data = mm[, -i])
      score_0 <- fit_0$stats["Score"]
      res$p_value[i] <- as.numeric(1 - pchisq(score_1 - score_0, df = 1)) 
    }
  }
  
  return(res |> tibble::rownames_to_column("variable"))
}
```



#### The basic two-group comparison {#sec-orm-basic}

We first perform the basic two-group comparison without any other covariates.
We compare the relative abundance of taxa between the people with and without
ADHD (indicated by the variable `patient_status`).

The code below works as follows.

 1. We extract the assay with relative abundances from the `tse` object
 2. The assay is transposed and transformed to a `data.frame`
 3. The function `run_orm` is run for each taxon
    i) The `.` indicates here the relative abundance of each taxon
    ii) The metadata including the variable `patient_status` is extracted from
    the `tse` object
    iii) The model formula indicates that we included only `patient_status` in
    our analysis
 4. The results for all taxa are binded to as one `data.frame`
 5. Lastly, we calculate p-values adjusted from multiple testing (`q-value`) by
 using the standard FDR correction by @Benjamini1995 (`method = "BH"`).


```{r}
#| label: run_daa_basic

res_daa_basic <- assay(tse, "relabundance") |>  # Extract relative abundances
  t() |>                                        # Transpose the assay
  as.data.frame() |>                            # Transform to a data.frame
  map(~ run_orm(                                # Run ORM for each taxon
    .,                                            
    metadata = colData(tse) |> as.data.frame(), # Extract the metadata
    formula = ~ patient_status)                 # Model formula. Now only 
                                                # patient_status
    ) |> 
  bind_rows(.id = "taxon") |>                   # Bind the results of all taxa
  mutate(
    q_value = p.adjust(p_value, method = "BH")) # Calculate q-values
                                                # (FDR adjusted p-values)
```

Next, we examine the results ordered by p-value.

```{r}
#| label: print_daa_basic

# The results for taxa with the smallest p-values
res_daa_basic |> arrange(p_value) |> head() |> kable()
```

In the output

- `variable` indicates the predictor variable. As `patient_status` is a
categorical variable, it also tells the group used as the non-reference level
(here `ADHD` as `Control` was set as the reference level).
- `estimate` is log(odds ratio). Most importantly, positive values indicate
abundance being higher in the non-reference group (here `ADHD`). Otherwise,
estimates can be a bit difficult to interpret for ORMs. 

If we use the standard criterion for statistical significance, namely q-value <
0.05, we can see that the difference is statistically significant only for the
genus named as [Ruminococcus]_gauvreauii_group. As the sign of the estimate is
_negative_ it means that the relative abundance is _higher in the control
group_.

This difference can be illustrated by a strip plot with boxplots. From those we
can clearly see how the relative abundances are higher in the control group.
 
```{r}
#| label: plot_daa_basic

# We use the function plotBoxpplot to illustrate the result for
# [Ruminococcus]_gauvreauii_group
plotBoxplot(tse,
            assay.type = "relabundance",
            features = "[Ruminococcus]_gauvreauii_group",
            x = "patient_status",
            colour.by = "patient_status",
            fill.by = "patient_status",
            apply.beeswarm = T)
```


#### DAA with additional covariates {#sec-orm-cov}

Here we show how to perform DAA with an additional covariate included in the
analysis. One usually wants to include a variable as a covariate if they suspect
that it may act as a _confounding_ variable, namely, a variable that causes a
spurious association between the variable of main interest and the abundances.

We use `cohort` as a covariate. We proceed as above except that we now have
`formula = ~ patient_status + cohort`. Furthermore, as there are now many
variables in the model we modify the code so that the multiplicity correction
for the p-values is made separately for each variable.  


```{r}
#| label: run_daa_cov

# Run orm for each taxon with cohort as an additional covariate
res_daa_cov <- assay(tse, "relabundance") |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_orm(
    .,
    metadata = colData(tse) |> as.data.frame(),
    formula = ~ patient_status + cohort)
  ) |> 
  bind_rows(.id = "taxon") |> 
  group_by(variable) |>    # p-values FDR adjusted separately for each variable.    
  mutate(q_value = p.adjust(p_value, method = "BH")) |> 
  ungroup()

res_daa_cov |> head() |> kable()
```

Note that there now appear results also for the variable `cohort`. As `cohort`
consists of three groups, namely, `Cohort_1` (the reference group), `Cohort_2`
and `Cohort_3`, there are results separately for the comparison between
`Cohort_2` vs. `Cohort_1` (`cohortCohort_2`) and for the comparison between
`Cohort_3` vs. `Cohort_1` (`cohortCohort_3`).

We can now examine how the inclusion of `cohort` as a covariate affects the
results for `patient_status`.

```{r}
#| label: print_daa_cov

# The results for patient_status for the taxa with the smallest p-values
res_daa_cov |> 
  filter(variable == "patient_statusADHD") |> 
  arrange(p_value) |> 
  head() |> 
  kable()
```

We can see that the results for [Ruminococcus]_gauvreauii_group and
_Erysipelatoclostridium_ are in fact clearer (e.g. smaller p-values) when
cohort is controlled for.

By examining strip plots by `patient_status` and `cohort`, we can see that
the abundances are overall higher in cohort 3, and therefore the difference
between the patients with and without ADHD seems clearer in cohorts 1 and 2.
This seems to explain the difference in the results compared to the basic case
(@sec-orm-basic).

```{r}
#| label: plot_daa_cov

# Plot abundance of [Ruminococcus]_gauvreauii_group by patient_status and cohort
plotBoxplot(tse,
            assay.type = "relabundance",
            features = "[Ruminococcus]_gauvreauii_group",
            x = "patient_status",
            colour.by = "patient_status",
            fill.by = "patient_status",
            facet.by = "cohort",
            apply.beeswarm = T)
```


#### DAA with an interaction {#sec-orm-ia}

Here we show a more complex version of DAA, namely, a case where the interaction
effect of two variables is analyzed. We give this example here for illustration
purposes but note that such complex analysis may not be advisable with a small
dataset such as our example dataset.

We analyze interaction between `patient_status` and `cohort`. We proceed as in
@sec-orm-cov above except that we now add also the interaction term into the
formula. We could thus write
`formula = ~ patient_status + cohort + patient_status:cohort` but we recommend
using the write the shorter version `formula = ~ patient_status * cohort`.
Using the `*` between the terms automatically adds the interaction term as well
as the main effect terms in the analysis. 

```{r}
#| label: run_daa_ia

# Run orm with interaction term for each taxon
res_daa_ia <- assay(tse, "relabundance") |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_orm(
    .,
    metadata = colData(tse) |> as.data.frame(),
    formula = ~ patient_status * cohort)
  ) |> 
  bind_rows(.id = "taxon") |> 
  group_by(variable) |>                                  
  mutate(q_value = p.adjust(p_value, method = "BH")) |> 
  ungroup()

res_daa_ia |> head() |> kable()
```

Note that the results object `res_daa_ia` now also includes the interaction
terms `patient_statusADHD:cohortCohort_2` and
`patient_statusADHD:cohortCohort_3`. 

We examine the term `patient_status:cohortCohort_3`. It indicates how the
differences between the ADHD and Control groups vary between cohorts 1 and 3.

```{r}
#| label: print_daa_ia

res_daa_ia |> 
  filter(variable == "patient_statusADHD:cohortCohort_3") |> 
  arrange(p_value) |> 
  head() |> 
  kable()
```

The smallest p-value is observed for _Coprobacter_. By examining the boxplots
of the relative abundances of _Coprobacter_ by `patient_status` and `cohort`, we
can see the interaction effect by observing that in cohort 1 the relative
abundance is higher in the ADHD group while in cohort 3 the abundance is higher
in the control group.

```{r}
#| label: plot_daa_ia

# Strip plots with boxplots for Coprobacter by patient_status and cohort
plotBoxplot(tse,
            assay.type = "relabundance",
            features = "Coprobacter",
            x = "patient_status",
            colour.by = "patient_status",
            fill.by = "patient_status",
            facet.by = "cohort",
            apply.beeswarm = T)
```


### Analyzing presence/absence of taxa with logistic regression

We show here how to perform differential presence/absence analysis (DPAA) with
logistic regression. As some versions of logistic regression perform badly in
the cases where, for instance, the prevelance of a taxon is zero in one
group we employ a robust version of logistic regression introduced by
@Firth1993. This version has been observed to perform well in DPAA in practice
[@Pelto2025].

It is also important to note that we will analyze the _observed_
presence/absence of taxa (i.e. observed non-zero/zero count) and not the
presence/absence of taxa in their ecosystem. Those concepts are not the same
because a taxon that is present in an ecosystem may be left undetected if, for
instance, its abundance is below the detection limit of the experimantal
workflow used in processing/analyzing the microbial samples [@Silverman2020].

We proceed as in the previous section @sec-daa. We first create a function that
runs logistic regression for presence/absence and returns the
effect size estimate and p-value for each variable in the model.

```{r}
#| label: function_logr

# The function uses the logistf() function from the logistf package 
library(logistf)

run_logr <- function(abundance, metadata, formula){
  
  # Create the design matrix of the model
  mm <- model.matrix(formula, metadata) |> 
    cbind(abundance) |> 
    tibble::as_tibble() |> 
    dplyr::select(-"(Intercept)")
  
  # Fit Firth logistic regression model
  fit <- logistf::logistf(abundance ~ ., data = mm)
  
  res <- data.frame(estimate = fit$coefficients,
                    p_value = fit$prob)
  
  return(res[-1, ] |> tibble::rownames_to_column("variable"))
}
```


#### The basic two-group comparison for presence/absence {#sec-logr-basic}

We first perform the basic two-group comparison without any other covariates.
We compare the presence/absence of taxa between the people with and without
ADHD.


```{r}
#| label: run_dpaa_basic

res_dpaa_basic <- assay(tse, "pa") |>           # Extract presence/absences
  t() |>                                        # Transpose the assay
  as.data.frame() |>                            # Transform to a data.frame
  map(~ run_logr(                               # Run ORM for each taxon
    .,                                            
    metadata = colData(tse) |> as.data.frame(), # Extract the metadata
    formula = ~ patient_status)                 # Model formula.
    ) |> 
  bind_rows(.id = "taxon") |>                   # Bind the results of all taxa
  mutate(
    q_value = p.adjust(p_value, method = "BH")) # Calculate q-values
                                                # (FDR adjusted p-values)
```

Next, we again examine the results sorted by p-value.

```{r}
#| label: print_dpaa_basic

# The results for taxa with the smallest p-values
res_dpaa_basic |> arrange(p_value) |> head() |> kable()
```

Now the `estimate` in the output is the familiar log(odds ratio) of logistic
regression. A positive value indicates the taxon being more prevalent in the
non-refrence group (here `ADHD`).

The smallest p-value and q-value is observed for genus _Erysipelatoclostridium_.
As the sign of the estimate is _negative_ is means that the genus is more
prevalent in the control group. This difference can be illustrated using
`plotBoxplot` function with the proportion bars added to specifically illustrate
the prevalence. From there we can clearly see that _Erysipelatoclostridium_ is
more prevalent in the control group.
 
```{r}
#| label: plot_dpaa_basic

# Illustrate the prevalence of Erysipelatoclostridium using plotBoxplot with
# proportion bar
plotBoxplot(tse,
            assay.type = "relabundance",
            features = "Erysipelatoclostridium",
            x = "patient_status",
            colour.by = "patient_status",
            fill.by = "patient_status",
            apply.beeswarm = T,
            add.proportion = T)
```


#### DPAA with additional covariates {#sec-logr-cov}

Here we show how to perform DPAA with an additional `cohort` included in the
analysis. We proceed as in the section @sec-orm-cov.

```{r}
#| label: dpaa_cov

# Run logistic regression for each taxon with cohort as an additional covariate
res_dpaa_cov <- assay(tse, "pa") |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_logr(
    .,
    metadata = colData(tse) |> as.data.frame(),
    formula = ~ patient_status + cohort)
  ) |> 
  bind_rows(.id = "taxon") |> 
  group_by(variable) |>                        # FDR adjust p-values separately          
  mutate(                                      # for each variable.
    q_value = p.adjust(p_value, method = "BH")
    ) |> 
  ungroup()

# The results for patient_status for the taxa with the smallest p-values
res_dpaa_cov |> 
  filter(variable == "patient_statusADHD") |> 
  arrange(p_value) |> 
  head() |> 
  kable()

# Strip plots with proportion bars to illustrate the revelence of
# Erysipelatoclostridium by patient_status and cohort
plotBoxplot(tse,
            assay.type = "relabundance",
            features = "Erysipelatoclostridium",
            x = "patient_status",
            colour.by = "patient_status",
            fill.by = "patient_status",
            facet.by = "cohort",
            apply.beeswarm = T,
            add.proportion = T)
```


#### DPAA with an interaction {#sec-logr-ia}

Again, we proceed similarly as in the section @sec-orm-ia.

```{r}
#| label: dpaa_ia

# Run logistic regression with an interaction term for each taxon
res_dpaa_ia <- assay(tse, "pa") |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_logr(
    .,
    metadata = colData(tse) |> as.data.frame(),
    formula = ~ patient_status * cohort)
  ) |> 
  bind_rows(.id = "taxon") |> 
  group_by(variable) |>                                  
  mutate(                                      
    q_value = p.adjust(p_value, method = "BH")
    ) |> 
  ungroup()

# Interaction results for patient_statusADHD:cohortCohort_3
res_daa_ia |> 
  filter(variable == "patient_statusADHD:cohortCohort_3") |> 
  arrange(p_value) |> 
  head() |> 
  kable()

# Cerate strip plots with proportion bars to illustrate the prevalence of
# Erysipelatoclostridium by patient_status and cohort
plotBoxplot(tse,
            assay.type = "relabundance",
            features = "Coprobacter",
            x = "patient_status",
            colour.by = "patient_status",
            fill.by = "patient_status",
            facet.by = "cohort",
            apply.beeswarm = T,
            add.proportion = T)
```

The figure illustrates how _Coprobacter_ is more prevalent in the ADHD group in
cohort 1 but more prevalent in the control group in cohort 3.


### Analyzing positive relative abundances with linear regression {#sec-daa_lm}

Relative abundances can also be analyzed using the standard linear regression
methods (e.g. the t-test) instead of ORM. This, however, requires in practice
some transformation to be applied for the abundances. A typical transformation
is log transformation. It is, however, problematic because zeros cannot be log
transformed. Instead, zeros must be treated somehow before the transformation
(e.g. by using pseudo-count).

Analyzing log transformed relative abundances can be advantageous, however, for
instance in the case where one wants to analyze only _positive_ (non-zero)
relative abundances. Such analysis can sometimes be biologically relevant,
especially when performed along with DPAA.

It should be noted, however, that especially in small datasets (like the
example dataset used here), the number of positive/non-zero abundances may be
very small for some taxa. Therefore, this type of analysis may not always be
reasonable. Furthermore, we note again that one should keep in mind that the
observed absence of a taxon does not necessarily mean that the taxon
is truly absent in a microbial sample.


#### The basic two-group comparison {#sec-daa-pos-basic}

Analysis of positive relative abundances can be easily performed by using e.g.
maaslin3 package [@Nickols2024].

```{r}
#| label: run_daa_pos_basic

library(maaslin3)

# We first perform DAA with maaslin3 function
obj_maaslin3_basic <- maaslin3(tse,
                               output = 'output',
                               formula = ~ patient_status,
                               median_comparison_abundance = FALSE,
                               verbosity = 'ERROR',
                               plot_associations = FALSE,
                               plot_summary_plot = FALSE)

# We then extract and transform the results to the same form as before 
res_daa_pos_basic <- obj_maaslin3_basic$fit_data_abundance$results |> 
  select(taxon = feature,
         variable = name,
         estimate = coef,
         p_value = pval_individual,
         q_value = qval_individual)
```

Then, we again examine the results ordered by p-value.

```{r}
#| label: print_daa_pos_basic

# The results for taxa with the smallest p-values
res_daa_pos_basic |> arrange(p_value) |> head() |> kable()
```


Plotting the abundance of _Coprobacter_ illustrates how the positive (non-zero)
relative abundances are clearly higher in the ADHD group. (We omit the boxplots
this time as they are affected by the zero values.)

```{r}
#| label: plot_daa_pos_basic

# Plot the result for Coprobacter
plotBoxplot(tse,
            assay.type = "relabundance",
            features = "Coprobacter",
            x = "patient_status",
            colour.by = "patient_status",
            fill.by = "patient_status",
            add.box = FALSE,
            apply.beeswarm = T)
```


#### Analysis with covariates {#sec-daa-pos-cov}

To include cohort as a covariate in the analysis, we proceed as above but we
just set `formula = ~ patient_status + cohort`. We also calculate the q-values
manually as `maaslin3` calculates them over all variables, and not separately
for each variable. The latter method is more sensible in most cases as one is
usually mainly interested in their "main variable", i.e., `patient_status`
here.

```{r}
#| label: daa_pos_cov

# Run maaslin3 function.
obj_maaslin3_cov <- maaslin3(tse,
                             output = 'output',
                             formula = ~ patient_status + cohort,
                             median_comparison_abundance = FALSE,
                             verbosity = 'ERROR',
                             plot_associations = FALSE,
                             plot_summary_plot = FALSE)

# Extract the results and calculate q-values 
res_daa_pos_cov <- obj_maaslin3_cov$fit_data_abundance$results |> 
  select(taxon = feature,
         variable = name,
         estimate = coef,
         p_value = pval_individual) |> 
  group_by(variable) |> 
  mutate(q_value = p.adjust(p_value, method = 'BH')) |> 
  ungroup()

# The results for patient_status for the taxa with the smallest p-values
res_daa_pos_cov |> 
  filter(variable == "patient_statusADHD") |> 
  arrange(p_value) |> 
  head() |> 
  kable()
```

```{r}
#| label: plot_daa_pos_cov

# Plot the result for Coprobacter
plotBoxplot(tse,
            assay.type = "relabundance",
            features = "Coprobacter",
            x = "patient_status",
            colour.by = "patient_status",
            fill.by = "patient_status",
            facet.by = "cohort",
            add.box = FALSE,
            apply.beeswarm = T,
            add.proportion = T)

```



#### Analysis with interaction {#sec-daa-pos-ia}

The analysis of positive relative abundances with interaction effect can be
performed as follows. (The interaction analysis is performed here only for
illustration purposes, but we note that it may not be advisable to conduct
such a complex analysis when the number of non-zero abundances is very low for
many taxa.) 

```{r}
#| label: daa_pos_ia

# Run maaslin3 function.
obj_maaslin3_ia <- maaslin3(tse,
                             output = 'output',
                             formula = ~ patient_status * cohort,
                             median_comparison_abundance = FALSE,
                             verbosity = 'ERROR',
                             plot_associations = FALSE,
                             plot_summary_plot = FALSE)

# Extract the results and calculate q-values 
res_daa_pos_ia <- obj_maaslin3_ia$fit_data_abundance$results |> 
  select(taxon = feature,
         variable = name,
         estimate = coef,
         p_value = pval_individual) |> 
  group_by(variable) |> 
  mutate(q_value = p.adjust(p_value, method = 'BH')) |> 
  ungroup()

# The results for the interacion between patient_status and cohort (when
# comparing Cohort_3 to Cohort_1) for the taxa with the smallest p-values
res_daa_pos_ia |> 
  filter(variable == "patient_statusADHD:cohortCohort_3") |> 
  arrange(p_value) |> 
  head() |> 
  kable()
```

The result was not even close to significant for any taxa. This is likely due to
the comlex nature of interaction anlaysis and the very small number of non-zero
abundances for many taxa. 


::: {.callout-tip}
## Additional resources
While we generally recommend the use of the above simple methods in DAA there
exist dozens of alternative methods. A (non-comprehensive) list of these methods
is given below.

- ALDEx2 [@Gloor2016]
- ANCOM [@Mandal2015]
- ANCOMBC [@ancombc2020]
- BIRDMAn [@Rahman2023]
- corncob [@Martin2021]
- DACOMP [@Brill2019]
- DESeq2 [@Love2014]
- eBay [@Liu2020]
- edgeR [@Chen2016]
- fastANCOM [@fastANCOM2022]
- fido [@Silverman2019]
- LDM [@Hu2020]
- lefser [@Khlebrodova2021]
- limma [@Ritchie2015]
- LinDA [@Zhou2022]
- MaAsLin2 [@Mallick2020]
- MaAsLin3 [@Nickols2024]
- metagenomeSeq [@Paulson2017]
- Omnibus [@Omnibus2018]
- RAIDA [@Sohn2015]
- sccomp [@mangiola2023]
- ZicoSeq [@Yang2022]
- ZINQ [@Ling2021]
:::

::: {.callout-tip icon=false}

## Exercises

**Goal:** The goal of these exercises it to learn how to perform differential
abundance and presence/absence analysis and to visualize the results.

**Exercise 1: Differential abundance analysis (DAA)**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Observe `colData`. Ensure that the data includes meaningful sample grouping.
If it does not, choose another dataset.

3. Agglomerate the data by genus and filter by a prevalence of 10%.

4. Run the basic two-group DAA and DPAA using ORM and logistic regression,
respectively. (You can also try DAA of positive abundances with maaslin3.)

5. Explore the results tables. Are there any taxa with significant
abundance/prevalence differences?

6. Run the same analyses now with an additional covariate. Do the results
differ? Why?

7. You can try also more complex analysis, e.g. with several covariates and/or
with interactions.

8. Visualize the results.

Useful functions:

`data()`, `colData`, `subsetByPrevalent()`

:::

