# Differential abundance {#sec-differential-abundance}

```{r}
#| label: setup
#| echo: false
#| results: asis

library(rebook)
chapterPreamble()
```

An important component of many microbiome studies is differential abundance
analysis (DAA). It is mainly used to identify differences in the abundances of
individual taxa (at any taxonomic level) between two or more groups, such as
treatment versus control groups.

The goal of DAA is thus to identify biomarkers of a certain phenotype or
condition, and gain understanding of a complex system by looking at its isolated
components. For example, the identification of a bacterial taxon that is more
abundant in healthy patients compared to diseased patients can lead to important
insights into the underlying mechanisms of the disease. In other words,
differentially abundant taxa can be involved in the dynamics of the disease,
which in turn helps to understand the system as a whole.

## Challenges in differential abundance analysis

As discussed in @sec-stat-challenges, microbiome data display unique
properties, such as compositionality and skewed/zero-inflated distributions.
Evidently, these properties need to be addressed properly also in DAA.
Unfortunately, there is currently no consensus on how that should be done.
Instead, there exist dozens of methods employing varying techniques and,
consequently, providing highly discordant results.

The discordancy of results provided by different DAA methods was recently well
illustrated by @Nearing2022. In this study, multiple DAA methods were applied to
38 different datasets and their results were compared to one another. Similar
findings were also made by @Yang2022 who evaluated DAA methods 106 real datasets
(along with simulated data). They found especially that on many datasets almost
any taxon was detected by at least one DAA method (see Figure 7 in their
article). They thus raised concerns of potential cherry-picking, that is, a
researcher would be likely to find the taxa in favour of their hypothesis by just
trying out several different DAA methods.

## Recommendations based on recent literature {#sec-daa-recommendations}

Usually, when a new DAA method is introduced, the authors demonstrate it to
perform better compared to some of the existing alternatives. That is often
achieved by employing simulated data and possibly one or more real datasets.
The authors may have, however, selected their simulation models,
example datasets, evaluation metrics, compared methods, etc. specifically to
show the introduced method in a favourable light. Assessing the performance of
different methods based solely on the studies where they were originally
introduced may therefore not be advisable.

A more neutral evaluation of the performance of different DAA methods can be
found from several independent benchmarking studies (e.g. @Weiss2017,
@Hawinkel2019, @Calgaro2020, @Wallen2021, @Nearing2022, @Yang2022,
@Cappellato2022, @Roche2022, @Swift2023, @Wirbel2024, @Pelto2025,
@Gamboa-Tuz2025). Unfortunately, these studies have ended up with very
discordant conclusions on the methods. However, if we look specifically at some
benchmarking studies that have used only real datasets (instead of simulated
data), we can see that analysis of relative abundances (TSS normalized counts)
with a simple non-parametric method or linear regression has
performed well.

- @Nearing2022 studied the performance of DAA methods using 38 datasets. They
evaluated the number of detected taxa (ASVs), consistency of the detected taxa
between methods, replication of results between studies and error rate on
datasets with randomly assigned groups. They emphasized conservatism
(low error rate) in their recommendations and ended up recommending
`r BiocStyle::Biocpkg("ALDEx2")` [@Fernandes2014] and
`r BiocStyle::Biocpkg("ANCOMBC")` [@Mandal2015]. Additionally, they recommended
`r BiocStyle::Biocpkg("Maaslin2")` [@Mallick2020] (i.e. analyzing arcsine
square root transformed relative abundances with linear regression) as a more
powerful option.

- In addition to simulations, @Yang2022 estimated the FDR of different DAA
methods on 106 real datasets with randomly created groups. They found that,
the observed FDR was at the nominal level when, for instance, analyzing
relative abundances either with the Wilcoxon test or with linear regression
(`r BiocStyle::Biocpkg("Maaslin2")`).

-	@Pelto2025 studied how the results from different DAA methods replicate
between random partitions of datasets and between datasets from independent
studies. They found that when considering consistency along with sensitivity,
the best performance was obtained by analyzing relative abundances with a
non-parametric method or (after log transformation) with linear
regression/t-test (`r BiocStyle::Biocpkg("Maaslin2")`) or analyzing
presence/absence of taxa with logistic regression.

-	@Gamboa-Tuz2025 employed three real datasets to benchmark different DAA
methods against biological ground truth. They found that methods employing the
TSS normalization (with e.g. the Wilcoxon test) and RNA-seq-derived methods
performed best.

We therefore recommend performing DAA by analyzing
**relative (TSS normalized) abundances by a non-parametric method**
(e.g., the Wilcoxon test or ordinal regression model) or, after a log
transformation, with **t-test/linear regression**. In addition to their
good performance on real data they have the benefit of being **intuitive and
easy to understand**.

We note, however, that the performance of DAA methods likely depends
on the characteristics of the analyzed data (@Weiss2017, @Calgaro2020). 
Consequently, while the methods recommended above seem to be good general
default methods, they may not necessarily be the optimal methods in many cases.
If one wants to evaluate which method may perform best specifically on their
data, one option to use the `r BiocStyle::Biocpkg("benchdamic")` package
by @Calgaro2022 (see vignette
[here](https://www.bioconductor.org/packages/release/bioc/vignettes/benchdamic/inst/doc/intro.html)).

Lastly, based on recent studies on classification accuracy of healthy vs.
non-healthy patients, it seems that a high proportion of information in
microbiome data is included in the mere **presence/absence** of taxa
(@Giliberti2022 @Karwowska2025). A viable, and even simpler, alternative to DAA
is thus **differential prevalence analysis** (see also @Pelto2025).

## Performing differential abundance analysis {#sec-daa}

Following the recommendations above, we show in this section how to perform DAA
by analyzing (log transformed) relative abundances with linear regression
models. We first illustrate the analysis in the simplest case where
two groups are compared, and no other covariates are included in the analysis.
After that we illustrate more complex analyses, namely, the analysis with
additional covariate(s) included and the analysis with interaction of two
variables. Furthermore, we illustrate how to perform differential prevalence
analysis and the differential analysis of positive (non-zero) abundances.

### Preparing the data {#sec-prepare-data}

We use the Tengeler2020 dataset from `r BiocStyle::Biocpkg("mia")` package.
The dataset includes samples from people with ADHD (N = 13) and without ADHD
(Control group, N = 14). The people are from three different cohorts.

```{r}
#| label: load_packages_and_data

# Load the needed packages
library(mia)
library(knitr)
library(dplyr)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Show patient status by cohort
table(tse$patient_status, tse$cohort) |>
    kable()
```

Before starting the analysis, we agglomerate the features by genus and filter
them by a prevalence threshold of 10%.
Furthermore, we create assays for relative abundances and
presence/absence of taxa. Lastly, we transform variable `patient_status` to a
factor with `Control` as the reference level. (It is a standard practice to use
the control group as the reference group in many statistical analyses.)

```{r}
#| label: prepare_data

# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse, rank = "Genus", prevalence = 0.10)

# Transform count data and create assay of relative abundances
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")

# Transform count data and create assay of presence/absence of taxa
tse <- transformAssay(tse, assay.type = "counts", method = "pa")

# Transform patient_status to a factor and set "Control" as the reference level
tse$patient_status <- factor(tse$patient_status, levels = c("Control", "ADHD"))

# We also transform cohort as a factor
tse$cohort <- factor(tse$cohort, levels = c("Cohort_1", "Cohort_2", "Cohort_3"))
```

### Simple two-group comparison {#sec-daa-basic}

Here we perform the simplest version of DAA, namely, the comparison of two
groups without any other covariates included in the analysis.
In our case this means comparing the relative abundances of genera between the
people with and without ADHD.

We employ the `Maaslin2` function from the `r BiocStyle::Biocpkg("Maaslin2")`
package [@Mallick2020]. The function

- transforms counts to relative abundances,
- replaces zeros by half of the smallest non-zero relative abundance,
(separately for each taxon),
- applies log~2~ transformation to the relative abundances, and
- runs a linear regression model for each taxon using the transformed relative
abundances as the response variable.

To run the function we need to provide it with the count matrix and metadata as
inputs. We also need to provide it with the variable indicating the compared 
groups, i.e. `patient_status` in our case. Furthermore, we need to specify the
folder for some output files. Lastly, we prevent `Maaslin2()` to create certain
plots. (Note that `Maaslin2()` prints a lot of some output
which is suppressed here for convenience.)

```{r}
#| label: run_daa_basic
#| results: hide

library(Maaslin2)

obj_daa_basic <- Maaslin2(
    input_data = assay(tse), # The count matrix
    input_metadata = colData(tse) |> as.data.frame(), # The metadata
    fixed_effects = "patient_status", # The predictor variable

    output = "output", # The name of the folder for output files.
    plot_heatmap = FALSE,
    plot_scatter = FALSE
)
```

Next, we extract the `data.frame` of results from the list object returned by
`Maaslin2()`. We also calculate the 95% confidence intervals and transform the
results to a slightly more convenient format. (The sample size `N` is used as
an approximation for the residual degrees of freedom (`df`) in the calculation
of confidence intervals.)

```{r}
#| label: results_daa_basic

# Extract the results from the list object and calculate the 95% CIs
res_daa_basic <- obj_daa_basic$results |>
    mutate(
        ci_lwr = coef - qt(.975, df = N) * stderr,
        ci_upr = coef + qt(.975, df = N) * stderr
    ) |>
    select(feature, name, coef, pval, qval, ci_lwr, ci_upr)

# Print the results for taxa with the smallest p-values
res_daa_basic |>
    arrange(pval) |>
    head() |>
    kable()
```

The columns of `res_daa_basic` are as follows:

- `feature` indicates the name of a taxon (here genus).
- `name` indicates the predictor variable. As `patient_status` is a
categorical variable, it also tells the group used as the _non_-reference level
(here `ADHD` as `Control` was set as the reference level).
- `coef` is the log~2~-fold change estimate, a standard way to express effect
size in DAA.
Most importantly, positive values indicate abundance being higher in the
non-reference group (here `ADHD`). More precisely, `2 ^ coef` would be
interpreted as the ratio of geometric means of the relative abundances
between the ADHD and Control groups.
- `pval` is the unadjusted p-value.
- `qval` (q-value) is the FDR adjusted p-value. This is generally used to define
statistical significance. Typically, q-values < .05 are considered
(statistically) significant.
- `ci_lwr` and `ci_upr` are the lower and upper bounds of the 95% confidence 
intervals for the log~2~-fold change.

Now, if we use the standard criterion for statistical significance, namely,
q-value < 0.05, we can see that the DAA result is statistically significant for
the genera named as _[Ruminococcus] gauvreauii group_, _Faecalibacterium_,
_Catabacter_, _[Clostridium] innocuum group_ and _Erysipelatoclostridium_.
(`r BiocStyle::Biocpkg("Maaslin2")` has some problems with special characters
such as `[` or `]`.)
Moreover, for instance, the positive log~2~-fold change estimate of
_Faecalibacterium_ tells that the average relative abundance of
_Faecalibacterium_ is higher in the ADHD group.

#### Visualizing the results {#sec-daa-viz}

While the numbers above provide summarized information about the group
differences for each taxon, it is often useful to further examine these
associations in more detail. That can be done, for instance, using strip
plots with boxplots. We show here how such plots can be created for the
significant taxa by using `plotBoxplot` function from
the `r BiocStyle::Biocpkg("miaViz")` package.

```{r}
#| label: daa-viz-1

library(miaViz)

# Create strip plots with boxplots (and prevalence bars)
plotBoxplot(
    tse,
    features = c(
        "[Ruminococcus]_gauvreauii_group",
        "Faecalibacterium",
        "Catabacter",
        "Erysipelatoclostridium",
        "[Clostridium]_innocuum_group"
    ),
    assay.type = "relabundance", # To plot relative abundances
    x = "patient_status", # The variable on the x (horizontal) axis
    facet.by = "rownames", # To create a separate facet for each feature
    scales = "free_y", # To allow different y-axis scales
    add.proportion = TRUE, # Add bars indicating prevalence
    point.offset = "quasirandom", # To plot points nicely
    point.size = 2, # The size of the points
    point.color = "black", # The color of the edges of the points
    point.alpha = 1, # The transparency of the edges of the points
    box.alpha = .20
) + # The transparency of the boxplots
    theme(strip.text = element_text(size = 7)) # The size of the facet labels
```

In each facet,

- the thicker horizontal line in each boxplot indicates the median,
- each box indicates the interquartile range (IQR), namely, the range between
the 25th (Q1) and 75th percentile (Q3),
- the lower whisker indicates the smallest value above $Q1 - 1.5 \times IQR$,
and
- the lower whisker indicates the largest value below $Q3 + 1.5 \times IQR$.
- The horizontal bars at the bottom indicate the prevalence of each genus in
each group.

The strip plots and boxplots clearly show, for instance, how the relative
abundance of _Faecalibacterium_ is higher in the ADHD group, as was indicated
by the positive the log~2~-fold estimate (`coef`) in the results table above.
Moreover, for instance, the prevalence bars of _Erysipelatoclostridium_
show how _Erysipelatoclostridium_ is more prevalent in the control group.

Lastly, we show how the results for all genera can be summarized visually
by plotting the log~2~-fold estimates and their 95% confidence
intervals.

```{r}
#| label: plot_daa_est_ci

# First arrange the genera by log2-fold estimate
top_features <- res_daa_basic[order(res_daa_basic[["coef"]], decreasing = TRUE), "feature"]
res_daa_basic[["genus_ord"]] <- factor(res_daa_basic[["feature"]], levels = top_features)

# Plot the log2-fold changes and the 95% confidence intervals
ggplot(res_daa_basic, aes(x = coef, y = genus_ord)) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_point() +
    geom_errorbarh(aes(xmin = ci_lwr, xmax = ci_upr), height = 0.2) +
    labs(x = "Log2-fold change (95% CI)") +
    theme_light() +
    theme(
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = 7),
        plot.title = element_text(hjust = 0.5)
    )
```

### DAA with additional covariates {#sec-daa-cov}

Here we show how to perform DAA with an additional covariate included in the
analysis. One usually wants to include a variable as a covariate if they suspect
that it may act as a _confounding_ variable, namely, a variable that causes a
spurious association between the variable of main interest and the abundances.

We use here `cohort` as a covariate. We proceed as above except that we now set
`fixed_effects = c("patient_status", "cohort")`. Furthermore, we now calculate
the q-values manually so that they are calculated separately for each variable
in the model (indicated by the variable `metadata` in `obj_daa_cov$results`).
This is because `Maaslin2()` calculates the q-values over all
variables in the model which is not what here as we are only interested in the
variable `patient_status`.

```{r}
#| label: daa_cov
#| results: hide

# Run Maaslin2 with patient_status and cohort as predictor variables
obj_daa_cov <- Maaslin2(
    input_data = assay(tse),
    input_metadata = colData(tse) |> as.data.frame(),
    fixed_effects = c("patient_status", "cohort"),
    output = "output",
    plot_heatmap = FALSE,
    plot_scatter = FALSE
)

# Extract the results and calculate the 95% CIs and q-values
res_daa_cov <- obj_daa_cov$results |>
    mutate(
        ci_lwr = coef - qt(.975, df = N) * stderr,
        ci_upr = coef + qt(.975, df = N) * stderr
    ) |>
    group_by(metadata) |>
    mutate(qval = p.adjust(pval, method = "fdr")) |>
    ungroup() |>
    select(feature, name, coef, pval, qval, ci_lwr, ci_upr)
```

```{r}
#| label: print_daa_cov_1

# Print a sample of the results
res_daa_cov |>
    arrange(feature, name) |>
    head() |>
    kable()
```

Note that there now appear results also for the variable `cohort`. As `cohort`
consists of three groups, namely, `Cohort_1` (the reference group), `Cohort_2`
and `Cohort_3`, there are results separately for the comparison between
`Cohort_2` vs. `Cohort_1` (`cohortCohort_2`) and for the comparison between
`Cohort_3` vs. `Cohort_1` (`cohortCohort_3`).

We can now examine how the inclusion of `cohort` as a covariate affects the
results for `patient_status`.

```{r}
#| label: print_daa_cov_2

# Print the results for taxa with the smallest p-values for patient_status
res_daa_cov |>
    filter(name == "patient_statusADHD") |>
    arrange(pval) |>
    head() |>
    kable()
```

When comparing these results to the results in @sec-daa-basic, we
can see that, for instance, the q-values of _Faecalibacterium_ and
_Catabacter_ are now somewhat smaller. Nevertheless, the results look overall
rather similar to the results without the covariate included. Therefore, cohort
does not seem to be a confounder causing spurious associations between 
patient status and the abundances of any taxa.

### DAA with an interaction {#sec-daa-ia}

Here we show a more complex version of DAA, namely, a case where the interaction
effect of two variables is analyzed. We give this example here for illustration
purposes but note that such complex analysis may not be advisable in a small
dataset such as our example dataset.

We analyze the interaction between `patient_status` and `cohort`.
We proceed as in @sec-daa-cov above except that we now add also the interaction
term into the formula. As `Maaslin2()` does not directly support the analysis
with interactions we must first create the interaction dummy variables in to the
`tse` object. The main effects `patient_status` and `cohort` as well as the
interaction dummy variables are then included in the `Maaslin2()` function.

```{r}
#| label: run_daa_ia
#| results: hide

# Create dummy variables for the interaction term
tse$patient_statusADHD_cohortCohort_2 <-
    ifelse(tse$patient_status == "ADHD" & tse$cohort == "Cohort_2", 1, 0)
tse$patient_statusADHD_cohortCohort_3 <-
    ifelse(tse$patient_status == "ADHD" & tse$cohort == "Cohort_3", 1, 0)

# Run Maaslin2 with patient_status, cohort and their interaction terms
obj_daa_ia <- Maaslin2(
    input_data = assay(tse),
    input_metadata = colData(tse) |> as.data.frame(),
    fixed_effects = c(
        "patient_status",
        "cohort",
        "patient_statusADHD_cohortCohort_2",
        "patient_statusADHD_cohortCohort_3"
    ),
    output = "output",
    plot_heatmap = FALSE,
    plot_scatter = FALSE
)

# Extract the results and calculate the 95% CIs and q-values
res_daa_ia <- obj_daa_ia$results |>
    mutate(
        ci_lwr = coef - qt(.975, df = N) * stderr,
        ci_upr = coef + qt(.975, df = N) * stderr,
        metadata = ifelse(
            metadata %in% c(
                "patient_statusADHD_cohortCohort_2",
                "patient_statusADHD_cohortCohort_3"
            ),
            "patient_satus:cohort",
            metadata
        )
    ) |>
    group_by(metadata) |>
    mutate(qval = p.adjust(pval, method = "fdr")) |>
    ungroup() |>
    select(feature, name, coef, pval, qval, ci_lwr, ci_upr)
```

```{r}
#| label: run_daa_ia_1

# Print a sample of the results
res_daa_ia |>
    arrange(feature, name) |>
    head() |>
    kable()
```

The results object `res_daa_ia` now also includes the interaction
terms `patient_statusADHD_cohortCohort_2` and
`patient_statusADHD_cohortCohort_3`. 

We examine the term `patient_status_cohortCohort_3`. It indicates how the
differences between the ADHD and Control groups vary between cohorts 1 and 3.

```{r}
#| label: print_daa_ia_2

# Print the results for the interaction term patient_statusADHD_cohortCohort_3
res_daa_ia |>
    filter(name == "patient_statusADHD_cohortCohort_3") |>
    arrange(pval) |>
    head() |>
    kable()
```

The smallest p-value is observed for _Coprobacter_. As the corresponding
log~2~-fold change
estimate is negative it means that the difference in the relative abundance of
_Coprobacter_ between the ADHD and Control groups is higher in Cohort 1 than
in Cohort 3.

This interaction can be better illustrated by plotting the relative
abundances of _Coprobacter_ by `patient_status` and `cohort`.

```{r}
#| label: daa-viz-2

# Plot the abundances of Coprobacter by patient_status and cohort
plotBoxplot(
    tse,
    features = c("Coprobacter"),
    assay.type = "relabundance", # To plot relative abundances
    x = "patient_status", # The variable on the x (horizontal) axis
    facet.by = "cohort", # To create a separate plot for each cohort
    scales = "fixed", # To have the same y-axis scale in each facet
    add.proportion = TRUE, # Add bars indicating prevalence
    point.offset = "quasirandom", # To plot points nicely
    point.size = 2, # The size of the points
    point.color = "black", # The color of the edges of the points
    point.alpha = 1, # The transparency of the edges of the points
    box.alpha = .20
) # The transparency of the boxplots
```

Here we can clearly see how the relative abundances of _Coprobacter_ are higher
in the ADHD group in Cohort 1 but they are higher in the Control group in 
Cohort 3. As the association is positive in Cohort 1 (the baseline cohort) and
negative in Cohort 3, sign of interaction estimate is negative (the association
becomes more negative when we move from the baseline cohort to a non-baseline
cohort).

Note that a similar plot could also be used to illustrate a possible confounding
effect.

### Differential prevalence analysis (DPA) {#sec-DPA}

We show here how to perform DPA with logistic regression. We show the case
where we compare subjects with and without ADHD while controlling for `cohort`. 
Due to the convenient implementation of logistic regression for DPA,
we employ here the `maaslin3()` function from the
`r BiocStyle::Biocpkg("maaslin3")` package [@Nickols2024].

To run the `maaslin3()` function, we need to provide it with

- a `tse` object,
- a model formula, in this case `~ patient_status + cohort + total`, and
- a folder for output files

Note that we also control for the sequencing depth/library size/total read
count (`total`) in the analysis. This is because the detection
(observed presence) of taxa may depend on the sequencing depth.

```{r}
#| label: run_dpa_cov

library(scater)
library(maaslin3)

# Calculate the sequencing depth (the total number of reads/counts) for each
# sample
tse <- addPerCellQC(tse)

# Run maaslin3 with patient_status, cohort and reads as the predictor variables
obj_DPA_cov <- maaslin3(
    tse,
    formula = ~ patient_status + cohort + total, # Model formula
    output = "output", # The name of the folder for some output files
    plot_associations = FALSE, # To prevent some unnecessary plots
    plot_summary_plot = FALSE,
    verbosity = "ERROR"
) # To prevent some unnecessary output
```

Next, we extract the `data.frame` of results from the list object returned by
`maaslin3()`. We remove the variable `total` from the output as it is merely a 
technical covariate that we are not interested in. We also calculate the 95%
confidence intervals. Furthermore, as with `r BiocStyle::Biocpkg("Maaslin2")`,
we calculate the q-values separately for `patient_status` and `cohort`.
Lastly, we transform the results to a slightly more convenient format.

```{r}
#| label: results_DPA_cov

# Extract the results and calculate the 95% CIs and q-values
res_DPA_cov <- obj_DPA_cov$fit_data_prevalence$results |>
    filter(name != "total") |>
    mutate(
        ci_lwr = coef - qnorm(.975) * stderr,
        ci_upr = coef + qnorm(.975) * stderr
    ) |>
    group_by(metadata) |>
    mutate(qval = p.adjust(pval_individual, method = "fdr")) |>
    ungroup() |>
    select(feature, name, coef, pval = pval_individual, qval, ci_lwr, ci_upr)

# Print the results for taxa with the smallest p-values for patient_status
res_DPA_cov |>
    filter(name == "patient_statusADHD") |>
    arrange(pval) |>
    head() |>
    kable()
```

The results are now in the same format as the DAA results in @sec-daa-cov. The
only difference is `coef` now indicates the log-odds ratio for the difference
in prevalences. The odds ratios (OR) would thus be calculated as
`or = exp(coef)`. Note also that the results are not provided for taxa that
were present in all samples (as it is not sensible to analyze presence/absence
differences for such taxa).

By examining the results we see that the smallest p-value was observed for
genus _Erysipelatoclostridium_. As the sign of the estimate is _negative_ is
means that the genus is more prevalent in the control group. (This is can be
seen from the prevalence bars of _Erysipelatoclostridium_ in the strip plots in
@sec-daa-viz.)

#### DPA with an interaction {#sec-logr-ia}

Contrary to `Maaslin2()`, the `maaslin3()` function supports also the analysis
with interaction terms. Therefore, in the analysis with interaction(s), one can
just add the interaction term(s) in to the model formula by
using the standard R syntax, e.g., 
`formula = ~ patient_status + cohort + patient_status:cohort + reads` or
alternatively `formula = ~ patient_status * cohort + reads`

### Analysis of positive (non-zero) abundances {#sec-daa-pos}

It may sometimes be biologically relevant to analyze only the positive
(non-zero) relative abundances of taxa. We thus briefly show here how to
perform such analysis using the `maaslin3()` function. Purely for
illustration purposes, we show here the analysis with an interaction term.

We proceed as in the previous sections. The difference
compared to DPA (@sec-DPA) is that we now have to
extract `fit_data_abundance` (and not
`fit_data_prevalence`) from the list object returned by `maaslin3()`. We also
need to set `median_comparison_abundance = FALSE` in the function call as we
want to estimate the differences of (positive) relative abundances and not make
any compositional adjustments.

```{r}
#| label: daa_pos_ia

# Run maaslin3 function.
obj_daa_pos_ia <- maaslin3(
    tse,
    output = "output",
    formula = ~ patient_status * cohort,
    median_comparison_abundance = FALSE,
    verbosity = "ERROR",
    plot_associations = FALSE,
    plot_summary_plot = FALSE
)

# Extract the results and calculate the 95% confidence intervals and q-values
res_daa_pos_ia <- obj_daa_pos_ia$fit_data_abundance$results |>
    filter(!is.na(pval_individual) & N_not_zero > 6) |>
    mutate(
        ci_lwr = coef - qt(.975, df = N_not_zero - 6) * stderr,
        ci_upr = coef + qt(.975, df = N_not_zero - 6) * stderr
    ) |>
    group_by(metadata) |>
    mutate(qval = p.adjust(pval_individual, method = "fdr")) |>
    ungroup() |>
    select(feature, name, coef,
        pval = pval_individual, qval, ci_lwr, ci_upr,
        N_not_zero
    )

# The results for the interaction between patient_status and cohort (when
# comparing Cohort_3 to Cohort_1) for the taxa with the smallest p-values
res_daa_pos_ia |>
    filter(name == "patient_statusADHD:cohortCohort_3") |>
    arrange(pval) |>
    head() |>
    kable()
```

The output is similar to the outputs in the previous sections. It, however,
now includes also `N_not_zero` indicating the number of samples with
positive (non-zero) relative abundances, that is, the sample size
used in the analysis for each taxon.

The results are interpreted in a similar manner as in normal DAA
(@sec-daa-ia). For instance, the estimates (`coef`) indicate the log~2~-fold
changes of positive relative abundances (as `maaslin3()` performs log~2~
transformation to the positive relative abundances).

### Alternative options {#sec-daa-alternative}

#### Advanced compositionality adjustments

We have shown above (sections -@sec-daa-basic, -@sec-daa-cov, -@sec-daa-ia and
-@sec-daa-pos) how to perform DAA by analyzing relative abundances
(TSS normalized counts) without any advanced adjustments for compositionality.
As explained in @sec-daa-recommendations, this choice was motivated by 
some recent benchmark studies that have employed only real data, especially the
study by @Gamboa-Tuz2025. 

We note, however, that in some benchmark studies such simple versions of DAA
have performed worse to some more advanced methods when compared on
_simulated_ data with strong compositional effects (e.g. @Roche2022, @Yang2022).
We therefore give below some options for incorporating more advanced
compositionality adjustments in DAA.

- In `r BiocStyle::Biocpkg("Maaslin2")`, the default normalization method (TSS)
can be replaced by TMM normalization [@Robinson2010] by including
`NORM = "TMM"` in the `Maaslin2()` function.
- In the analysis of positive abundances with `r BiocStyle::Biocpkg("maaslin3")`,
the adjustment for compositional bias can be enabled by setting
`median_comparison_abundance = TRUE` in `maaslin3()` function.
- One can use DAA methods that are specifically designed to account for
the compositionality of microbiome data. Such methods include, for instance,
  - `r BiocStyle::Biocpkg("ANCOMBC")` [@ancombc2020],
  [vignette](https://www.bioconductor.org/packages/devel/bioc/vignettes/ANCOMBC/inst/doc/ANCOMBC.html).
Note that we refer here to the older version of ANCOM-BC. There is also a
newer version (ANCOM-BC2) available, but some problems have been observed in its
performance (e.g., @Pelto2025).
  - `LinDA` [@Zhou2022] from the `r BiocStyle::CRANpkg("MicrobiomeStat")` package,
  [reference manual](https://cran.r-project.org/web/packages/MicrobiomeStat/MicrobiomeStat.pdf).
  - `ZicoSeq` [@Yang2022] from the `r BiocStyle::CRANpkg("GUniFrac")` package,
  [vignette](https://cran.r-project.org/web/packages/GUniFrac/vignettes/ZicoSeq.html).

#### A non-parametic alternative to MaAsLin2

As MaAsLin2 analyzes log transformed relative abundances, it needs to replace
the zeros by some positive numbers before the transformation.
If one wants to avoid this somewhat arbitrary imputation, we suggest performing
DAA by using non/semi-parametric ordinal regression models (ORM). ORM can be
seen as an extension of the familiar Wilcoxon test in the sense that it provides
p-values similar to those of the Wilcoxon test in the simple two group
comparison, but it can also incorporate covariates (and interaction terms)
(@Harrell2001, @Liu2017). ORM has also performed well in one benchmark study
[@Pelto2025].

As there does not currently exist a good implementation of ORM for
DAA one must use a custom function. Such custom function with application
examples is given in [@sec-extra_daa]. We also note that, in contrast to
`Maaslin2()` function, the custom function provides confidence intervals and
q-values (calculated for the variable of interest) automatically in its output.

#### A more powerful alternative to MaAsLin3 in DPA
  
While the `r BiocStyle::Biocpkg("maaslin3")` package provides a convenient
implementation of logistic regression for microbial DPA, we note, that it may
provide somewhat conservative results, that is, it rarely provides false
positive results but it may fail to detect some true effects. To obtain higher
statistical power, we recommend using a version of logistic regression
introduced by @Firth1993. As there currently exists no convenient
implementation of Firth logistic regression for DPA, one needs to use a custom
function. Such custom function with application examples is given in the
[@sec-extra_daa].

::: {.callout-tip icon=false}

## Exercises

**Goal:** The goal of these exercises it to learn how to perform differential
abundance analysis and how to visualize the results.

**Exercises: Differential abundance analysis (DAA)**

1. Load any of the example data sets mentioned in [@sec-example-data].

2. Observe `colData`. Ensure that the data includes a meaningful way to group
samples
(e.g., Region, Treatment, Diagnosis). If it does not, choose another dataset.

3. Agglomerate the data to the genus level and filter by a prevalence of 10%.

4. Run the basic two-group DAA.

5. Explore the results tables. Are there any taxa with significant
abundance differences?

6. Visualize the results (for significant taxa) with strip plots and boxplots.
Can you "see" how the abundances of (possible)
significant taxa differ between the groups?

7. Run the same analysis now with an additional covariate(s). Do the results
differ? Why?

8. You can try also more complex analysis, e.g. with several covariates and/or
with interactions.

9. Bonus 1: Try also differential prevalence analysis (DPA) and/or the analysis
of positive abundances with your data.

10. Bonus 2: Try to run DAA and DPA with ordinal regression model and Firth 
logistic regression with the custom functions provided in the extra materials.
See how the results differ from the results obtained with `Maaslin2()` and
`maaslin3()`.

Useful functions:

`SummarizedExperiment::colData()`, `mia::subsetByPrevalent()`,
`Maaslin2::Maaslin2()`, `maaslin3::maaslin3()`

:::
