# Differential abundance {#sec-differential-abundance}

```{r echo=FALSE, results="asis"}
#| label: setup

library(rebook)
chapterPreamble()
```

An important component of many microbiome studies is differential abundance
analysis (DAA). It is used to identify differences in the abundances of
individual taxa (at any taxonomic level) between two or more groups, such as
treatment versus control groups.

The goal of DAA is to identify biomarkers of a certain phenotype or condition,
and gain understanding of a complex system by looking at its isolated
components. For example, the identification of a bacterial taxon that is more
abundant in healthy patients compared to diseased patients can lead to important
insights into the underlying mechanisms of the disease. In other words,
differentially abundant taxa can be involved in the dynamics of the disease,
which in turn helps to understand the system as a whole.


## Challenges in differential abundance analysis

As discussed in Section [@sec-stat-challenges], microbiome data display unique
properties, such as compositionality and skewed/zero-inflated distributions.
Evidently, these properties need to be addressed properly also in DAA.
Unfortunately, there is currently no consensus on how that should be done.
Instead, there exist dozens of methods employing varying techniques and,
consequently, providing highly discordant results.

The discordancy of results provided by different DAA methods was recently well
illustrated by @Nearing2022. In this study, multiple DAA methods were applied to
38 different datasets and their results were compared to one another. Similar
findings were also made by @Yang2022 who evaluated DAA methods 106 real datasets
(along with simulated data). They found especially that on many datasets almost
any taxon was detected by at least one DAA method (see Figure 7 in their
article). They thus raised concerns of potential cherry-picking, that is, a
researcher is likely to find the taxa in favor of their hypothesis by just
trying out several different DAA methods.


## Recommendations based on recent literature

Usually, when a new DAA method is introduced, the authors demonstrate it to
perform better to some of the existing alternatives. That is often achieved by
employing simulated data and possibly one or more real datasets. The authors may
have, however, selected their simulation models, example datasets, evaluation
metrics, compared methods, etc. specifically to show the introduced method in a
favorable light. Assessing the performance of different methods based solely on
the studies where they were originally introduced may therefore not be
advisable.

A more neutral evaluation of the performance of different DAA methods can be
found from several independent benchmarking studies (e.g. @Weiss2017,
@Hawinkel2019, @Calgaro2020, @Wallen2021, @Nearing2022, @Cappellato2022,
@Roche2022, @Swift2023). Unfortunately, these studies have ended up with very
discordant conclusions on the methods. However, three very recent benchmark
studies employing mostly real datasets, have ended up with somewhat concordant
conclusions.

-	@Wirbel2024 produced realistic test datasets by first creating random
case-control groups on real datasets (from healthy patients) and then implanting
differential abundance signals to such data. They found that only t-test/linear
regression, the Wilcoxon test, limma [@Ritchie2015], and fastANCOM
[@fastANCOM2022] properly controlled false discoveries at relatively high
sensitivity.

-	@Pelto2025 studied how the results from different DAA methods replicate
between random partitions of datasets and between datasets from independent
studies. They found that when considering consistency along with sensitivity,
the best performance was obtained by analyzing relative (TSS normalized)
abundances with a non-parametric method or (after log transformation) with
linear regression/t-test or analyzing presence/absence of taxa with logistic
regression.

-	@Gamboa-Tuz2025 employed three real datasets to benchmark different DAA
methods against biological ground truth. They found that methods employing the
TSS normalization and RNA-seq-derived methods performed best.

In conclusion, it seems that simple methods perform better (or similarly) than
more elaborate methods. An additional benefit of using simple methods is that
they are intuitive and easy to understand. We therefore recommend performing DAA
by analyzing **relative (TSS normalized) abundances by a non-parametric method**
(e.g. ordinal regression model) or, after a log transformation,
by **t-test/linear regression**.

Furthermore, based on recent studies on classification accuracy of healthy vs.
non-healthy patients, it seems that a high proportion of information in
microbiome data is included in the mere **presence/absence** of taxa
(@Giliberti2022 @Karwowska2025). A viable, and even simpler, alternative to DAA
is thus **differential presence/absence analysis** (DPAA, see also @Pelto2025).

Lastly, we note that the performance of a DAA method may depend
on the characteristics of the analyzed data (@Weiss2017). Consequently,
there surely exist cases where some other methods perform better than the
recommended simple methods. If one wants to evaluate which method may perform
best specifically on their data, one option to use the benchdamic package by
@Calgaro2022 (see vignette
[here](https://www.bioconductor.org/packages/release/bioc/vignettes/benchdamic/inst/doc/intro.html)).


## Performing differential abundance and presence/absence analysis

In this section we show how to perform differential abundance or
presence/absence analysis as recommended above. We illustrate the basic case of
comparing abundances or presence/absences of taxa between two experimental
groups. Furthermore, we illustrate how to add covariates in the analysis and
how to examine the interaction effect of two variables.

We use the Tengler2020 dataset from mia package. The dataset includes samples
from people with ADHD (N = 13) and without ADHD (Control group, N = 14). The
people are from three different cohorts.

```{r}
#| label: load_packages_and_data

library(tidyverse)
library(mia)
library(knitr)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Show patient status by cohort
table(tse$patient_status, tse$cohort) |>
  kable()
```


### Preparing the data

Before starting the analysis, it is recommended to reduce the size and
complexity of the data to make the results more reproducible. For this
purpose, we agglomerate the features by genus and filter them by a prevalence
threshold of 10%. Furthermore, we create assays for relative abundances and
presence/absence of taxa. Lastly, we transfrom variable `patient_status` to a
factor with `Control` as the reference level. (It is a standard practice to use
the control group as the reference group in many statistical analyses.)

```{r}
#| label: prepare_data 

# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse, rank = "Genus", prevalence = 0.10)

# Transform count data and create assay of relative abundances
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")

# Transform count data and create assay of presence/abundance of taxa
tse <- transformAssay(tse, assay.type = "counts", method = "pa")

# Transform patient_status to a factor and set "Control" as the refrence level
tse$patient_status <- factor(tse$patient_status, levels = c('Control', 'ADHD'))

# We also transform cohort as a factor
tse$cohort <- factor(tse$cohort, levels = c('Cohort_1', 'Cohort_2', 'Cohort_3'))
```


### Performing differential abundance analysis {#sec-daa}

We show here how to perform DAA by analyzing relative abundances with linear
regression models. (Note that e.g. t-test, ANOVA and ANCOVA are special cases of
linear regression models.)

The advantages of linear regression models are, for instance,
their robustness to assumptions, simplicity and interpretability. However,
as linear regression models assume normality of the response variable the
relative abundances (having usually very skewed distributions) need to be
transformed before the analysis. The choice of transformation adds
arbitrariness to the analysis, however. Furthermore, when using the log
transformation the zero abundances must be treated somehow, for instance, by
adding an arbitrary pseudo-count to the counts (before transforming to relative
abundances). The arbitrariness related to the transformation can be considered
as one of the main disadvantages of employing linear regression models in DAA.
Nevertheless, linear regression models have been observed to perform well in
practice in DAA (@Wirbel2024, @Pelto2025), so we will employ them here.

We employ the `Maaslin2` function from the MaAsLin2 package [@Mallick2020]. The
function takes the untransformed count matrix as input. It first replaces
the zero counts by half of the smallest non-zero count separately for each
taxon. Next, it transforms the counts to relative abundances and then
applies $log_{2}$ transformation to them. Lastly, it runs
a linear regression model for each taxon using the transformed relative
abundances as the response variable.


#### The basic two-group comparison {#sec-daa-basic}

We first perform the basic two-group comparison without any other covariates
(i.e. t-test). We compare the relative abundance of taxa between the people
with and without ADHD.

The `Maaslin2` function takes the count matrix and metadata as inputs.
We also need to provide it with the variable indicating the groups,
i.e. `patient_status` in our case. Furthermorem we need to specify the folder
for some ouput files, even though we do not need the files here. Lastly, we
prevent Maaslin2 to create certain plots. (`Maaslin2` prints a lot of some
rather unnecessary output which is supresseed here for conveniency.)

```{r}
#| label: run_daa_basic
#| results: hide

library(Maaslin2)

obj_daa_basic <- Maaslin2(
  input_data = assay(tse),                          # The count matrix
  input_metadata = colData(tse) |> as.data.frame(), # The metadata
  fixed_effects = "patient_status",                 # The predictor variable
  
  output = 'output', # The name of the folder for ouput files.
  plot_heatmap = FALSE,
  plot_scatter = FALSE)
```

Next, we extract the data.frame of results from the list object returned by
`Maaslin2`. We also calculate the 95% confidence intervals and transform the
results to a slightly more convenient format. (The `N - 2` in the calculation
of the confidence intervals indicates the residual degrees of freedom. Usually
you can use the sample size `N` there as a good approximation.)

```{r}
#| label: results_daa_basic

# Extract the results from the list object and calculate the 95% CIs
res_daa_basic <- obj_daa_basic$results |> 
  mutate(ci_lwr = coef - qt(.975, df = N - 2) * stderr,
         ci_upr = coef + qt(.975, df = N - 2) * stderr) |>
  select(feature, name, coef, pval, qval, ci_lwr, ci_upr)

# Print the results for taxa with the smallest p-values
res_daa_basic |> arrange(pval) |> head() |> kable()
```

The columns of `res_daa_basic` are as follows:

- `feature` indicates the taxon (here genus) name.
- `name` indicates the predictor variable. As `patient_status` is a
categorical variable, it also tells the group used as the _non_-reference level
(here `ADHD` as `Control` was set as the reference level).
- `coef` is the log~2~-fold change estimate, a standard way to express effect
size in DAA.
Most importantly, positive values indicate abundance being higher in the
non-reference group (here `ADHD`). More precisely, `2 ^ coef` would be
interpreted here as the ratio of geometric means of the relative abundances
between the ADHD and Control groups.
- `pval` is the unadjusted p-value.
- `qval` (q-value) is the FDR adusted p-value. This is generally used to define
statisitcal significance. Typically, q-values < .05 are considered as
(statistically) significant.
- `ci_lwr` and `ci_upr` are the lower and upper bounds of the 95% confidence 
intervals for the log~2~-fold change.

Thus, if we use the standard criterion for statistical significance, namely,
q-value < 0.05, we can see that the difference is statistically significant for
the genera named as _Ruminococcus gauvreauii group_, _Faecalibacterium_,
_Catabacter_, _Clostridium innocuum group_ and _Erysipelatoclostridium_.
Furthermore, as the log~2~-fold change estimate is positive, for instance, for
_Faecalibacterium_ it means that the average relative abundance of
_Faecalibacterium_ is higher in the ADHD group.


#### DAA with additional covariates {#sec-daa-cov}

Here we show how to perform DAA with an additional covariate included in the
analysis. One usually wants to include a variable as a covariate if they suspect
that it may act as a _confounding_ variable, namely, a variable that causes a
spurious association between the variable of main interest and the abundances.

We use here `cohort` as a covariate. We proceed as above except that we now set
`fixed_effects = c("patient_status", "cohort")`. Furthermore, we now calculate
the q-values manually so that they are calculated separately for each variable
in the model. This is because `Maaslin2` calculates the q-values over all
variables in the model which is usually _not_ what we want as we are usually
mainly interested only in one variable (here `patient_status`).


```{r}
#| label: daa_cov
#| results: hide

# Run Maaslin2 with patient_status and cohort as predictor variables
obj_daa_cov <- Maaslin2(
  input_data = assay(tse),                          
  input_metadata = colData(tse) |> as.data.frame(),
  fixed_effects = c("patient_status", "cohort"),
  output = 'output',
  plot_heatmap = FALSE,
  plot_scatter = FALSE)

# Extract the results and calculate the 95% CIs and q-values
res_daa_cov <- obj_daa_cov$results |> 
  mutate(ci_lwr = coef - qt(.975, df = N - 4) * stderr,
         ci_upr = coef + qt(.975, df = N - 4) * stderr) |>
  group_by(name) |> 
  mutate(qval = p.adjust(pval, method = "fdr")) |>
  ungroup() |>
  select(feature, name, coef, pval, qval, ci_lwr, ci_upr)
```

```{r}
#| label: print_daa_cov_1

# Print a sample of the results
res_daa_cov |> arrange(feature, name) |> head() |> kable()
```

Note that there now appear results also for the variable `cohort`. As `cohort`
consists of three groups, namely, `Cohort_1` (the reference group), `Cohort_2`
and `Cohort_3`, there are results separately for the comparison between
`Cohort_2` vs. `Cohort_1` (`cohortCohort_2`) and for the comparison between
`Cohort_3` vs. `Cohort_1` (`cohortCohort_3`).

We can now examine how the inclusion of `cohort` as a covariate affects the
results for `patient_status`.

```{r}
#| label: print_daa_cov_2

# Print the results for taxa with the smallest p-values for patient_status
res_daa_cov |> 
  filter(name == "patient_statusADHD") |> 
  arrange(pval) |> 
  head() |> 
  kable()
```

When comparing these results to the results without the covariate included, we
can see that, for instance, the q-value of _Faecalibacterium_ and
_Catabacter_ are now somewhat smaller. Nevertheless, the results overally look
rather similar to the results without the covariate included. Therefore, cohort
does not not seem to be a confounder causing spurious associations between 
patient status and the abundances of any taxa.


#### DAA with an interaction {#sec-daa-ia}

Here we show a more complex version of DAA, namely, a case where the interaction
effect of two variables is analyzed. We give this example here for illustration
purposes but note that such complex analysis may not be advisable with a small
dataset such as our example dataset.

We analyze interaction between `patient_status` and `cohort`. We proceed as in
@sec-daa-cov above except that we now add also the interaction term into the
formula. As `Maaslin2` does not directly support the analysis with interactions
we must first create the interaction dummy variables in to the
metadata. The main effects `patient_status` and `cohort` as well as the
interaction dummy variables are then included in the `Maaslin2` function.

```{r}
#| label: run_daa_ia
#| results: hide

# Create dummy variables for the interaction term
tse$patient_statusADHD_cohortCohort_2 <- 
  ifelse(tse$patient_status == "ADHD" & tse$cohort == "Cohort_2", 1, 0)
tse$patient_statusADHD_cohortCohort_3 <- 
  ifelse(tse$patient_status == "ADHD" & tse$cohort == "Cohort_3", 1, 0)

# Run Maaslin2 with patient_status, cohort and their interaction terms
obj_daa_ia <- Maaslin2(
  input_data = assay(tse),                          
  input_metadata = colData(tse) |> as.data.frame(),
  fixed_effects = c("patient_status",
                    "cohort",
                    "patient_statusADHD_cohortCohort_2",
                    "patient_statusADHD_cohortCohort_3"),
  output = 'output',
  plot_heatmap = FALSE,
  plot_scatter = FALSE)

# Extract the results and calculate the 95% CIs and q-values
res_daa_ia <- obj_daa_ia$results |> 
  mutate(ci_lwr = coef - qt(.975, df = N - 6) * stderr,
         ci_upr = coef + qt(.975, df = N - 6) * stderr) |>
  group_by(name) |> 
  mutate(qval = p.adjust(pval, method = "fdr")) |>
  ungroup() |>
  select(feature, name, coef, pval, qval, ci_lwr, ci_upr)
```

```{r}
#| label: run_daa_ia_1
#| 
# Print a sample of the results
res_daa_ia |> arrange(feature, name) |> head() |> kable()
```

The results object `res_daa_ia` now also includes the interaction
terms `patient_statusADHD_cohortCohort_2` and
`patient_statusADHD_cohortCohort_3`. 

We examine the term `patient_status_cohortCohort_3`. It indicates how the
differences between the ADHD and Control groups vary between cohorts 1 and 3.

```{r}
#| label: print_daa_ia_2

# Print the results for the interaction term patient_statusADHD_cohortCohort_3
res_daa_ia |> 
  filter(name == "patient_statusADHD_cohortCohort_3") |> 
  arrange(pval) |> 
  head() |> 
  kable()
```

The smallest p-value is observed for _Coprobacter_. As the log~2~-fold change
estimate is negative it means that the difference in the relative abundance of
_Coprobacter_ between the ADHD and Control groups is higher in Cohort 1 than
in Cohort 3.


#### Ordinal regression model as an alternative to linear regression model

If on wants to avoid the use of transformations needed with linear regression
models, we suggest performing DAA using non/semi-parametric ordinal regression
models (ORM). As there does not currently exist a good implementation of ORM for
DAA one has to use a custom function. Such custom function with application
examples is given in the [@sec-extras].


### Performing differential presence/absence (prevalence) analysis {#sec-dpaa}

We show here how to perform differential presence/absence analysis (DPAA) with
logistic regression. Due to a convenient implementation of logistic regression
for DPAA, we employ here the maaslin3 package [@Nickols2024].

We note, however, that the version of logistic regression used maaslin3 provides
rather conservative results, that is, it rarely provides false positive results
but, importantly,
_it may **not** detect some clear differences in presence/absence_. For a better
performance, we recommend using a version of logistic regression introduced by
@Firth1993. As there currently exist no convenient implenetation of Firth
logistic regression for DPAA, one needs to use a custom function. Such custom
function with application examples is given in the [@sec-extras]. 

It is also important to note that we will analyze the _observed_
presence/absence of taxa (i.e. observed non-zero/zero count) and not the
presence/absence of taxa in their ecosystem. Those concepts are not the same
because a taxon that is present in an ecosystem may be left undetected if, for
instance, its abundance is below the detection limit of the experimantal
workflow used in processing/analyzing the microbial samples [@Silverman2020].


#### The basic two-group comparison for presence/absence {#sec-dpaa-basic}

We first perform the basic two-group comparison without any other covariates.
We compare the presence/absence of taxa between the people with and without
ADHD.

The `maaslin3` function takes the `tse` object as input. We also need to provide
it with the model formula, in this case `~ patient_status`. Furthermore, we need
to specify the folder for some ouput files, even though we do not need the files
here. Lastly, we prevent `maaslin3` to create certain plots and print some
unnecessary output.

```{r}
#| label: run_dpaa_basic

library(maaslin3)

# Run maaslin3 with patient_status as the predictor variable
obj_dpaa_basic <- maaslin3(
  tse,                          
  formula = ~ patient_status, # Model formula
  output = 'output',          # The name of the folder for some ouput files.
  plot_associations = FALSE,  # Prevent some unnecessary plots
  plot_summary_plot = FALSE,
  verbosity = 'ERROR')        # Prevent some unnecessary output
```

Next, we extract the data.frame of results from the list object returned by
`maaslin3`. We also calculate the 95% confidence intervals and transform the
results to a slightly more convenient format.

```{r}
#| label: results_dpaa_basic

# Extract and transform the results to the same form as before 
res_dpaa_basic <- obj_dpaa_basic$fit_data_prevalence$results |> 
  mutate(ci_lwr = coef - qnorm(.975) * stderr,
         ci_upr = coef + qnorm(.975) * stderr) |>
  select(feature, name, coef, pval = pval_individual, qval = qval_individual,
         ci_lwr, ci_upr)

# The results for taxa with the smallest p-values
res_dpaa_basic |> arrange(pval) |> head() |> kable()
```

The results are now in the same format as the DAA results in @sec-daa-basic. The
only difference is `coef` now indicates the log-odds ratio for the difference
in prevalences. The odds ratios (OR) are thus calculated as `or = exp(coef)`.
Note also that the results are not provided for taxa that were present in all
samples (as it is not sensible to analyze presence/absence differences for
such taxa).

By examining the results we see that the smallest p-value was observed for
genus _Erysipelatoclostridium_. As the sign of the estimate is _negative_ is
means that the genus is more prevalent in the control group.


#### DPAA with additional covariates {#sec-dpaa-cov}

Here we show how to perform DPAA with `cohort` included as an additional
covariate in the analysis. We proceed as above but we now set the formula to
`~ patient_status + cohort`. We also calculate the q-values manually as in 
@sec-daa-cov.

```{r}
#| label: dpaa_cov

# Run maaslin3 with patient_status and cohort as predictor variables
obj_dpaa_cov <- maaslin3(
  tse,                          
  formula = ~ patient_status + cohort,
  output = 'output',          
  plot_associations = FALSE,
  plot_summary_plot = FALSE,
  verbosity = 'ERROR')

# Extract the results and calculate the 95% CIs and q-values
res_dpaa_cov <- obj_dpaa_cov$fit_data_prevalence$results |> 
  mutate(ci_lwr = coef - qnorm(.975) * stderr,
         ci_upr = coef + qnorm(.975) * stderr) |>
  group_by(name) |> 
  mutate(qval = p.adjust(pval_individual, method = "fdr")) |>
  ungroup() |>
  select(feature, name, coef, pval = pval_individual, qval, ci_lwr, ci_upr)
```

The `res_dpaa_cov` corresponds to the `res_daa_cov` in @sec-daa-cov.

```{r}
#| label: print_dpaa_cov

# Print the results for taxa with the smallest p-values for patient_status
res_dpaa_cov |> 
  filter(name == "patient_statusADHD") |> 
  arrange(pval) |> 
  head() |> 
  kable()
```

By examining the results of taxa with the smallest p-values we see that,
overall, the results are rather similar to the results without the covariate.


#### DPAA with an interaction {#sec-logr-ia}

Next, we show how to perform DPAA with interaction terms. Contrary to
`Maaslin2`, the `maaslin3` function supports also the analysis with interaction
terms. Therefore, we can just add an intercation term in the model formula by
using the stndard R syntax
`formula = ~ patient_status + cohort + patient_status:cohort` or
alternatively `formula = ~ patient_status * cohort`. The latter is usually more
convenient.

```{r}
#| label: dpaa_ia

# Run maaslin3 with patient_status, cohort and their interaction
obj_dpaa_ia <- maaslin3(
  tse,                          
  formula = ~ patient_status * cohort,
  output = 'output',          
  plot_associations = FALSE,
  plot_summary_plot = FALSE,
  verbosity = 'ERROR')

# Extract the results and calculate the 95% CIs and q-values
res_dpaa_ia <- obj_dpaa_ia$fit_data_prevalence$results |> 
  mutate(ci_lwr = coef - qnorm(.975) * stderr,
         ci_upr = coef + qnorm(.975) * stderr) |>
  group_by(name) |> 
  mutate(qval = p.adjust(pval_individual, method = "fdr")) |>
  ungroup() |>
  select(feature, name, coef, pval = pval_individual, qval, ci_lwr, ci_upr)
```

As in @sec-daa-ia we examine the term `patient_status:cohortCohort_3`. It
indicates how the differences in presence/absences between the ADHD and Control
groups vary between cohorts 1 and 3.

```{r}
#| label: print_dpaa_ia

# Print the results for the interaction term patient_statusADHD:cohortCohort_3
res_dpaa_ia |> 
  filter(name == "patient_statusADHD:cohortCohort_3") |> 
  arrange(pval) |> 
  head() |> 
  kable()
```

As, for instance, the log odd ratio estimate (`coef`) is negative for
_Barnesiella_ it indicates that the difference in the prvelance of
_Barnesiella_ between the ADHD and Control groups is higher in Cohort 1 than
in Cohort 3. (None of the results are close to significant after the FDR
adjustnment, however.)


### Analyzing positive (non-zero) abundances {#sec-daa-pos}

It may sometimes be biologically relevant to analyze only the positive
relative abundances of taxa. We show here how to perform such analysis.

It should be noted, however, that especially in small datasets (like the
example dataset used here), the number of positive/non-zero abundances may be
very small for some taxa. Therefore, this type of analysis may not always be
reasonable. Furthermore, we note again that one should keep in mind that the
observed absence of a taxon does not necessarily mean that the taxon
is truly absent in a microbial sample.

The analysis of positive relative abundances can be easily performed by using
maaslin3 package [@Nickols2024]. It first applies $log_{2}$ transformation to
the positive relative abundances and performs linear regression analysis for
each taxon. It should be especially noted that log transformation is very
convenient here as there are no zeros that would require some special treatment.


#### The basic two-group comparison {#sec-daa-pos-basic}

We proceed as in sections @sec-daa-basic and @sec-dpaa-basic. The difference
copmpared to DPAA is that we now have to extract `fit_data_abundance` (and not
`fit_data_prevalence`) from the list object returned by `maaslin3`. We also
need to set `median_comparison_abundance = FALSE` in the function call as we
want to estimate the differences of (positive) relative abundances and not make
any compositinal adjustments.

```{r}
#| label: daa_pos_basic

# Run maaslin3 with patient_status as the predictor variable
obj_daa_pos_basic <- maaslin3(
  tse,
  output = 'output',
  formula = ~ patient_status,
  median_comparison_abundance = FALSE,
  verbosity = 'ERROR',
  plot_associations = FALSE,
  plot_summary_plot = FALSE)

# Extract the results and calculate also the 95% confidence intervals
res_daa_pos_basic <- obj_daa_pos_basic$fit_data_abundance$results |> 
  mutate(ci_lwr = coef - qt(.975, df = N_not_zero - 2) * stderr,
         ci_upr = coef + qt(.975, df = N_not_zero - 2) * stderr) |>
  select(feature, name, coef, pval = pval_individual, qval = qval_individual,
         ci_lwr, ci_upr, N_not_zero)

# Print the results for taxa with the smallest p-values
res_daa_pos_basic |> arrange(pval) |> head() |> kable()
```

The results are interpreted in a similar manner as in DAA (section
@sec-daa-basic). `N_not_zero` indicates the number of samples with
positive (non-zero) relative abundances. Note that the results are not provided
for taxa with too few positive abundances.


#### Analysis with covariates {#sec-daa-pos-cov}

To include cohort as a covariate in the analysis, we proceed as above but we
just set `formula = ~ patient_status + cohort`. Moreover, we again calculate the q-values manually as explained in section @sec-daa-cov. (The results for taxa
with too few positive abundances are also filtered out.)

```{r}
#| label: daa_pos_cov

# Run maaslin3 with patient_status and cohort as the predictor variables
obj_daa_pos_cov <- maaslin3(
  tse,
  output = 'output',
  formula = ~ patient_status + cohort,
  median_comparison_abundance = FALSE,
  verbosity = 'ERROR',
  plot_associations = FALSE,
  plot_summary_plot = FALSE)

# Extract the results and alculate the 95% confidence intervals and q-values 
res_daa_pos_cov <- obj_daa_pos_cov$fit_data_abundance$results |> 
  filter(!is.na(pval_individual) & N_not_zero > 4) |>
  mutate(ci_lwr = coef - qt(.975, df = N_not_zero - 4) * stderr,
         ci_upr = coef + qt(.975, df = N_not_zero - 4) * stderr) |>
  group_by(name) |> 
  mutate(qval = p.adjust(pval_individual, method = "fdr")) |>
  ungroup() |>
  select(feature, name, coef, pval = pval_individual, qval, ci_lwr, ci_upr,
         N_not_zero)

# The results for patient_status for the taxa with the smallest p-values
res_daa_pos_cov |> 
  filter(name == "patient_statusADHD") |> 
  arrange(pval) |> 
  head() |> 
  kable()
```

The results are interpreted in a similar manner as in DAA (section
@sec-daa-cov).


#### Analysis with interaction {#sec-daa-pos-ia}

The analysis of positive relative abundances with interaction effect can be
performed as follows.

```{r}
#| label: daa_pos_ia

# Run maaslin3 function.
obj_daa_pos_ia <- maaslin3(
  tse,
  output = 'output',
  formula = ~ patient_status * cohort,
  median_comparison_abundance = FALSE,
  verbosity = 'ERROR',
  plot_associations = FALSE,
  plot_summary_plot = FALSE)

# Extract the results and calculate the 95% confidence intervals and q-values 
res_daa_pos_ia <- obj_daa_pos_ia$fit_data_abundance$results |> 
  filter(!is.na(pval_individual) & N_not_zero > 6) |>
  mutate(ci_lwr = coef - qt(.975, df = N_not_zero - 6) * stderr,
         ci_upr = coef + qt(.975, df = N_not_zero - 6) * stderr) |>
  group_by(name) |> 
  mutate(qval = p.adjust(pval_individual, method = "fdr")) |>
  ungroup() |>
  select(feature, name, coef, pval = pval_individual, qval, ci_lwr, ci_upr,
         N_not_zero)

# The results for the interacion between patient_status and cohort (when
# comparing Cohort_3 to Cohort_1) for the taxa with the smallest p-values
res_daa_pos_ia |> 
  filter(name == "patient_statusADHD:cohortCohort_3") |> 
  arrange(pval) |> 
  head() |> 
  kable()
```

Again, the results are interpreted in a similar manner as in DAA (section
@sec-daa-ia).


## Visualizing results {#sec-daa-viz}

In this section we briefly show how to visualize the results of the above
analyses.

### Visualizing abundances and prevalences

While DAA and DPAA provide information on which taxa how much their abundances
differ between the examined groups, some relevant infromation on _how_ the
abundances differ may not be seen from the result tables. It is therefore also
important to visualize the distributions of abundances of the (significant)
taxa. Moreover, for instance, the effect of covariates or the interaction terms
may be easier to comprehend by visualizing the abundances than by just trying
to interpret the estimates and p-values.

We thus show here how visualize the distributions of abundances by employing
the `plotBoxplot` function from the miaViz package.

#### The basic two group comparisons {#sec-daa-viz-1}

We first visualize the relative abundances of the taxa that had the q-values 
$< 0.05$ in the basic two group comparisons in DAA (section @sec-daa-basic).

```{r}
#| label: daa-viz-print-1

# The taxa with q-values < .05 in DAA
res_daa_basic |> arrange(pval) |> filter(qval < .05) |>  kable()
```

We illustrate the distribution of the relative abundances and presence/absences
of these genera by strip plots with boxplots and bars indicating prevalence.
To specifically illustrate the differences between the ADHD and Control groups
we group and color the abundances by `patient_status`.

```{r}
#| label: daa-viz-1

library(miaViz)

# Create strip plots with boxplots and bars indicating prevalence
plotBoxplot(
  tse,
  features = c("[Ruminococcus]_gauvreauii_group",
               "Faecalibacterium",
               "Catabacter",
               "Erysipelatoclostridium",
               "[Clostridium]_innocuum_group"),
  
  assay.type = "relabundance",   # To plot relative abundances
  x = "patient_status",          # The variable on the x (horizontal) axis
  color.by = "patient_status",   # The variable for the color of the points
  fill.by = "patient_status",    # The boxplots
  facet.by = "rownames",         # To create a separate plot for each feature
  scales = 'free_y',             # To allow different y-axis scales
  add.proportion = TRUE,         # Add bars indicating prevalence
  point.offset = "quasirandom",  # To plot points nicely
  point.size = 2,                # The size of the points
  point.alpha = .50,             # The transparency of the points
  box.alpha = .20) +             # The transparency of the boxplots
  theme(strip.text = element_text(size = 6)) # The size of the facet labels
```

In the boxplots,

- the thicker horizontal line indicates the median,
- the box indicates the interquartile range (IQR), namely, the range between
the 25th (Q1) and 75th percentile (Q3),
- the lower whisker indicates the smallest value above $Q1 - 1.5 \times IQR$,
and
- the lower whisker indicates the largest value below $Q3 + 1.5 \times IQR$.

The strip plots and boxplots show clearly how, for instance, the relative
abundance of the
genus named as _[Ruminococcus] gauvreauii group_ is higher in the Control group
as indicated by the negative sign of the log~2~-fold estimate (`coef`). The
bars indicating the prevalence (below the boxplots) also show how the prevalence
of _[Ruminococcus] gauvreauii group_ is higher in the Control group.
Correspondingly, for instance, the relative abundances of _Faecalibacterium_
are higher in the ADHD group as indicated by the positive sign of their
log~2~-fold estimates.

The result for _Erysipelatoclostridium_ in DPAA, namely, the small q-value with
negative log(OR) estimate, is clearly illustrated by the prevelence bars. They
show a high prevelence of _Erysipelatoclostridium_ in the Control group and low
prevelance in the ADHD group.


#### More complex analyses {#sec-daa-viz-2}

By plotting the abundances of a taxon by two predictor variables, we can
illustrate the confounding and interaction effects. We show here how to
illustrate the significant interaction of `patient_status` and `cohort` for
_Coprobacter_ (see section @sec-daa-ia).

```{r}
#| label: daa-viz-print-2

# The significant results for the interaction in DAA
res_daa_ia |> 
  filter(name %in% c("patient_statusADHD_cohortCohort_3",
                     "patient_statusADHD_cohortCohort_2")) |>
  arrange(pval) |> 
  filter(qval < .05) |> 
  kable()
```

```{r}
#| label: daa-viz-2

# Plot the abundances of Odoribacter by patient_status and cohort
plotBoxplot(
  tse,
  features = c("Coprobacter"),
  assay.type = "relabundance",   # To plot relative abundances
  x = "patient_status",          # The variable on the x (horizontal) axis
  color.by = "patient_status",   # The variable for the color of the points
  fill.by = "patient_status",    # The boxplots
  facet.by = "cohort",           # To create a separate plot for each cohort
  scales = 'fixed',              # To have the same y-axis scale in each facet
  add.proportion = TRUE,         # Add bars indicating prevalence
  point.offset = "quasirandom",  # To plot points nicely
  point.size = 2,                # The size of the points
  point.alpha = .50,             # The transparency of the points
  box.alpha = .20)               # The transparency of the boxplots
```

Here we can clearly see how the relative abundances of _Coprobacter_ are higher
in the ADHD group in Cohort 1 but they are higher in the Control group in 
Cohort 3. As the association is positive in Cohort 1 (the baseline cohort) and
negative in Cohort 3, sign of interaction estimate is negative (the association
becomes more negative when we move from the baseline cohort to a non-baseline
cohort).

### Summarizing results by plotting estimates and confidence intervals

The results of an analysis can be conveniently visualized by plotting the
log~2~-fold changes (or log ORs in DPAA) and the 95% confidence intervals.

```{r}
#| label: plot_daa_est_ci

# Correct the names of certain taxa and reorder the taxa by the log2-fold
# change estimates
res_viz_daa_basic <- res_daa_basic |> 
  mutate(taxon = str_replace(feature, "X\\.", "\\["),
         taxon = str_replace(taxon, "\\._", "\\]_"),
         taxon = fct_reorder(taxon, coef))

# Plot the log2-fold changes and the 95% confidence intervals
ggplot(res_viz_daa_basic, aes(x = coef, y = taxon)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point() +
  geom_errorbarh(aes(xmin = ci_lwr, xmax = ci_upr), height = 0.2) +
  labs(x = "Log2-fold change (95% CI)") +
  theme_light() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_text(size = 7),
        plot.title = element_text(hjust = 0.5))
```



::: {.callout-tip}
## Additional resources
While we generally recommend the use of the above simple methods in DAA there
exist dozens of alternative methods. A (non-comprehensive) list of these methods
is given below.

- ALDEx2 [@Gloor2016]
- ANCOM [@Mandal2015]
- ANCOMBC [@ancombc2020]
- BIRDMAn [@Rahman2023]
- corncob [@Martin2021]
- DACOMP [@Brill2019]
- DESeq2 [@Love2014]
- eBay [@Liu2020]
- edgeR [@Chen2016]
- fastANCOM [@fastANCOM2022]
- fido [@Silverman2019]
- LDM [@Hu2020]
- lefser [@Khlebrodova2021]
- limma [@Ritchie2015]
- LinDA [@Zhou2022]
- MaAsLin2 [@Mallick2020]
- MaAsLin3 [@Nickols2024]
- metagenomeSeq [@Paulson2017]
- Omnibus [@Omnibus2018]
- RAIDA [@Sohn2015]
- sccomp [@mangiola2023]
- ZicoSeq [@Yang2022]
- ZINQ [@Ling2021]
:::

::: {.callout-tip icon=false}

## Exercises

**Goal:** The goal of these exercises it to learn how to perform differential
abundance and presence/absence analysis and to visualize the results.

**Exercise 1: Differential abundance analysis (DAA)**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Observe `colData`. Ensure that the data includes meaningful sample grouping.
If it does not, choose another dataset.

3. Agglomerate the data by genus and filter by a prevalence of 10%.

4. Run the basic two-group DAA and DPAA. You can also run DAA for positive
abundances.

5. Explore the results tables. Are there any taxa with significant
abundance/prevalence differences?

6. Run the same analyses now with an additional covariate. Do the results
differ? Why?

7. You can try also more complex analysis, e.g. with several covariates and/or
with interactions.

8. Visualize the results.

9. Bonus: Try to run DAA and DPAA with ordinal regression model and Firth 
logistic regression with the custom functions provided in the extra materials.
See how the results differ from the results obtained with Maaslin2 and maaslin3.

Useful functions:

`data()`, `colData`, `subsetByPrevalent()`

:::

## These will come to Chapter C - Extra material

### Analyzing relative abundances with ordinal regression model {#sec-orm}

We show here how to perform DAA by analyzing relative abundances with ordinal
regression model (ORM). We have chosen ORM as it makes few assumptions on the
distributions of abundances while it can incorporate covariates (including
interaction terms) in the analysis. As ORM performs very similarly to the
familiar non-parametric Wilcoxon test in the basic two group comparison it can
also be seen as an extension of the Wilcoxon test (@Harrell2001, @Liu2017).

By using ORM to model the skewed and/or zero inflated distributions of microbial
abundances, we can avoid the strong parametric assumptions (e.g. negative
binomial distribution) made by some methods. Furthermore, we do not need to
transform abundances. This good because the choice of the transformation may be
somewhat arbitrary while it may affect the DAA results. Moreover, ORM has been
shown to perform well in practice in DAA when considering its sensitivity and
the consistency of its results (@Pelto2025).

We first create a function that (for each taxon) runs ORM and returns the effect
size estimate and p-value (based on score test) for each variable in the model.

```{r}
#| label: function_orm

# The function uses the orm() function from the rms package 
library(rms)

run_orm <- function(abundance, metadata, formula){
  
  # Create the design matrix of the model
  mm <- model.matrix(formula, metadata) |> 
    cbind(abundance) |> 
    tibble::as_tibble() |> 
    dplyr::select(-"(Intercept)")
  
  inds <- 1:(ncol(mm) - 1)
  vars <- colnames(mm)[inds]
  
  # Fit ordinal regression model with all variables
  # (Intercepts are automatically added)
  fit_1 <- rms::orm(abundance ~ ., data = mm)
  score_1 <- fit_1$stats["Score"]
  
  res <- data.frame(estimate = fit_1$coefficients[vars], p_value = NA)
  
  # Calculate p-value based on score test for each variable
  if(length(inds) == 1){
    res$p_value <- fit_1$stats["Score P"]
  }else{
    for(i in inds){
      fit_0 <- rms::orm(abundance ~ ., data = mm[, -i])
      score_0 <- fit_0$stats["Score"]
      res$p_value[i] <- as.numeric(1 - pchisq(score_1 - score_0, df = 1)) 
    }
  }
  
  return(res |> tibble::rownames_to_column("variable"))
}
```

We use the same Tengler2020 dataset as in @sec-differential-abundance.

```{r}
#| label: orm_prepare

library(tidyverse)
library(mia)
library(knitr)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse, rank = "Genus", prevalence = 0.10)

# Transform count data and create assay of relative abundances
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")

# Transform patient_status to a factor and set "Control" as the refrence level
tse$patient_status <- factor(tse$patient_status, levels = c('Control', 'ADHD'))

# We also transform cohort as a factor
tse$cohort <- factor(tse$cohort, levels = c('Cohort_1', 'Cohort_2', 'Cohort_3'))
```

#### The basic two-group comparison {#sec-orm-basic}

We compare the relative abundance of taxa between the people with and without
ADHD (indicated by the variable `patient_status`).

The code below works as follows.

 1. We extract the assay with relative abundances from the `tse` object
 2. The assay is transposed and transformed to a `data.frame`
 3. The function `run_orm` is run for each taxon
    i) The `.` indicates here the relative abundance of each taxon
    ii) The metadata including the variable `patient_status` is extracted from
    the `tse` object
    iii) The model formula indicates that we included only `patient_status` in
    our analysis
 4. The results for all taxa are binded to as one `data.frame`
 5. Lastly, we calculate p-values adjusted from multiple testing (`q-value`) by
 using the standard FDR correction by @Benjamini1995 (`method = "BH"`).


```{r}
#| label: run_orm_basic

res_daa_basic <- assay(tse, "relabundance") |>  # Extract relative abundances
  t() |>                                        # Transpose the assay
  as.data.frame() |>                            # Transform to a data.frame
  map(~ run_orm(                                # Run ORM for each taxon
    .,                                            
    metadata = colData(tse) |> as.data.frame(), # Extract the metadata
    formula = ~ patient_status)                 # Model formula. Now only 
                                                # patient_status
    ) |> 
  bind_rows(.id = "taxon") |>                   # Bind the results of all taxa
  mutate(
    q_value = p.adjust(p_value, method = "BH")) # Calculate q-values
                                                # (FDR adjusted p-values)

# The results for taxa with the smallest p-values
res_daa_basic |> arrange(p_value) |> head() |> kable()
```

In the output

- `variable` indicates the predictor variable. As `patient_status` is a
categorical variable, it also tells the group used as the non-reference level
(here `ADHD` as `Control` was set as the reference level).
- `estimate` is log(odds ratio). Most importantly, positive values indicate
abundance being higher in the non-reference group (here `ADHD`). Otherwise,
estimates can be a bit difficult to interpret for ORMs.

#### More complex analyses {#sec-orm-complex}

Analysis with covariates and analysis of interaction terms can be performed in
by just using different `formula` in the `run_orm` function, e.g.
`formula = patient_status + cohort` or `formula = patient_status * cohort`.


### DPAA with Firth logistic regression {#sec-firth}

As some versions of logistic regression perform badly in
the cases where, for instance, the prevelance of a taxon is zero in one
group we show here how to perform DPAA with a robust version of logistic
regression introduced by @Firth1993. This version has been observed to perform
well in DPAA in practice [@Pelto2025].

We first create a function that runs Firth logistic regression for
presence/absence and returns the effect size estimate and p-value for each
variable in the model.

```{r}
#| label: function_firth

# The function uses the logistf() function from the logistf package 
library(logistf)

run_firth <- function(abundance, metadata, formula){
  
  # Create the design matrix of the model
  mm <- model.matrix(formula, metadata) |> 
    cbind(abundance) |> 
    tibble::as_tibble() |> 
    dplyr::select(-"(Intercept)")
  
  # Fit Firth logistic regression model
  fit <- logistf::logistf(abundance ~ ., data = mm)
  
  res <- data.frame(estimate = fit$coefficients,
                    p_value = fit$prob)
  
  return(res[-1, ] |> tibble::rownames_to_column("variable"))
}
```

We use the same Tengler2020 dataset as in @sec-differential-abundance.

```{r}
#| label: firth_prepare

library(tidyverse)
library(mia)
library(knitr)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse, rank = "Genus", prevalence = 0.10)

# Transform count data and create assay of presence/absences
tse <- transformAssay(tse, assay.type = "counts", method = "pa")

# Transform patient_status to a factor and set "Control" as the refrence level
tse$patient_status <- factor(tse$patient_status, levels = c('Control', 'ADHD'))

# We also transform cohort as a factor
tse$cohort <- factor(tse$cohort, levels = c('Cohort_1', 'Cohort_2', 'Cohort_3'))
```

#### The basic two-group comparison {#sec-firth-basic}

We perform here the basic two-group comparison without any other covariates.
We compare the presence/absence of taxa between the people with and without
ADHD.

```{r}
#| label: run_firth_basic

res_firth_basic <- assay(tse, "pa") |>           # Extract presence/absences
  t() |>                                        # Transpose the assay
  as.data.frame() |>                            # Transform to a data.frame
  map(~ run_firth(                               # Run ORM for each taxon
    .,                                            
    metadata = colData(tse) |> as.data.frame(), # Extract the metadata
    formula = ~ patient_status)                 # Model formula.
    ) |> 
  bind_rows(.id = "taxon") |>                   # Bind the results of all taxa
  mutate(
    q_value = p.adjust(p_value, method = "BH")) # Calculate q-values
                                                # (FDR adjusted p-values)

# The results for taxa with the smallest p-values
res_firth_basic |> arrange(p_value) |> head() |> kable()
```

In the output

- `variable` indicates the predictor variable. As `patient_status` is a
categorical variable, it also tells the group used as the non-reference level
(here `ADHD` as `Control` was set as the reference level).
- `estimate` is the familiar log(odds ratio). A positive value indicates the
taxon being more prevalent in the non-refrence group (here `ADHD`).

#### More complex analyses {#sec-firth-complex}

Analysis with covariates and analysis of interaction terms can be performed in
by just using different `formula` in the `run_firth` function, e.g.
`formula = patient_status + cohort` or `formula = patient_status * cohort`.

