# Differential abundance {#sec-differential-abundance}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

An important component of many microbiome studies is Differential Abundance Analysis (DAA). It is used to identify differences in the abundances of individual taxa (at any taxonomic level) between two or more groups, such as treatment versus control groups.

The goal of DAA is to identify biomarkers of a certain phenotype or condition, and gain understanding of a complex system by looking at its isolated components. For example, the identification of a bacterial taxon that is more abundant in healthy patients compared to diseased patients can lead to important insights into the underlying mechanisms of the disease. In other words, differentially abundant taxa can be involved in the dynamics of the disease, which in turn helps to understand the system as a whole.


## Challenges in differential abundance analysis

As discussed in Section [@sec-stat-challenges], microbiome data display unique properties, such as compositionality and skewed/zero-inflated distributions. Evidently, these properties need to be addressed properly also in DAA. Unfortunately, there is currently no consensus on how that should be done. Instead, there exist dozens of methods employing varying techniques and, consequently, providing highly discordant results.

The discordancy of results provided by different DAA methods was recently well illustrated by @Nearing2022. In this study, multiple DAA methods were applied to 38 different datasets and their results were compared to one another. Similar findings were also made by @Yang2022 who evaluated DAA methods 106 real datasets (along with simulated data). They found especially that on many datasets almost any taxon was detected by at least one DAA method (see Figure 7 in their article). They thus raised concerns of potential cherry-picking, that is, a researcher is likely to find the taxa in favor of their hypothesis by just trying out several different DAA methods.


## Recommendations based on recent literature

Usually, when a new DAA method is introduced, the authors demonstrate it to perform better to some of the existing alternatives. That is often achieved by employing simulated data and possibly one or more real datasets. The authors may have, however, selected their simulation models, example datasets, evaluation metrics, compared methods, etc. specifically to show the introduced method in a favorable light. Assessing the performance of different methods based solely on the studies where they were originally introduced may therefore not be advisable.

A more neutral evaluation of the performance of different DAA methods can be found from several independent benchmarking studies (e.g. @Weiss2017, @Hawinkel2019, @Calgaro2020, @Wallen2021, @Nearing2022, @Cappellato2022, @Roche2022, @Swift2023). Unfortunately, these studies have ended up with very discordant conclusions on the methods. However, three very recent benchmark studies employing mostly real datasets, have ended up with somewhat concordant conclusions.

-	@Wirbel2024 produced realistic test datasets by first creating random case-control groups on real datasets (from healthy patients) and then implanting differential abundance signals to such data. They found that only t-test/linear regression, the Wilcoxon test, limma [@Ritchie2015], and fastANCOM [@fastANCOM2022] properly controlled false discoveries at relatively high sensitivity.

-	@Pelto2025 studied how the results from different DAA methods replicate between random partitions of datasets and between datasets from independent studies. They found that when considering consistency along with sensitivity, the best performance was obtained by analyzing relative (TSS normalized) abundances with a non-parametric method or (after log transformation) with linear regression/t-test or analyzing presence/absence of taxa with logistic regression.

-	@Gamboa-Tuz2025 employed three real datasets to benchmark different DAA methods against biological ground truth. They found that methods employing the TSS normalization and RNA-seq-derived methods performed best.

In conclusion, it seems that simple methods perform better (or similarly) than more elaborate methods. An additional benefit of using simple methods is that they are intuitive and easy to understand. We therefore recommend performing DAA by analyzing **relative (TSS normalized) abundances by a non-parametric method** (e.g. ordinal regression model) or, after log transformation, by **t-test/linear regression**.

Furthermore, based on recent studies on classification accuracy of healthy vs. non-healthy patients, it seems that a high proportion of information in microbiome data is included in the mere **presence/absence** of taxa (@Giliberti2022 @Karwowska2025). A viable, and even simpler, alternative to DAA is thus **differential presence/absence analysis** (DPAA, see also @Pelto2025).

Lastly, we note that the performance a DAA method generally depends to some degree on the characteristics of the analyzed data (@Weiss2017). Consequently, there surely exist cases where some other methods perform better than the recommended simple methods. If one wants to evaluate which method may perform best specifically on their data, one option to use the benchdamic package by @Calgaro2022 (see vignette [here](https://www.bioconductor.org/packages/release/bioc/vignettes/benchdamic/inst/doc/intro.html)).


## Performing differential abundance and presence/absence analysis

In this section we show how to perform differential abundance or presence/absence analysis as recommended above. We illustrate the basic case of comparing abundances or presence/absences of taxa between two experimental groups. Furthermore, we illustrate how to add covariates in the analysis and how to study interaction between two variables.

We use the Tengler2020 dataset from mia package. The dataset included samples from people with ADHD (N = 13) and without ADHD (Control group, N = 14). The people are from three different cohorts.

```{r import-daa-data}
library(tidyverse)
library(mia)
library(knitr)
library(miaViz)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Show patient status by cohort
table(tse$patient_status, tse$cohort) |>
  kable()
```


#### Preparing the data

Before starting the analysis, it is recommended to reduce the size and
complexity of the data to make the results more reproducible. For this
purpose, we agglomerate the features by genus and filter them by a prevalence
threshold of 10%.

```{r prep-daa-data}
# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse, rank = "Genus", prevalence = 10/100)
```

We also the control group as the refence group for `patient_status`. It is a standard practice to use the control group as the reference group in many statistical analyses. 

```{r}
tse$patient_status <- factor(tse$patient_status, levels = c('Control', 'ADHD'))
```

### Analyzing relative abundances with ordinal regression model {#sec-daa}

We show here how to perform DAA by analyzing relative abundances with ordinal regression model (ORM). We have chosen ORM as it makes few assumptions on the distributions of abundances while it can incorporate covariates (including interaction terms) in the analysis. As ORM performs very similarly to the familiar non-parametric Wilcoxon test in the basic two group comparison it can be seen as a generalization of the Wilcoxon test (@Harrell2001, @Liu2017).

By using ORM to model the skewed and/or zero inflated distributions of microbial abundances, we can avoid the strong parametric assumptions (e.g. negative binomial distribution) made by some methods. Furthermore, we do not need to transform abundances. This good because the choice of the transformation may be somewhat arbitrary while it may affect the DAA results. Moreover, ORM has been shown to perform well in practice in DAA when considering its sensitivity and the consistency of its results (@Pelto2025).

We first create a function that (for each taxon) runs ORM and returns the effect size estimate and p-value (based on score test) for each variable in the model.

```{r function-orm}
# The function uses the orm() function from the rms package 
library(rms)

run_orm <- function(abundance, metadata, formula){
  
  #Create the design matrix of the model
  mm <- model.matrix(formula, metadata) |> 
    cbind(abundance) |> 
    tibble::as_tibble() |> 
    dplyr::select(-'(Intercept)')
  
  inds <- 1:(ncol(mm) - 1)
  vars <- colnames(mm)[inds]
  
  #Fit ordinal regression model with all variables (Intercepts are automatically added)
  fit_1 <- rms::orm(abundance ~ ., data = mm)
  score_1 <- fit_1$stats['Score']
  
  res <- data.frame(estimate = fit_1$coefficients[vars], p_value = NA)
  
  #Calculate p-value based on score test for each variable
  if(length(inds) == 1){
    res$p_value <- fit_1$stats['Score P']
  }else{
    for(i in inds){
      fit_0 <- rms::orm(abundance ~ ., data = mm[, -i])
      score_0 <- fit_0$stats['Score']
      res$p_value[i] <- as.numeric(1 - pchisq(score_1 - score_0, df = 1)) 
    }
  }
  
  return(res |> tibble::rownames_to_column('variable'))
}
```

Next, we transform count assay to relative abundances.

```{r transform-rel}
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")
```


#### The basic two-group comparison {#sec-orm-basic}

We first perform the basic two-group comparison without any other covariates. We compare the relative abundances between the people with and without ADHD as indicated by the variable `patient_status`.

The code below works as follows.

 1. We extract the assay with relative abundances from the `tse` object
 2. The assay is transposed and transformed to a `data.frame`
 3. The function `run_orm` is run for each taxon
    i) The `.` indicates here the relative abundance of each taxon
    ii) The metadata including the variable `patient_status` is extracted from the `tse` object
    iii) The model formula indicates that we included only `patient_status` in our analysis
 4. The results for all taxa are binded to as one `data.frame`
 5. Lastly, we calculate p-values adjusted from multiple testing (`q-value`) by using the standard FDR correction by @Benjamini1995 (`method = 'BH'`).


```{r run-orm-basic}
res_daa_basic <- assay(tse, 'relabundance') |>  # Extract the assay of relative abundances
  t() |>                                        # Transpose the assay
  as.data.frame() |>                            # Transform to a data.frame
  map(~ run_orm(                                # Run ORM for each taxon
    .,                                            
    metadata = colData(tse) |> as.data.frame(), # Extract the metadata
    formula = ~ patient_status)                 # Model formula. Includes now only patient_status
    ) |> 
  bind_rows(.id = "taxon") |>                   # Bind the results of all taxa
  mutate(
    q_value = p.adjust(p_value, method = 'BH')) # Calculate q-values (multiplicity adjusted p-values)
```

Next, we examine the results ordered by p-value.

```{r}
# The results for taxa with the smallest p-values
res_daa_basic |> arrange(p_value) |> head() |> kable()
```

In the output

- `variable` indicates the predictor variable. As `patient_status` is a categorical variable, it also tells the group used as the non-reference level (here `ADHD` as `Control` was set as the reference level).
- `estimate` is log(odds ratio). Most importantly, positive values indicate abundance being higher in the non-reference group (here `ADHD`). Otherwise, estimates can be a bit difficult to interpret for ORMs. 

If we use the standard criterion for statistical significance, namely q-value < 0.05, we can see that the difference is statistically significant only for the genus named as [Ruminococcus]_gauvreauii_group. As the sign of the estimate is _negative_ it means that the relative abundance is _higher in the control group_.

This difference can be illustrated by histograms. From there we can clarly see how the relative abundances are higher in the control group.
 
```{r}
# First create a dataset for plotting
data_fig <- colData(tse) |> 
  as.data.frame() |> 
  cbind(assay(tse, 'relabundance') |>  t())

# Plot histograms by patient_status
ggplot(data_fig, aes(x = `[Ruminococcus]_gauvreauii_group`)) +
  geom_histogram(bins = 50) +
  facet_grid(patient_status ~ .) +
  labs(x = 'Relative abudance of [Ruminococcus]_gauvreauii_group' ) +
  theme_light(base_size = 12)
```
 

#### DAA with additional covariates {#sec-orm-cov}

Here we show how to perform DAA with an additional covariate included in the analysis. One usually wants to include a variable as a covariate if they suspect that it may act as a _confounding_ variable, namely, a variable that causes a spurious association between the variable of main interest and the abundances.

We use `cohort` as a covariate. We proceed as above except that we now have `formula = ~ patient_status + cohort`. Furthermore, as there are now many variables in the model we modify the code so that the multiplicity correction for the p-values is made separately for each variable.  


```{r run-orm-cov}
res_daa_cov <- assay(tse, 'relabundance') |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_orm(
    .,
    metadata = colData(tse) |> as.data.frame(),
    formula = ~ patient_status + cohort)
  ) |> 
  bind_rows(.id = "taxon") |> 
  group_by(variable) |>                        # Group by variable to correct p-values separately          
  mutate(                                      # for each variable.
    q_value = p.adjust(p_value, method = 'BH')
    ) |> 
  ungroup()

res_daa_cov |> head() |> kable()
```

Note that there now appear results also for the variable `cohort`. As `cohort` consists of three groups, namely, `Cohort_1` (the reference group), `Cohort_2` and `Cohort_3`, there are results separately for the comparison between `Cohort_2` vs. `Cohort_1` (`cohortCohort_2`) and for the comparison between `Cohort_3` vs. `Cohort_1` (`cohortCohort_3`).

We can now examine how the inclusion of `cohort` as a covariate affects the results for `patient_status`.

```{r}
# The results for the variable patient_status for the taxa with the smallest p-values
res_daa_cov |> 
  filter(variable == 'patient_statusADHD') |> 
  arrange(p_value) |> 
  head() |> 
  kable()
```

We can see that the results for [Ruminococcus]_gauvreauii_group and _Erysipelatoclostridium_ are in fact more clear (e.g. smaller p-values) when cohort is controlled for.

By examining the histograms by `patient_status` and `cohort`, we can see that the abundances are overall higher in cohort 3, and therefore the difference between the patients with and without ADHD seems clearer in cohorts 1 and 2. This seems to explain the difference in the results compared to the basic case (@sec-orm-basic).

```{r}
# Plot histograms by patient_status and cohort
ggplot(data_fig, aes(x = `[Ruminococcus]_gauvreauii_group`)) +
  geom_histogram(bins = 50) +
  facet_grid(patient_status ~ cohort) +
  labs(x = 'Relative abudance of [Ruminococcus]_gauvreauii_group') +
  theme_light(base_size = 12)
```


#### DAA with an interaction {#sec-orm-ia}

Here we show a more complex version of DAA, namely, a case when an interaction between two variables evaluated. We give this example here for illustration purposes but note that such complex analysis may not be advisable with a small dataset such as our example dataset.

We analyze interaction between `patient_status` and `cohort`. We proceed as in @sec-orm-cov above except that we now add also the interaction term into the formula. We could thus write `formula = ~ patient_status + cohort + patient_status:cohort` but recommend using the write the shorter version `formula = ~ patient_status * cohort`. Using the `*` between the terms automatically adds the interaction term as well as the main effect terms in the analysis. 

```{r run-orm-ia}
res_daa_ia <- assay(tse, 'relabundance') |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_orm(
    .,
    metadata = colData(tse) |> as.data.frame(),
    formula = ~ patient_status * cohort)
  ) |> 
  bind_rows(.id = "taxon") |> 
  group_by(variable) |>                                  
  mutate(                                      
    q_value = p.adjust(p_value, method = 'BH')
    ) |> 
  ungroup()

res_daa_ia |> head() |> kable()
```

Note that the results object `res_daa_ia` now also includes the interaction terms `patient_statusADHD:cohortCohort_2` and `patient_statusADHD:cohortCohort_3`. 

We examine the term `patient_status:cohortCohort_3`. It indicates how the differences between the ADHD and Control groups vary between cohorts 1 and 3.

```{r}
res_daa_ia |> 
  filter(variable == 'patient_statusADHD:cohortCohort_3') |> 
  arrange(p_value) |> 
  head() |> 
  kable()
```

The smallest p-value is observed for _Coprobacter_. By examining the histograms of relative abundances of _Coprobacter_ by `patient_status` and `cohort`, we can see that in cohort 1 the relative abundance is higher in the ADHD group while in cohort 3 the abundance is higher in the control group. That is the interaction effect. 

```{r}
# Plot histograms by patient_status and cohort
ggplot(data_fig, aes(x = `Coprobacter`)) +
  geom_histogram(bins = 50) +
  facet_grid(patient_status ~ cohort) +
  labs(x = 'Relative abudance of Coprobacter') +
  theme_light(base_size = 12)
```


### Analyzing presence/absence of taxa with logistic regression

We show here how to perform differential presence/absence analysis (DPAA) with logistic regression. When we analyze only the presence absence rather basic methods could be used. We note, however, that some versions of logistic regression perform badly in cases where, for instance, a taxon absent in all samples of one group (prevalence = 0) or when it is present in all samples of a group (prevalence = 1). We therefore employ a robust version of logistic regression introduced by @Firth1993. This version of logistic regression has been observed to perform well in DPAA [@Pelto2025].

We proceed as in the previous section @sec-daa. We first create a function that (for each taxon) runs logistic regression for presence/absence and returns the effect size estimate and p-value for each variable in the model.

```{r function-logr}
# The function uses the logistf() function from the logistf package 
library(logistf)

run_logr <- function(abundance, metadata, formula){
  
  #Create the design matrix of the model
  mm <- model.matrix(formula, metadata) |> 
    cbind(abundance) |> 
    tibble::as_tibble() |> 
    dplyr::select(-'(Intercept)')
  
  #Fit Firth logistic regression model
  fit <- logistf::logistf(abundance ~ ., data = mm)
  
  res <- data.frame(estimate = fit$coefficients,
                    p_value = fit$prob)
  
  return(res[-1, ] |> tibble::rownames_to_column('variable'))
}
```

Next, we transform count assay to presence/absences.

```{r transform-pa}
tse <- transformAssay(tse, assay.type = "counts", method = "pa")
```


#### The basic two-group comparison for presence/absence {#sec-logr-basic}

We first perform the basic two-group comparison without any other covariates. We compare the presence/absence of taxa between the people with and without ADHD.


```{r run-logr-basic}
res_dpaa_basic <- assay(tse, 'pa') |>           # Extract the assay of presence/absences
  t() |>                                        # Transpose the assay
  as.data.frame() |>                            # Transform to a data.frame
  map(~ run_logr(                               # Run ORM for each taxon
    .,                                            
    metadata = colData(tse) |> as.data.frame(), # Extract the metadata
    formula = ~ patient_status)                 # Model formula. Includes now only patient_status
    ) |> 
  bind_rows(.id = "taxon") |>                   # Bind the results of all taxa
  mutate(
    q_value = p.adjust(p_value, method = 'BH')) # Calculate q-values (multiplicity adjusted p-values)
```

Next, we examine the results sorted by p-value.

```{r}
# The results for taxa with the smallest p-values
res_dpaa_basic |> arrange(p_value) |> head() |> kable()
```

Now the `estimate` in the output is the familiar log(odds ratio) of logistic regression. A positive value indicates the taxon being more prevalent in the non-refrence group (here `ADHD`).

The smallest p-value and q-value is observed for genus _Erysipelatoclostridium_. As the sign of the estimate is _negative_ is means that the genus is more prevalent in the control group. This difference can be illustrated e.g. by bar plots. From there we can clearly see that _Erysipelatoclostridium_ is more prevalent in the control group.
 
```{r}
# First create a dataset for plotting
data_fig_pa <- colData(tse) |> 
  as.data.frame() |> 
  cbind(assay(tse, 'pa') |> 
          t() |>
          as.data.frame() |> 
          mutate_all(~ ifelse(. == 1, 'Present', 'Absent')))

# A bar plot on presence/absence of Erysipelatoclostridium
ggplot(data_fig_pa, aes(x = `Erysipelatoclostridium`)) +
  geom_bar() +
  facet_grid(patient_status ~ .) +
  theme_light()
```


#### DPAA with additional covariates {#sec-logr-cov}

Here we show how to perform DPAA with an additional `cohort` included in the analysis. We proceed as in the section @sec-orm-cov.

```{r run-logr-cov}
res_dpaa_cov <- assay(tse, 'pa') |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_logr(
    .,
    metadata = colData(tse) |> as.data.frame(),
    formula = ~ patient_status + cohort)
  ) |> 
  bind_rows(.id = "taxon") |> 
  group_by(variable) |>                        # Group by variable to correct p-values separately          
  mutate(                                      # for each variable.
    q_value = p.adjust(p_value, method = 'BH')
    ) |> 
  ungroup()

# The results for the variable patient_status for the taxa with the smallest p-values
res_dpaa_cov |> 
  filter(variable == 'patient_statusADHD') |> 
  arrange(p_value) |> 
  head() |> 
  kable()

# Bar plots for presence/absence of Erysipelatoclostridium by patient_status and cohort
ggplot(data_fig_pa, aes(x = `Erysipelatoclostridium`)) +
  geom_bar() +
  facet_grid(patient_status ~ cohort) +
  theme_light()
```


#### DPAA with an interaction {#sec-logr-ia}

Again, we proceed similarly as in the section @sec-orm-ia.

```{r run-logr-ia}
res_dpaa_ia <- assay(tse, 'pa') |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_logr(
    .,
    metadata = colData(tse) |> as.data.frame(),
    formula = ~ patient_status * cohort)
  ) |> 
  bind_rows(.id = "taxon") |> 
  group_by(variable) |>                                  
  mutate(                                      
    q_value = p.adjust(p_value, method = 'BH')
    ) |> 
  ungroup()

# Interaction results for patient_statusADHD:cohortCohort_3
res_daa_ia |> 
  filter(variable == 'patient_statusADHD:cohortCohort_3') |> 
  arrange(p_value) |> 
  head() |> 
  kable()

# Bar plots for presence/absence of Coprobacter by patient_status and cohort
ggplot(data_fig_pa, aes(x = `Coprobacter`)) +
  geom_bar() +
  facet_grid(patient_status ~ cohort) +
  theme_light()
```

The figure illustrates how _Coprobacter_ is more prevalent in the ADHD group in cohort 1 but more prevalent in the control group in cohort 3.


### DAA with linear regression/t-test {#sec-daa_lm}

Relative abundances can also be analyzed using the standard linear regression methods (e.g. the t-test) instead of ORM. This, however, requires in practice some transformation to be applied for the abundances. A typical transformation is log transformation. It is, however, problematic because zeros cannot be log transformed. Instead, zeros must be treated somehow before the transformation (e.g. by using pseudo-count).

Analyzing log transformed relative abundances can be advantageous, however, for instance in the case where one wants to analyze only _positive_ (non-zero) relative abundances. Such analysis can sometimes be biologically relevant, especially when combined with analysis of presence/absence of taxa.

Again, we first create a function to run a linear model for each taxon. When one analyzes only positive relative abundances the sample size for each taxon is the number of non-zero abundances for the taxon. The function thus also returns this number `n_non_zero`. As there may be taxa with too few non-zero abundances to fit certain model the function does not return any result for such taxa.   


```{r function-lm}
run_lm <- function(abundance, metadata, formula){
  
  #Create the design matrix of the model
  mm <- model.matrix(formula, metadata) |> 
    cbind(abundance) |> 
    tibble::as_tibble() |> 
    dplyr::filter(abundance > -40) |> 
    dplyr::select(-'(Intercept)')
  
  #Fit linear model
  fit <- lm(abundance ~ ., data = mm)
  
  res <- coef(summary(fit)) |> 
    as.data.frame() |> 
    tibble::rownames_to_column('variable') |> 
    mutate(n_non_zero = nrow(mm)) |> 
    filter(fit$rank == ncol(mm)) |>  
    dplyr::select(variable, estimate = Estimate, p_value = 'Pr(>|t|)', n_non_zero)
  
  return(res[-1, ])
}
```


#### Analyzing _positive_ relative abundances with linear regression

We first log transform relative abundances using base 2 logarithm transformation. An extremely small pseudo-count (`pseudocount = 2 ^ -41`) is added to relative abundances in the last step. It has essentially no effect on the results but it is needed to perform log transformation with the `transformAssay()` function. Log transformed zeros are automatically filtered out from the analysis by the `run_lm` function.

```{r}
tse <- tse |> 
  transformAssay(assay.type = "counts", method = "relabundance") |> 
  transformAssay(assay.type = "relabundance", method = "log2", pseudocount = 2 ^ -41)
```

One can then proceed as with ORM. For instance, the basic comparison between two groups can be performed as follows. Note that as we use the $log_{2}$ transformation the estimates are now $log_{2}$-fold changes.

```{r}
res_daa_pos_basic <- assay(tse, 'log2') |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_lm(.,
                metadata = colData(tse) |> as.data.frame(),
                formula = ~ patient_status)
  ) |> 
  bind_rows(.id = "taxon") |> 
  mutate(q_value = p.adjust(p_value, method = 'BH'))

res_daa_pos_basic |> arrange(p_value) |> head() |> kable()
```

#### Performing DAA with linear regression

Lastly, we show an example of how to analyze log transformed relative abundances with a linear model. To include zeros in the analysis we add a pseudo-count to count data before transforming to relative abundances. The relative abundances are then log transformed (using base 2 logarithm).

```{r}
tse_lm <- tse |> 
  transformAssay(assay.type = "counts", method = "relabundance", pseudocount = T) |> 
  transformAssay(assay.type = "relabundance", method = "log2")
```

We can then perform for instance the basic two-group comparison as before. (Note the use of `tse_lm` instead of `tse`.)

```{r}
res_daa_basic_lm <- assay(tse_lm, 'log2') |> 
  t() |> 
  as.data.frame() |> 
  map(~ run_lm(.,
               metadata = colData(tse) |> as.data.frame(),
               formula = ~ patient_status)
  ) |> 
  bind_rows(.id = "taxon") |> 
  mutate(q_value = p.adjust(p_value, method = 'BH'))

res_daa_basic_lm |> arrange(p_value) |> head() |> kable()
```

We can see that for many taxa the p-values are rather similar to the corresponding p-values provided by ORM in section @sec-orm-basic.



::: {.callout-tip}
## Additional resources
While we generally recommend the use of the above simple methods in DAA there exist dozens of alternative methods. A (non-comprehensive) list of these methods is given below.

- ALDEx2 [@Gloor2016]
- ANCOM [@Mandal2015]
- ANCOMBC [@ancombc2020]
- BIRDMAn [@Rahman2023]
- corncob [@Martin2021]
- DACOMP [@Brill2019]
- DESeq2 [@Love2014]
- eBay [@Liu2020]
- edgeR [@Chen2016]
- fastANCOM [@fastANCOM2022]
- fido [@Silverman2019]
- LDM [@Hu2020]
- lefser [@Khlebrodova2021]
- limma [@Ritchie2015]
- LinDA [@Zhou2022]
- MaAsLin2 [@Mallick2020]
- MaAsLin3 [@Nickols2024]
- metagenomeSeq [@Paulson2017]
- Omnibus [@Omnibus2018]
- RAIDA [@Sohn2015]
- sccomp [@mangiola2023]
- ZicoSeq [@Yang2022]
- ZINQ [@Ling2021]
:::

::: {.callout-tip icon=false}

## Exercises

**Goal:** The goal of these exercises it to learn how to perform differential
abundance and presence/absence analysis and to visualize the results.

**Exercise 1: Differential abundance analysis (DAA)**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Observe `colData`. Ensure that the data includes meaningful sample grouping.
If it does not, choose another dataset.

3. Agglomerate the data by genus and filter by a prevalence of 10%.

4. Run the basic two-group DAA and DPAA using ORM and logistic regression, respectively. (You can also try DAA with linear regression.)

5. Explore the results tables. Are there any taxa with significant abundance/prevalence differences?

6. Run the same analyses now with an additional covariate. Do the results differ? Why?

7. You can try also more complex analysis, e.g. with several covariates and/or with interactions.

8. Visualize the results.

Useful functions:

`data()`, `colData`, `subsetByPrevalent()`

:::
