# Differential abundance analysis {#sec-extra_daa}

```{r}
#| label: setup
#| echo: false

knitr::opts_chunk$set(eval = TRUE, warning = FALSE, message = FALSE)
```

### Differential abundance analysis with ordinal regression model {#sec-orm}

We show here how to perform DAA by analyzing relative abundances with ordinal
regression model (ORM). We have chosen ORM as it makes few assumptions on the
distributions of abundances while it can incorporate covariates (including
interaction terms) in the analysis. As ORM performs very similarly to the
familiar non-parametric Wilcoxon test in the basic two group comparison it can
also be seen as an extension of the Wilcoxon test (@Harrell2001, @Liu2017).

By using ORM to model the skewed and/or zero inflated distributions of microbial
abundances, we can avoid the strong parametric assumptions (e.g. negative
binomial distribution) made by some methods. Furthermore, we do not need to
transform abundances. This good because the choice of the transformation may be
somewhat arbitrary while it may affect the DAA results. Moreover, ORM has been
shown to perform well in practice in DAA when considering its sensitivity and
the consistency of its results (@Pelto2025).

We first create a function that (for each taxon) runs ORM and returns the effect
size estimate and p-value (based on score test) for each variable in the model.

```{r}
#| label: function_orm

# The function uses the orm() function from the rms package
library(rms)

run_orm <- function(abundance, metadata, formula) {
    # Create the design matrix of the model
    mm <- model.matrix(formula, metadata) |>
        cbind.data.frame(abundance)
    mm[["(Intercept)"]] <- NULL

    inds <- 1:(ncol(mm) - 1)
    vars <- colnames(mm)[inds]

    # Fit ordinal regression model with all variables
    # (Intercepts are automatically added)
    fit_1 <- orm(abundance ~ ., data = mm)
    score_1 <- fit_1$stats["Score"]

    res <- data.frame(estimate = fit_1$coefficients[vars], p_value = NA)

    # Calculate p-value based on score test for each variable
    if (length(inds) == 1) {
        res$p_value <- fit_1$stats["Score P"]
    } else {
        for (i in inds) {
            fit_0 <- orm(abundance ~ ., data = mm[, -i])
            score_0 <- fit_0$stats["Score"]
            res$p_value[i] <- as.numeric(1 - pchisq(score_1 - score_0, df = 1))
        }
    }

    res[["variable"]] <- rownames(res)
    rownames(res) <- NULL
    return(res)
}
```

We use the same Tengler2020 dataset as in @sec-differential-abundance.

```{r}
#| label: orm_prepare_data

library(mia)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse, rank = "Genus", prevalence = 0.10)

# Transform count data and create assay of relative abundances
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")

# Transform patient_status to a factor and set "Control" as the reference level
tse$patient_status <- factor(tse$patient_status, levels = c("Control", "ADHD"))

# We also transform cohort as a factor
tse$cohort <- factor(tse$cohort, levels = c("Cohort_1", "Cohort_2", "Cohort_3"))
```

#### The basic two-group comparison {#sec-orm-basic}

We compare the relative abundance of taxa between the people with and without
ADHD (indicated by the variable `patient_status`).

The code below works as follows.

 1. We extract the assay with relative abundances from the `tse` object
 2. The assay is transposed and transformed to a `data.frame`
 3. The function `run_orm` is run for each taxon
    i) The `.` indicates here the relative abundance of each taxon
    ii) The metadata including the variable `patient_status` is extracted from
    the `tse` object
    iii) The model formula indicates that we included only `patient_status` in
    our analysis
 4. The results for all taxa are binded to as one `data.frame`
 5. Lastly, we calculate p-values adjusted from multiple testing (`q-value`) by
 using the standard FDR correction by @Benjamini1995 (`method = "BH"`).


```{r}
#| label: run_orm_basic

library(purrr)
library(dplyr)
library(knitr)

res_daa_basic <- assay(tse, "relabundance") |> # Extract relative abundances
    t() |> # Transpose the assay
    as.data.frame() |> # Transform to a data.frame
    map(
        ~ run_orm( # Run ORM for each taxon
            .,
            metadata = colData(tse) |> as.data.frame(), # Extract the metadata
            formula = ~patient_status # Model formula. Now only patient_status
        )
    ) |>
    bind_rows(.id = "taxon") |> # Bind the results of all taxa
    mutate(
        q_value = p.adjust(p_value, method = "BH")
    ) # Calculate q-values
# (FDR adjusted p-values)

# The results for taxa with the smallest p-values
res_daa_basic |>
    arrange(p_value) |>
    head() |>
    kable()
```

In the output

- `variable` indicates the predictor variable. As `patient_status` is a
categorical variable, it also tells the group used as the non-reference level
(here `ADHD` as `Control` was set as the reference level).
- `estimate` is log(odds ratio). Most importantly, positive values indicate
abundance being higher in the non-reference group (here `ADHD`). Otherwise,
estimates can be a bit difficult to interpret for ORMs.

#### More complex analyses {#sec-orm-complex}

Analysis with covariates and analysis of interaction terms can be performed in
by just using different `formula` in the `run_orm` function, e.g.
`formula = patient_status + cohort` or `formula = patient_status * cohort`.

### Analysis of presence/absence with Firth logistic regression {#sec-firth}

As some versions of logistic regression perform badly in
the cases where, for instance, the prevalence of a taxon is zero in one
group we show here how to perform DPAA with a robust version of logistic
regression introduced by @Firth1993. This version has been observed to perform
well in DPAA in practice [@Pelto2025].

We first create a function that runs Firth logistic regression for
presence/absence and returns the effect size estimate and p-value for each
variable in the model.

```{r}
#| label: function_firth

# The function uses the logistf() function from the logistf package
library(logistf)

run_firth <- function(abundance, metadata, formula) {
    # Create the design matrix of the model
    mm <- model.matrix(formula, metadata) |>
        cbind.data.frame(abundance)
    mm[["(Intercept)"]] <- NULL

    # Fit Firth logistic regression model
    fit <- logistf(abundance ~ .,
        data = mm,
        control = logistf.control(maxit = 1000)
    )

    res <- data.frame(
        estimate = fit$coefficients,
        p_value = fit$prob
    )

    res <- res[!rownames(res) %in% c("(Intercept)"), , drop = FALSE]
    res[["variable"]] <- rownames(res)
    rownames(res) <- NULL

    return(res)
}
```

We use the same Tengler2020 dataset as in @sec-differential-abundance.

```{r}
#| label: firth_prepare_data

library(mia)
library(scater)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Agglomerate by genus and subset by prevalence
tse <- subsetByPrevalent(tse, rank = "Genus", prevalence = 0.10)

# Transform count data and create assay of presence/absences
tse <- transformAssay(tse, assay.type = "counts", method = "pa")

# Transform patient_status to a factor and set "Control" as the reference level
tse$patient_status <- factor(tse$patient_status, levels = c("Control", "ADHD"))

# We also transform cohort as a factor
tse$cohort <- factor(tse$cohort, levels = c("Cohort_1", "Cohort_2", "Cohort_3"))

# Calculate the sequencing depth (the total number of reads/counts) for each
# sample (needed when analyzing presence/absence of taxa)
tse <- addPerCellQC(tse)
```

#### The basic two-group comparison {#sec-firth-basic}

We perform here the basic two-group comparison while controlling for sequencing
depth (`reads`) as in section @sec-dpaa. We compare the presence/absence of taxa
between the people with and without ADHD.

```{r}
#| label: run_firth_basic

library(purrr)
library(dplyr)
library(knitr)

res_firth_basic <- assay(tse, "pa") |> # Extract presence/absences
    t() |> # Transpose the assay
    as.data.frame() |> # Transform to a data.frame
    map(
        ~ run_firth( # Run logistic regression
            ., # for each taxon
            metadata = colData(tse) |> as.data.frame(), # Extract the metadata
            formula = ~ patient_status + total # Model formula
        )
    ) |>
    bind_rows(.id = "taxon") |> # Bind the results of all taxa
    filter(variable != "total") |> # Remove the variable total
    mutate(
        q_value = p.adjust(p_value, method = "BH")
    ) # Calculate q-values
# (FDR adjusted p-values)

# The results for taxa with the smallest p-values
res_firth_basic |>
    arrange(p_value) |>
    head() |>
    kable()
```

In the output

- `variable` indicates the predictor variable. As `patient_status` is a
categorical variable, it also tells the group used as the non-reference level
(here `ADHD` as `Control` was set as the reference level).
- `estimate` is the familiar log(odds ratio). A positive value indicates the
taxon being more prevalent in the non-reference group (here `ADHD`).

#### More complex analyses {#sec-firth-complex}

Analysis with covariates and analysis of interaction terms can be performed in
by just using different `formula` in the `run_firth` function, e.g.
`formula = patient_status + cohort + reads` or
`formula = patient_status * cohort + reads`.
