# Multi-omics prediction and classification {#sec-multi-omics-integration}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

This chapter illustrates how we can create a classification model that utilizes
multiomics data. In contrast [@sec-machine_learning] shows classification models
that utilizes single omic data.

In multiview data analysis, there are two main types of approaches: early fusion
and late fusion. \underline{Early fusion} combines all datasets into a single
representation, which then serves as input for a supervised learning model. In
contrast, \underline{late fusion} builds individual models for each data view
and combines their predictions using a second-level model as the final
predictor. However, these traditional paradigms treat the data views in
isolation and do not allow for interactions or dependencies between them. A more
advanced method, called cooperative learning [@Ding2022], which is also known as
\underline{intermediate fusion}, combines the best of both worlds by encouraging
predictions from different data views to align through an agreement parameter
($\rho$).

## Import and preprocess data

We use the publicly available type 2 diabetes (T2D) data from the 
`curatedMetagenomicData` package [@Lloyd-Price2019], where we aim to 
predict T2D disease status based on both taxonomic (species abundances) and 
functional (pathway abundances) profiles. Note that the dataset is the same as
used in [@sec-machine_learning] for single omic ML model.

```{r}
#| label: load-pkg-data

library(curatedMetagenomicData)
library(dplyr)
library(mia)

tse1 <- sampleMetadata |>
  filter(study_name == "QinJ_2012") |>
  returnSamples("relative_abundance", rownames = "short")

tse2 <- sampleMetadata |>
  filter(study_name == "QinJ_2012") |>
  returnSamples("pathway_abundance", rownames = "short")

# Create a MAE object
mae <- MultiAssayExperiment(
    ExperimentList(microbiota = tse1, pathway = tse2)
    )
mae
```

Further data pre-processing is necessary to handle near-zero-variance features
in this dataset. A combination of variance and prevalence filtering
is applied to the dataset to include only meaningful features.

```{r}
#| label: preprocess

# Subset by prevalence
mae[[1]] <- subsetByPrevalent(
    mae[[1]], assay.type = "relative_abundance", prevalence = 0.1,
    include.lowest = TRUE)
mae[[2]] <- subsetByPrevalent(
    mae[[2]], assay.type = "pathway_abundance", prevalence = 0.1,
    include.lowest = TRUE)

# Subset those features that have near zero variance
rowData(mae[[1]])[["sd"]] <- rowSds(assay(mae[[1]], "relative_abundance"))
rowData(mae[[2]])[["sd"]] <- rowSds(assay(mae[[2]], "pathway_abundance"))
mae[[1]] <- mae[[1]][ rowData(mae[[1]])[["sd"]] > 0.001, ]
mae[[2]] <- mae[[2]][ rowData(mae[[2]])[["sd"]] > 0.001, ]

# Transform the data with CLR
mae[[1]] <- transformAssay(
  mae[[1]], assay.type = "relative_abundance", method = "clr",
  pseudocount = TRUE)
mae[[2]] <- transformAssay(
  mae[[2]], assay.type = "pathway_abundance", method = "clr",
  pseudocount = TRUE)
```

Ultimately, `r nrow(mae[[1]])+nrow(mae[[1]])` features are retained, consisting
of `r nrow(mae[[1]])` pathways and `r nrow(mae[[2]])` species.

We randomly select 20% of the samples for the validation set, ensuring they are
not used in training. This approach provides a more robust estimate of how well
the model generalizes to unseen data.

```{r}
#| label: select_validation

set.seed(3747)
colData(mae)[["validation_set"]] <- sample(
    c(TRUE, FALSE), size = ncol(mae[[1]]), replace = TRUE, prob = c(0.2, 0.8))
```

Now, we are ready to fit the model using the `IntegratedLearner` package, which
supports early, late, and intermediate fusion methods [@Mallick2024]. Under the
hood, it leverages the `SuperLearner` package. The process begins by fitting
"base learners" for each layer, with separate models trained for each omic
dataset. Then, a "meta learner" integrates the outputs from these individual
models to make the final predictions.

```{r}
#| label: fit_model

# For some reason Java Virtual Machine is often incorrectly configured. This
# makes sure that enough memory is assigned to JVM (5GB).
options(java.parameters = "-Xmx5g")

library(IntegratedLearner)
library(SuperLearner)

# Fit the model
fit <- IntegratedLearnerFromMAE(
    mae,
    # Select experiments and assays
    experiment = names(experiments(mae)),
    assay.type = c("clr", "clr"),
    # Select columns from sample metadata
    outcome.col = "disease",
    valid.col = "validation_set",
    # Options for base and meta models
    base_learner = "SL.bartMachine",
    meta_learner = "SL.nnls.auc",
    # The outcome is binary
    family = binomial()
)
```
The outputs include the following models:

    - Individual models: Trained separately for each omic dataset.
    - Stacked model (late fusion): Combines the predictions from each individual model into a single meta-model.
    - Concatenated model (early fusion): Trains a model where all features are merged before training.

We can summarize the model performance using a Receiver Operating Characteristic (ROC) plot.

```{r}
#| label: visualize_result

library(tidyverse)
library(cowplot)

p <- IntegratedLearner:::plot.learner(fit)
```

Based on the ROC plot described below, we observe that the AUC is `r fit$AUC.test[[2]]` when
considering only the pathway abundance data in the model, and `r fit$AUC.test[[1]]` for the model
including only the species abundance data. The AUC increases to `r fit$AUC.test[[4]]` when using
the early fusion model and reaches `r fit$AUC.test[[3]]` with the late fusion model.
Overall, most integrated
classifiers outperform individual layers in distinguishing between T2D and
healthy controls.

When using the BART machine, in addition to point predictions, we can also generate:

    1. Credible intervals for all observations,
    2. Estimated layer weights in the meta-model, and
    3. Feature importance scores.

For visualization tips on how to plot these and more, refer to the
[IntegratedLearner vignette](http://htmlpreview.github.io/?https://github.com/himelmallick/IntegratedLearner/blob/master/vignettes/IntegratedLearner.html).





```{r}
#| label: capture output

library(mcmcplots)

capture.output(temp <- caterplot(t(weighted_post_samples), add = FALSE))

```



Finally, we visualize the results for the top 20 features for each layer based
on the intermediate fusion.

```{r}

```


## Prediction of fat and prebiotics in diet of rats

As an example we would build a random forest classifier based on the 
HintikkaxOData multi-omics profiles to classify fat and prebiotics in diet of rats. 

### With validation set

```{r valid-set}
# load required libraries 

library(mia)
library(SuperLearner)
library(IntegratedLearner)
library(tidyverse)
library(cowplot)

# load multi-omic data 

data("HintikkaXOData")
mae <- HintikkaXOData

# Add validation set info to colData 


colData(mae)[["valid"]] <- sample(c(TRUE, FALSE), nrow(colData(mae)), 
                                  replace = TRUE)

# Run with validation set 


fit1 <- IntegratedLearnerFromMAE(
    mae,
    experiment = c(1, 2, 3),
    assay.type = c("counts", "nmr", "signals"),
    outcome.col = "Fat",
    valid.col = "valid",
    folds = 10,
    na.rm = TRUE,
    base_learner = 'SL.randomForest',
    meta_learner = 'SL.nnls.auc',
    verbose = TRUE,
    family = binomial()
)

```

IntegratedLearner offers easily accessible and interpretable summary outputs 
including 1) computation time, 2) per-layer AUC/ R2 scores on training , 3) AUC/ R2
metrics for stacked and concatenated model if run_stacked=TRUE and r
un_concat=TRUE and 4) estimated per-layer weights from meta learner in stacked 
model.

We can visualize the classification performance by constructing layer-wise ROC 
curves for both train and test set using plot.learner() function that takes 
IntegratedLearner object as input.

```{r plot fit1}

# Summary outputs 

plot <- IntegratedLearner:::plot.learner(fit1)

```

The stacked model achieves superior accuracy to individual layers and concatenation 
model in independent validation, indicating that the stacked multi-omics classifier 
leads to a competitive or superior cross-validated and independent validation 
classification accuracy than its single-omics counterparts.

### Without validataion set

The default base model recommendation for IntegratedLearner is Bayesian additive 
regression trees or BART (base_learner='SL.BART'). Using BART as base-model yields 
uncertainty estimates (i.e. credible intervals) of the prediction and model 
parameters in addition to reporting a small set of interpretable features for 
follow-up experiments (i.e. feature importance scores).

```{r no_valid set}

# Run without validation set 

fit2 <- IntegratedLearnerFromMAE(
    mae,
    experiment = c(1, 2, 3),
    assay.type = c("counts", "nmr", "signals"),
    outcome.col = "Diet",
    folds = 10,
    na.rm = TRUE # The data has missing values that are removed
)

```

```{r plot fit2}

# Summary outputs 

plot <- IntegratedLearner:::plot.learner(fit2)

```

As before, the multi-omics stacked prediction model leads to a better 
cross-validated accuracy than its single-omics counterparts and concatenation 
model.

When base_learner='SL.BART', in addition to point predictions, we can also 
generate 1) credible intervals for all observations, 2) estimated layer weights 
in the meta model and 3) feature importance scores.

```{r weights fit2}
weights <- fit2$weights

dataX <- fit2$X_train_layers
dataY <- fit2$Y_train


post_samples <- vector("list", length(weights))
names(post_samples) <- names(dataX)

for(i in seq_along(post_samples)){
post_samples[[i]] <- bartMachine::bart_machine_get_posterior(
    fit2$model_fits$model_layers[[i]], dataX[[i]])$y_hat_posterior_samples
}

weighted_post_samples <-Reduce('+', Map('*', post_samples, weights))
rownames(weighted_post_samples) <- rownames(dataX[[1]])
names(dataY) <- rownames(dataX[[1]])
```

```{r capture output}

library(mcmcplots)
capture.output(temp <- caterplot(t(weighted_post_samples), add = FALSE))

```

We show below the 68% and 95% credible intervals obtained from stacked model for 
all 40 observations. The filled circle indicates the posterior median and empty 
circle indicates the true value of the observation.

```{r plot capture}

temp <- caterplot(t(weighted_post_samples),
          horizontal = FALSE, add = FALSE)
points(dataY[temp])
title(main ="", xlab = "Observations", ylab = "Fat and prebiotics in rats",
       line = NA, outer = FALSE)

```
