# Multi-omics Integration {#sec-multi-omics-integration}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

In multiview data analysis, there are two main types of approaches: early fusion
and late fusion. \underline{Early fusion} combines all datasets into a single
representation, which then serves as input for a supervised learning model. In
contrast, \underline{late fusion} builds individual models for each data view
and combines their predictions using a second-level model as the final
predictor. However, these traditional paradigms treat the data views in
isolation and do not allow for interactions or dependencies between them. A more
advanced method, called cooperative learning [@Ding2022], which is also known as
\underline{intermediate fusion}, combines the best of both worlds by encouraging
predictions from different data views to align through an agreement parameter
($\rho$).

## Prediction of fat and prebiotics in diet of rats

As an example we would build a random forest classifier based on the 
HintikkaxOData multi-omics profiles to classify fat and prebiotics in diet of rats. 

### With validation set

```{r valid-set}

###########################
# load required libraries #
###########################
library(mia)
library(SuperLearner)
library(IntegratedLearner)
library(tidyverse)
library(cowplot)

########################
# load multi-omic data #
########################
data("HintikkaXOData")
mae <- HintikkaXOData

######################################
# Add validation set info to colData #
######################################

colData(mae)[["valid"]] <- sample(c(TRUE, FALSE), nrow(colData(mae)), 
                                  replace = TRUE)

###########################
# Run with validation set #
###########################

fit1 <- IntegratedLearnerFromMAE(
    mae,
    experiment = c(1, 2, 3),
    assay.type = c("counts", "nmr", "signals"),
    outcome.col = "Fat",
    valid.col = "valid",
    folds = 10,
    na.rm = TRUE,
    base_learner = 'SL.randomForest',
    meta_learner = 'SL.nnls.auc',
    verbose = TRUE,
    family = binomial()
)

```

IntegratedLearner offers easily accessible and interpretable summary outputs 
including 1) computation time, 2) per-layer AUC/ R2 scores on training , 3) AUC/ R2
metrics for stacked and concatenated model if run_stacked=TRUE and r
un_concat=TRUE and 4) estimated per-layer weights from meta learner in stacked 
model.

We can visualize the classification performance by constructing layer-wise ROC 
curves for both train and test set using plot.learner() function that takes 
IntegratedLearner object as input.

```{r plot fit1}
###################
# Summary outputs #
###################

plot <- IntegratedLearner:::plot.learner(fit1)

```

The stacked model achieves superior accuracy to individual layers and concatenation 
model in independent validation, indicating that the stacked multi-omics classifier 
leads to a competitive or superior cross-validated and independent validation 
classification accuracy than its single-omics counterparts.

### Without validataion set

The default base model recommendation for IntegratedLearner is Bayesian additive 
regression trees or BART (base_learner='SL.BART'). Using BART as base-model yields 
uncertainty estimates (i.e. credible intervals) of the prediction and model 
parameters in addition to reporting a small set of interpretable features for 
follow-up experiments (i.e. feature importance scores).

```{r no_valid set}

##############################
# Run without validation set #
##############################

fit2 <- IntegratedLearnerFromMAE(
    mae,
    experiment = c(1, 2, 3),
    assay.type = c("counts", "nmr", "signals"),
    outcome.col = "Diet",
    folds = 10,
    na.rm = TRUE # The data has missing values that are removed
)

```

```{r plot fit2}

###################
# Summary outputs #
###################

plot <- IntegratedLearner:::plot.learner(fit2)

```

As before, the multi-omics stacked prediction model leads to a better 
cross-validated accuracy than its single-omics counterparts and concatenation 
model.

When base_learner='SL.BART', in addition to point predictions, we can also 
generate 1) credible intervals for all observations, 2) estimated layer weights 
in the meta model and 3) feature importance scores.

```{r weights fit2}
weights <- fit2$weights

dataX <- fit2$X_train_layers
dataY <- fit2$Y_train


post_samples <- vector("list", length(weights))
names(post_samples) <- names(dataX)

for(i in seq_along(post_samples)){
post_samples[[i]] <- bartMachine::bart_machine_get_posterior(
    fit2$model_fits$model_layers[[i]], dataX[[i]])$y_hat_posterior_samples
}

weighted_post_samples <-Reduce('+', Map('*', post_samples, weights))
rownames(weighted_post_samples) <- rownames(dataX[[1]])
names(dataY) <- rownames(dataX[[1]])
```

```{r capture output}

library(mcmcplots)
capture.output(temp <- caterplot(t(weighted_post_samples), add = FALSE))

```

We show below the 68% and 95% credible intervals obtained from stacked model for 
all 40 observations. The filled circle indicates the posterior median and empty 
circle indicates the true value of the observation.

```{r plot capture}

temp <- caterplot(t(weighted_post_samples),
          horizontal = FALSE, add = FALSE)
points(dataY[temp])
title(main ="", xlab = "Observations", ylab = "Fat and prebiotics in rats",
       line = NA, outer = FALSE)

```
