# Machine learning {#sec-machine_learning}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```
## Background

Machine learning (ML) is a branch of artificial intelligence. While there are
several definitions of machine learning, it generally involves using
computational methods (i.e., machines) to identify patterns in data 
(i.e., learning).

Machine learning can be divided into supervised and unsupervised learning.
Supervised ML is used to predict outcomes based on labeled data, whereas
unsupervised ML is used to discover unknown patterns and structures within
the data.

### Unsupervised machine learning

"Unsupervised" means that the labels (e.g., patient status) are not known by the
model during its training, and patterns are learned based solely on the data,
such as an abundance table.

Common tasks in unsupervised machine learning include dimension reduction and
clustering that were already discussed in [@sec-community-similarity] and
[@sec-clustering], respectively.

### Supervised machine learning

"Supervised" means that a model is trained on observations paired with a known
output (e.g., patient status: healthy or diseased). During training, the
model learns patterns from a portion of the data and it is then evaluated on its
ability to generalize to new data. This process involves splitting the collected
data into a training and testing sets, commonly in 80/20 ratio, although other
proportions can be used depending on the size and applications of the dataset.

Training is usually enhanced with cross-validation to improve the model's
robustness. However, when the dataset is small, splitting it into training and
test sets might not be feasible. In such cases, cross-validation alone can be
used to provide a rough estimate of the model's performance. While this approach
is not as reliable as having a separate test set, it can still give valuable
insights into how well the model might perform on new data. 

Common tasks for supervised machine learning includes classification (e.g.,
predict categorical variables) and regression (e.g., predicting continuous
variables). The focus of this chapter are classification tasks, but note that
some ML models discussed below can also be used for regression.

::: {.callout-note}
## Note: ML in multi-omics data analysis
ML applications for the integration of multi-omic datasets is covered in
[@sec-cross-correlation], [@sec-multiassay_ordination] and
[@sec-multi-omics-integration]
:::

## Setup

The fecal microbiome data from 
[Qin, J. et al 2012](https://pubmed.ncbi.nlm.nih.gov/23023125/) [@qin2012_t2d] 
will be used to illustrate how to deploy two supervised machine learning 
algorithms commonly used to classify observations into outcomes of clinical
relevance. In this case, we are interesting in classifying subjects in two
groups: Type II diabetes or Healthy (encoded as T2D and healthy in the metadata).

The R package `mikropml`[@mikropmlSoftware;@mikropmlApplied] will be used to
achieve that goal. This package was developed to offer a user-friendly interface
to supervised ML algorithms implemented in the `caret` package.

[Here](https://www.schlosslab.org/mikropml/articles/introduction.html#the-methods-we-support)
is a list of models well supported by mikropml and
[here](https://topepo.github.io/caret/available-models.html)
a complete list of other models available in caret that *might* work when
called through mikropml.

```{r}
#| label: superML1 - Setup
#| output: FALSE

library(tidyverse)
library(curatedMetagenomicData)

samples <- sampleMetadata[ sampleMetadata[["study_name"]] == "QinJ_2012", ]

tse <- returnSamples(
    samples, 
    dataType = "relative_abundance",
    counts = TRUE, 
    rownames = "short"
    )

table(tse[["disease"]])
##
## healthy     T2D 
##     193     170
```

## Data preprocessing

Before applying any ML algorithms, the data must be preprocessed. This speeds 
up the training of the models by reducing the amount of features analysed, a 
desirable outcome when working with high-dimensional microbiome data. 

Besides faster performance, common pre-processing steps have biological
justifications. For instance:

* **Collapse highly correlated features:** It's common for two or more taxa to 
show high correlation in their abundances within a community. Thus, removing or
collapsing correlated taxa allows the model to analyse them as one group.
* **Remove features with near-zero variance:** Features that don't vary across
groups can hardly help in discerning between them, as they don't hold any
biologically relevant information. Additionally, these variables can effectively 
become of zero variance under certain data splits.
* **Remove features with low prevalence:** Microbiome data is sparse,
and taxa present in just a few samples of each group hardly provide useful 
information for their classification. 

::: {.callout-important}
## Important: Data leakage is a common pitfall.

Data leakage occurs when information from the test set influences the model
training process, leading to overly optimistic performance estimates. This
can happen, for example, if preprocessing steps like scaling are applied to
both the training and test data together, allowing the model to indirectly
"see" the test data during training. Fortunately, there are questions we can
ask to ourselves to void data leakage [@bernett2024_dataLeakage] 
:::

```{r}
#| label: superML2 - preprocessing
#| output: FALSE

library(mia)

# Preprocessing step 1 - 
# Keep taxa present in more than 10% of samples
tse_prev <- subsetByPrevalent(tse, 
                              assay.type = "relative_abundance",
                              prevalence = 10/100)

# calculate all available alpha diversity measures
vars_before <- colnames(colData(tse))
tse_prev <- addAlpha(tse_prev, assay.type = "relative_abundance")

# By comparing variables, we get which indices were calculated
index <- colnames(colData(tse_prev))[!colnames(colData(tse_prev)) %in% vars_before]

# Apply CLR transform
tse_prev <- transformAssay(
        tse_prev, 
        assay.type = "relative_abundance", 
        method = "clr", 
        MARGIN = "cols",
        pseudocount = TRUE
        )

# Get assay
assay <- assay(tse_prev, "clr")
# Transpose assay
assay <- t(assay)
# Add alpha diversity measures
assay <- cbind(assay, colData(tse_prev)[, index, drop = FALSE])

# Preprocessing step 2 - 
# Drop variables with zero variance
ind <- sapply(assay, sd, na.rm = TRUE)
ind <- is.na(ind) | ind == 0
assay <- assay[, !ind, drop = FALSE]

# Convert `assay` into data.frame
assay <- as.data.frame(assay)


# Preprocessing step 3 -
# Group predictors (taxa or alpha diversities) with perfect correlation
m <- cor(assay)
cor_df <- data.frame(row = rownames(m)[as.vector(row(m))],
                     col = colnames(m)[as.vector(col(m))],
                     cor = c(m)) |>
        filter(row > col & abs(cor) == 1)
##
##                      row                    col  cor
## simpson_lambda_dominance gini_simpson_diversity   -1
##       relative_dominance          dbp_dominance    1
##        observed_richness         chao1_richness    1

# Group names of perfectly correlated variables 
cor_groups <- paste(cor_df$row, cor_df$col, sep = "|")
# Remove one of the perfectly correlated variables from the dataset
assay <- within(assay, rm(list = cor_df[["row"]]))
# Change name of remaining variable to the variables of the group
to_rename <- which(colnames(assay) %in% cor_df[["col"]])
colnames(assay)[to_rename] <- cor_groups

# Get labels, i.e., known outcome
labels <- tse_prev[["disease"]]
#  Combine data into single data.frame
df <- assay
df[["diagnosis"]] <- labels
```

The three preprocessing strategies discussed above were applied to this dataset. 
Notice that pearson correlation was used as the abundances are CLR transformed.
In the case of counts, other correlation metrics might be appropriate. 

::: {.callout-note}
## Note: Preprocessing strategies must follow research goals

The code above should be considered as a reference. Preprocessing strategies can
impact the performance of ML models, and thus the steps used should always be
tailored to satisfy specific research goals.
:::

## Algorithms

After the preprocessing is done, different supervised ML models can be used to
classify samples. In this section, two of the most used supervised ML algorithms
are discussed: 

* Random Forest
* XGBoost

Other resources are suggested for an introduction to the mathematical
underpinnings of these -and other- models [@james2013_ISLR].

### Random forest 

Random Forests (RF) is a consensus algorithm that deploys and combines the
outputs of multiple decision trees.

When applied to classification problems, each tree predicts the class of an
observation (e.g., healthy or T2D) based on the values of the features (e.g.,
microbial taxa). Each tree finds the best split of the data by reducing how
mixed the classes are.

Thus, RF algorithms combine the outputs of multiple decision trees through
majority voting. Since each tree is trained on a random subset of the data and
features, RF reduces overfitting and improves generalization to new data.


```{r}
#| label: superML3 - random forest
library(mikropml)

# Run random forest
rf <- run_ml(dataset = df, 
             method = "rf", 
             outcome_colname = "diagnosis", 
             seed = 1,
             kfold = 2, 
             cv_times = 2, 
             training_frac = .8, 
             find_feature_importance = FALSE
             )
```

Below are some metrics about the model's performance in the testing step.

```{r}
#| label: superML3.1 - random forest outputs

# Show model's performance metrics
rf[["performance"]] |> knitr::kable()
```

Two preliminary assessments are good indicate of how good the model performed.
First, the value in the AUC column, which in this case shows a relatively
good performance, specially considering that predicting outcomes based on
complex microbiome data is typically challenging. 

Second, the relationship between `AUC` and `cv_metric_AUC`. Whenever the first 
is lower than the latter, it suggests an overfitting of the model to the 
training data, which indicates a poor ability of the model to generalize to 
new data. Thus, the opposite scenario is usually what researchers are interested
in.

::: {.callout-tip}
## Tip: Always use multiple performance metrics

Although this chapter focuses in the area under the receiver operator 
characteristic curve (AUC), it is often a good idea to look at different 
metrics to accurately assess the model's performance.
:::

### XGBoost

Extreme Gradient Boosting (XGBoost) is a consensus algorithm that sequentially
builds decision trees, each new tree improving the performance in the
classifications made by the previous.

Again, each tree predicts the class of an (e.g., healthy or T2D) based on
feature values (e.g., microbial taxa). However, instead of 'voting' like in
Random Forest, XGBoost improves the performance by assigning higher weights (or
penalties) to misclassified observations.

```{r}
#| label: superML4 - XGBoost

xgb <- run_ml(dataset = df, 
       method = "xgbTree", 
       outcome_colname = "diagnosis", 
       seed = 1997,
       kfold = 2, 
       cv_times = 2, 
       training_frac = .8,
       find_feature_importance = FALSE
       )

# Show metrics of the testing step
xgb[["performance"]] |> knitr::kable()
```

It can be noted that the XGBoost model build using the seed on the example
performs slightly better than the RF model built above. XGBoost is a more 
complex model usually described as one of the best performing models for 
tabular data [@SHWARTZZIV202284]. However, it is difficult to make strong 
conclusions from models trained with low number of kfolds and cross-validations

## Inspect model's performance

It was already shown how to obtain the performance of the model. However,
those results are valid for a particular 80/20 split  of train and test
data(determined by the seed used), and thus it represents the performance of
the model in one of multiple scenarios. 

Often, researchers are interested in generating multiple splits, calculate the
performance of the model of each on them, and then look at the variability of 
the model. This gives a more accurate assessment of the model's performance.

Below the RF model is applied two more times and then the performance of the 
model in the three iterations is explored. Since implementing a model multiple
times can be time and resource consuming, it is only performed two more times.
Ideally, this should be done multiple times. Fortunately, the package provides
[documentation](https://www.schlosslab.org/mikropml/articles/parallel.html) on
the matter.

```{r}
#| label: superML5 - Evaluate the model

# Inspect the variability of model's performance
multiple_rf <- map(.x = 2:3,
                   .f = function(seed){
                           run_ml(dataset = df, 
                                  method = "rf", 
                                  outcome_colname = "diagnosis", 
                                  seed = seed,
                                  kfold = 2, 
                                  cv_times = 2, 
                                  training_frac = .8)
                           }
                   )

# Join the 2 new rf outputs with the first output generated
rf_list <- multiple_rf
rf_list[[3]] <- rf
```

### AUC

While AUC values of 0.5 indicate that the model is not better than a random
classification of observations, values closer to 1 indicate classification
performances closer to perfection (i.e., no false positives or negatives).

```{r}
#| label: superML5.1 - Plot AUCs

# Join model's performance df of each iteration of `run_ml`
rf_performance_df <- map(.x = rf_list,
                         .f = pluck, 
                         "performance") |> 
        list_rbind() |> 
        # Get training and test metrics
        select(seed, 
               method, 
               training = cv_metric_AUC, 
               test = AUC) |> 
        # Prepare data for plotting
        pivot_longer(cols = c(training, test),
                     names_to = "split",
                     values_to = "AUC") |> 
        mutate(split = factor(x = split, 
                              levels = c("training",
                                         "test"))) |>
        mutate(mean_AUC = mean(AUC), .by = split)

# Plot the performance of each iteration
rf_performance_df |> 
        ggplot(aes(x = split, y = AUC, fill = split)) + 
        geom_point(size = 4,
                   shape = 21,
                   show.legend = FALSE,
                   alpha = 0.65) +
        stat_summary(geom = "pointrange",
                     fun.data = mean_se, 
                     color = "black", 
                     show.legend = FALSE) +
        scale_y_continuous(limits = c(0.5, 1)) + 
        theme_classic()

```

Although the results above are based only on three splits, the plot suggests
that the model performs consistently well, with a relatively high AUC and low 
variability across iterations.

### ROC

Different receiver-operator characteristic (ROC) curves can have the same area
under the curve. Thus, plotting ROC curves can give complementary information 
to what AUC provides.

Below there is an example where the three RF models are used to calculate and 
show the mean sensitivity (and its associated variability).

```{r}
#| label: superML5.2 - Plot ROCs

# Calculate mean ROC using all RF models
rf_roc_data <- map(
        .x = rf_list,
        .f = function(model){
                calc_model_sensspec(model[["trained_model"]],
                                    model[["test_data"]],
                                    "diagnosis") |> 
                        mutate(seed = model[["performance"]][["seed"]]) |> 
                        as_tibble(rownames = "sample")
                }
        ) |> 
        list_rbind() |> 
        calc_mean_roc()

# Plot ROC
rf_roc_data |> 
        plot_mean_roc()

```

Single ROC curves -generated from one 80/20 split- have a classic stair-like 
shape. However, the code above calculate the mean of three splits, hence the
title "**Mean** Sensitivity" for the Y axis, and the occasional falls in mean
sensitivity along different values of Specificity. 

## Interpretation of the model

Besides the performance of the model, often times researchers are interested in
understanding what are the features that are affecting the model performance,
and how they do so, a property of ML models called interpretability.

ML models have different degrees of interpretability. One way to understand how
different features (e.g., microbial taxa) affect model's performance is by 
randomly permutating the values of that feature across samples and then 
evaluating the changes in a performance metric.

Below there is code showing how to estimate feature importance after the model
is trained and tested. 

::: {.callout-note}
Estimating feature importance can be slow, and thus the number of permutations
used in the examples below is just a fraction of the recommended.
:::

```{r}
#| label: superML6 - Interpret the model

# Estimate features importance
rf_feat_imp <- get_feature_importance(
        trained_model = rf[["trained_model"]], 
        test_data = rf[["test_data"]],
        outcome_colname = "diagnosis",
        perf_metric_function = multiClassSummary, 
        perf_metric_name = "AUC", 
        class_probs = TRUE, 
        method = "rf", 
        seed = 1, 
        nperms = 4 # 1/25 of default to speed calculations up
        )

# Prepare data for plotting
top_imp_features_df <- rf_feat_imp |> 
        slice_max(order_by = perf_metric_diff, 
                  n = 10) |> 
        mutate(feat = fct_reorder2(feat, 
                                   feat, 
                                   perf_metric_diff, 
                                   .desc = FALSE))

# Plot feature's effect in models performance
top_imp_features_df |> 
        ggplot(aes(x = perf_metric_diff, y = feat)) +
        geom_col() +
        scale_x_continuous(
                limits = c(0, 
                           top_imp_features_df$perf_metric_diff[1] * 1.1),
                expand = c(0, 0)) +
        labs(x = "Change in performance",
             y = "Features") +
        theme_classic()
```

This plot shows the features with the highest effect on model's performance,
or in other words, which features have most of the information for
predicting the outcome. Notice that within the most influential features
there are some grouped variables.

Although these results were generated with just a few permutations and thus
should be taken with caution, they illustrate an important aspect of
interpreting microbiome data: often, these plots are interpreted with a
reductionist mindset, making researchers conclude that a single microbial
feature is responsible of a clinical output of interest. However, notice that
more often than not, microbes interact with one another, and thus the effect
of one simple taxa doesn't reflect ecosystem-level properties of the
microbiome.