# Mediation {#sec-mediation}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

Mediation analysis is used to study the effect of an independent variable (X) on
the outcome variable (Y) through a third variable, known as mediator (M). Mathematically, this relationship is described as follows:

Y ~ X * M

The significance of a mediator is typically quantified in terms of the mediated
effect, that is, the portion of the association between X and Y that can be
explained through M. This corresponds to the difference between the total effect
and the direct effect and is called Average Causal Mediated Effect (ACME):

ACME = Total Effect - Average Direct Effect

Put nice illustration of mediation, use a DAG.

The microbiome has the potential to mediate the effects of multiple environmental
stimuli on human health. For example, the impact of antibiotics on humour largely
depends on microbiome depletion. In this case, we talk about complete mediation.
On the other hand, diet affects the organism directly via nutrition and
metabolism, but in part also delivered through the microbiome. In this case, we
talk about partial mediation.

We demonstrate a basic mediation analysis and visualisation with the hitchip1006
dataset from miaTime package.

```{r}
library(mia)
library(miaViz)

# Load dataset
data(hitchip1006, package = "miaTime")
tse <- hitchip1006
```

```{r}
# Convert BMI variable to numeric
tse$bmi_group <- as.numeric(tse$bmi_group)

# Agglomerate features by family (merely to speed up execution)
tse <- agglomerateByRank(tse, rank = "Phylum")

# Apply clr transformation to counts assay
tse <- transformAssay(tse,
                      method = "clr",
                      pseudocount = 1)
```

## Feature mediation

```{r}
# Analyse mediated effect of nationality on BMI via alpha diversity
# 100 permutations were done to speed up execution, but ~1000 are recommended 
med_df <- getMediation(tse,
                       outcome = "bmi_group",
                       treatment = "nationality",
                       mediator = "diversity",
                       covariates = c("sex", "age"),
                       treat.value = "Scandinavia",
                       control.value = "CentralEurope",
                       boot = TRUE, sims = 100)
 
# Visualise model statistics for 1st mediator
plotMediation(med_df, layout = "forest")
```

what is the meaning of ACME, ADE and total effect?

## Assay mediation

```{r}
# Analyse mediated effect of nationality on BMI via clr-transformed features
# 100 permutations were done to speed up execution, but ~1000 are recommended     
tse <- addMediation(tse, name = "assay_mediation",
                    outcome = "bmi_group",
                    treatment = "nationality",
                    assay.type = "clr",
                    covariates = c("sex", "age"),
                    treat.value = "Scandinavia",
                    control.value = "CentralEurope",
                    boot = TRUE, sims = 100,
                    p.adj.method = "fdr")

# Show results for first 5 mediators
head(metadata(tse)$assay_mediation, 5)
```

```{r}
plotMediation(tse, "assay_mediation", layout = "heatmap")
```


## Mediation of reduced dimensions

```{r}
# Perform ordination
tse <- runPCA(tse, name = "PCA",
              assay.type = "clr",
              ncomponents = 3)

# Analyse mediated effect of nationality on BMI via MDS components 
# 100 permutations were done to speed up execution, but ~1000 are recommended       
tse <- addMediation(tse, name = "reddim_mediation",
                    outcome = "bmi_group",
                    treatment = "nationality",
                    dimred = "PCA",
                    covariates = c("sex", "age"),
                    treat.value = "Scandinavia",
                    control.value = "CentralEurope",
                    boot = TRUE, sims = 100,
                    p.adj.method = "fdr")

# Show results for first 5 mediators
head(metadata(tse)$reddim_mediation, 5)
```

```{r}
p1 <- plotMediation(tse, "reddim_mediation", layout = "forest")
p2 <- plotLoadings(tse, "PCA", ncomponents = 3, n = 8, layout = "heatmap")

library(patchwork)

p1 / p2
```

