# Multimodal Mediation Analysis {#sec-MSEA}

```{r}
#| label: setup
#| echo: false
#| results: asis
remove(list = ls())
invisible(gc())
library(rebook)
chapterPreamble()
```

Building upon the concepts and workflows presented in the previous chapter on 
mediation analysis, here we demonstrate how to perform **multimodal mediation** 
**analysis** to identify potential mediators across multiple omics layers. 
**Multimodal mediation analysis** examines whether and how an exposure (e.g., 
treatment, diet) affects an outcome (e.g., disease, phenotype) **through** 
**intermediate variables (mediators) across multiple omics layers** 
**simultaneously** (e.g., microbiome, metabolomics). It identifies which 
features across modalities act as mediators, quantifies indirect (mediated) 
effects while adjusting for covariates, and compares mediation strength across 
layers, providing mechanistic insights into **how exposures impact outcomes** 
**via molecular pathways**.

```{r}
#| label: fig_multimodal_mediation
#| fig-cap: Directed acyclic graph illustrating multimodal mediation where an exposure affects multiple mediators across modalities (microbial species and metabolites), which in turn affect the outcome. A direct path from exposure to outcome is also included (dashed). While gut microbial species biologically influence metabolite levels, for tractability in high-dimensional mediation analysis we assume species and metabolites act as conditionally independent parallel mediators given exposure, acknowledging this as an approximation
#| echo: false

library(ggdag)
library(ggplot2)

# Define DAG: exposure -> 4 parallel mediators -> outcome, plus a direct path
dag <- dagify(
    outcome ~ sp1 + sp2 + met1 + met2 + exposure,
    sp1 ~ exposure,
    sp2 ~ exposure,
    met1 ~ exposure,
    met2 ~ exposure,
    exposure = "exposure",
    outcome = "outcome",
    labels = c(
        exposure = "Exposure",
        outcome = "Outcome",
        sp1 = "Species 1",
        sp2 = "Species 2",
        met1 = "Metabolite 1",
        met2 = "Metabolite 2"
    ),
    coords = list(
        x = c(exposure = 0, sp1 = 1, sp2 = 1, met1 = 1, met2 = 1, outcome = 2),
        y = c(exposure = 0, sp1 = 1.5, sp2 = 0.5, met1 = -0.5, met2 = -1.5, outcome = 0)
    )
)

# Flag the direct exposure -> outcome edge so it renders dashed
tidy_dag <- tidy_dagitty(dag)
tidy_dag$data$direct <- with(
    tidy_dag$data, name == "exposure" & !is.na(to) & to == "outcome"
)

ggplot(tidy_dag, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_dag_edges(aes(edge_linetype = direct)) +
    geom_dag_point(color = "lightblue") +
    geom_dag_text(aes(label = label), color = "black", size = 3) +
    scale_edge_linetype_manual(values = c("solid", "dashed"), guide = "none") +
    theme_dag()
```

In this chapter, we demonstrate multimodal mediation analysis using microbiome 
data from the iHMP IBDMDB study from the R/Bioconductor package 
curatedMetagenomicData. [@Lloyd-Price2019]. Unlike conventional mediation 
analysis, which typically focuses on a single mediator or a small set of 
mediators, we performed multimodal mediation analysis with R/Bioconductor 
package \[multimedia\] [@Jiang2025] here, which can handle many potential 
mediators across multiple data modalities. For example, species-level taxonomic 
abundances, pathways abundances, and metabolomic profiles can all serve as 
potential mediators. This method allows us to estimate and rank each mediator’s 
contribution while accounting for correlations between mediators in a high- 
dimensional setting.

In this example, time point (baseline vs. post-treatment) is used as the 
treatment, and microbiome-based dysbiosis scores are used as the outcome. We 
will first focus on the relative abundances of gut microbial species as 
mediators, followed by pathways abundances.

Generally, we will proceed through the following key steps:

1.  Prepare the microbiome taxonomic data from the iHMP IBDMDB project.

2.  Calculate dysbiosis scores as the outcome.

3.  Define the mediation analysis data structure.

4.  Fit the multimodal mediation model using the R package multimedia.

5.  Interpret both the overall indirect effects and the mediator-specific 
indirect effects.

6.  Visualize the results using forest plots, histograms, and rankings to 
prioritize findings.

7.  Repeat the process for the iHMP microbial pathways.

## Performing multimodal mediation analysis for iHMP species relative abundance {#sec-relative-abundance}

We begin by loading the relative abundance data and calculating dysbiosis scores 
for each sample based on Bray–Curtis dissimilarity from a healthy reference set. 
This represents deviation from a healthy microbiome, where higher dysbiosis 
scores indicate greater microbial imbalance. We then subset to IBD subjects with 
samples collected at both baseline (visit 1) and post-treatment (visit 21).

We first load the packages and define small plotting helpers used below.

```{r}
#| label: load_libraries_and_helpers
#| message: false

library(curatedMetagenomicData)
library(SummarizedExperiment)
library(dplyr)
library(vegan)
library(tidyverse)
library(multimedia)
library(stringr)
library(ggplot2)
library(mia)
library(miaViz)
```

We pull the iHMP IBDMDB species relative-abundance table directly from
`curatedMetagenomicData` and apply a minimal preprocessing pipeline: convert
to relative abundances, compute per-sample dysbiosis scores (median
Bray–Curtis distance from the healthy reference), subset to IBD subjects
with paired baseline (visit 1) and post-treatment (visit 21) samples,
define the treatment factor, and standardize the assay for mediation.

```{r}
#| label: load_iHMP_relative_demo_data
#| message: false

# Import iHMP species relative abundance data
tse <- curatedMetagenomicData(
    "HMP_2019_ibdmdb.relative_abundance",
    rownames = "short", dryrun = FALSE
)[[1]]

# Convert the assay to relative abundances in [0, 1]
tse <- transformAssay(
    tse,
    assay.type = "relative_abundance", method = "relabundance"
)

# Dysbiosis: median Bray-Curtis distance to healthy references (excluding self)
diss <- as.matrix(getDissimilarity(
    tse,
    method = "bray", assay.type = "relabundance", na.rm = TRUE
))
healthy <- tse$disease == "healthy"
tse$dysbiosis <- sapply(seq_len(ncol(tse)), function(i) {
    median(diss[i, healthy & seq_len(ncol(tse)) != i], na.rm = TRUE)
})

# Keep IBD subjects with paired baseline (visit 1) and post-treatment (visit 21)
tse <- tse[, tse$disease != "healthy" & tse$visit_number %in% c(1, 21)]
paired <- names(which(table(tse$subject_id) == 2))
tse <- tse[, tse$subject_id %in% paired]

# Treatment factor and standardized mediator assay
tse$treatment <- factor(
    ifelse(tse$visit_number == 1, "T0", "T1"),
    levels = c("T0", "T1")
)
tse <- transformAssay(
    tse,
    assay.type = "relabundance", method = "standardize", name = "scaled"
)

# Use syntactically valid mediator names in model formulas
rownames(tse) <- rownames(tse) |>
    str_remove(".*s__") |>
    str_remove_all("\\[|\\]") |>
    str_replace_all("[: \\.,]", "_") |>
    make.names(unique = TRUE)
```

The mediation analysis components are defined as following: 

Treatment: time point (baseline vs. post-baseline)
Outcome: dysbiosis score
Mediators: scaled species abundances

These are bundled into a mediation data object (exper), which is passed into the 
multimedia function to estimate path models for the treatment-to-mediator (α 
path), mediator-to-outcome (β path), and the combined indirect effect (α×β 
path).

We now fit the multimodal mediation model and inspect the overall indirect and 
direct effects to understand whether microbiome composition mediates the 
treatment's effect on dysbiosis.

```{r}
#| label: overall_mediation_analysis
#| message: false

# Use indices for mediators
medi_idx <- seq_len(nrow(tse))

# Create the mediation data object
exper <- mediation_data(
    tse,
    outcomes = "dysbiosis",
    treatments = "treatment",
    mediators = medi_idx
)

# Fit the multimodal mediation model and estimate effects
mdl <- multimedia(exper)
res <- estimate(mdl, exper)

# Summarize overall indirect and direct effects
summary(res)
print(indirect_overall(res, exper))
print(direct_effect(res, exper))
```

Mediator-specific indirect effects are then extracted and ranked by absolute 
effect size.

```{r}
#| label: specific_mediation
#| message: false

effects_by_mediator <- indirect_pathwise(res, exper)

# Rank mediators by absolute indirect effect size
top_mediators <- effects_by_mediator %>%
    arrange(desc(abs(indirect_effect))) %>%
    head(20)

print(top_mediators)
```

To quantify uncertainty around the indirect effects, we use non-parametric 
bootstrap resampling to compute **95% confidence intervals** for both the 
overall and mediator-specific indirect effects.

```{r}
#| label: visualization_for_species_overall_indirect_effects
#| message: false

# Bootstrap the overall indirect effect
set.seed(12345)
boot_overall <- bootstrap(mdl, exper, c(indirect = indirect_overall), B = 10)

# Summarize the bootstrap distribution and percentile interval
summary(boot_overall$indirect)
quantile(boot_overall$indirect$indirect_effect, probs = c(0.025, 0.975))

# Plot the bootstrap distribution
ggplot(boot_overall$indirect, aes(indirect_effect)) +
    geom_histogram(bins = 20, fill = "#69b3a2", color = "black") +
    theme_classic() +
    labs(
        x = "Overall Indirect Effect",
        y = "Frequency",
        title = "Bootstrap Distribution of Overall Indirect Effect (B = 100)"
    )
```

Finally, we visualize the **mediator-specific indirect effects** using a forest 
plot, displaying point estimates and 95% CIs to highlight the most influential 
mediators.

```{r}
#| label: visualization_for_species_specific_indirect_effects
#| message: false

# Define mediator-specific indirect effects for bootstrap resampling
indirect_each <- function(mdl, exper) {
    res <- estimate(mdl, exper)
    alpha <- sapply(res@mediation@estimates, function(m) coef(m)["treatmentT1"])
    names(alpha) <- sub("\\.treatmentT1$", "", names(alpha))
    beta <- coef(res@outcome@estimates$dysbiosis)[names(alpha)]
    indirect <- alpha * beta
    names(indirect) <- names(alpha)
    indirect <- indirect[!is.na(indirect)]
    return(indirect)
}

# Bootstrap mediator-specific indirect effects
set.seed(12345)
boot_each <- bootstrap(mdl, exper, c(indirect = indirect_each), B = 100)
summary(boot_each$indirect)

# Calculate percentile intervals and bootstrap means
lower_upper <- apply(
    boot_each$indirect,
    2,
    function(x) quantile(x, c(0.025, 0.975), na.rm = TRUE)
)
means <- apply(boot_each$indirect, 2, mean, na.rm = TRUE)

# Collect estimates and interval limits in one data frame
summary_df <- data.frame(
    mediator = colnames(boot_each$indirect),
    estimate = means,
    lower = lower_upper[1, ],
    upper = lower_upper[2, ]
)
summary_df <- summary_df[-1, ]

# Flag intervals that do not cross zero
summary_df$significant <- with(summary_df, lower > 0 | upper < 0)

# Remove mediators with degenerate bootstrap intervals
summary_df <- summary_df %>%
    filter((upper - lower) != 0)

# Prioritize significant mediators, then sort by effect size
summary_df$rank <- ifelse(summary_df$significant, 1, 2)
summary_df <- summary_df[order(summary_df$rank, -abs(summary_df$estimate)), ]

# Add absolute effects for forest plot ordering
summary_df <- summary_df %>%
    mutate(abs_estimate = abs(estimate))

# Plot mediator-specific effects with confidence intervals
plotForest(
    summary_df,
    effect.var = "estimate",
    ci.lower.var = "lower",
    ci.upper.var = "upper",
    pval.var = NULL,
    id.var = "mediator",
    order.by = "abs_estimate",
    colour.by = "significant"
) +
    labs(
        x = "Observed Indirect Effect with Bootstrap CI",
        y = "Mediator",
        title = "Forest Plot of Mediation-specific Indirect Effects for Species"
    )
```

Based on the forest plot results, we can see that the Agathobaculum 
butyriciproducens and Escherichia coli pathways has significant indirect 
mediation effects.

We also inspect the top species mediators with the default mediator-level 
visualization.

```{r}
#| label: default_visualization_plot_mediators_species
#| message: false

# Select the top mediators for the default mediator-level plot
top_meds <- summary_df %>%
    arrange(desc(abs(estimate))) %>%
    slice_head(n = 12) %>%
    pull(mediator)

# Create the effect table expected by plot_mediators
ie_pw_fast <- tibble(
    outcome = "dysbiosis",
    mediator = top_meds,
    direct_setting = levels(tse$treatment)[1],
    contrast = paste(levels(tse$treatment)[1], "-", levels(tse$treatment)[2]),
    indirect_effect = summary_df$estimate[match(top_meds, summary_df$mediator)]
)

# Visualize the selected mediators across samples
plot_mediators(ie_pw_fast, exper, n_panels = 12)
```
Mediators are standardized (z-scored) for comparability across features; hence values may be negative. Besides, many microbial features are sparse; several mediators exhibit near-zero values for most samples, producing vertical bands.

Next we repeat the mediation analysis with the iHMP pathway abundances using the 
same setup as before, using the time point as the treatment, dysbiosis score as 
the outcome, but pathways abundances as the mediation variables instead of the 
relative species abundance.

## Performing multimodal mediation analysis for pathways abundances {#sec-pathways-abundances}

We then repeat the same minimal preprocessing on the iHMP pathway abundance
table (paired visits 1 and 27 for pathways).

```{r}
#| label: load_iHMP_pathway_demo_data
#| message: false

# Import iHMP pathway abundance data
tse <- curatedMetagenomicData(
    "HMP_2019_ibdmdb.pathway_abundance",
    rownames = "short", dryrun = FALSE
)[[1]]

# Drop samples with any NAs; keep only top-level pathways (rows without "|"
# strata) and remove the UNMAPPED / UNINTEGRATED rows; scale counts to [0, 1]
tse <- tse[, colSums(is.na(assay(tse))) == 0]
rows_to_keep <- !grepl("\\|", rownames(tse))
rows_to_keep[which(rows_to_keep)[1:2]] <- FALSE
tse <- tse[rows_to_keep, ]
assay(tse) <- assay(tse) / 100

# Dysbiosis: median Bray-Curtis distance to healthy references (excluding self)
diss <- as.matrix(vegdist(t(assay(tse)), method = "bray", na.rm = TRUE))
healthy <- tse$disease == "healthy"
tse$dysbiosis <- sapply(seq_len(ncol(tse)), function(i) {
    median(diss[i, healthy & seq_len(ncol(tse)) != i], na.rm = TRUE)
})

# Keep IBD subjects with paired baseline (visit 1) and post-treatment (visit 27)
tse <- tse[, tse$disease != "healthy" & tse$visit_number %in% c(1, 27)]
paired <- names(which(table(tse$subject_id) == 2))
tse <- tse[, tse$subject_id %in% paired]

# Treatment factor and standardized mediator assay
tse$treatment <- factor(
    ifelse(tse$visit_number == 1, "T0", "T1"),
    levels = c("T0", "T1")
)
tse <- transformAssay(
    tse,
    assay.type = "pathway_abundance", method = "standardize", name = "scaled"
)

# Syntactically valid mediator names for model formulas
rownames(tse) <- rownames(tse) |>
    str_remove_all("\\[|\\]") |>
    str_replace_all("[: \\.,]", "_") |>
    make.names(unique = TRUE)
```

Next, we continue to fit the multimodal mediation model, and extract the overall 
and mediation-specific indirect effects.

```{r}
#| label: mediation_analysis
#| message: false

# Use indices for mediators
medi_idx <- seq_len(nrow(tse))

# Create the mediation data object
exper <- mediation_data(
    tse,
    outcomes = "dysbiosis",
    treatments = "treatment",
    mediators = medi_idx
)

# Fit the multimodal mediation model and estimate effects
mdl <- multimedia(exper)
res <- estimate(mdl, exper)

# Summarize overall indirect and direct effects
summary(res)
print(indirect_overall(res, exper))
print(direct_effect(res, exper))

effects_by_mediator <- indirect_pathwise(res, exper)

# Rank mediators by absolute indirect effect size
top_mediators <- effects_by_mediator %>%
    arrange(desc(abs(indirect_effect))) %>%
    head(20)

print(top_mediators)
```

The non-parametric bootstrap resampling is also used to compute **95% CIs** for 
both the overall and mediator-specific pathway indirect effects.

```{r}
#| label: visualization_for_pathway_overall_indirect_effects
#| message: false

# Bootstrap the overall indirect effect
set.seed(1234)
boot_overall <- bootstrap(mdl, exper, c(indirect = indirect_overall), B = 100)

# Summarize the bootstrap distribution and percentile interval
summary(boot_overall$indirect)
quantile(boot_overall$indirect$indirect_effect, probs = c(0.025, 0.975))

# Plot the bootstrap distribution
ggplot(boot_overall$indirect, aes(indirect_effect)) +
    geom_histogram(bins = 20, fill = "#69b3a2", color = "black") +
    theme_classic() +
    labs(
        x = "Overall Indirect Effect",
        y = "Frequency",
        title = "Bootstrap Distribution of Overall Indirect Effect (B = 100)"
    )
```

We visualize the **mediator-specific indirect effects** using a forest plot for 
pathway abundance as well, with the point estimates and 95% CIs.

```{r}
#| label: visualization_for_pathway_specific_indirect_effects
#| message: false

# Define mediator-specific indirect effects for bootstrap resampling
indirect_each <- function(mdl, exper) {
    res <- estimate(mdl, exper)
    alpha <- sapply(res@mediation@estimates, function(m) coef(m)["treatmentT1"])
    names(alpha) <- sub("\\.treatmentT1$", "", names(alpha))
    beta <- coef(res@outcome@estimates$dysbiosis)[names(alpha)]
    indirect <- alpha * beta
    names(indirect) <- names(alpha)
    indirect <- indirect[!is.na(indirect)]
    return(indirect)
}

# Bootstrap mediator-specific indirect effects
set.seed(1234)
boot_each <- bootstrap(mdl, exper, c(indirect = indirect_each), B = 100)

# Calculate percentile intervals and bootstrap means
lower_upper <- apply(
    boot_each$indirect,
    2,
    function(x) quantile(x, c(0.025, 0.975), na.rm = TRUE)
)
means <- apply(boot_each$indirect, 2, mean, na.rm = TRUE)

# Collect estimates and interval limits in one data frame
# (plotForest reads rownames by default, so no mediator column needed)
summary_df <- data.frame(
    estimate = means,
    lower = lower_upper[1, ],
    upper = lower_upper[2, ],
    row.names = colnames(boot_each$indirect)
)

summary_df <- summary_df[-1, ]

# Flag intervals that do not cross zero
summary_df$significant <- with(summary_df, lower > 0 | upper < 0)

# Remove mediators with degenerate bootstrap intervals
summary_df <- summary_df[summary_df$upper != summary_df$lower, ]

# Restore readable pathway labels for plotting
rownames(summary_df) <- rownames(summary_df) |>
    str_replace_all("PWY0\\.|PWY\\.", "PWY-") |>
    str_replace_all("[_\\.]+", " ") |>
    str_squish()

# Add absolute effects for forest plot ordering
summary_df <- summary_df %>%
    mutate(abs_estimate = abs(estimate))

# Plot mediator-specific effects with confidence intervals
plotForest(
    summary_df,
    effect.var = "estimate",
    ci.lower.var = "lower",
    ci.upper.var = "upper",
    pval.var = NULL,
    order.by = "abs_estimate",
    colour.by = "significant"
) +
    labs(
        x = "Observed Indirect Effect with Bootstrap CI",
        y = "Mediator",
        title = "Forest Plot of Mediation-specific Indirect Effects for Pathway"
    )
```

Based on the forest plot results, we can see that the tRNA charging pathway has 
significant indirect mediation effects on the dysbiosis scores.
