# Ordination-based multiassay analysis {#sec-multiassay_ordination}

```{r}
#| label: setup
#| echo: false
#| results: asis

library(rebook)
chapterPreamble()
```

In this chapter, we continue from where we left off in the previous section
(referenced as [@sec-cross-correlation]). Specifically, we introduce an
analytical method called Multi-Omics Factor Analysis. This method is described
as "ordination-based", meaning it involves techniques that reduce the
dimensionality of the data while preserving as much variability as possible.
Ordination methods are often used in multivariate analysis to visualize
patterns, trends, and relationships in complex data sets, making them
particularly useful for multi-omics data where multiple types of biological
data are integrated and analyzed together.

See [@sec-community-similarity] for more information on ordination methods.

## Multi-Omics Factor Analysis {#sec-mofa}

Multi-Omics Factor Analysis (MOFA) is an unsupervised method for integrating
multi-omic data sets in a downstream analysis [@Argelaguet2018]. It could be
seen as a generalization of principal component analysis. Yet, with the ability
to infer a latent (low-dimensional) representation, shared among the multiple
(-omics) data sets in hand.

We use the `r BiocStyle::Biocpkg("MOFA2")` package for analysis.

```{r}
#| label: mofa_data1

library(mia)

# Import dataset
data("HintikkaXOData", package = "mia")

# Store as mae
mae <- HintikkaXOData
```

The `mae` object could be used straight to create the MOFA model. Yet,
we transform our assays since the model assumes normality per
default, and Gaussian model is recommended
(see [MOFA2 FAQ](https://biofam.github.io/MOFA2/faq.html)). However, Poisson and
Bernoulli distribution models are also available.

```{r}
#| label: mofa2
#| message: false
#| warning: false

library(MOFA2)

# For simplicity, reduce samples to two categories: high vs low fat diet
mae$Diet <- ifelse(
    mae$Diet %in% c("High-fat", "High-fat + XOS"),
    "High-fat", "Low-fat"
)

# Agglomerate microbiome data
mae[[1]] <- agglomerateByPrevalence(mae[[1]], rank = "Genus")

# Apply CLR transformation to microbiome counts assay
mae[[1]] <- transformAssay(mae[[1]], method = "clr", pseudocount = TRUE)

# Standardize microbiome CLR assay
mae[[1]] <- transformAssay(
    mae[[1]],
    assay.type = "clr",
    method = "standardize",
    MARGIN = "rows",
    name = "clr-z"
)

# Apply log10 transformation to metabolome NMR assay
mae[[2]] <- transformAssay(mae[[2]], assay.type = "nmr", method = "log10")

# Standardize metabolome log10 assay
mae[[2]] <- transformAssay(
    mae[[2]],
    assay.type = "log10",
    method = "standardize",
    name = "log10-z"
)

# Standardize biomarker assay
mae[[3]] <- transformAssay(
    mae[[3]],
    assay.type = "signals",
    method = "standardize",
    MARGIN = "rows",
    name = "z"
)
```

Note that agglomeration by taxonomic rank might produce duplicate feature names,
such as "uncultured". If present, they can be made unique as follows:

```{r}
#| label: mofa3
#| results: false

# Check if any feature names are duplicated
mae[[1]] |>
    rownames() |>
    anyDuplicated()

# If duplicated, make feature names unique
rownames(mae[[1]]) <- getTaxonomyLabels(mae[[1]], make.unique = TRUE)
```

Then, we prepare the MOFA2 model and define the data and training parameters
available as additional argument, which will otherwise be set to default values.
The full parameter list can be found with `?mofa2`.

```{r}
#| label: mofa4

# Build MOFA2 model
model <- mofa2(
    mae,
    experiments = c("microbiota", "metabolites", "biomarkers"),
    assays = c("clr-z", "log10-z", "z"),
    groups = "Diet",
    num_factors = 5
)

# View model
model
```

The model is then and trained with `run_mofa`.

```{r}
#| label: mofa5
#| results: false

# Some systems may require the specification `use_basilisk = TRUE`
# so it has been added to the following code
model <- run_mofa(model, use_basilisk = TRUE)
```

The explained variance is visualized with the `plot_variance_explained()`
function.

```{r}
#| label: mofa6
#| messsage: false
#| warning: false
#| fig-height: 8
#| fig-width: 10

library(patchwork)
library(ggplot2)

plot_list <- plot_variance_explained(
    model,
    x = "view", y = "factor",
    plot_total = TRUE
)

wrap_plots(plot_list, nrow = 2) +
    plot_annotation(
        title = "Variance Explained per factor and assay",
        theme = theme(plot.title = element_text(hjust = 0.5))
    )
```

From the plot, we can observe that the microbiota accounts for most of the
variability in the data. Biomarkers do not significantly explain the
variability. The variability in the microbiota is primarily captured in factors
1 and 2, while metabolites are primarily represented in factor 2.

We can then visualize the top weights for microbiota and metabolites for
the first two factors to analyze co-varying features.

```{r}
#| label: mofa7
#| warning: false
#| message: false
#| fig-height: 10
#| fig-width: 10

plot_list <- lapply(
    c("microbiota", "metabolites"),
    plot_top_weights,
    object = model,
    factors = c(1, 2),
    nfeatures = 10
)

wrap_plots(plot_list, ncol = 1) & theme(text = element_text(size = 8))
```

From the visualization, we can see that glycerol, lactate, and choline are
positively associated with factor 2, while acetate is negatively associated
among the metabolites. In terms of microbiota, Ruminococcaceae and
Ruminodostridium, for instance, are positively associated with the same factor.
This indicates that these features co-vary in the data, suggesting a positive
association with glycerol, lactate, and choline, and a negative association
with acetate and these microbes.

More tutorials and examples of using the package are found at
[MOFA2 tutorials](https://biofam.github.io/MOFA2/tutorials.html).

::: {.callout-tip icon=false}
## Exercises

**Goal:** After completing these exercises, you should be able to run and
interpret MOFA results.

**Exercise 1: MOFA**

1. Load any of the example datasets in 
`r BiocStyle::Biocpkg("MultiAssayExperiment")` format mentioned in
[@sec-example-data].

2. Agglomerate microbiota data to prevalent phyla.

3. Apply CLR transformation to microbiota data.

4. Apply relative transformation and log10 transformation to other metabolite or
other omic.

5. Standardize each feature (within each dataset) by centering to mean 0 and
scaling to unit variance so that all variables are on a comparable scale.

6. Create a MOFA2 object from the `MultiAssayExperiment`.

7. Run MOFA.

8. Visualize the explained variance. How much variance is explained in total,
and which omic accounts the most variance?

9. For each omic, visualize the features with the largest weights. Can you
identify features that are co-varying?

10. If you completed exercises in [@sec-cross-correlation], did you get similar
results? Do the co-varying features also exhibited correlation?

Useful functions:

`utils::data()`, `mia::agglomerateByPrevalence`, `mia::transformAssay()`,
`MOFA2::mofa2()`, `MOFA2::plot_variance_explained()`,
`MOFA2::plot_top_weights()`, `mia::getCrossAssociation()`,
`ComplexHeatmap::Heatmap()`

:::
