# Multivariate Mediation Analysis {#sec-MSEA}

```{r}
#| label: setup
#| echo: false
#| results: asis
remove(list = ls())
invisible(gc())
library(rebook)
chapterPreamble()
```

Building upon the concepts and workflows presented in the previous chapter on
mediation analysis, here we demonstrate how to perform **multivariate**
**mediation analysis** to identify potential mediators when many candidate
features are measured simultaneously within a single omic layer.
**Multivariate mediation analysis** examines whether and how an exposure
(e.g., treatment, diet) affects an outcome (e.g., disease, phenotype)
**through a high-dimensional set of intermediate variables (mediators)**,
such as the hundreds of species-level abundances or pathway abundances
in a microbiome study. It identifies which features act as mediators,
quantifies indirect (mediated) effects while adjusting for covariates,
and ranks mediation strength across features, providing mechanistic
insights into **how exposures impact outcomes via microbial features and**
**functional pathways**.

```{r}
#| label: fig_multivariate_mediation
#| fig-cap: Directed acyclic graph illustrating multivariate mediation where an exposure affects multiple mediators across modalities (microbial species and metabolites), which in turn affect the outcome. A direct path from exposure to outcome is also included (dashed). While gut microbial species biologically influence metabolite levels, for tractability in high-dimensional mediation analysis we assume species and metabolites act as conditionally independent parallel mediators given exposure, acknowledging this as an approximation
#| echo: false

library(ggdag)
library(ggraph)
library(ggplot2)

# Define DAG: exposure -> 4 parallel mediators -> outcome, plus a direct path
dag <- dagify(
    outcome ~ sp1 + sp2 + met1 + met2 + exposure,
    sp1 ~ exposure,
    sp2 ~ exposure,
    met1 ~ exposure,
    met2 ~ exposure,
    exposure = "exposure",
    outcome = "outcome",
    labels = c(
        exposure = "Exposure",
        outcome = "Outcome",
        sp1 = "Species 1",
        sp2 = "Species 2",
        met1 = "Metabolite 1",
        met2 = "Metabolite 2"
    ),
    coords = list(
        x = c(exposure = 0, sp1 = 1, sp2 = 1, met1 = 1, met2 = 1, outcome = 2),
        y = c(exposure = 0, sp1 = 1.5, sp2 = 0.5, met1 = -0.5, met2 = -1.5, outcome = 0)
    )
)

# Flag the direct exposure -> outcome edge so it renders dashed
tidy_dag <- tidy_dagitty(dag)
tidy_dag$data$direct <- with(
    tidy_dag$data, name == "exposure" & !is.na(to) & to == "outcome"
)

ggplot(tidy_dag, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_dag_edges(aes(edge_linetype = direct)) +
    geom_dag_point(color = "lightblue") +
    geom_dag_text(aes(label = label), color = "black", size = 3) +
    scale_edge_linetype_manual(values = c("solid", "dashed"), guide = "none") +
    theme_dag()
```

In this chapter, we demonstrate multivariate mediation analysis using microbiome
data from the iHMP IBDMDB study from the R/Bioconductor package
curatedMetagenomicData. [@Lloyd-Price2019]. Unlike conventional mediation
analysis, which typically focuses on a single mediator or a small set of
mediators, we performed multivariate mediation analysis with the R/Bioconductor
package \[multimedia\] [@Jiang2025] here, which can handle many potential
mediators within a single omic layer. For example, species-level taxonomic
abundances or pathway abundances can each serve as a set of parallel mediators.
This method allows us to estimate and rank each mediator's contribution while
accounting for correlations between mediators in a high-dimensional setting.

In this example, time point (baseline vs. post-treatment) is used as the
treatment, and microbiome-based dysbiosis scores are used as the outcome. We
will first focus on the relative abundances of gut microbial species as
mediators, followed by pathways abundances.

Generally, we will proceed through the following key steps:

1.  Prepare the microbiome taxonomic data from the iHMP IBDMDB project.

2.  Calculate dysbiosis scores as the outcome.

3.  Define the mediation analysis data structure.

4.  Fit the multivariate mediation model using the R package multimedia.

5.  Interpret both the overall indirect effects and the mediator-specific
indirect effects.

6.  Visualize the results using forest plots, histograms, and rankings to
prioritize findings.

7.  Repeat the process for the iHMP microbial pathways.

## Performing multivariate mediation analysis for iHMP species relative abundance {#sec-relative-abundance}

We use a pre-processed subset of the iHMP IBDMDB study, shipped with `miaViz`
as example data (`tse_relative`). This subset already contains the
`relabundance` and `scaled` assays, per-sample dysbiosis scores computed from
Bray–Curtis distance to healthy references, only IBD subjects with paired
baseline (visit 1) and post-treatment (visit 21) samples, and a `treatment`
factor with levels `T0` and `T1`. See `?miaViz::multimedia-datasets` for
details, and `inst/scripts/multimedia_relative_data_prep.R` in the miaViz
source for the full preparation code.

We first load the packages used below.

```{r}
#| label: load_libraries_and_helpers
#| message: false

library(mia)
library(miaViz)
library(multimedia)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r}
#| label: load_iHMP_relative_demo_data
#| message: false

data(tse_relative)
tse <- tse_relative
```

The mediation analysis components are defined as following:

Treatment: time point (baseline vs. post-baseline)
Outcome: dysbiosis score
Mediators: scaled species abundances

These are bundled into a mediation data object (exper), which is passed into the
multimedia function to estimate path models for the treatment-to-mediator (α
path), mediator-to-outcome (β path), and the combined indirect effect (α×β
path).

We now fit the multivariate mediation model and inspect the overall indirect and
direct effects to understand whether microbiome composition mediates the
treatment's effect on dysbiosis.

```{r}
#| label: overall_mediation_analysis
#| message: false

# Create the mediation data object
exper <- mediation_data(
    tse,
    outcomes = "dysbiosis",
    treatments = "treatment",
    mediators = rownames(tse)
)

# Fit the multivariate mediation model and estimate effects
mdl <- multimedia(exper)
res <- estimate(mdl, exper)

# Summarize overall indirect and direct effects
summary(res)
print(indirect_overall(res, exper))
print(direct_effect(res, exper))
```

Mediator-specific indirect effects are then extracted and ranked by absolute
effect size.

```{r}
#| label: specific_mediation
#| message: false

effects_by_mediator <- indirect_pathwise(res, exper)

# Rank mediators by absolute indirect effect size
top_mediators <- effects_by_mediator |>
    arrange(desc(abs(indirect_effect))) |>
    head(20)

print(top_mediators)
```

To quantify uncertainty around the indirect effects, we use non-parametric
bootstrap resampling to compute **95% confidence intervals** for both the
overall and mediator-specific indirect effects. To keep the book build time
short, we use only `B = 10` bootstrap iterations below; for real applications
we recommend `B >= 500` (ideally 1000 or more) to obtain stable percentile
confidence intervals, so the CI widths shown here should be interpreted as
illustrative rather than definitive.

```{r}
#| label: visualization_for_species_overall_indirect_effects
#| message: false

set.seed(12345)

# Bootstrap the overall indirect effect
boot_overall <- bootstrap(mdl, exper, c(indirect = indirect_overall), B = 10)

# Summarize the bootstrap distribution and percentile interval
summary(boot_overall$indirect)
quantile(boot_overall$indirect$indirect_effect, probs = c(0.025, 0.975))

# Plot the bootstrap distribution
ggplot(boot_overall$indirect, aes(indirect_effect)) +
    geom_histogram(bins = 20, fill = "#69b3a2", color = "black") +
    ggtitle("Bootstrap Distribution of Overall Indirect Effect (B = 10)") +
    labs(x = "Overall Indirect Effect", y = "Frequency") +
    theme_classic()
```

Finally, we visualize the **mediator-specific indirect effects** using a forest
plot, displaying point estimates and 95% CIs to highlight the most influential
mediators.

```{r}
#| label: visualization_for_species_specific_indirect_effects
#| message: false

# Define mediator-specific indirect effects for bootstrap resampling
indirect_each <- function(mdl, exper) {
    res <- estimate(mdl, exper)
    alpha <- sapply(res@mediation@estimates, function(m) coef(m)["treatmentT1"])
    names(alpha) <- sub("\\.treatmentT1$", "", names(alpha))
    beta <- coef(res@outcome@estimates$dysbiosis)[names(alpha)]
    indirect <- alpha * beta
    names(indirect) <- names(alpha)
    indirect <- indirect[!is.na(indirect)]
    return(indirect)
}

set.seed(12345)

# Bootstrap mediator-specific indirect effects
boot_each <- bootstrap(mdl, exper, c(indirect = indirect_each), B = 10)

# Remove bootstrap index
boot_each <- boot_each$indirect[, -1]

summary(boot_each)

# Calculate percentile intervals and bootstrap means
lower_upper <- apply(
    boot_each,
    MARGIN = 2,
    quantile,
    probs = c(0.025, 0.975),
    na.rm = TRUE
)

# Collect estimates and interval limits in one data frame
summary_df <- data.frame(
    mediator = colnames(boot_each),
    lower = lower_upper[1, ],
    upper = lower_upper[2, ]
)

summary_df$effect <- apply(boot_each, 2, mean, na.rm = TRUE)

summary_df <- summary_df |>
    mutate(
        # Flag intervals that do not cross zero
        significant = lower > 0 | upper < 0,
        # Add absolute effects for forest plot ordering
        abs_effect = abs(effect)
    ) |>
    # Remove mediators with degenerate bootstrap intervals
    filter(upper - lower > 0)

# Plot mediator-specific effects with confidence intervals
plotForest(
    summary_df,
    id.var = "mediator",
    label.by = "CI",
    order.by = "abs_effect"
) +
    ggtitle("Forest Plot of Mediation-specific Indirect Effects for Species") +
    labs(x = "Observed Indirect Effect with Bootstrap CI", y = "Mediator")
```

Based on the forest plot results, we can see that the Agathobaculum
butyriciproducens and Escherichia coli pathways has significant indirect
mediation effects.

We also inspect the top species mediators with the default mediator-level
visualization.

```{r}
#| label: default_visualization_plot_mediators_species
#| message: false

# Select the top mediators for the default mediator-level plot
top_meds <- summary_df |>
    arrange(desc(abs_effect)) |>
    slice_head(n = 12) |>
    pull(mediator)

# Create the effect table expected by plot_mediators
ie_pw_fast <- tibble(
    outcome = "dysbiosis",
    mediator = top_meds,
    direct_setting = levels(tse$treatment)[1],
    contrast = paste(levels(tse$treatment)[1], "-", levels(tse$treatment)[2]),
    indirect_effect = summary_df$effect[match(top_meds, summary_df$mediator)]
)

# Visualize the selected mediators across samples
plot_mediators(ie_pw_fast, exper, n_panels = 12)
```

Mediators are standardized (z-scored) for comparability across features; hence values may be negative. Besides, many microbial features are sparse; several mediators exhibit near-zero values for most samples, producing vertical bands.

Next we repeat the mediation analysis with the iHMP pathway abundances using the
same setup as before, using the time point as the treatment, dysbiosis score as
the outcome, but pathways abundances as the mediation variables instead of the
relative species abundance.
