---
title: "Multimodal Mediation Analysis"
format: html
editor: visual
author: Yihan Liu, Himel Mallick
---

# **Multimodal Mediation Analysis** {#sec-MSEA}

```{r}
#| label: setup
#| echo: false
#| results: asis
remove(list = ls())
invisible(gc())
library(rebook)
chapterPreamble()

```

Building upon the concepts and workflows presented in the previous chapter on 
mediation analysis, here we demonstrate how to perform **multimodal mediation** 
**analysis** to identify potential mediators across multiple omics layers. 
**Multimodal mediation analysis** examines whether and how an exposure (e.g., 
treatment, diet) affects an outcome (e.g., disease, phenotype) **through** 
**intermediate variables (mediators) across multiple omics layers** 
**simultaneously** (e.g., microbiome, metabolomics). It identifies which 
features across modalities act as mediators, quantifies indirect (mediated) 
effects while adjusting for covariates, and compares mediation strength across 
layers, providing mechanistic insights into **how exposures impact outcomes** 
**via molecular pathways**.

```{r}
#| label: fig_multimodal_mediation
#| fig-cap: Directed acyclic graph illustrating multimodal mediation where an exposure affects multiple mediators across modalities (microbial species and metabolites), which in turn affect the outcome. A direct path from exposure to outcome is also included (dashed). While gut microbial species biologically influence metabolite levels, for tractability in high-dimensional mediation analysis we assume species and metabolites act as conditionally independent parallel mediators given exposure, acknowledging this as an approximation
#| echo: false

library(DiagrammeR)

grViz("
digraph multimodal_mediation {
  graph [layout = dot, rankdir = LR]

  node [fontname = Helvetica, fontsize = 12]

  Exposure [shape=box, style=filled, fillcolor=lightblue, color=black]
  Outcome [shape=box, style=filled, fillcolor=palegreen, color=black]
  
  M1 [label='Species 1', shape=ellipse, style=filled, fillcolor=lightyellow, color=black]
  M2 [label='Species 2', shape=ellipse, style=filled, fillcolor=lightyellow, color=black]
  M3 [label='Metabolite 1', shape=ellipse, style=filled, fillcolor=lightpink, color=black]
  M4 [label='Metabolite 2', shape=ellipse, style=filled, fillcolor=lightpink, color=black]
  
  Exposure -> M1
  Exposure -> M2
  Exposure -> M3
  Exposure -> M4
  M1 -> Outcome
  M2 -> Outcome
  M3 -> Outcome
  M4 -> Outcome
  Exposure -> Outcome [style=dashed]
}
")
```

In this chapter, we demonstrate multimodal mediation analysis using microbiome 
data from the iHMP IBDMDB study from the R/Bioconductor package 
curatedMetagenomicData. [@Lloyd-Price2019]. Unlike conventional mediation 
analysis, which typically focuses on a single mediator or a small set of 
mediators, we performed multimodal mediation analysis with R/Bioconductor 
package \[multimedia\] [@Jiang2025] here, which can handle many potential 
mediators across multiple data modalities. For example, species-level taxonomic 
abundances, pathways abundances, and metabolomic profiles can all serve as 
potential mediators. This method allows us to estimate and rank each mediator’s 
contribution while accounting for correlations between mediators in a high- 
dimensional setting.

In this example, time point (baseline vs. post-treatment) is used as the 
treatment, and microbiome-based dysbiosis scores are used as the outcome. We 
will first focus on the relative abundances of gut microbial species as 
mediators, followed by pathways abundances.

Generally, we will proceed through the following key steps:

1.  Prepare the microbiome taxonomic data from the iHMP IBDMDB project.

2.  Calculate dysbiosis scores as the outcome.

3.  Define the mediation analysis data structure.

4.  Fit the multimodal mediation model using the R package multimedia.

5.  Interpret both the overall indirect effects and the mediator-specific 
indirect effects.

6.  Visualize the results using forest plots, histograms, and rankings to 
prioritize findings.

7.  Repeat the process for the iHMP microbial pathways.

### Performing multimodal mediation analysis for iHMP species relative abundance {#sec-relative-abundance}

We begin by loading the relative abundance data and calculating dysbiosis scores 
for each sample based on Bray–Curtis dissimilarity from a healthy reference set. 
This represents deviation from a healthy microbiome, where higher dysbiosis 
scores indicate greater microbial imbalance. We then subset to IBD subjects with 
samples collected at both baseline (visit 1) and post-treatment (visit 21).

We first load the packages and define small plotting helpers used below.

```{r}
#| label: load_libraries_and_helpers
#| message: false

library(curatedMetagenomicData)
library(SummarizedExperiment)
library(dplyr)
library(vegan)
library(tidyverse)
library(multimedia)
library(stringr)
library(ggplot2)
library(mia)
library(miaViz)

plot_indirect_histogram <- function(boot_indirect, title) {
    histogram_data <- SummarizedExperiment(
        assays = list(
            indirect_effect = matrix(
                boot_indirect$indirect_effect,
                nrow = 1,
                dimnames = list("indirect_effect", seq_len(nrow(boot_indirect)))
            )
        )
    )
  
    miaViz::plotHistogram(histogram_data, assay.type = "indirect_effect") +
        labs(
            x = "Overall Indirect Effect",
            y = "Frequency",
            title = title
        )
}

plot_indirect_forest <- function(summary_df, title) {
    forest_data <- summary_df %>%
        mutate(abs_estimate = abs(estimate))
  
    miaViz::plotForest(
        forest_data,
        effect.var = "estimate",
        ci.lower.var = "lower",
        ci.upper.var = "upper",
        pval.var = NULL,
        id.var = "mediator",
        order.by = "abs_estimate",
        colour.by = "significant"
    ) +
        labs(
            x = "Observed Indirect Effect with Bootstrap CI",
            y = "Mediator",
            title = title
        )
}

clean_pathway_names <- function(pathways) {
    pathways <- gsub("PWY0\\.", "PWY0-", pathways)
    pathways <- gsub("PWY\\.", "PWY-", pathways)
    pathways <- gsub("__", ": ", pathways)
    pathways <- gsub("_\\.", " ", pathways)
    pathways <- gsub("\\.", " ", pathways)
    pathways <- gsub("_", " ", pathways)
    pathways <- gsub("\\s+", " ", pathways)
    trimws(pathways)
}

```

The processed demo data are loaded directly from the local data file prepared 
for this chapter.

```{r}
#| label: load_iHMP_relative_demo_data
#| message: false

load("/Users/lyh_ciel/Library/CloudStorage/OneDrive-YaleUniversity/WCM/OMA/multimedia_data_demo.rda")
tse <- tse_relative

```

The mediation analysis components are defined as following: 

-   Treatment: time point (baseline vs. post-baseline)
-   Outcome: dysbiosis score
-   Mediators: scaled species abundances

These are bundled into a mediation data object (exper), which is passed into the 
multimedia function to estimate path models for the treatment-to-mediator (α 
path), mediator-to-outcome (β path), and the combined indirect effect (α×β 
path).

We now fit the multimodal mediation model and inspect the overall indirect and 
direct effects to understand whether microbiome composition mediates the 
treatment's effect on dysbiosis.

```{r}
#| label: overall_mediation_analysis
#| message: false

medi_idx <- seq_len(nrow(se_relative))

exper <- mediation_data(
    se_relative,
    outcomes = "dysbiosis",
    treatments = "treatment",
    mediators = medi_idx
)

mdl <- multimedia(exper)
res <- estimate(mdl, exper)

summary(res)
print(indirect_overall(res, exper))
print(direct_effect(res, exper))

```

Mediator-specific indirect effects are then extracted and ranked by absolute 
effect size.

```{r}
#| label: specific_mediation
#| message: false

effects_by_mediator <- indirect_pathwise(res, exper)

top_mediators <- effects_by_mediator %>%
    arrange(desc(abs(indirect_effect))) %>%
    head(20)

print(top_mediators)

```

To quantify uncertainty around the indirect effects, we use non-parametric 
bootstrap resampling to compute **95% confidence intervals** for both the 
overall and mediator-specific indirect effects.

```{r}
#| label: visualization_for_species_overall_indirect_effects
#| message: false

set.seed(12345)
boot_overall <- bootstrap(mdl, exper, c(indirect = indirect_overall), B = 10)
summary(boot_overall$indirect)
quantile(boot_overall$indirect$indirect_effect, probs = c(0.025, 0.975))

plot_indirect_histogram(
    boot_overall$indirect,
    "Bootstrap Distribution of Overall Indirect Effect (B = 100)"
)

```

Finally, we visualize the **mediator-specific indirect effects** using a forest 
plot, displaying point estimates and 95% CIs to highlight the most influential 
mediators.

```{r}
#| label: visualization_for_species_specific_indirect_effects
#| message: false

indirect_each <- function(mdl, exper) {
    res <- estimate(mdl, exper)
    alpha <- sapply(res@mediation@estimates, function(m) coef(m)["treatmentT1"])
    names(alpha) <- sub("\\.treatmentT1$", "", names(alpha))
    beta <- coef(res@outcome@estimates$dysbiosis)[names(alpha)]
    indirect <- alpha * beta
    names(indirect) <- names(alpha)
    indirect <- indirect[!is.na(indirect)]
    return(indirect)
}

set.seed(12345)
boot_each <- bootstrap(mdl, exper, c(indirect = indirect_each), B = 100)
summary(boot_each$indirect)

lower_upper <- apply(
    boot_each$indirect,
    2,
    function(x) quantile(x, c(0.025, 0.975), na.rm = TRUE)
)
means <- apply(boot_each$indirect, 2, mean, na.rm = TRUE)

summary_df <- data.frame(
    mediator = colnames(boot_each$indirect),
    estimate = means,
    lower = lower_upper[1, ],
    upper = lower_upper[2, ]
)
summary_df <- summary_df[-1, ]

summary_df$significant <- with(summary_df, lower > 0 | upper < 0)
summary_df <- summary_df %>%
    filter((upper - lower) != 0)
summary_df$rank <- ifelse(summary_df$significant, 1, 2)
summary_df <- summary_df[order(summary_df$rank, -abs(summary_df$estimate)), ]

plot_indirect_forest(
    summary_df,
    "Forest Plot of Mediation-specific Indirect Effects for Species"
)

```

Based on the forest plot results, we can see that the Agathobaculum 
butyriciproducens and Escherichia coli pathways has significant indirect 
mediation effects.

We also inspect the top species mediators with the default mediator-level 
visualization.

```{r}
#| label: default_visualization_plot_mediators_species
#| message: false

top_meds <- summary_df %>%
    dplyr::arrange(dplyr::desc(abs(estimate))) %>%
    dplyr::slice_head(n = 12) %>%
    dplyr::pull(mediator)

ie_pw_fast <- tibble::tibble(
    outcome = "dysbiosis",
    mediator = top_meds,
    direct_setting = levels(tse$treatment)[1],
    contrast = paste(levels(tse$treatment)[1], "-", levels(tse$treatment)[2]),
    indirect_effect = summary_df$estimate[match(top_meds, summary_df$mediator)]
)

plot_mediators(ie_pw_fast, exper, n_panels = 12)


```
Mediators are standardized (z-scored) for comparability across features; hence values may be negative. Besides, many microbial features are sparse; several mediators exhibit near-zero values for most samples, producing vertical bands.

Next we repeat the mediation analysis with the iHMP pathway abundances using the 
same setup as before, using the time point as the treatment, dysbiosis score as 
the outcome, but pathways abundances as the mediation variables instead of the 
relative species abundance.

### Performing multimodal mediation analysis for pathways abundances {#sec-pathways-abundances}

Let's try to use the pathway abundance data from the iHMP data to do the 
analysis as well.

```{r}
#| label: load_iHMP_pathway_demo_data
#| message: false

tse <- tse_pathway

```

Next, we continue to fit the multimodal mediation model, and extract the overall 
and mediation-specific indirect effects.

```{r}
#| label: mediation_analysis
#| message: false

medi_idx <- seq_len(nrow(tse))

exper <- mediation_data(
    tse,
    outcomes = "dysbiosis",
    treatments = "treatment",
    mediators = medi_idx
)

mdl <- multimedia(exper)
res <- estimate(mdl, exper)
summary(res)
print(indirect_overall(res, exper))
print(direct_effect(res, exper))

effects_by_mediator <- indirect_pathwise(res, exper)

top_mediators <- effects_by_mediator %>%
    arrange(desc(abs(indirect_effect))) %>%
    head(20)

print(top_mediators)

```

The non-parametric bootstrap resampling is also used to compute **95% CIs** for 
both the overall and mediator-specific pathway indirect effects.

```{r}
#| label: visualization_for_pathway_overall_indirect_effects
#| message: false

set.seed(1234)
boot_overall <- bootstrap(mdl, exper, c(indirect = indirect_overall), B = 100)
summary(boot_overall$indirect)
quantile(boot_overall$indirect$indirect_effect, probs = c(0.025, 0.975))

plot_indirect_histogram(
    boot_overall$indirect,
    "Bootstrap Distribution of Overall Indirect Effect (B = 100)"
)

```

We visualize the **mediator-specific indirect effects** using a forest plot for 
pathway abundance as well, with the point estimates and 95% CIs.

```{r}
#| label: visualization_for_pathway_specific_indirect_effects
#| message: false

indirect_each <- function(mdl, exper) {
    res <- estimate(mdl, exper)
    alpha <- sapply(res@mediation@estimates, function(m) coef(m)["treatmentT1"])
    names(alpha) <- sub("\\.treatmentT1$", "", names(alpha))
    beta <- coef(res@outcome@estimates$dysbiosis)[names(alpha)]
    indirect <- alpha * beta
    names(indirect) <- names(alpha)
    indirect <- indirect[!is.na(indirect)]
    return(indirect)
}

set.seed(1234)
boot_each <- bootstrap(mdl, exper, c(indirect = indirect_each), B = 100)

lower_upper <- apply(
    boot_each$indirect,
    2,
    function(x) quantile(x, c(0.025, 0.975), na.rm = TRUE)
)
means <- apply(boot_each$indirect, 2, mean, na.rm = TRUE)
summary_df <- data.frame(
    mediator = colnames(boot_each$indirect),
    estimate = means,
    lower = lower_upper[1, ],
    upper = lower_upper[2, ]
)

summary_df <- summary_df[-1, ]
summary_df$significant <- with(summary_df, lower > 0 | upper < 0)
summary_df <- summary_df[summary_df$upper != summary_df$lower, ]

pwy <- clean_pathway_names(rownames(summary_df))
rownames(summary_df) <- pwy
summary_df$mediator <- pwy

plot_indirect_forest(
    summary_df,
    "Forest Plot of Mediation-specific Indirect Effects for Pathway"
)

```

Based on the forest plot results, we can see that the tRNA charging pathway has 
significant indirect mediation effects on the dysbiosis scores.
