# Network comparison {#sec-network-comparison}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

```{r, echo=FALSE}
# The networks shown in this chapter take several minutes to generate.
# Therefore, the network objects have been stored in a folder and are loaded
# here to reduce the time needed to knit the book.
networks_diet_file  <- system.file(
    "extdata", "networks_diet.RData", package = "OMA")
load(networks_diet_file)
```

This chapter assumes that you are already familiar with how to construct and
analyze a single microbial network, which is explained in
[@sec-network-learning].

Since microbial interactions are likely to change between conditions, such as
between patients and healthy individuals or between different environmental
states, identifying network differences between groups is often an integral
secondary analysis step. Differences can be detected visually by plotting the
networks side by side, or quantitatively using differential network analysis
tools.

Two approaches for comparing networks between two conditions are considered
in this chapter:

- **Differential network analysis**, which analyzes differences in network
metrics and network structure.
- **Differential association analysis**, which focuses on differences in the
strength of individual associations.

See [@sec-netcomp-methods] for further details on these two approaches
and the methods used in this chapter.

Here we use [NetCoMi](https://github.com/stefpeschel/NetCoMi)
[@peschel2021netcomi] for network comparison, which includes several
**differential network analysis approaches** as well as functionality for
*differential association analysis**, i.e., generating a differential network.
How to install NetCoMi from GitHub is explained in [@sec-network-learning].

The **PeerJ data set** [@potbhare2022skin] containing skin microbial profiles
for 58 subjects is again used in this chapter. The dataset also includes
information on the subjects' geographic location, gender, age and diet. Whether
the **skin microbiome differs** between people with **different diets** is an
interesting question that will be explored in this chapter.

## Data preparation

We perform the same data preprocessing steps as in [@sec-network-learning].

```{r load_packages, message=FALSE, warning=FALSE}
library(NetCoMi)
library(mia)
```

```{r load_data}
data("peerj13075", package = "mia")
tse <- peerj13075
```

```{r preprocessing}
# Agglomerate to genus level
tse <- agglomerateByRank(tse, rank = "genus")

# Add relative abundances
tse <- transformAssay(
    tse,
    assay.type = "counts",
    method = "relabundance",
    MARGIN = "cols")

# Filter by prevalence
tse <- subsetByPrevalent(
    tse,
    prevalence = 0.2,
    detection = 0,
    assay.type = "relabundance")

# Add log10-transformed abundances
tse <- transformAssay(tse, method = "log10", pseudocount = 1)

# Add clr-transformed abundances
tse <- transformAssay(tse, method = "clr", pseudocount = 1)
```

Based on "Diet", the `tse` object is then split into two groups: One
with mixed diet subjects, and one with vegetarian subjects. Both subsets
have nearly the same sample size and are therefore comparable.

```{r split_data}
table(tse$Diet)
tse_list <- splitOn(tse, group = "Diet", use.names = TRUE, by = "cols")
```

## Network learning and analysis

The approach starts again with network construction and analysis, but this
time we pass the two data sets to `netConstruct()` to perform a network
comparison.

The `rep.num` argument is set to 10 to perform only 10 repetitions in the
model selection approach. This speeds up the permutation tests performed
later, and has a negligible effect for this data set.

```{r netConstruct_diet, eval=FALSE}
spring_net_diet <- netConstruct(
    data = tse_list$Mixed,
    data2 = tse_list$Veg,
    taxRank = "genus",
    filtTax = "highestFreq",
    filtTaxPar = list(highestFreq  = 100),
    measure = "spring",
    measurePar = list(
        nlambda = 20,
        rep.num = 10,
        thresh = 0.05,
        Rmethod = "approx"),
    sparsMethod = "none",
    dissFunc = "signed",
    verbose = 3,
    seed = 13075)
```

All network measures are now computed for both networks. Also, both GCMs are
plotted together with a third matrix containing the differences between the
GCMs and significance codes that express if the differences are significantly
different from zero.

```{r netAnalyze_diet, fig.width=10, fig.height=10}
spring_netprops_diet <- netAnalyze(
    spring_net_diet,
    clustMethod = "cluster_fast_greedy",
    hubPar = "eigenvector",
    normDeg = FALSE)
```

In both of the networks, some graphlet correlations are significantly different
from zero. However, none of the correlations are significantly different
between the groups.

```{r summary_diet}
summary(spring_netprops_diet, groupNames = c("Mixed diet", "Vegetarian"))
```

For each centrality measure, the five nodes with the highest centrality
in each group are plotted by default.

We notice some differences in the network properties. The differential
network analysis performed in the next section will show if the differences
are significant.

## Differential network analysis

### Visual comparison

We start with a visual comparison of the two networks using NetCoMi's plot
function. The same configuration as in [@sec-network-learning] is used.

```{r network_plot_diet_difflay, fig.width=17, fig.height=8}
plot(spring_netprops_diet,
     repulsion = 0.97,
     rmSingles = TRUE,
     labelScale = FALSE,
     nodeSize = "eigenvector",
     nodeSizeSpread = 2,
     nodeColor = "cluster",
     sameColThresh = 2,
     hubBorderCol = "darkgray",
     cexNodes = 2,
     edgeTranspHigh = 20,
     title1 = "Mixed diet",
     title2 = "Vegetarian",
     showTitle = TRUE,
     cexTitle = 2,
     mar = c(1, 4, 4, 4))

# Overlay a transparent plot on which the legend is plotted
par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), mar=c(0, 0, 0, 0), new=TRUE)
plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')

legend(-0.2, -0.9, cex = 1.5, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"),
       bty = "n", horiz = TRUE)
```

The layout is computed separately for each network, making it difficult to
visually compare certain associations. It is therefore recommended to use the
same layout for both groups (argument `sameLayout`). Instead of simply copying
one layout to the other network, we set `layoutGroup` to "union". This ensures
that the nodes are placed as optimally as possible for both networks.

```{r network_plot_diet_samelay, fig.width=17, fig.height=8}
plot(spring_netprops_diet,
     sameLayout = TRUE,
     repulsion = 0.95,
     rmSingles = "inboth",
     labelScale = FALSE,
     nodeSize = "eigenvector",
     nodeSizeSpread = 2,
     nodeColor = "cluster",
     sameColThresh = 2,
     hubBorderCol = "darkgray",
     cexNodes = 2,
     edgeTranspHigh = 20,
     title1 = "Mixed diet",
     title2 = "Vegetarian",
     showTitle = TRUE,
     cexTitle = 2,
     mar = c(1, 4, 4, 4))

# Add legend
par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), mar=c(0, 0, 0, 0), new=TRUE)
plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
legend(-0.2, -0.8, cex = 1.7, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"),
       bty = "n", horiz = TRUE)
```

**A few notes:**

-   Differences in the edge weights can now be seen at first glance.
For example, Serratia and Citrobacter are strongly associated in the mixed diet
group, but not at all in the vegetarian group.
-   Clusters must share at least two nodes (`sameColThresh` argument) to be
colored equally in both networks, which is why the color of some clusters
differs between the groups.
-   The clustering generally differs markedly. In particular, the cluster
assignment of many of the nodes in the largest connected component differs
between the two groups.

As in [@sec-network-learning], we also generate a network plot using phylum
names to color the nodes and mclr-transformed abundances to scale node sizes.

```{r, message=FALSE, warning=FALSE}
library(RColorBrewer)
```

```{r network_plot_diet_phyla, fig.width=17, fig.height=8}
# Generate vector with phylum names for node coloring
phyla <- as.factor(rowData(tse)$phylum)
names(phyla) <- rowData(tse)$genus

# Create color vector
colvec <- RColorBrewer::brewer.pal(length(levels(phyla)), "Set3")

p_diet <- plot(
    spring_netprops_diet,
    sameLayout = TRUE,
    repulsion = 0.95,
    rmSingles = "inboth",
    labelScale = FALSE,
    nodeSize = "clr",
    nodeColor = "feature",
    featVecCol = phyla,
    colorVec =  colvec,
    nodeTransp = 20,
    sameColThresh = 2,
    highlightHubs = FALSE,
    cexNodes = 2,
    edgeTranspHigh = 20,
   title1 = "Mixed diet",
   title2 = "Vegetarian",
   showTitle = TRUE,
   cexTitle = 2,
   mar = c(1, 4, 4, 4))

# Add legends
# Colors used in the legend should be equally transparent as in the plot
col_transp <- colToTransp(colvec, 20)

par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), mar=c(0, 0, 0, 0), new=TRUE)
plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
legend(-0.15, -0.8, cex = 1.7, title = "estimated correlation:",
       legend = c("+","-"), lty = 1, lwd = 3, col = c("#009900","red"),
       bty = "n", horiz = TRUE)
legend(-0.15, 1.3, cex = 1.7, pt.cex = 2.5, title = "Phylum:",
       legend=levels(phyla), col = col_transp, bty = "n", pch = 16)
```

### Quantitative comparison

`netCompare()` enables a quantitative network comparison using comparative
measures such as Jaccard's Index, Adjusted Rand Index, and permutation tests.

To test for statistical significance of differences in network properties, we
perform permutation tests with 1000 permutations. Multiple CPU cores are used
to save run time. The association matrices estimated for all permutations are
stored in an external file. We will reuse them later when performing
differential association analysis. They could also be used to rerun
`netCompare()` with different parameter settings.

Note that unless running on a cluster with considerably more CPU cores, a
network comparison with permutation tests may take several hours. You should
test the code below with a small number of permutations to make sure it works
before applying it to your data.

```{r netcomp_diet_permute, eval=FALSE}
spring_netcomp_diet <- netCompare(
    spring_netprops_diet,
    permTest = TRUE,
    nPerm = 1000,
    cores = 6,
    seed = 13075,
    storeAssoPerm = TRUE,
    fileStoreAssoPerm = "general/network_data/spring_assoPerm",
    verbose = TRUE)
```

```{r summary_netcomp_diet, eval=FALSE}
summary(spring_netcomp_diet,
        groupNames = c("Mixed diet", "Vegetarian"),
        numbNodes = 5)
## 
## Comparison of Network Properties
## ----------------------------------
## CALL: 
## netCompare(x = spring_netprops_diet, permTest = TRUE, verbose = TRUE, 
##     nPerm = 1000, cores = 19, seed = 13075, storeAssoPerm = TRUE, 
##     fileStoreAssoPerm = "general/network_data/spring_assoPerm")
## 
## ______________________________
## Global network properties
## `````````````````````````
## Largest connected component (LCC):
##                          Mixed diet   Vegetarian    abs.diff.     p-value    
## Relative LCC size             0.220        0.256        0.037    0.838162    
## Clustering coefficient        0.104        0.224        0.120    0.450549    
## Modularity                    0.515        0.477        0.038    0.801199    
## Positive edge percentage     80.000      100.000       20.000    0.000999 ***
## Edge density                  0.131        0.124        0.007    0.910090    
## Natural connectivity          0.073        0.063        0.011    0.812188    
## Vertex connectivity           1.000        1.000        0.000    1.000000    
## Edge connectivity             1.000        1.000        0.000    1.000000    
## Average dissimilarity*        0.960        0.962        0.001    0.945055    
## Average path length**         2.228        2.246        0.019    0.984016    
## 
## Whole network:
##                          Mixed diet   Vegetarian    abs.diff.     p-value    
## Number of components         50.000       52.000        2.000     0.93706    
## Clustering coefficient        0.075        0.165        0.090     0.46254    
## Modularity                    0.752        0.668        0.084     0.24975    
## Positive edge percentage     88.571      100.000       11.429     0.02398 *  
## Edge density                  0.011        0.011        0.000     0.97303    
## Natural connectivity          0.013        0.013        0.000     0.89011    
## -----
## p-values: one-tailed test with null hypothesis diff=0
##  *: Dissimilarity = 1 - edge weight
## **: Path length = Units with average dissimilarity
## 
## ______________________________
## Jaccard index (similarity betw. sets of most central nodes)
## ```````````````````````````````````````````````````````````
##                     Jacc   P(<=Jacc)     P(>=Jacc)    
## degree             0.560      0.9944       0.01637 *  
## betweenness centr. 0.294      0.4777       0.71860    
## closeness centr.   0.560      0.9944       0.01637 *  
## eigenvec. centr.   0.560      0.9944       0.01637 *  
## hub taxa           0.111      0.1431       0.97399    
## -----
## Jaccard index in [0,1] (1 indicates perfect agreement)
## 
## ______________________________
## Adjusted Rand index (similarity betw. clusterings)
## ``````````````````````````````````````````````````
##         wholeNet       LCC
## ARI        0.367     0.035
## p-value    0.000     0.563
## -----
## ARI in [-1,1] with ARI=1: perfect agreement betw. clusterings
##                    ARI=0: expected for two random clusterings
## p-value: permutation test (n=1000) with null hypothesis ARI=0
## 
## ______________________________
## Graphlet Correlation Distance
## `````````````````````````````
##         wholeNet         LCC  
## GCD       1.5980      2.1770  
## p-value   0.4226      0.7143  
## -----
## GCD >= 0 (GCD=0 indicates perfect agreement between GCMs)
## p-value: permutation test with null hypothesis GCD=0
## 
## ______________________________
## Centrality measures
## - In decreasing order
## - Centrality of disconnected components is zero
## ````````````````````````````````````````````````
## Degree (unnormalized):
##                      Mixed diet Vegetarian abs.diff. adj.p-value  
## Siccibacter                   0          5         5           1  
## Aeromonas                     1          5         4           1  
## Citrobacter                   4          0         4           1  
## Escherichia/Shigella          2          5         3           1  
## Streptococcus                 3          0         3           1  
## 
## Betweenness centrality (normalized):
##                      Mixed diet Vegetarian abs.diff. adj.p-value  
## Escherichia/Shigella      0.118      0.532     0.414           1  
## Aeromonas                 0.000      0.368     0.368           1  
## Serratia                  0.419      0.100     0.319           1  
## Siccibacter               0.000      0.300     0.300           1  
## Streptococcus             0.279      0.000     0.279           1  
## 
## Closeness centrality (normalized):
##               Mixed diet Vegetarian abs.diff. adj.p-value  
## Siccibacter        0.000      0.727     0.727      0.2676  
## Citrobacter        0.713      0.000     0.713      0.9799  
## Streptococcus      0.615      0.000     0.615      0.2676  
## Escherichia        0.587      0.000     0.587      0.9799  
## Xenorhabdus        0.000      0.527     0.527      0.2676  
## 
## Eigenvector centrality (normalized):
##                      Mixed diet Vegetarian abs.diff. adj.p-value  
## Siccibacter               0.000      0.888     0.888      0.6553  
## Citrobacter               0.810      0.000     0.810      1.0000  
## Escherichia               0.610      0.000     0.610      1.0000  
## Serratia                  1.000      0.416     0.584      1.0000  
## Escherichia/Shigella      0.387      0.969     0.582      1.0000  
## 
## _________________________________________________________
## Significance codes: ***: 0.001, **: 0.01, *: 0.05, .: 0.1
```

**Interpreting some results:**

- Of the global network measures, only the positive edge percentage is 
significantly different between the groups.
- Jaccard index: Only for degree, closeness, and eigenvector centrality
the probability P(\>=Jacc) is significant, meaning that the sets of the most
central nodes are quite similar for these three measures. The Jaccard index for
the hub nodes, on the other hand, is low because the two networks share only
one hub node ("Erwinia").
- As indicated by some similarities in the clusterings, the adjusted Rand
index (ARI) of the whole network is significantly different from zero and thus
from random clustering. The ARI of the largest connected component (LCC),
however, is close to zero due to the different clusterings in the LCC.
- The two GCD values are not significantly different from zero. Both GCD values 
are different from zero, which indicates some differences in the overall network 
structures. However, these differences are not significant.
- All nodes are also tested for significantly different centrality values (only 
the five nodes with the highest absolute difference are shown in the summary). 
None of the differences are significant, as none of the p-values are smaller 
than 0.1, which could be due to the small number of permutations.

## Differential association analysis

The `diffnet()` function provides statistical tests to assess whether the
associations themselves are significantly different between the two groups.
`NetCoMi` also provides a plot function to generate a differential network,
where two nodes are connected if they are differentially associated between
the groups.

Since we have already computed the permutation association matrices before,
we can reuse them here (argument `fileLoadAssoPerm`).

The local false discovery rate is controlled at level 0.2 to account for
multiplicity.

```{r diffnet_diet, eval=FALSE, message=FALSE, results='hide'}
spring_diffnet <- diffnet(
    spring_net_diet,
    diffMethod = "perm",
    fileLoadAssoPerm = "general/network_data/spring_assoPerm",
    adjust = "lfdr")
```

```{r hist_pvals, out.width='80%'}
sum(spring_diffnet$pAdjustVec < 0.05)
sum(spring_diffnet$pvalsVec < 0.05)
```

Some of the unadjusted p-values are below the usual 5% significance level.
However, none of the differences remain significant after adjusting for
multiple testing so that the differential network would be empty.

To demonstrate the interpretation of a differential network, we set `adjust`
to "none", which is actually statistically incorrect.

```{r diffnet_diet_unadjusted, message=FALSE, results='hide'}
spring_diffnet_unadj <- diffnet(
    spring_net_diet,
    pvalsVec = spring_diffnet$pvalsVec,
    diffMethod = "perm",
    alpha = 0.05,
    adjust = "none")
```

The `diffnet` object is now plotted using NetCoMi's `plot()` function. 

```{r diffnet_plot, fig.width=20, fig.height=14}
plot(spring_diffnet_unadj,
     cexLabels = 1,
     cexNodes = 1.3,
     cexLegend = 2.5,
     cexTitle = 3,
     mar = c(3,2,5,15),
     legendGroupnames = c("Mixed diet", "Vegetarian"),
     legendPos = c(1.2,1.5),
     legendArgs = list(lwd = 4),
     fade = FALSE)
```

Edge colors represent the direction of the associations in the two groups.
For example, if two OTUs are positively correlated in the mixed diet group and
uncorrelated in the vegetarian group (such as Serratia and Citrobacter), the
edge color is dark green.

```{r, eval=FALSE, echo=FALSE}
save(spring_diffnet, spring_net_diet, spring_netcomp_diet,
     file = "general/network_data/networks_diet.RData")
```


## Network comparison methods {#sec-netcomp-methods}

While many approaches exist for the detection of **differential correlations**,
e.g. [@yu2019new; @mckenzie2016dgca; @siska2017differential], the literature on
the more general case of **differential association** detection is scarce.
@bhuva2019differential compare various methods in a simulation study, which
again includes many differential correlation approaches, but also more
general methods such as latent differential graphical models.
@gill2010statistical introduce an approach to analyze whether the connectivity
of individual nodes is different between two groups using permutation tests,
which is applicable to any kind of association. @he2019statistical propose a
test to infer the differential network structure for two conditional
dependence networks.

Performing **differential network analysis** is challenging because network
measures do not follow classical statistical distributions.
@shojaie2021differential provide an overview of differential network
analysis methods, but focus only on changes in edge sets.
@lichtblau2017comparative compare differential network analysis methods
that incorporate multiple local and global network measures.
@jardim2019bionetstat present a tool "BioNetStat" for differential analysis
of biological networks, which is able to compare certain network measures
between groups.

The [`NetCoMi`](https://github.com/stefpeschel/NetCoMi) package used for
network comparison in this chapter includes the following
**differential network analysis approaches:**:

- **Permutation approach** to test global network measures (e.g.,
transitivity, connectivity, or average path length) as well as centrality
measures for group differences.
- **Jaccard** index to assess the similarity between sets of most central nodes
- **Adjusted Rand index** to assess the similarity between clusterings
- **Graphlet Correlation Distance** (GCM)

See [@peschel2021netcomi] for an explanation of the first three approaches.
The GCM was proposed by @yaverouglu2014revealing.

Two methods (Fisher's z-test [@fisher1970statistical] and the Discordant
method [@siska2016discordant]) are available for identifying differential
correlations, and **permutation tests** for the more general case of
identifying **differential associations**. See [@peschel2021netcomi] for
details. NetCoMi offers also a function for plotting a differential network.
