
# Extra material {#sec-extras}

```{r}
#| label: setup

knitr::opts_chunk$set(eval=TRUE, warning=FALSE, message=FALSE)
```

## `phyloseq` vs TreeSE cheatsheet

This section has a cheatsheet for translating common functions in phyloseq
to TreeSE/mia with example code.

```{r}
#| label: download_libs

# Download libraries
library(mia)
library(phyloseq)
library(dplyr)
library(ggplot2)
```

Start by loading data as a `phyloseq` object "phy" and as `TreeSE` object "tse".

```{r}
#| label: load_data

# Loading example data
# Using GlobalPatterns dataset

data(package = "phyloseq", "GlobalPatterns") # phyloseq object
phy <- GlobalPatterns # Rename
phy # Check the phyloseq object

data(package = "mia", "GlobalPatterns") # TreeSE object
tse <- GlobalPatterns # Rename
tse # Check the tse object
```

### Accessing different types of data in `phyloseq` versus `TreeSE`

Often microbiome datasets contain three different types of tables, one which
defines the microbes' taxonomy from domain to species level, one that
describes sample level information like whether the sample is from a
healthy or a diseased person, and one that has the abundances of taxa from
mapping, like an OTU table.

There are slightly different names for these tables in phyloseq and tse,
but they can be retrieved from the phyloseq and tse containers in analogous
ways.

**Accessing the table of taxonomic names: tax_table = rowData**

phyloseq and TreeSE objects' taxonomy tables can be accessed with tax_table and
rowData commands.

```{r}
#| label: access_data

phy_taxtable <-
    tax_table(phy)  |> # Access the phyloseq taxonomic name table
    data.frame() # Make into a data frame

tse_taxtable <- rowData(tse) |> # Same for tse
    data.frame()

```

**Accessing sample data: sample_data = colData**

Sample data can be accessed with `sample_data` and `colData` commands.

```{r}
#| label: access_sampledata

phy_sampledata <-
    sample_data(phy) |> data.frame()

tse_sampledata <-
    colData(tse) |> data.frame()
```

**Accessing operational taxonomic unit (OTU) abundance objects: otu_table = assay**

OTU tables can be accessed with otu_table and assay commands. The assay can
also hold other types of information like taxa abundances from shotgun
metagenomic annotation, or functional gene abundances.

```{r}
#| label: access_otutable

phy_otutable <-
    otu_table(phy) |> data.frame()

tse_otutable <-
    assay(tse) |> data.frame()
```

### Building phyloseq objects vs TreeSE objects: phyloseq = TreeSummarizedExperiment

After learning how to access various data types from TreeSE, let's see how
creating TreeSE objects compares to creating phyloseq objects. We will use
the vanilla dataframes we created from the phyloseq object to demonstrate
making both types of data objects. These are identical to the equivalent tse
dataframes but for demonstration we will use ones created from phy.

Let's start by checking what we have.

```{r}
#| label: inspect_data

phy_otutable |> head()
phy_sampledata |> head()
phy_taxtable |> head()
```

Ok, these are all normal data frames which could come from upstream
bioinformatics, like OTU tables that come from 16S analysis, and taxonomy
tables.

Let's demo how to create the TreeSE object, how it compares to creating
phyloseq and how assay in TreeSE compares to otu_table in phyloseq.

```{r}
#| label: build_dataobjects


# Create phyloseq object
OTU_phy <- otu_table(as.matrix(phy_otutable), taxa_are_rows = TRUE) # Make OTU table
TAX_phy <- tax_table(as.matrix(phy_taxtable)) # Make TAX table
SAMPLE_phy <- sample_data(phy_sampledata) # Make sample data table

phy <- phyloseq(OTU_phy, TAX_phy, SAMPLE_phy) # Combine into phyloseq object
phy # Inspect

```

Let's start by checking our otu table, and see if it is counts or already
normalized. We will use the same data frame extracted from the phy object
as before.

```{r}
#| label: check_otutable

# Check if we have counts or normalized data

phy_otutable |> head()
```

We have counts!

Since TreeSEs can hold many different versions of the OTU table, most commonly
either relative abundances or counts, we will need to give our assay
(which corresponds to otu_table in Phyloseq) a name and
list the different types of assays or transformations we have. In this
example we only have one item 'counts' in the list.

Let's convert the data frame to a matrix and make the list of assays.

```{r}
#| label: create_tse

# Create TreeSE
counts <- as.matrix(phy_otutable) # Convert to a matrix
assays <- SimpleList(counts = counts)
tse <- TreeSummarizedExperiment(
    assays = assays,
    colData = phy_sampledata,
    rowData = phy_taxtable
)
```

Let's check the different assay names we have.

```{r}
#| label: check_assays
 
assayNames(tse)
```


### Handling different OTU table normalizations in phyloseq vs TreeSE

Adding the assays as a list might seem inconvenient if you only have one type of
OTU table (`counts` in our example), but let's see why it is actually very
convenient to be able to hold multiple assays in one data object. 


Here we'll show an example of how to add relative abundances and CLR normalized
OTU tables to your tse assays.

With phyloseq you would need three different phyloseq objects, each taking up
7.7 MB of memory, whilst the tse with the three assays takes up only 18.3 MB.

```{r}
#| label: transform_assay

# Add another assay that holds the relative abundance normalized OTU table
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")
assays(tse) # Let's check

# With phyloseq you would need to have two different phyloseq objects
phy_relab  = transform_sample_counts(phy, function(x)
  x / sum(x))

# Let's add clr transformed data just for the fun of it :)
tse <- transformAssay(
    tse,
    assay.type = "counts",
    method = "clr",
    pseudocount = 1)

assays(tse) # Let's check

# With phyloseq you would need to have a third phyloseq object.
# phy_CLR <- microbiome::transform(phy, 'clr') # Example, don't run
```