# Taxonomic information {#sec-taxonomic-information}

```{r}
#| label: setup
#| echo: false
#| results: asis

library(rebook)
chapterPreamble()
```

```{r}
#| label: load_pkg_data
#| message: false

library(mia)
data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns
```

Taxonomic information is a key part of analyzing microbiome data, and
without it, any type of data analysis probably would not make much
sense. However, the degree of detail of taxonomic information differs
depending on the dataset and annotation data used.

Therefore, the `r BiocStyle::Biocpkg("mia")` package expects a loose assembly
of taxonomic information and assumes certain key aspects:

- Taxonomic information is given as character vectors or factors in
the `rowData` of a `r BiocStyle::Biocpkg("SummarizedExperiment")` object.
- The columns containing the taxonomic information must be named
`domain`, `kingdom`, `phylum`, `class`, `order`, `family`, `genus`,
`species` or with a capital first letter.
- The columns must be given in the order shown above.
- A column can be omitted, but the order must remain.

In this chapter, we will refer to the co-abundant groups as CAGs, which
are clusters of taxonomic features that co-vary across samples.

## Assigning taxonomic information

There are a number of methods to assign taxonomic information. We like
to give a short introduction about the methods available without ranking
one over the other. This has to be your choice based on the result for
the individual dataset.

### DADA2

The `r BiocStyle::Biocpkg("dada2")` package [@Callahan2016dada2] implements the
`assignTaxonomy()` function, which takes as input the ASV sequences
associated with each row of data and a training dataset. For more
information visit the
[dada2 homepage](https://benjjneb.github.io/dada2/assign.html).

### DECIPHER

The `r BiocStyle::Biocpkg("DECIPHER")` package [@R_DECIPHER] implements the
`IDTAXA` algorithm to assign either taxonomic information or function
information. For `r BiocStyle::Biocpkg("mia")`, only the first option is of
interest for now and more information can be found on the
[DECIPHER website](http://www2.decipher.codes/Classification.html).

## Functions to access taxonomic information

### Check taxonomy ranks in data

`checkTaxonomy()` checks whether the taxonomic information is usable for
`r BiocStyle::Biocpkg("mia")`.

```{r}
#| label: check_tax

checkTaxonomy(tse)
```

Since the `rowData` can contain other data, `taxonomyRanks()` will
return the columns that `r BiocStyle::Biocpkg("mia")` assumes to contain the
taxonomic information.

```{r}
#| label: tax_ranks

taxonomyRanks(tse)
```

This can then be used to subset the `rowData` to columns as needed.

```{r}
#| label: subset_rowData

rowData(tse)[, taxonomyRanks(tse)]
```

`taxonomyRankEmpty()` checks for empty values in the given `rank` and
returns a logical vector of `length(x)`.

```{r}
#| label: check_empty

all(!taxonomyRankEmpty(tse, rank = "Kingdom"))
table(taxonomyRankEmpty(tse, rank = "Genus"))
table(taxonomyRankEmpty(tse, rank = "Species"))
```

### Get taxonomy labels

`getTaxonomyLabels()` is a multi-purpose function, which turns taxonomic
information into a character vector of `length(x)`

```{r}
#| label: get_taxa_labels1

getTaxonomyLabels(tse) |> head()
```

By default, this will use the lowest non-empty information to construct
a string with the following scheme `level:value`. If all levels are the
same, this part is omitted, but can be added by setting
`with.rank = TRUE`.

```{r}
#| label: get_taxa_labels2

phylum <- !is.na(rowData(tse)$Phylum) &
    vapply(data.frame(apply(
        rowData(tse)[, taxonomyRanks(tse)[3:7]], 1L, is.na
    )), all, logical(1))
getTaxonomyLabels(tse[phylum, ]) |> head()
getTaxonomyLabels(tse[phylum, ], with.rank = TRUE) |> head()
```

By default the return value of `getTaxonomyLabels()` contains only
unique elements by passing it through `make.unique`. This step can be
omitted by setting `make.unique = FALSE`.

```{r}
#| label: get_taxa_labels3

getTaxonomyLabels(tse[phylum, ], with.rank = TRUE, make.unique = FALSE) |> head()
```

To apply the loop resolving function `resolveLoop()` from the
`r BiocStyle::Biocpkg("TreeSummarizedExperiment")` package
[@R_TreeSummarizedExperiment] within `getTaxonomyLabels()`, set
`resolve.loops = TRUE`.

### Get information on specific features

The function `getUnique()` gives a list of unique features for the
specified taxonomic rank.

```{r}
#| label: getUnique

getUnique(tse, rank = "Phylum") |> head()
```

With `mapTaxonomy()`, you can search information on certain taxonomic
features from the taxonomy table. For instance, we can check all the
features that matches with "Escherichia".

```{r}
#| label: mapTaxonomy

mapTaxonomy(tse, taxa = "Escherichia")
```

## Prune taxonomy tree {#sec-update-tree}

Subsetting is explained in detail in [@sec-treese_subsetting]. However,
if you've already subsetted your data, you may have noticed that the
taxonomy tree does not automatically update when using the `[]`
operators. Although the linkages between rows and tree nodes remain
correct, the tree retains its original, complex structure. You may be
wondering how to update the tree to reflect the newly simplified data.

`r BiocStyle::Biocpkg("mia")` package functions `subsetBy*` and
`agglomerateBy*` (see [@sec-agglomeration]) include an `update.tree`
parameter to handle this adjustment. However, when using `[]`, tree pruning
must be done as an additional step.

Let's start by selecting 5 arbitrary rows.

```{r}
#| label: subset_tse

tse_sub <- tse[1:5, ]
tse_sub
```

Even though we have only 5 rows, the tree still retains its original
number of tips. To align the tree with the subsetted data, we can use
the `r BiocStyle::Biocpkg("TreeSummarizedExperiment")`´s `subsetByLeaf()`
function, which allows us to select specific tips from the tree, effectively
pruning it to match the current subset of data.

```{r}
#| label: subset_tree

tse_sub <- subsetByLeaf(tse_sub, rowLeaf = rownames(tse_sub))
tse_sub
```

Now, we can see that the taxonomy tree has a simpler structure,
including only the selected leaves.

## Generate a hierarchy tree on the fly {#sec-fly-tree}

A hierarchy tree shows mapping between the taxonomic levels in taxonomic
rank table (included in `rowData`), rather than the detailed
phylogenetic relations. Usually, a phylogenetic tree refers to the
latter which is why we here call the generated tree a "hierarchy tree".

To create a hierarchy tree, `getHierarchyTree()` uses the information
and returns a `phylo` object. Duplicate information from the `rowData`
is removed.

```{r}
#| label: getHierarchyTree

getHierarchyTree(tse)
```

```{r}
#| label: addHierarchyTree

tse <- addHierarchyTree(tse)
tse
```

The implementation is based on the `toTree()` function from the
`r BiocStyle::Biocpkg("TreeSummarizedExperiment")` package
[@R_TreeSummarizedExperiment].

## Set taxonomy ranks

If your data includes taxonomy ranks that are not included by default in
`r BiocStyle::Biocpkg("mia")`, you can set the ranks manually. By doing so,
`r BiocStyle::Biocpkg("mia")` will be able to detect and utilize these
taxonomy ranks from your data as expected.

Get default ranks of `r BiocStyle::Biocpkg("mia")`.

```{r}
#| label: get_default_ranks

getTaxonomyRanks()
```

Set ranks to your own ranks. Remember that the order is meaningful.

```{r}
#| label: get_ranks
#| echo: false
#| message: false

ranks <- getTaxonomyRanks()
```

```{r}
#| label: set_ranks

# Set ranks
setTaxonomyRanks(c("test", "test2", "apple"))

# Get ranks
getTaxonomyRanks()
```

```{r}
#| label: set_default_ranks
#| echo: false
#| message: false

# Set default ranks back
setTaxonomyRanks(ranks)
```

## Visualizing phylogeny

The `plotRowTree()` function from the `r BiocStyle::Biocpkg("miaViz")` package
provides a convenient way to visualize the phylogeny.

```{r}
#| label: plot_phylogeny1

library(miaViz)

plotRowTree(tse)
```

By default, the visualized phylogeny has only a limited amount of
information. To enrich the visualization, we can prune the tree to genus
level and visualize to which phyla these genera belong to.

```{r}
#| label: plot_phylogeny2

# Agglomerate data and prune tree
tse <- agglomerateByRank(tse, rank = "Genus")

# Get top features
top <- getTop(tse, top = 50L)

# Plot
plotRowTree(
    tse[top, ],
    tip.colour.by = "Phylum"
)
```

In addition to nodes and tips, we can also color edges based on feature
metadata. Below, we color edges based on phyla and illustrate the mean
abundance of genera by coloring tips. Moreover, we switch to a dendrogram
layout with the help of the `r BiocStyle::Biocpkg("scater")` package.

```{r}
#| label: plot_phylogeny3

library(scater)

# Calculate mean abundance
tse <- addPerFeatureQC(tse)
rowData(tse)[["log_mean"]] <- log(rowData(tse)[["mean"]])

# Plot
plotRowTree(
    tse[top, ],
    layout = "dendrogram",
    edge.colour.by = "Phylum",
    tip.colour.by = "log_mean"
)
```

::: {.callout-tip icon="false"}
## Exercises

**Goal:** Learn how to handle taxonomy information.

**Exercise 1: Taxonomy information**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Ensure that the data includes `rowData` and `rowTree`. If not,
select some other dataset.

3. List the available taxonomic ranks in the data.

4. List the unique taxonomic features from specific taxonomy rank.

5. Subset data by taking, e.g., the first 10 rows.

6. How many nodes does the phylogeny include?

7. Prune the tree to correspond the subsetted data. How many nodes are
left after pruning?

8. Visualize the phylogeny.

9. Save the data to another variable. Calculate the hierarchy tree, and
visualize it. How does the tree differ from phylogeny?

Useful functions:

`utils::data()`, `SummarizedExperiment::rowData()`,
`TreeSummarizedExperiment::rowTree()`, `mia::taxonomyRanks()`,
`mia::getUnique()`, `TreeSummarizedExperiment::rowTree(tse)$Nnode`,
`TreeSummarizedExperiment::subsetByLeaf()`, `miaViz::plotRowTree()`,
`mia::addHierarchyTree()`

:::
