# Time series {#sec-time-series}

```{r setup, echo = FALSE, results = "asis"}
library(rebook)
chapterPreamble()
```

```{r include = FALSE}
# global knitr options
knitr::opts_chunk$set(
  message = FALSE,
  fig.width = 10,
  dpi = 300,
  dev = "png",
  dev.args = list(type = "cairo-png")
)
```

Microbial communities are dynamic, constantly evolving over time. However,
despite temporal fluctuations driven by environmental and other factors, the
microbiome retains its core characteristics [@Schmidt2018].

Compared to cross-sectional studies, a longitudinal study design allows
researchers to control time-related factors, enabling a more focused
investigation of specific research questions. However, longitudinal studies are
usually more complex than cross-sectional. Analyzing such data includes specific
aspects that need to be taken into account such as the dependence between
samples collected at multiple time points. Since these samples come from the
same set of subjects, they are paired, allowing each patient to serve as their
own control.

Most of the methods introduced earlier in this book can be applied to temporal
studies with appropriate modifications and considerations.

In this chapter, we use an example dataset from probiotic/placebo intervention
study. The dataset includes gut microbiota samples from 22 subjects and 2 time
points. The first time point was sampled before introducing a treatment.

```{r}
#| label: load_data

library(microbiomeDataSets)

mae <- peerj32()
tse <- getWithColData(mae, 1L)
tse[["time_point"]] <- tse[["time"]] |> as.factor()
```

## Alpha diversity

[@sec-alpha-diversity] introduced methods and main idea of alpha diversity.
Here we calculate Shannon diversity index.

```{r}
#| label: alpha1

library(mia)

tse <- addAlpha(tse, index = "shannon")
```

In addition to visualizing the diversity in respect to sample groups, we want
to analyse how much the diversity changes between time points. That is why
we calculate the difference between time points for each subject.

```{r}
#| label: alpha2

library(dplyr)

df <- colData(tse) |>
    as.data.frame() |>
    # Sort based on time
    arrange(time) |>
    # Group based on subject so that we calculate the difference between each
    # subject's samples
    group_by(subject) |>
    # Calculate difference
    mutate(diff = diff(shannon)) |>
    ungroup() |>
    DataFrame()

# Sort to original order and add difference info to colData
df <- df[match(tse[["sample"]], df[["sample"]]), , drop = FALSE]
rownames(df) <- colnames(tse)
colData(tse) <- df

colData(tse) |> head()
```

Now we can continue to visualize the alpha diversities. First we create a simple
boxplot showing Shannon index in respect to groups and time.

```{r}
#| label: alpha3

library(ggplot2)

# Get colData
df <- colData(tse) |>
    as.data.frame()

# Create a second df which is sorted
df2 <- df |>
    arrange(subject, time_point, group)

# Create a jitter
jitter_position <- position_jitterdodge(
    dodge.width = 0.75,
    jitter.width = 0,
    jitter.height = 0
    )

# Create a plot
p <- ggplot(mapping = aes(x = time_point, y = shannon, fill = group)) +
    # We disable outlier plotting as we plot points with jitter. Otherwise
    # outliers would be duplicated
    geom_boxplot(data = df, outlier.shape = NA) +
    # Add jitter plot
    geom_point(data = df2, aes(
        group = interaction(subject, group)),
        position = jitter_position)

p
```

The boxplot above shows alpha diversity in each group and time point. In the
dataset, each subject was sampled in each time point, however, from the plot
we cannot see which samples correspond to which subject. That is why we 
might be interested on linking the sample-pairs to highlight
the subject-specific deviation between time points.

```{r}
#| label: alpha4

# Add line to link subjects
p_path <- p +
    geom_line(
      data = df2,
      aes(group = interaction(subject, group)),
      position = jitter_position,
      alpha = 0.3
      )
p_path
```

We can further refine the plot by adding the magnitude of change with a color.

```{r}
#| label: alpha5

# Add line linking subjects
p_path <- p +
    geom_line(
      data = df2,
      aes(group = interaction(subject, group), color = diff),
      position = jitter_position
      ) +
  scale_color_gradient2(
      low="blue", mid="white", high="red", 
      limits = c(-max(abs(df[["diff"]])), max(abs(df[["diff"]]))))

p_path
```

The visualization does not show clear effect of treatment. We can still test
if we can get support on these null findings with statistical tests. The
simplest test might be the Wilcoxon test that we have to apply for each
treatment group separately. Moreover, we have to apply the paired version of
test as the samples are from same subjects.

We first test if placebo treatment had effect on alpha diversity.

```{r}
#| label: alpha6

# Get data on placebo samples
df <- tse[, tse[["group"]] == "Placebo"] |> colData()
# Get first and second time points separately
before <- df[df[["time_point"]] == 1, "shannon"]
after <- df[df[["time_point"]] == 2, "shannon"]

# Apply paired Wilcoxon test
wilcox.test(before, after, paired = TRUE)
```

Based on results, alpha diversity in second time point equals to first time
point for the placebo group, which is expected result. We can then do the same
test for group that got probiotics.

```{r}
#| label: alpha7

# Get data on placebo samples
df <- tse[, tse[["group"]] == "LGG"] |> colData()
# Get first and second time points separately
before <- df[df[["time_point"]] == 1, "shannon"]
after <- df[df[["time_point"]] == 2, "shannon"]

# Apply paired Wilcoxon test
wilcox.test(before, after, paired = TRUE)
```

Wilcoxon test confirms the results also for probiotics group: probiotic
treatment do not increase or decrease diversity of microbioal community.

More elegant test would be model-based approach. Below, we fit a simple linear
mixed effects model to examine the impact of time and treatment group on
Shannon diversity. In this model, both time and group are treated as fixed
effects, meaning that we assume their influence on Shannon diversity is
consistent across all subjects. Additionally, we include subject as a random
effect to account for the repeated measurements from the same individuals.

If you are not familiar with R formulas, here’s an explanation of how to
interpret the components:

- The left-hand side of the formula represents the predicted variable
(in this case, shannon diversity).

- The `+` sign in the formula means that we are adding the terms on the
right-hand side of the formula together (i.e., combining them as separate
effects).

- The `*` sign indicates an interaction between the terms on either side. This
means that we are not only testing the individual effects of time and group, but
also how these two factors interact with each other to influence Shannon
diversity. Specifically, it allows us to assess how the effect of treatment
group differs between time point 1 and time point 2.

```{r}
#| label: alpha8

library(lme4)
library(lmerTest)

# Get sample metadata
df <- colData(tse) |>
    as.data.frame()

# Fit the model
model <- lmer(shannon ~ time * group + (1 | subject), data = df)

# Display summary with p-values
summary(model)
```

The results indicate that none of the variables in the model show statistically
significant effects. Specifically, "time" represents the overall effect of time
on Shannon diversity, while "groupPlacebo" represents the overall difference in
Shannon diversity between the Placebo group and the LGG group, averaged across
both time points.

The most interesting variable in the results is the "time:groupPlacebo"
interaction term. This term investigates whether the effect of time on Shannon
diversity differs between the Placebo group and the LGG group when comparing the
second time point to the first one.

A significant interaction would indicate that the change in diversity over time
is different between the two groups. However, in this case, the interaction is
not statistically significant, suggesting that the treatment effect did not
differ significantly between the Placebo group and the LGG group between the two
time points.

## Time divergence

[@sec-divergence] introduced divergence which means that we calculate
dissimilarity between samples. We can do the same and calculate dissimilarity
between time points for each subject. Below we calculate Bray-Curtis
dissimilarity between consecutive time points.

```{r}
#| label: divergence1

library(miaTime)

tse <- transformAssay(tse, method = "relabundance")
tse <- addStepwiseDivergence(
    tse,
    time.col = "time",
    group = "subject",
    assay.type = "relabundance",
    method = "bray"
)
```

Then we can visualize the results with a box plot.

```{r}
#| label: divergence2

library(scater)

plotColData(tse, x = "group", y = "divergence", show_boxplot = TRUE)
```

Figure above summarizes the divergence. Each point represent a subject:
dissimilarity between samples from the two time points. Divergence between
consecutive time points seems to be equal between treatment groups.

## Ordination

Ordination plots, e.g., Principal Coordinate Analysis (PCoA) plot, is a common
visualization technique to find patterns from the data. Below, we perform PCoA
and visualize the results as showed in [@#sec-community-similarity].

Note that it is important to sort the data based on time so that the arrows that
we add in the next step have correct direction.

```{r}
#| label: pcoa

# Sort data based on time
tse <- tse[, order(tse[["time"]])]

# Perform PCoA
tse <- addMDS(tse, assay.type = "relabundance", method = "bray")
# Create ordination plot
p <- plotReducedDim(
    tse, dimred = "MDS", colour_by = "group",
    # Add other fields that are used later to add paths
    other_fields = c("subject", "time")
)
p
```

After initializing a scatter plot, we can add arrows that show how the microbial
communities evolved over time.

```{r}
#| label: pcoa2

# Add arrows to connect samples from subjects
p <- p + geom_path(
    mapping = aes(group = subject, ),
    arrow = arrow(length = unit(0.2, "cm")),
    alpha = 0.5
)
p
```

From the PCoA plot, we cannot observe clear patterns. As a final conclusion, it
seems that probiotics did not have effect on microbiome compared to placebo.
