# Time series {#sec-time-series}

```{r setup, echo = FALSE, results = "asis"}
library(rebook)
chapterPreamble()
```

```{r include = FALSE}
# global knitr options
knitr::opts_chunk$set(
  message = FALSE,
  fig.width = 10,
  dpi = 300,
  dev = "png",
  dev.args = list(type = "cairo-png")
)
```

Microbial communities are dynamic, constantly evolving over time. Despite
temporal fluctuations driven by environmental and other factors, the
microbiome retains its core characteristics [@Schmidt2018].

Compared to cross-sectional studies, a longitudinal study design allows
researchers to control time-related factors, enabling a more focused
investigation of specific research questions. However, longitudinal studies are
usually more complex than cross-sectional. Analyzing such data includes specific
aspects that need to be taken into account such as the dependence between
samples collected at multiple time points. Since these samples come from the
same set of subjects, they are paired, allowing each patient to serve as their
own control.

Most of the methods introduced earlier in this book can be applied to temporal
studies with appropriate modifications and considerations.

In this chapter, we use an example dataset from probiotic/placebo intervention
study.

```{r}
#| label: load_data

library(microbiomeDataSets)

mae <- microbiomeDataSets::peerj32()
tse <- getWithColData(mae, 1L)
tse[["time_point"]] <- tse[["time"]] |> as.factor()
```

## Alpha diversity

[@sec-alpha-diversity] introduced methods and main idea of alpha diversity.
Here we calculate Shannon diversity index.

```{r}
#| label: alpha1

library(mia)

tse <- addAlpha(tse, index = "shannon")
```

In addition to visualizing the diversity in respect to sample groups, we want
to visualize how much the diversity changes between time points. That is why
we calculate the difference for each subject.

```{r}
#| label: alpha2

library(dplyr)

df <- colData(tse) |>
    as.data.frame() |>
    # Sort based on time
    arrange(time) |>
    # Group based on subject so that we calculate the difference between each
    # subject's samples
    group_by(subject) |>
    # Calculate difference
    mutate(diff = diff(shannon)) |>
    ungroup() |>
    DataFrame()

# Sort to original order and add difference info to colData
df <- df[match(tse[["sample"]], df[["sample"]]), , drop = FALSE]
rownames(df) <- colnames(tse)
colData(tse) <- df

colData(tse) |> head()
```

Now we can continue to visualize the alpha diversities. First we create a simple
boxplot showing Shannon index in respect to groups and time.

```{r}
#| label: alpha3

library(ggplot2)

# Get colData
df <- colData(tse) |>
    as.data.frame()

# Create a second df which is sorted
df2 <- df |>
    arrange(subject, time_point, group)

# Create a jitter
jitter_position <- position_jitterdodge(
    dodge.width = 0.75,
    jitter.width = 0,
    jitter.height = 0
    )

# Create a plot
p <- ggplot(mapping = aes(x = time_point, y = shannon, fill = group)) +
    # We disable outlier plotting as we plot points with jitter. Otherwise
    # outliers would be duplicated
    geom_boxplot(data = df, outlier.shape = NA) +
    # Add jitter plot
    geom_point(data = df2, aes(
        group = interaction(subject, group)),
        position = jitter_position)

p
```

Additionally, we might be interested on linking the sample-pairs to visualize
their difference.

```{r}
#| label: alpha4

# Add line linking subjects
p_path <- p +
    geom_line(
      data = df2,
      aes(group = interaction(subject, group)),
      position = jitter_position
      )
p_path
```

```{r}
# Add line linking subjects
p_path <- p +
    geom_line(
      data = df2,
      arrow = arrow(length = unit(2, "mm")),
      aes(group = interaction(subject, group), color = diff),
      position = jitter_position
      ) +
  scale_color_gradient2(
      low="blue", mid="white", high="red", 
      limits = c(-max(abs(df[["diff"]])), max(abs(df[["diff"]]))))

p_path
```


MIten alfa-diversiteetti eroaa? Viiva potilainen näytteidenn välillä.
Väritä, onko punaine tai sininen, muutoksen suunta.

Tee nopea testi, jolla testataan, onko paired samples wilcoxon test, merkitsevä ero


## Time divergence

Analysoi näytteiden välinen ero, miten visualisoida?

Box plot, divergence y-askelilla, x -askelilla ryhmä

## Ordination

Viiva pootlaiden välillä, nuoli

In datasets including time points, researcher might be interested on how similar
samples are within a subject or patient and how the microbial communities
evolve. This can be done for instance by connecting points of each subject.

```{r}
library(miaTime)
library(scater)
data("temporalMicrobiome20")
tse <- temporalMicrobiome20

tse <- transformAssay(tse, method = "clr", pseudocount = 1)

tse <- runPCA(tse, assay.type = "clr")

p <- plotReducedDim(tse, "PCA")
```

Lisää tehtäviä. Lisää esimerkkivastauksiin chunk

