# Time series {#sec-time-series}

```{r setup, echo = FALSE, results = "asis"}
library(rebook)
chapterPreamble()
```

```{r include = FALSE}
# global knitr options
knitr::opts_chunk$set(
  message = FALSE,
  fig.width = 10,
  dpi = 300,
  dev = "png",
  dev.args = list(type = "cairo-png")
)
```

Microbial communities are dynamicm, constantly evolving over time. Despite
temporal fluctuations driven by environmental and other factors, the
microbiome retains its core characteristics [@Schmidt2018].

Compared to cross-sectional studies, a longitudinal study design allows
researchers to control time-related factors, enabling a more focused
investigation of specific research questions. However, longitudinal studies are
usually more complex than cross-sectional. Analyzing such data includes specific
aspects that need to be taken into account such as the dependence between
samples collected at multiple time points. Since these samples come from the
same set of subjects, they are paired, allowing each patient to serve as their
own control.

Most of the methods introduced earlier in this book can be applied to temporal
studies with appropriate modifications and considerations.

In this chapter, we use an example dataset from probiotic/placebo intervention
study.

```{r}
#| label: load_data

library(microbiomeDataSets)

mae <- microbiomeDataSets::peerj32()
tse <- getWithColData(mae, 1L)
tse[["time_point"]] <- tse[["time"]] |> as.factor()
```

## Alpha diversity

[@sec-alpha-diversity] introduced methods and main idea of alpha diversity.
Here we calculate

MIten alfa-diversiteetti eroaa? Viiva potilainen näytteidenn välillä.
Väritä, onko punaine tai sininen, muutoksen suunta.

Tee nopea testi, jolla testataan, onko paired samples wilcoxon test, merkitsevä ero

```{r}
#| label: alpha1

library(mia)

tse <- addAlpha(tse, index = "shannon")
```

```{r}
library(dplyr)
df <- colData(tse)
df <- df |> as.data.frame() |> arrange(time) |> group_by(subject) |> mutate(diff = diff(shannon)) |> ungroup() |> DataFrame()


library(ggplot2)
p <- ggplot(df, aes(x = time_point, y = shannon, group = interaction(time_point, group))) +
  geom_boxplot(outlier.shape = NA) +
  geom_point()
p
```


## Time divergence

Analysoi näytteiden välinen ero, miten visualisoida?

Box plot, divergence y-askelilla, x -askelilla ryhmä

## Ordination

Viiva pootlaiden välillä, nuoli

In datasets including time points, researcher might be interested on how similar
samples are within a subject or patient and how the microbial communities
evolve. This can be done for instance by connecting points of each subject.

```{r}
library(miaTime)
library(scater)
data("temporalMicrobiome20")
tse <- temporalMicrobiome20

tse <- transformAssay(tse, method = "clr", pseudocount = 1)

tse <- runPCA(tse, assay.type = "clr")

p <- plotReducedDim(tse, "PCA")
```

Lisää tehtäviä. Lisää esimerkkivastauksiin chunk

