# Transformation {#sec-assay-transform}

```{r}
#| label: setup
#| echo: false
#| results: asis

library(rebook)
chapterPreamble()
```

Data transformations are common in (microbial) ecology [@Legendre2001].
They are used to mitigate technical biases in the data, to obtain more
interpretable values, to enhance the comparability of samples/features or
to make data compatible with the assumptions of certain statistical methods.

Examples include transforming feature counts into relative abundances
(i.e., "normalising as proportions"), or with compositionality-aware
transformations such as the centered log-ratio transformation (clr).

## Characteristics of microbiome data to inform data transformations {#sec-stat-challenges}

Transformations are important in working with microbiome data due to
various unique characteristics of sequencing data. For example, the above
mentioned examples of feature transformations are commonly used with
microbiome data following the inherently proportional nature of sequencing
reads: Due to the nature of sequencing technology, read counts of taxa do not
represent real counts of microbes in the original sample, and raw counts of
taxa per sample are thus not comparable between samples. 

This **Compositionality** means that a change in the absolute abundance of
one taxon will lead to apparent variations in the relative abundances of other
taxa in the same sample. If neglected, such properties may cause significant
bias in the results of statistical tests. The above-mentioned transformation
processes are one method developed to overcome these issues by making taxon
abundances comparable across samples.

In addition to this compositionality, other special characteristics of
microbiome sequencing data are high variability and zero-inflation.
**High variability** means that abundance of taxa often varies by several
orders of magnitude from sample to sample. **Zero-inflation** means that
typically more than 70% of the taxa-per-sample values are zeros, which could
be due to either physical absence (structural zeros) or insufficient sequencing
coverage (sampling zeros).

## Common transformation methods

Let us now summarize some commonly used transformations in microbiome data
science; further details and benchmarkings are available in the
references.

- **alr**: The additive log ratio transformation is part of a broader Aitchison
family of transformations with 'clr' and 'rclr'. Compared to them the biggest
difference is that it selects a single feature or component as a reference
and expresses all other features as log-ratios relative to it.
@Greenacre2021 provides guidance on choosing an appropriate reference
feature.

- **clr**: Centered log ratio transformation [@Aitchison1986] is used to
reduce data skewness and compositionality bias in relative
abundances, while bringing the data onto a logarithmic scale. This
transformation is frequently applied in microbial ecology as it enhances
comparability of relative differences between samples [@Gloor2017]. However,
the resulting transformed values are difficult to interpret directly, and
it can only be applied to positive values, not zeros. A usual solution for
for making values non-zero is to add pseudocount, which adds another type
of bias in the data as true taxon absences are not taken into account.
This is mitigated by the method of robust clr transformation
(see rclr below).

- **hellinger**: Hellinger transformation is equal to the square root of
relative abundances. This ecological transformation can be useful
when the focus is on how species proportions vary across samples, rather
than on absolute counts.

- **log, log2, log10**: Logarithmic transformations, used e.g. to
reduce data skewness. With compositional data, the clr (or rclr)
transformation is often preferred.

- **pa**: Presence/Absence transformation ignores abundances and only
indicates whether the given feature is detected above the given
threshold (default: 0). This simple transformation is relatively
widely used in ecological research, whenever we have reasons to be interested
in which taxa are present in which samples more so than their abundances. An
example of such case is the process of inferring microbial ecological
association networks from covariation of taxa across samples. Here, microbes
that are rarely seen in the same samples are expected to have a negative
ecological association (either competitive exclusion with each other or
different ecological niche preferences). Conversely, looking for negative
correlations in their relative abundances would not be a meaningful measure
of ecological associations. Moreover, it has been used in machine learning
classification models with a good performance [@Giliberti2022; @Karwowska2024].

- **rank**: Rank transformation replaces each feature abundance value by its
rank. It is similar to relative rank transformation (rrank) which uses relative
ranks. This has use, for instance, in non-parametric statistics.
   
- **rclr**: The robust clr (rclr) is similar to regular clr (see above) but
allows data with zeroes and avoids the need to add pseudocount
[@Keshavan2010; @Martino2019].
   
- **relabundance**: Relative transformation, also known as normalising as
proportions, total sum scaling (TSS) and compositional transformation.
This converts counts into proportions (at the scale [0, 1]) that sum up to 1.
Much of the currently available taxonomic abundance data from
high-throughput assays (16S, metagenomic sequencing) is
compositional by nature, even if the data is provided as "counts"
[@Gloor2017].

- **standardize**: Standardize (or z-score) transformation scales data to zero
mean and unit variance. This is used to bring features (or samples) to more
comparable levels in terms of mean and scale of the values. This can enhance
visualization and interpretation of the data.

- Other available transformations include Chi square (`chi.square`),
Frequency transformation (`frequency`), and make margin sum of
squares equal to one (`normalize`).

Transformations on abundance assays can be performed with the
`transformAssay()` function from the `r BiocStyle::Biocpkg("mia")` package,
keeping both the original and the transformed
assay(s) in the data object. The function applies sample-wise or
column-wise transformation when `MARGIN = 'cols'`, feature-wise or
row-wise transformation when `MARGIN = 'rows'`. A complete list of
available transformations and parameters, is available in the function
[help](https://microbiome.github.io/mia/reference/transformAssay.html).

::: {.callout-important}

`Pseudocount` is a small non-negative value (e.g., 1) added to the normalized
data to avoid taking the logarithm of zero. Its value can have a significant
impact on the results when applying a logarithm transformation to normalized
data, as the logarithm transformation is a nonlinear operation that can
fundamentally change the data distribution [@Costea2014].

`Pseudocount` should be chosen consistently across all normalization
methods being compared, for example, by setting it to a value smaller
than the minimum abundance value before transformation. Some tools, like
`r BiocStyle::Biocpkg("ANCOMBC")`, take into account the effect of the
`pseudocount` by performing sensitivity tests using multiple pseudocount
values. See [@sec-differential-abundance].
:::

## Rarefaction {#sec-rarefaction}

Another approach to control uneven sampling depths is to apply rarefaction with `rarefyAssay()`, which resamples the samples to an equal number of reads. This remains controversial, however, and strategies to mitigate the information loss in rarefaction have been proposed [@Schloss_2024a; @Schloss_2024b]. Moreover, this practice has been discouraged for the analysis of differentially abundant microorganisms [@McMurdie_and_Holmes_2014].

Rarefaction can be performed iteratively by using the `niter` parameter in `rarefyAssay()`. This creates multiple rarefied versions of the data, which can help account for the stochasticity introduced by random subsampling. The resulting rarefied assays can then be used for downstream analyses such as alpha and beta diversity calculations.

### Using rarefaction with alpha diversity

When calculating alpha diversity indices, you can apply rarefaction iteratively and then compute diversity metrics across the rarefied replicates. The `addAlpha()` function can work with rarefied data:

```{r}
#| label: rarefaction-alpha
#| eval: false

# Load example data
library(mia)
data("Tengeler2020")
tse <- Tengeler2020

# Get minimum read depth for rarefaction
min_reads <- min(colSums(assay(tse, "counts")))

# Perform iterative rarefaction
tse <- rarefyAssay(
  tse,
  method = "subsample",
  sample = min_reads,
  niter = 100
)

# Calculate alpha diversity on rarefied data
tse <- addAlpha(
  tse,
  assay_name = "counts_rarefied",
  sample = min_reads,
  niter = 100
)
```

### Using rarefaction with beta diversity

Similarly, rarefaction can be applied before calculating beta diversity and performing ordination. The `addMDS()` function can utilize rarefied data for more robust distance calculations:

```{r}
#| label: rarefaction-beta
#| eval: false

# Perform MDS ordination on rarefied data
tse <- addMDS(
  tse,
  assay_name = "counts_rarefied",
  method = "bray",
  niter = 100
)
```

### Function comparison

**`addAlpha()` vs `getAlpha()`**: Both functions calculate alpha diversity indices, but `addAlpha()` stores the results directly into the `colData` of the TreeSummarizedExperiment object, while `getAlpha()` returns the diversity values as a separate vector or matrix. Use `addAlpha()` when you want to keep all data together in one object, and `getAlpha()` when you need the diversity values for immediate use in other calculations.

**`runMDS()` vs `addMDS()`**: The `runMDS()` function calculates multidimensional scaling coordinates and returns them as a separate matrix, whereas `addMDS()` calculates the MDS coordinates and stores them directly into the `reducedDim` slot of the TreeSummarizedExperiment object. Using `addMDS()` is generally preferred as it maintains all results within the same data object, making downstream analyses and visualization more straightforward.


## Transformations in practice

Below, we apply relative transformation to counts table.

```{r}
#| label: transf1

# Load example data
library(mia)
data("Tengeler2020")
tse <- Tengeler2020

# Transform counts assay to relative abundances
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")
```

Get the values in the resulting assay, and view some of the first
entries of it with the `head` command.

```{r}
#| label: transf2

assay(tse, "relabundance") |> head()
```

In 'pa' transformation, abundance table is converted to a
presence/absence table that ignores abundances and only indicates
whether the given feature is detected in the sample.

```{r}
#| label: transf3

tse <- transformAssay(tse, method = "pa")
assay(tse, "pa") |> head()
```

You can now view the entire list of abundance assays in your data object
with:

```{r}
#| label: transf4

assays(tse)
```

A common question is whether the centered log-ratio (clr) transformation
should be applied directly to raw counts or if a prior transformation,
such as conversion to relative abundances, is necessary.

In theory, the clr transformation is scale-invariant, meaning it does
not matter whether it is applied to raw or relative abundances, as long
as the relative scale of abundances remains the same. However, in
practice, there are some differences due to the introduction of a
pseudocount, which can introduce bias.

There is no single correct answer, but the following considerations may
help:

- Data imputation should typically be applied to raw abundances,
regardless of the microbial profiling pipeline used or whether the
obtained abundances are counts or relative abundances.

- Once a pseudocount has been added, it makes no difference whether
one first converts to relative abundances before applying clr or
applies clr directly to the adjusted counts.

- Since applying clr directly to raw counts is the simpler approach,
it is generally recommended.

- One might also consider using robust clr instead.

```{r}
#| label: transf5

tse <- transformAssay(
    x = tse,
    assay.type = "counts",
    method = "clr",
    pseudocount = TRUE,
    name = "clr"
)
```

To incorporate phylogenetic information, one can apply the phylogenetic
isometric log-ratio (`r BiocStyle::Biocpkg("philr")`) transformation
[@Silverman2017]. Unlike standard transformations,
`r BiocStyle::Biocpkg("philr")` accounts for the genetic relationships
between taxonomic features. This is important because closely related
species often share similar properties, which traditional
transformations fail to capture.

```{r}
#| label: transf6

tse <- transformAssay(tse, method = "philr", MARGIN = 1L, pseudocount = TRUE)
```

Unlike other transformations, `r BiocStyle::Biocpkg("philr")` outputs a table
where rows represent nodes of phylogeny. These new features do not match with
features of `r BiocStyle::Biocpkg("TreeSummarizedExperiment")` which is why
this new dataset is stored into `altExp`.

```{r}
#| label: transf7

altExp(tse, "philr")
```

::: callout-tip
## Summary

Microbiome data is characterized by the following features:

- Compositionality
- High variability
- Zero-inflation

[OSCA book](https://bioconductor.org/books/3.18/OSCA.basic/normalization.html#normalization-transformation)
provides additional information on normalization from the perspective of
single-cell analysis.

:::

::: {.callout-tip icon="false"}
## Exercises

**Goal:** The goal is to learn how to apply different transformations.

**Exercise 1: Transform data**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Visualize counts with a histogram. Describe the data distribution.
Is there a lot of zeros?

3. Transform the counts assay into relative abundances and store it
into the `TreeSE` as an assay named `relabund`.

4. Similarly, perform a CLR transformation on the counts assay with a
`pseudocount` of 1 and add it to the `TreeSE` as a new assay.

5. List the available assays by name.

6. Visualize the CLR-transformed data with a histogram. Compare the
distribution with distribution of counts data.

7. Access the CLR-assay and store it to variable. Select a subset
of its first 100 features and 10 samples, and print the abundance table.
Explore the data.

8. Agglomerate the data with `agglomerateByRanks()` then apply transformations
to the stored alternative experiments created. Use the option
`altexp = altExpnames(tse)`.

9. If the data has a phylogenetic tree, agglomerate the data to the family
level and perform the phILR transformation. Where was the transformed data
stored? Compare the feature names with original data. Why do the names differ?

Useful functions:

`utils::data()`, `miaViz::plotHistogram()`, `mia::transformAssay()`,
`mia::assayNames()`, `SummarizedExperiment::assay()`,
`mia::agglomerateByRanks()`, `mia::altExp()`, `BiocGenerics::rownames()`

:::
